---
params:
   task_nr: 19
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("2021_%s",
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}

code em {
font-weight:bold
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
	if (knitr::is_latex_output()) {
		sprintf("\\texttt{%s}", as.character(x))
	} else if (knitr::is_html_output()) {
		sprintf("<tt>%s</tt>", as.character(x))
	} else {
		x
	}
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(tibble)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(digest)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2021/day/", params$task_nr)
puzzle <- GET(base_url,
				  session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	xml_find_all("///article") %>% 
	lapply(as.character)

puzzle_data <- local({
	lines <- GET(paste0(base_url, "/input"),
					 session_cookie) %>% 
		content(encoding = "UTF-8") %>% 
		str_split("\n") %>% 
		`[[`(1L) %>% 
		Filter(nzchar, .)
	idx <- lines %>% 
		str_which("^---")
	map2(idx + 1L,
		  c(tail(idx, -1L) - 1L, length(lines)),
		  ~ lines[.x:.y] %>% 
		  	str_split(",") %>% 
		  	do.call(rbind, .) %>% 
		  	{matrix(as.numeric(.), nrow(.))})
})
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1L]])
```


### Solution

In a first step we have to find out which beacons any 2 scanners have in common. Since 
we do not have a reference point, we will calculate all distances between each beacon 
and match according to this distance. This works nicely, as the distance is
invariant to any rotation (we do not stretch).

If, for instance, beacons `1` and `2` are `123.45` units apart for scanner `1`, 
and beacons `4` and `14` are also `123.45` units apart, we can conclude that 
beacons map as follows between scanner `1` and `2`: `1->4` and `2->14`.

Once we established the common beacons, we have to apply all possible rotations
and check if the difference between the common beacons of scanner 1 and the 
beacons of the rotated scanner 2 became constant across all beacons. This yields
the offset between scanner 1 and 2.

We apply this idea to the initial list of scanners to get all scanners which have 
beacons in common with scanner 1. These scanners are removed from the list of
unmatched scanners. If the list of unmatched scanners is not empty yet, we continue 
with the first element in the list of matched scanners after scanner 1. In this way
we will eventually match all scanners. All we have to do then is to align all 
scanners by recursively applying the rotation and adding the offset and count the 
unique beacons.

```{r solution-a}
rotate <- function(alpha, beta, gamma) {
	c(cos(beta) * cos(gamma), 
	  cos(beta) * sin(gamma), 
	  -sin(beta), 
	  sin(alpha) * sin(beta) * cos(gamma) - cos(alpha) * sin(gamma), 
	  sin(alpha) * sin(beta) * sin(gamma) + cos(alpha) * cos(gamma),
	  sin(alpha) * cos(beta), 
	  cos(alpha) * sin(beta) * cos(gamma) + sin(alpha) * sin(gamma),
	  cos(alpha) * sin(beta) * sin(gamma) - sin(alpha) * cos(gamma), 
	  cos(alpha) * cos(beta)) %>% 
		round(0) %>% 
		as.integer() %>% 
		matrix(3, 3)
}

angle <- 0:3 * pi / 2
par <- expand.grid(alpha = angle, beta = angle, gamma = angle) %>% 
	as_tibble()
## remove identical rotation matrices
rot_mat <- par %>% 
	pmap(rotate) %>% 
	`[`(!duplicated(.))

match_scanners <- function(coords1, coords2) {
	## get pairwise distances
	d1 <- dist(coords1)
	d2 <- dist(coords2)
	
	## find out how many matching distances there are per beacon
	nr_matches <- d1 %>% 
		as.matrix() %>% 
		{`dim<-`(. %in% d2, dim(.))} %>% 
		colSums()
	
	if (sum(rel_beacons <- nr_matches == 11L) >= 12L) {
		## if we have at least 12 matching beacons...
		## ... reduce first scanner results to these
		c1 <- coords1[rel_beacons, ]
		## ... recalculate the distances for the matching beacons
		rel_dist <- dist(c1)
		
		## for each of these distances get row and col in the second
		## distance matrix
		match_coords <- map(rel_dist[seq(1L, nrow(c1) - 1L)],
								  ~ which(`dim<-`(as.matrix(d2) == .x, 
								  					 dim(as.matrix(d2))), 
								  		  arr.ind = TRUE))
		## however, for each matching we will get 2 positions (matrix is symmetric)
		## chose the one which occurs everywhere and bring it to long format
		match_coords <- match_coords %>% 
			reduce(~ inner_join(as_tibble(.x),
									  as_tibble(.y),
									  "row")) %>% 
			pivot_longer(everything())
		
		## now match_coords$value contains the beacons in order for the second
		## scanner
		c2 <- coords2[match_coords$value, ]
		stopifnot(isTRUE(all.equal(dist(c1), dist(c2), check.attributes = FALSE)))
		
		## c1 and c2 are not the common beacons and also aligned
		## find proper rotation such that diff is constant
		
		for (rm in rot_mat) {
			offset <- c1 - c2 %*% rm
			if (sum(!duplicated(offset)) == 1L) {
				## we have one single offset
				break;
			}
		}
		stopifnot(all.equal(c1, t(t(c2 %*% rm) + offset[1, ])))
		res <- tibble(base = digest(coords1),
						  self = digest(coords2),
						  orig = list(coords2), 
						  offset = list(matrix(offset[1, ], 1, 3)),
						  rotation = list(rm))
	} else {
		## no matching beacons
		res <- tibble(base = list(), self = list(), orig = list(), 
						  offset = list(), rotation = list())
	}
	res
}

remove_from_list <- function(lst, el) {
	dupes <- !duplicated(c(list(el), lst))
	lst[tail(dupes, -1L)]
}

rebase <- function(res) {
	get_value <- function(hash, what) {
		res %>% filter(self == hash) %>% 
			pull(what) %>% 
			`[[`(1L)
	}
	## each scanner is relative to another scanner, that is
	## self->base, we follow these path until we reach the
	## baseline
	lkp <- res %>% 
		pull(base) %>% 
		set_names(res %>% pull(self))
	base0 <- lkp[lkp == names(lkp)]
	new_offsets <- new_rotations <- vector("list", nrow(res)) %>% 
		set_names(names(lkp))
	for (hash in names(lkp)) {
		orig_hash <- hash
		offset <- get_value(hash, "offset")
		rm <- get_value(hash, "rotation")
		while (hash != base0) {
			## we follow the path at each offset is already in terms of
			## coordinates of the new base, that is we have first to
			## rotate the old offsets to align them to the new base
			## and then add the offset of this base (which is already in
			## coorindates of the base of this base)
			hash <- lkp[hash]
			offset <- (offset %*% get_value(hash, "rotation")) + 
				get_value(hash, "offset")
			## also rotate the rotation matrix such that in the end offset
			## and rotation are relative to base0
			rm <- rm %*% get_value(hash, "rotation")
		}
		new_offsets[[orig_hash]] <- offset
		new_rotations[[orig_hash]] <- rm
	}
	res %>% 
		mutate(offset = new_offsets,
				 rotation = new_rotations)
}

count_unique_beacons <- function(res) {
	## we need to pass the rebased offsets and rotations,
	## then all we have to is to rotate and add the offset
	shift <- function(orig, offset, rotation, ...) {
		t(t(orig %*% rotation) + c(offset))
	}
	## and count the number of unique beacons
	pmap(res, shift) %>% 
		do.call(rbind, .) %>% 
		unique() %>% 
		nrow()
}

get_all_offsets <- function(scanner_data = puzzle_data) {
	## start wiht the first scanner
	matched <- head(scanner_data, 1L)
	## all other scanners are unmatched
	unmatched <- tail(scanner_data, -1L)
	offsets <- vector("list", length(scanner_data))
	idx <- 1L
	coords0 <- tail(matched, 1L)[[1L]]
	offsets[[1L]] <- tibble(base = digest(coords0), self = digest(coords0),
									orig = list(coords0), 
									offset = list(matrix(rep(0, 3), 1, 3)), 
									rotation = list(diag(3)))
	while (length(unmatched)) {
		## while there are unmatched sensors, pop the last matched sensor
		coords0 <- tail(matched, 1L)[[1L]]
		matched <- head(matched, -1L)
		for (coords1 in unmatched) {
			## and loop through all unmatched sensors
			res <- match_scanners(coords0, coords1)
			if (nrow(res)) {
				## if we find a match
				idx <- idx + 1L
				offsets[[idx]] <- res
				## store the results and rmeove the matched sensor from 
				## the unmatched list
				## N.B. to later identify sensors we hash them to be able
				## to walk back eventually to base0 
				matched <- c(matched, list(coords1))
				unmatched <- remove_from_list(unmatched, coords1)
			}
		}	
	}
	do.call(rbind, offsets)
}

coords <- get_all_offsets() %>%
	rebase()

coords %>% 
	count_unique_beacons()
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

As we calculated already the relative positions of the scanners for part 1, all we
need to do is to get all pairwise distances.

```{r solution-b, eval = length(puzzle) > 1}
outer(coords$offset, 
		coords$offset, 
		\(x, y) map2_dbl(x, y, ~ sum(abs(.x - .y)))) %>% 
	max()
```

