---
params:
   task_nr: 19
   year: 2018
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}

.page-content table tr pre {
  border-radius: revert;
  background-color: revert;
  color: revert;
  padding: revert;
  border: revert;
}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(kableExtra)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  ip <- str_extract(text_block[[1L]], "\\d+") %>% 
    as.integer()
  ops <- text_block[-1L] %>% 
    str_split(" ") %>% 
    do.call(rbind, .) %>% 
    set_colnames(c("op", "A", "B", "C")) %>% 
    as_tibble() %>% 
    mutate(across(A:C, as.integer))
  list(ip = ip, ops = ops)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

Whenever we change register `r puzzle_data$ip` we implement a jump. `seti n . 4` 
instructions yield an absolute jump to instruction `n + 1`, while `addr j 4 4` (or 
equivalently `addr 4 j 4`) allow for skipping `r[j]` instructions. With this information 
we can deparse the code as follows:

```{r deparse-ops, echo = FALSE}
puzzle_data$ops %>% 
  mutate(Instruction = paste(op, A, B, C),
         Flow = c(
           "Jump to 18",
           "",
           "<--+",
           "<--|--+",
           "   |  |",
           "   |  |",
           "   |  |",
           "   |  |",
           "   |  |",
           "   |  |",
           "   |  |",
           "---|--+",
           "   |",
           "   |",
           "   |",
           "---+",
           "exit",
           rep("", 8L),
           "if r0 = 1 jump to 28",
           "else jump to 2",
           rep("", 8L),
           "jump to 2"),
         "Pseudo code" = c(
           "",
           "r3 = 1",
           "r1 = 1",
           "r2 = r3 * r1",
           rep("", 3L),
           "if (r2 == r5) r0+=r3",
           "r1++",
           rep("", 2L),
           "until r1 > r5",
           "r3++",
           rep("", 2L),
           "until r3 > r5",
           rep("", 20L)
         ), .keep = "none") %>% 
  mutate(across(-Instruction, ~ paste0("<pre>", .x, "</pre>"))) %>% 
  kbl(escape = FALSE, row.names = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE) %>% 
  column_spec(2:3, monospace = TRUE)
```

That is, instructions 18-27 form an initialization block, where we get a starting value
for `r5`. If `r0` equals 1 we add another initialization block in instructions 28-36.
After this setup block, we run a double loop where we look for all divisors of `r5` and 
sum them up. This can be tremendously simplified by **not** looping over 1 to `r5` but by
implementing this directly. This allows us to speed up the code significantly.


```{r get-solution-1}
add <- function(a, b, c, reg, immediate = FALSE) {
  if (immediate) {
    reg[c + 1L] <- reg[a + 1L] + b
  } else {
    reg[c + 1L] <- reg[a + 1L] + reg[b + 1L]
  }
  reg
}

mul <- function(a, b, c, reg, immediate = FALSE) {
  if (immediate) {
    reg[c + 1L] <- reg[a + 1L] * b
  } else {
    reg[c + 1L] <- reg[a + 1L] * reg[b + 1L]
  }
  reg
}

set <- function(a, b, c, reg, immediate = FALSE) {
  if (immediate) {
    reg[c + 1L] <- a
  } else {
    reg[c + 1L] <- reg[a + 1L]
  }
  reg
}

gt <- function(a, b, c, reg) {
  reg[c + 1L] <- reg[a + 1L] > reg[b + 1L]
  reg
}

eq <- function(a, b, c, reg) {
  reg[c + 1L] <- reg[a + 1L] == reg[b + 1L]
  reg
}

all_ops <- list(
  addr = \(...) add(..., immediate = FALSE),
  addi = \(...) add(..., immediate = TRUE),
  mulr = \(...) mul(..., immediate = FALSE),
  muli = \(...) mul(..., immediate = TRUE),
  setr = \(...) set(..., immediate = FALSE),
  seti = \(...) set(..., immediate = TRUE),
  gtrr = \(...) gt(...),
  eqrr = \(...) eq(...)
)

get_sum_of_divisors <- function(n) {
  div <- seq(1L, ceiling(sqrt(n)))
  sum(div[n %% div == 0L]) + 
    sum(n / div[n %% div == 0L & div ^ 2 != n])
}

apply_ops <- function(bg_dump = puzzle_data, r0 = 0) {
  ops <- bg_dump$ops
  ip <- bg_dump$ip + 1L
  ptr <- 0L
  reg <- rep(0L, 6)
  reg[1] <- r0
  i <- 1L
  while (between(ptr, 0L, nrow(ops) - 1L)) {
    if (ptr == 1L) {
      return(get_sum_of_divisors(reg[6L]))
    }
    op <- ops %>% 
      slice(ptr + 1L) %>% 
      as.list()
    reg[ip] <- ptr
    reg <- all_ops[[op$op]](op$A, op$B, op$C, reg)
    ptr <- reg[ip] + 1L
  }
  reg[1L]
}

apply_ops(puzzle_data, 0L)

```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For the second part we simply have to activate the second initialization block and run 
otherwise the same procedure.

```{r get-solution-2}
apply_ops(puzzle_data, 1L)
```
