---
params:
   task_nr: 17
   year: 2020
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(ravetools)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    str_split("") %>% 
    do.call(rbind, .)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We embed the matrix into a 3d array with enough space around to account for the expansion.
In each iteration we add at most one row/col/slice at either sides. Thus, if the current
dimensions are `(m, n, k)` the maximum dimensions needed are `(m + 2i, n + 2i, k +2i)`
where `i` is the number of iterations. 

Once we embedded the starting slice into a 3d array, we use convolution to calculate the
numbers of active neighbors for each element. To simplify this calculation, we encode the
data by `0` and `1` for active and inactive cubes respectively.


```{r get-solution-1}
evolve_cubes_3d <- function(state, cycles) {
  dd <- dim(state)
  slice <- matrix(0L, dd[1L] + 2* cycles, dd[2L] + 2* cycles)
  slice[(cycles + 1L):(cycles + dd[1L]), (cycles + 1L):(cycles + dd[2L])] <- 
    as.integer(state == "#")
  cube <- array(0L, c(dd, 1L) + 2* cycles)
  cube[, , cycles + 1L] <- slice
  kernel <- array(1L, rep(3L, 3L))
  kernel[2L, 2L, 2L] <- 0L
  for (i in seq_len(cycles)) {
    nbs <- convolve_volume(cube, kernel) %>% 
      round(0L)
    cube[cube == 1L & (nbs < 2L | nbs > 3L)] <- 0L
    cube[cube == 0L & nbs == 3L] <- 1L
  }
  sum(cube)
}

evolve_cubes_3d(puzzle_data, 6L)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]] %>% str_replace_all(fixed("*"), "\\*"))
```

### Solution

We still use the same logic as before but in 4d this time. However, since we cannot rely
on an existing convolution function we need to write an R based function for this, where 
we will use shifted arrays. 

```{r get-solution-2}

shift4d <- function(x, di = 0L, dj = 0L, dk = 0L, dl = 0L) {
  dd <- dim(x)
  stopifnot(length(dd) == 4L)
  out <- array(0L, dd)
  srcI <- max(1L, 1L + di):min(dd[1L], dd[1L] + di)
  srcJ <- max(1L, 1L + dj):min(dd[2L], dd[2L] + dj)
  srcK <- max(1L, 1L + dk):min(dd[3L], dd[3L] + dk)
  srcL <- max(1L, 1L + dl):min(dd[4L], dd[4L] + dl)
  
  dstI <- max(1L, 1L - di):min(dd[1L], dd[1L] - di)
  dstJ <- max(1L, 1L - dj):min(dd[2L], dd[2L] - dj)
  dstK <- max(1L, 1L - dk):min(dd[3L], dd[3L] - dk)
  dstL <- max(1L, 1L - dl):min(dd[4L], dd[4L] - dl)
  out[dstI, dstJ, dstK, dstL] <- x[srcI, srcJ, srcK, srcL]
  out
}

convolve_hypercube<- function(hypercube) {
  dd <- dim(hypercube)
  stopifnot(length(dd) == 4L)

  N <- -1:1
  res <- array(0L, dd)

  for (di in N) {
    for (dj in N) {
      for (dk in N) {
        for (dl in N) {
          if (!(di == 0L && dj == 0L && dk == 0L && dl == 0L)) {
            res <- res + shift4d(hypercube, di, dj, dk, dl)
          }
        }
      }
    }
  }
  res
}


evolve_cubes_4d <- function(state, cycles) {
  dd <- dim(state)
  slice <- matrix(0L, dd[1L] + 2 * cycles, dd[2L] + 2 * cycles)
  slice[(cycles + 1L):(cycles + dd[1L]), (cycles + 1L):(cycles + dd[2L])] <- 
    as.integer(state == "#")
  hypercube <- array(0L, c(dd, 1L, 1L) + 2 * cycles)
  hypercube[, , cycles + 1L, cycles + 1L] <- slice
  for (i in seq_len(cycles)) {
    nbs <- convolve_hypercube(hypercube) 
    hypercube[hypercube == 1L & (nbs < 2L | nbs > 3L)] <- 0L
    hypercube[hypercube == 0L & nbs == 3L] <- 1L
  }
  sum(hypercube)
}

evolve_cubes_4d(puzzle_data, 6L)
```
