---
params:
   task_nr: 5
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(knitr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

puzzle_data <- local({
  res <- GET(paste0(base_url, "/input"),
             session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    str_split("\n") %>% 
    extract2(1L)
  rules <- str_subset(res, "\\d+\\|\\d+") %>% 
    str_split(fixed("|")) %>% 
    map(~ setNames(as.integer(.x), c("x", "y")) %>% as.list() %>% as_tibble()) %>% 
    list_rbind()
  updates <- str_subset(res, "(\\d+,\\d+,?)+") %>% 
    str_split(fixed(",")) %>% 
    map(as.integer)
  list(rules = rules, updates = updates)
})
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

First, we define a function which loops through an update and checks if all relevant rules
are obeyed:

```{r check-compliance}
is_valid_order <- function(update, rules = puzzle_data$rules) {
  check_number <- function(solution, el) {
    ancestors <- solution$ancestors
    forbidden <- rules %>% 
      filter(x == el) %>% 
      pull(y)
    res <- !any(ancestors %in% forbidden)
    ret <- list(ancestors = head(ancestors, -1L), res = solution$res && res)
    if (!res) {
      done(ret)
    } else {
      ret
    }
  }
  reduce(rev(update), 
         check_number, .init = list(ancestors = head(update, -1L), res = TRUE)) %>% 
    extract2("res")
}
```

Now we can filter all valid rules, extract their middle point and sum it

```{r get-solution-1}
keep(puzzle_data$updates, is_valid_order) %>% 
  map_int(~ .x[ceiling(length(.x) / 2)]) %>% 
  sum()
```


## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

The idea of the sorting algorithm is as follows:

1. Start with an empty vector.
1. Take the first element and place it into the vector,
1. Consecutively add new elements `el`. For each element look at all rules, which apply, 
   find the indices of all the elements which `el` must precede and place it before the
   element with the smallest index.
1. Continue until all elements were put inot the list.

```{r sort-update}
sort_update <- function(update, rules = puzzle_data$rules) {
  insert_element <- function(sorted_list, element) {
    n <- length(sorted_list)
    ancestors <- rules %>% 
      filter(x == element) %>% 
      pull(y)
    idx <- which(sorted_list %in% ancestors)
    if (length(idx) == 0L) {
      res <- c(sorted_list, element)
    } else if(min(idx) == 1L) { 
      res <- c(element, sorted_list)
    } else {
      idx <- min(idx)
      res <- c(head(sorted_list, idx - 1L), element, tail(sorted_list, n - idx + 1L))
    }
    res
  }
  reduce(update, insert_element, .init = integer(0))
}
```

Now that we can properly sort the updates, we simply have to apply this sorting to 
invalid updates and sum their respective middle points:

```{r get-solution-2}
discard(puzzle_data$updates, is_valid_order) %>% 
  map_int(function(.x) {
    sx <- sort_update(.x, puzzle_data$rules)
    sx[ceiling(length(sx) / 2)]
  }) %>% 
  sum()
```
