---
params:
   task_nr: 3
   year: 2015
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2015"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(tibble)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(stringi)
library(knitr)
library(cli)
library(bit64)
library(igraph)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    str_split("") %>% 
    extract2(1L)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We use imaginary numbers to encode movement. The real part corresponds to horizontal 
movements and the imaginary part to vertical movements. Wiht this encoding, we simply
have to form the cumulative sum and count unique values.


```{r get-solution-1}
dirs <- c("<" = complex(real = -1, imaginary = 0),
          ">" = complex(real = 1, imaginary = 0),
          "^" =  complex(real = 0, imaginary = 1),
          "v" = complex(real = 0, imaginary = -1))

path <- c(
  complex(real = 0, imaginary = 0),
  dirs[puzzle_data] %>% 
    cumsum()
)

path %>% 
  unique() %>% 
  length()

```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For the second part we follow the same idea, but split the movements into the odd and even
positions.

```{r get-solution-2}
santa <- puzzle_data[seq(1L, length(puzzle_data), by = 2L)]
robo <- puzzle_data[seq(2L, length(puzzle_data), by = 2L)]

path <- c(complex(real = 0, imaginary = 0),
          dirs[santa] %>% cumsum(),
          dirs[robo] %>% cumsum())
path %>% 
  unique() %>% 
  length()
```
