---
params:
   task_nr: 10
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(knitr)
library(cli)
library(bit64)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) > 1L) {
    text_block <- paste(text_block, collapse = "\n")
  }
  text_block %>% 
    unlist() %>% 
    str_split("\n") %>% 
    extract2(1L) %>% 
    keep(nzchar) %>% 
    str_split("") %>% 
    map(as.integer) %>% 
    do.call(rbind, .) 
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```
```{r get_example_data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We use a simple backtracking algorithm, which walks the map and counts all valid paths.
A valid path starts at `0`, ends at `9` and visits the numbers `0` to `9` consecutively. 
If we can reach the same goal via several paths we still count it only once. However,
as in the second path we need to count all paths, we allow for coutning all paths already 
in his function to avoid re-inventing the wheel. For easier handling we store *x* and *y*
coordinates of the nodes visited again as complex numbers, with the real and the imaginary
part corresponding to the *x* and *y* coordinates respectively,


```{r get-solution-1}
get_id <- function(pos) {
  paste(pos, collapse = "/")
}

get_all_trails <- function(trail_map, store_all_paths) {
  dd <- dim(trail_map)
  starting_pos <- which(trail_map == 0L, arr.ind = TRUE)
  dirs <- list(
    up = cbind(-1, 0),
    right = cbind(0, 1),
    down = cbind(1, 0),
    left = cbind(0, -1)
  )
  all_paths <- vector("list", nrow(starting_pos))
  names(all_paths) <- apply(starting_pos, 1, get_id)
  find_path <- function(pos, current_height, start, current_path) {
    if (current_height == 9L) {
      traf <- if (store_all_paths) identity else unique
      all_paths[[get_id(start)]] <<- 
        c(all_paths[[get_id(start)]],
          complex(real = pos[1], imaginary = pos[2])) %>% 
        traf()
    } else {
      for (dir in dirs) {
        new_pos <- pos + dir
        if (all(new_pos >= c(1L, 1L) & new_pos <= dd)) {
          ## valid position
          if (trail_map[new_pos] == current_height + 1L) {
            Recall(new_pos, 
                   current_height + 1L, 
                   start, 
                   c(current_path, complex(real = pos[1], imaginary = pos[2])))
          }
        }
      }
    }
  }
  for (start_idx in 1:nrow(starting_pos)) {
    start <- starting_pos[start_idx, , drop = FALSE]
    find_path(start, 0L, start, complex(0))
  }
  all_paths %>% 
    lengths() %>% 
    sum()
}

get_all_trails(puzzle_data, FALSE)
```


## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

This time we do not filter out already visited end points.

```{r get-solution-2}
get_all_trails(puzzle_data, TRUE)
```
