---
params:
   task_nr: 15
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(knitr)
library(cli)
library(bit64)
library(ggplot2)
library(gganimate)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      unlist() %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      head(-1L)
  }
  split_idx <- which(!nzchar(text_block))
  warehouse <- text_block[seq(1L, split_idx - 1L)]
  movements <- text_block[seq(split_idx + 1L, length(text_block))]
  list(
    warehouse = str_split(warehouse, "") %>% 
      do.call(rbind, .),
    movements = str_split(movements, "") %>% 
      unlist()
  )
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r get_example_data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

The solution is rather straight forward. Let the robot move and on each step determine 
where he will move to:
1. If the new spot is empty, simply move to it.
1. If it is a wall, stay where you are.
1. If it is a crate, check for the first empty spot behind the crates and push the crates.


```{r get-solution-1, cache = TRUE, message = FALSE}
show_warehouse <- function(warehouse) {
  cat(
    apply(
      warehouse, 1, paste, collapse = ""
    ) %>% 
      paste(collapse = "\n"),
    "\n\n"
  )
}

plot_warehouse <- function(warehouse, direction) {
  if (!is.list(warehouse)) {
    warehouse <- list(warehouse)
  }
  stopifnot(length(warehouse) == length(direction))
  coords <- map(seq_along(warehouse), ~ tibble(
    x = rep(1:ncol(warehouse[[.x]]), each = nrow(warehouse[[.x]])),
    y = rep(nrow(warehouse[[.x]]):1, ncol(warehouse[[.x]])),
    value = c(warehouse[[.x]])
  ) %>% 
    mutate(fill = case_when(
      value == "@" ~ "firebrick",
      value %in% c("O", "[", "]") ~ "burlywood",
      value == "#" ~ "gray70",
      TRUE ~ NA_character_
    ),
    start = value == "@",
    dir = direction[[.x]],
    vj = case_when(
      dir == "^" ~ .75, 
      dir == "v" ~ .35,
      TRUE ~ .5),
    frame = .x)
  ) %>% 
    list_rbind()
  res <- ggplot(coords, aes(x = x, y = y)) + 
    geom_tile(aes(fill = fill), linewidth = .5, color = "black") +
    scale_fill_identity() + 
    theme_void() +
    coord_fixed() +
    geom_text(data = coords %>% filter(start),
              aes(label = dir, vjust = vj), 
              color = "white", size = 15) +
    geom_text(aes(label = label), 
              data = coords %>% 
                select(x, y) %>% 
                mutate(label = max(y) + 1 - y, x = 0) %>% 
                distinct(),
              color = "#007ACC",
              hjust = .5) +
    geom_text(aes(label = label), 
              data = coords %>% 
                select(x, y) %>% 
                mutate(label = max(y) + 1L - y, x = max(x) + 1) %>% 
                distinct(),
              color = "#007ACC",
              hjust = .5) +
    geom_text(aes(label = label), 
              data = coords %>% 
                select(x, y) %>% 
                mutate(label = x, y = 0) %>% 
                distinct(),
              color = "#007ACC",
              vjust = .5) +
    geom_text(aes(label = label), 
              data = coords %>% 
                select(x, y) %>% 
                mutate(label = x, y = max(y) + 1) %>% 
                distinct(),
              color = "#007ACC",
              vjust = .5) +
    labs(subtitle = "Step {frame}") +
    transition_manual(frame) +
    theme(
      plot.subtitle = element_text(color = "#007ACC"),
      plot.background = element_blank(),
      panel.background = element_blank()
    )
                                     
  invisible(res)
}

push_boxes <- function(warehouse, movements, do_print = FALSE, save_plot = FALSE) {
  dirs <- rbind(
    cbind(-1L, 0L),
    cbind(0L, 1L),
    cbind(1L, 0L),
    cbind(0L, -1L)
  ) %>% 
    set_rownames(c("^", ">", "v", "<"))
  current_pos <- which(warehouse == "@", arr.ind = TRUE)
  i <- 1
  all_warehouses <- list(warehouse)
  move <- function(current_pos, direction) {
    dir <- dirs[direction, , drop = FALSE]
    new_pos <- current_pos + dir
    new_field <- warehouse[new_pos]
    if (new_field == ".") {
      ## field is empty => move
      warehouse[rbind(current_pos, new_pos)] <<- c(".", "@")
    } else if (new_field == "#") {
      ## field is a wall => do not move
      new_pos <- current_pos
    } else if (new_field == "O") {
      ## field is a crate
      ## get distance to border
      if (any(dir < 0)) {
        ## we go left / up
        ref_point <- cbind(1L, 1L)
      } else {
        ref_point <- dim(warehouse)
      }
      distance_border <- abs(c((ref_point - current_pos) / dir))
      distance_border <- distance_border[is.finite(distance_border)]
      if (distance_border >= 3) {
        ## if we are at least 3 fields away from border 
        ## (for "# o @" there cannot be an empty space)
        ## all fields next to the current position until the border   
        idx <- map(seq(2L, distance_border - 1L), ~ current_pos + .x * dir) %>% 
          do.call(rbind, .)
        ## free spaces in the same row / col
        free_spaces <- warehouse[idx] == "."
        ## walls in the same row / col
        walls <- warehouse[idx] == "#"
        ## first free space
        first_free_space <- which.max(free_spaces)
        closest_wall <- which.max(walls)
        if (any(free_spaces) && (!any(walls) || first_free_space < closest_wall)) {
          ## if there is at least one free space _after_ a wall 
          warehouse[idx[seq(1L, first_free_space), , drop = FALSE]] <<- "O"
          warehouse[new_pos] <<- "@"
          warehouse[current_pos] <<- "."
        } else {
          ## if there's none, don't move
          new_pos <- current_pos
        }
      } else {
        new_pos <- current_pos
      }
    }
    if (do_print) {
      msg <- paste0("Step ", i, ": ", direction)
      cat(msg, "\n")
      show_warehouse(warehouse)
    }
    i <<- i + 1
    if (save_plot) {
      all_warehouses <<- c(all_warehouses, list(warehouse))
    }
    new_pos
  }
  reduce(movements, move, .init = current_pos)
  if (save_plot) {
    anim <- animate(plot_warehouse(all_warehouses, c("S", movements)), fps = 4, 
                    nframes = length(all_warehouses))
    
  } else {
    anim <- NULL
  }
  list(warehouse = warehouse, anim = anim)
}

calculate_gps <- function(warehouse) {
  pos <- which(warehouse == "O" | warehouse == "[", arr.ind = TRUE)
  sum((pos[, "row"] - 1L) * 100L + (pos[, "col"] - 1L))
}

solution1 <- push_boxes(puzzle_data$warehouse,
                        puzzle_data$movements, save_plot = FALSE)

calculate_gps(solution1$warehouse)
```

For the fun of it, here is a GIF showing the robot at work for a smaller warehouse:

```{r show-animation-1, echo = FALSE, message = FALSE}
example_data <- list(warehouse = structure(c("#", "#", "#", "#", "#", "#", "#", "#", 
                                             "#", ".", "#", ".", ".", ".", ".", "#", 
                                             "#", ".", "@", ".", "#", ".", ".", "#", 
                                             "#", "O", ".", ".", ".", ".", ".", "#", 
                                             "#", ".", "O", "O", "O", "O", ".", "#", 
                                             "#", "O", ".", ".", ".", ".", ".", "#", 
                                             "#", ".", ".", ".", ".", ".", ".", "#", 
                                             "#", "#", "#","#", "#", "#", "#", "#"), 
                                           dim = c(8L, 8L)), 
                     movements = c("<", "^", "^", ">", ">", ">", "v", "v", "<", "v", ">",
                                   ">", "v", "<", "<"))
ex1 <- push_boxes(example_data$warehouse, example_data$movements, save_plot = TRUE)
ex1$anim
```



## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We have to adapt our code slightly. Left right pushes work the same way as before, we just
have to make sure that we mark the new fields with the proper crate markers.

For upward and downward pushes, we have, however, to identify the current block. That is 
all crates which are vertically connected. We do that via a backtracking algorithm. If 
there is a wall above (below) any of the crates which we want to push, we return `NA`, 
thus, effectively avoiding a push.


```{r get-solution-2, cache = TRUE}
enlarge_warehouse <- function(warehouse) {
  dd <- dim(warehouse)
  new_warehouse <- matrix(".",  dd[1L], 2L * dd[2L])
  idxs <- which(warehouse != "." , arr.ind = TRUE)
  for (i in seq_len(nrow(idxs))) {
    idx <- idxs[i, , drop = FALSE]
    pos_1 <- pos_2 <- idx
    pos_1[2L] <- 2L * pos_1[2L] - 1L
    pos_2[2L] <- 2L * pos_2[2L]
    pos <- rbind(pos_1, pos_2)
    if (warehouse[idx] == "O") {
      new_warehouse[pos] <- c("[", "]")
    } else if (warehouse[idx] == "@") {
      new_warehouse[pos[1L, , drop = FALSE]] <- "@" 
    } else {
      new_warehouse[pos] <- warehouse[idx]
    }
  }
  new_warehouse
}

push_big_boxes <- function(warehouse, movements, do_print = FALSE, save_plot = FALSE) {
  dirs <- rbind(
    cbind(-1L, 0L),
    cbind(0L, 1L),
    cbind(1L, 0L),
    cbind(0L, -1L)
  ) %>% 
    set_rownames(c("^", ">", "v", "<"))
  current_pos <- which(warehouse == "@", arr.ind = TRUE)
  i <- 1
  all_warehouses <- list(warehouse)
  move <- function(current_pos, direction) {
    dir <- dirs[direction, , drop = FALSE]
    new_pos <- current_pos + dir
    new_field <- warehouse[new_pos]
    if (new_field == ".") {
      ## field is empty => move
      warehouse[rbind(current_pos, new_pos)] <<- c(".", "@")
    } else if (new_field == "#") {
      ## field is a wall => do not move
      new_pos <- current_pos
    } else if (new_field %in% c("[", "]")) {
      ## field is a crate
      if (direction %in% c("<", ">")) {
        ## get distance to border
        if (any(dir < 0)) {
          ## we go left / up
          ref_point <- cbind(1L, 1L)
        } else {
          ref_point <- dim(warehouse)
        }
        distance_border <- abs(c((ref_point - current_pos) / dir))
        distance_border <- distance_border[is.finite(distance_border)]
        ## left / right
        if (distance_border >= 5) {
          ## if we are at least 5 fields away from border 
          ## (for "# # [ ] @" there cannot be an empty space)
          ## all fields next to the current position until the border   
          idx <- map(seq(1L, distance_border - 1L), ~ current_pos + .x * dir) %>% 
            do.call(rbind, .)
          ## free spaces in the same row / col
          free_spaces <- warehouse[idx] == "."
          ## walls in the same row / col
          walls <- warehouse[idx] == "#"
          ## first free space
          first_free_space <- which.max(free_spaces)
          closest_wall <- which.max(walls)
          if (any(free_spaces) && (!any(walls) || first_free_space < closest_wall)) {
            ## if there is at least one free space _after_ a wall 
            
            block_idx <- seq(1L, first_free_space)
            stopifnot(length(block_idx) %% 2L == 1L)
            block <- rep(c("[", "]"), (length(block_idx) - 1L) / 2L)
            block_idx <- idx[tail(block_idx, -1L), , drop = FALSE]
            warehouse[block_idx[order(block_idx[, 2]), , drop = FALSE]] <<- block
            warehouse[new_pos] <<- "@"
            warehouse[current_pos] <<- "."
          } else {
            ## if there's none, don't move
            new_pos <- current_pos
          }
        } else {
          new_pos <- current_pos
        }
      } else {
        ## up / down
        get_block <- function(current_pos) {
          res <- NULL
          if (warehouse[current_pos] %in% c("[", "]")) {
            ## we hit (yet another) a box
            left_part <- right_part <- current_pos
            if (warehouse[current_pos] == "[") {
              right_part[2L] <- right_part[2L] + 1L
            } else {
              left_part[2L] <- left_part[2L] - 1L
            }
            me <- rbind(left_part, right_part)
            res <- rbind(me,
                         Recall(left_part + dir),
                         Recall(right_part + dir))
          } else if (warehouse[current_pos] == "#") {
            ## there is a wall in the way so we cannot move
            res <- NA
          } else if (warehouse[current_pos] == ".") {
            ## there is an empty space so this box part could be moved
            res <- NULL
          }
          res
        }
        block_idx <- get_block(new_pos)
        if (any(is.na(block_idx[, 1]))) {
          ## block cannot be moved b/c a wall is blocking at some point
          new_pos <- current_pos
        } else {
          block <- warehouse[block_idx]
          new_block_idx <- t(t(block_idx) + c(dir))
          warehouse[block_idx] <<- "."
          warehouse[new_block_idx] <<- block
          warehouse[new_pos] <<- "@"
          warehouse[current_pos] <<- "."
        }
      }
    }
    if (do_print) {
      msg <- paste0("Step ", i, ": ", direction)
      cat(msg, "\n")
      show_warehouse(warehouse)
    }
    i <<- i + 1
    all_warehouses <<- c(all_warehouses, list(warehouse))
    new_pos
  }
  reduce(movements, move, .init = current_pos)
  if (save_plot) {
    anim <- animate(plot_warehouse(all_warehouses, c("S", movements)), fps = 4, 
                    nframes = length(all_warehouses))
    
  } else {
    anim <- NULL
  }
  list(warehouse = warehouse, anim = anim)
}

solution2 <- enlarge_warehouse(puzzle_data$warehouse) %>% 
  push_big_boxes(puzzle_data$movements)
calculate_gps(solution2$warehouse)
```

And again a small visualization:

```{r show-animation-2, echo = FALSE, message = FALSE}
ex2 <- push_big_boxes(enlarge_warehouse(example_data$warehouse), 
                  example_data$movements, save_plot = TRUE)
ex2$anim
```


