---
params:
   task_nr: 11
   year: 2020
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    str_split("") %>% 
    do.call(rbind, .)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We simulate the process by counting occupied seats in each direction. For a fast 
calculation we operate on the whole matrix and build the sum by shifting the matrix in 
each of the 8 directions.

```{r get-solution-1}
simulate_seats <- function(seats) {
  nr <- nrow(seats)
  nc <- ncol(seats)
  unstable <- TRUE
  nbs <- vector("list", 8L)
  k <- 1
  for (i in -1:1) {
    for (j in -1:1) {
      if (i == 0L && j == 0) {
        next
      }
      nbs[[k]] <- list(from = list(r = max(1L, 1L - i):min(nr, nr - i),
                                   c = max(1L, 1L - j):min(nc, nc - j)),
                       to = list(r = max(1L, 1 + i):min(nr, nr + i),
                                 c = max(1, 1 + j):min(nc, nc + j))
                       )
      k <- k + 1L
    }
  }
  count_neighbors <- function(x) {
    new <- matrix(0L, nr, nc)
    for (il in nbs) {
      new[il$to$r, il$to$c] <- new[il$to$r, il$to$c] + x[il$from$r, il$from$c]
    }
    new
  }
  while (unstable) {
    prev <- seats
    nbs_cnt <- count_neighbors(seats == "#")
    seats[seats == "L" & nbs_cnt == 0L] <- "#"
    seats[seats == "#" & nbs_cnt >= 4L] <- "L"
    unstable <- !identical(prev, seats)
  }
  sum(seats == "#")
}

simulate_seats(puzzle_data)

```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]] %>% str_replace_all(fixed("*"), "\\*"))
```

### Solution

To implement the line-of-sight we pre-compute for each position the next seat ignoring
the floor. Then we iterate over each elemt in the matrix and use the pre-computed indices 
to get the neighbors.

```{r get-solution-2}
get_neighbors <- function(seats) {
  nr <- nrow(seats)
  nc <- ncol(seats)
  idx <- function(coord) {
    (coord[, 2L] - 1L) * nr + coord[, 1L]
  }
  dirs <- cbind(
    rep(-1:1, each = 3L),
    rep(-1:1, 3L)
  )[-5L, , drop  = FALSE]
  nbs <- vector("list", nr * nc)
  for (r in seq_len(nr)) {
    for (c in seq_len(nc)) {
      if (seats[r, c] != ".") {
        vis <- integer(0L)
        for (di in seq_len(8L)) {
          d <- dirs[di, , drop = FALSE]
          start <- cbind(r, c)
          nb <- start + d
          in_bounds <- nb[, 1L] >= 1L & nb[, 1L] <= nr & nb[, 2L] >= 1L & nb[, 2L] <= nc
          found <- FALSE
          while (in_bounds && !found) {
            if (seats[nb] != ".") {
              vis <- c(vis, idx(nb))
              found <- TRUE
            } else {
              nb <- nb + d
              in_bounds <- nb[, 1L] >= 1L & nb[, 1L] <= nr & 
                nb[, 2L] >= 1L & nb[, 2L] <= nc
            }
          }
          nbs[[idx(start)]] <- unname(vis)
        }        
      }
    }
  }
  nbs
}

simulate_seats_los <- function(seats) {
  nr <- nrow(seats)
  nc <- ncol(seats)
  unstable <- TRUE
  nbs <- get_neighbors(seats)
  r <- 1L
  while (unstable) {
    prev <- seats
    seat_idx <- which(seats != ".")
    for (idx in seat_idx) {
      nbs_cnt <- sum(prev[nbs[[idx]]] == "#")
      if (prev[idx] == "L" && nbs_cnt == 0L) {
        seats[idx] <- "#"
      } else if (prev[idx] == "#" && nbs_cnt >= 5L) {
        seats[idx] <- "L"
      }
    }
    unstable <- !identical(prev, seats)
  }
  sum(seats == "#")
}

simulate_seats_los(puzzle_data)
```
