---
params:
   task_nr: 7
   year: 2025
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L)
  }
  text_block[nzchar(text_block)] %>% 
    str_split("") %>% 
    do.call(rbind, .)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We walk from the source downwards, marking the path with `0` and splitters with `1`. 
Whenever we encounter a splitter, we add the two new paths to the matrix of positions to 
continue walking downwards. In the end, we sum up all the splitters that we have 
encountered.

```{r get-solution-1}
mark_path <- function(tachyon) {
  pos <- which(tachyon == "S", arr.ind = TRUE)
  path <- array(NA_integer_, dim(tachyon))
  down <- c(1L, 0L)
  sides <- rbind(
    "<" = c(0L, -1L),
    ">" = c(0L, 1L)
  )
  nr <- nrow(tachyon)
  path[pos] <- 0L
  while(nrow(pos) > 0L) {
    new_pos <- t(t(pos) + down)
    out_of_space <- new_pos[, 1L] > nr
    new_pos <- new_pos[!out_of_space, , drop = FALSE]
    path[new_pos] <- 0L
    is_splitter <- tachyon[new_pos] == "^"
    if (any(is_splitter)) {
      splits <- new_pos[is_splitter, , drop = FALSE]
      splits <- rbind(
        t(t(splits) + sides[1L, ]),
        t(t(splits) + sides[2L, ])
      )
      splits <- splits[!duplicated(splits), , drop = FALSE]
      path[new_pos[is_splitter, , drop = FALSE]] <- 1L
      path[splits] <- 0L
      new_pos <- rbind(new_pos[!is_splitter, , drop = FALSE],
                       splits)
    }
    pos <- new_pos
  }
  path
}

mark_path(puzzle_data) %>% 
  sum(na.rm = TRUE)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We use the marked path from part 1. We start in the last row, marking each end with `1`.
The we walk upwards, adding the number of paths from the row below. If there is a splitter
in the current row, we add the paths from the left and right side together. In the end, we 
return  the number of paths that lead to the source.

```{r get-solution-2}
count_timelines <- function(tachyon) {
  path <- mark_path(tachyon)
  timelines <- array(0, dim(tachyon))
  nr <- nrow(tachyon)
  nc <- ncol(tachyon)
  for (i in nr:1) {
    pos <- cbind(row = i, col = which(coalesce(path[i, ], -1L) == 0L))
    if (i == nr) {
      timelines[pos] <- 1
    } else {
      timelines[pos] <- timelines[t(t(pos) + c(1L, 0L))]
    }
    is_splitter <- coalesce(path[i, ], 0L) == 1L
    if (any(is_splitter)) {
      splitter_col <- which(is_splitter)
      left <- cbind(row = i, col = pmax(splitter_col - 1L, 1L))
      right <- cbind(row = i, col = pmin(splitter_col + 1L, nc))
      timelines[cbind(i, splitter_col)] <- timelines[left] + timelines[right]
    }
  }
  timelines[which(tachyon == "S", arr.ind = TRUE)]
}    
count_timelines(puzzle_data) %>% 
  print(digit = 12L)

```
