---
params:
   task_nr: 18
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("2021_%s",
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })      
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
	width: 100%;
}

.back {
	font-size: 2rem;
}

.day-desc {
	border: 2px dotted;
	border-radius: 6px;
	border-color: var(--blockquote-border-color);
	padding: 0 20px;
}

code em {
	font-weight:bold
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
	if (knitr::is_latex_output()) {
		sprintf("\\texttt{%s}", as.character(x))
	} else if (knitr::is_html_output()) {
		sprintf("<tt>%s</tt>", as.character(x))
	} else {
		x
	}
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(tibble)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(R6)
library(glue)
library(data.tree)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2021/day/", params$task_nr)
puzzle <- GET(base_url,
				  session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	xml_find_all("///article") %>% 
	lapply(as.character)

puzzle_data <- GET(paste0(base_url, "/input"),
						 session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	str_split("\n") %>% 
	`[[`(1L) %>% 
	head(-1L)

```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

To solve this puzzle, we have to create a proper data structure first. We rely on 
[R6](https://r6.r-lib.org/articles/Introduction.html) and use a tree inspired structure,
where each element contains references to its childs and its parent. A child can be either 
a regular number, `NULL` or another Snailfish node. With this recursive structure we can
easily traverse in either direction. Additionally, we tell nodes whether they are a left
or a right node, which is necessary to find the neighbors for exploding nodes.

This data structure is definitely an overkill and due to the involvement of environments
(the very basis of `R6`) a rather slow solution.

```{r data-structure}
Snailfish <- R6Class(
	"Snailfish",
	active = list(
		position = function(value) {
			if (missing(value)) {
				private$.pos  
			} else {
				private$.pos <- value
			}
		},
		parent = function(value) {
			if (missing(value)) {
				private$.parent
			} else {
				stopifnot(inherits(value, "Snailfish") || is.null(value))
				private$.parent <- value
			}
		}
	),
	private = list(
		.left     = NULL,
		.right    = NULL,
		.parent   = NULL,
		.depth    = 0L,
		.pos      = 0L
	)
)

Snailfish$set(
	"public",
	"is_regular",
	function(dir = allowed_dirs) {
		allowed_dirs <- c("left", "right", "both")
		dir <- match.arg(dir, allowed_dirs)
		if (dir == "both") {
			dir <- allowed_dirs[1:2]
		}
		dir <- glue(".{dir}")
		res <- !map_lgl(mget(dir, private), inherits, "Snailfish")
		names(res) <- str_remove(names(res), "^\\.")
		res
	}
)

Snailfish$set(
	"public",
	"set_depth",
	function(start = 0L) {
		private$.depth <- start
		left <- self$left()
		right <- self$right()
		if (!self$is_regular("left")) {
			left$set_depth(start + 1L)
		}
		if (!self$is_regular("right")) {
			right$set_depth(start + 1L)
		}
		invisible(self)
	}
)

Snailfish$set(
	"public",
	"set_value",
	function(value, dir = c("left", "right")) {
		dir <- match.arg(dir)
		private[[glue(".{dir}")]] <- value
		invisible(self)
	}
)

Snailfish$set(
	"public",
	"get_value",
	function(dir = c("left", "right")) {
		dir <- match.arg(dir)
		stopifnot(self$is_regular(dir))
		private[[glue(".{dir}")]]
	}
)

Snailfish$set(
	"public",
	"child",
	function(dir = c("left", "right")) {
		dir <- match.arg(dir)
		switch(dir,
				 left  = if (!self$is_regular("left")) private$.left else NULL,
				 right = if (!self$is_regular("right")) private$.right else NULL)
	}
)

Snailfish$set(
	"public",
	"left",
	function() {
		self$child("left")
	}
)

Snailfish$set(
	"public",
	"right",
	function() {
		self$child("right")
	}
)

Snailfish$set(
	"public",
	"depth",
	function() {
		private$.depth
	}
)


Snailfish$set(
	"public",
	"neighbor",
	function(dir = c("left", "right")) {
		dir <- match.arg(dir)
		el <- self
		pa <- self$parent
		while (!is.null(pa) && el$position == dir) {
			el <- pa
			pa <- pa$parent  
		}
		if (!is.null(pa)) {
			new_dir <- dir
			while (!pa$is_regular(new_dir)) {
				pa <- pa$child(new_dir)
				new_dir <- switch(dir,
										left = "right",
										right = "left")
			}
			pa <- list(node = pa, dir = new_dir)
		}
		pa
	}
)

Snailfish$set(
	"public",
	"find_next_node",
	function() {
		find_next_node <- function(node) {
			res <- NULL
			if (!is.null(node)) {
				if (node$depth() == 4L) {
					## explosion
					res <- list(node = node, event = "explode")
				} else {
					## check if split should happen at current stage
					if (node$is_regular("left") && node$get_value("left") >= 10) {
						res <- list(node = node, event = "split", dir = "left")
					}
					## ... however there could be some explosions somewhere else
					new_res <- Recall(node$left())
					if (!is.null(new_res)) {
						if (is.null(res) || 
							 (res$event != "explode" && 
							  new_res$event == "explode")) {
							res <- new_res
						} 
					}
					if (is.null(res) && node$is_regular("right") && 
						 node$get_value("right") >= 10) {
						res <- list(node = node, event = "split", dir = "right")
					}
					new_res <- Recall(node$right())
					if (!is.null(new_res)) {
						if (is.null(res) || 
							 (res$event != "explode" && 
							  new_res$event == "explode")) {
							res <- new_res
						}
					}
				}
			}
			res
		}
		find_next_node(self$root())
	}
)

Snailfish$set(
	"public",
	"root",
	function() {
		pa <- self
		while (pa$position != "root") {
			pa <- pa$parent
		}
		pa
	}
)
Snailfish$set(
	"public",
	"addr",
	function() {
		root <- self$root()
		addr <- Node$new("Snailfish")	
		create_tree <- function(node, parent) {
			if (!is.null(node)) {
				me <- parent$AddChild(obj_addr(node))
				if (!node$is_regular("left")) {
					Recall(node$left(), me)
				} else {
					me$AddChild(paste("L:", as.character(node$get_value("left"))))
				}
				if (!node$is_regular("right")) {
					Recall(node$right(), me)
				} else {
					me$AddChild(paste("R:", as.character(node$get_value("right"))))
				}
			}
		}
		create_tree(root, addr)
		print(addr)
	}
)

Snailfish$set(
	"public",
	"explode",
	function() {
		root <- self$root()
		add_value <- function(dir) {
			nb <- self$neighbor(dir)
			if (!is.null(nb)) {
				new_val <- self$get_value(dir)
				old_val <- nb$node$get_value(nb$dir)
				nb$node$set_value(old_val + new_val, nb$dir)
			}
		}
		add_value("left")
		add_value("right")
		self$parent$set_value(0L, self$position)
		self$parent <- NULL
		invisible(root)
	}
)

Snailfish$set(
	"public",
	"split",
	function(dir) {
		val <- self$get_value(dir) / 2L
		new_pair <- Snailfish$new(list(floor(val), ceiling(val)),
										  self,
										  self$depth() + 1L,
										  pos = dir)
		self$set_value(new_pair, dir)
		invisible(self$root())
	}
)

Snailfish$set(
	"public",
	"reduce",
	function(verbose = FALSE, ...) {
		root <- self$root()
		next_node <- root$find_next_node()
		while(!is.null(next_node)) {
			if (next_node$event == "explode") {
				next_node$node$explode()
				if (verbose) {
					wh <- capture.output(next_node$node)
					cat("after exploding @", wh, ":\t\t", sep = "")
					print(root, ...)
				}
			} else if (next_node$event == "split") {
				wh <- next_node$node$get_value(next_node$dir)
				next_node$node$split(next_node$dir)
				if (verbose) {
					cat("after splitting @", wh, ":\t\t", sep = "")
					print(root, ...)
				}
			}
			next_node <- root$find_next_node()
		}
		invisible(root)
	}
)

Snailfish$set(
	"public",
	"to_list",
	function() {
		to_list <- function(node) {
			res <- NULL
			if (node$is_regular("left")) {
				res <- node$get_value("left")
			} else {
				res <- Recall(node$left())
			}
			if (node$is_regular("right")) {
				res <- list(res, node$get_value("right"))
			} else {
				res <- list(res, Recall(node$right()))
			}
			res
		}
		to_list(self)
	}
)

Snailfish$set(
	"public",
	"magnitude",
	function() {
		sum <- function(node) {
			res <- 0
			if (!is.null(node)) {
				if (node$is_regular("left")) {
					res <- 3 * node$get_value("left")
				}
				if (node$is_regular("right")) {
					res <- res + 2 * node$get_value("right")
				}
				res <- res + 3 * Recall(node$left()) + 2 * Recall(node$right())
			}
			res
		}
		sum(self)
	}
)

Snailfish$set(
	"public",
	"initialize",
	function(lst, parent = NULL, depth = 0L, pos = "root", str = NULL) {
		stopifnot(is.null(str) || length(str) == 1L)
		if (!xor(missing(lst), is.null(str))) {
			warning("both 'lst' and 'str' are given - ",
					  "'lst' will be overridden by 'str'")
		}
		if (!is.null(str)) {
			lst <- str %>% 
				str_replace_all("\\[", "list(") %>% 
				str_replace_all("\\]", ")") %>% 
				parse(text = .) %>% 
				eval()
		}
		stopifnot(is.list(lst) && length(lst) == 2L)
		self$parent <- parent
		private$.depth  <- depth
		private$.pos    <- pos
		left <- lst[[1L]]
		right <- lst[[2L]]
		if (is.list(left)) {
			private$.left <- Snailfish$new(left, self, depth + 1L, "left")
		} else {
			private$.left <- left
		}
		if (is.list(right)) {
			private$.right <- Snailfish$new(right, self, depth + 1L, "right")
		} else {
			private$.right <- right
		}
	}
)

Snailfish$set(
	"public",
	"print",
	function(show_depth = FALSE, show_pos = FALSE, ...) {
		left <- if (self$is_regular("left")) 
			private$.left else
				capture.output(print(private$.left, show_depth, show_pos, ...))
		right <- if (self$is_regular("right")) 
			private$.right else
				capture.output(print(private$.right, show_depth, show_pos, ...))
		depth <- if (show_depth) glue("{private$.depth}") else ""
		pos <- if (show_pos) switch(private$.pos,
											 left = "L",
											 right = "R",
											 root = "~") else ""
		sep <- if (show_depth | show_pos) ": " else ""
		mod <- glue("{depth}{pos}{sep}")
		cat("[", mod, left, ",", right, "]", "\n", sep = "")
		invisible(self)
	}
)

`%+%` <- function(e1, e2) {
	new_sf <- list(e1$to_list(),
						e2$to_list())
	Snailfish$new(new_sf)
}
```


```{r solution-a, cache = TRUE, message = FALSE}
sf0 <- Snailfish$new(str = puzzle_data[[1L]])

res <- reduce(puzzle_data[-1L], function(.x, .y) {
	sf <- Snailfish$new(str = .y)
	message(glue("Adding Snailfish {capture.output(sf)}"))
	sum <- .x %+% sf
	sum$reduce()
}, .init = sf0)

res$magnitude()
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

To solve this puzzle we simply calculate all magnitudes by brute force.

```{r solution-b, eval = length(puzzle) > 1, message = FALSE}
cache_file <- here::here("2021", "Task-18", "pairwise_distance.rds")
force <- FALSE
if (!file.exists(cache_file) || force) {
	idx <- seq_along(puzzle_data)
	res <- matrix(NA, length(idx), length(idx))
	
	for (i in idx) {
		for (j in idx) {
			.x <- Snailfish$new(str = puzzle_data[[i]])
			.y <- Snailfish$new(str = puzzle_data[[j]])
			message(glue("Adding Snailfishs {i} and {j}: ",
							 "{capture.output(.x)} and {capture.output(.y)}"))
			sum <- .x %+% .y
			res[i, j] <- sum$reduce()$magnitude()
		}
	}
	
	saveRDS(res, cache_file)
} else {
	res <- readRDS(cache_file)
}
max(res[-seq(1, prod(dim(res)), ncol(res) + 1)])
```
