---
params:
   task_nr: 8
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
	width: 100%;
}

.back {
	font-size: 2rem;
}

.day-desc {
	border: 2px dotted;
	border-radius: 6px;
	border-color: var(--blockquote-border-color);
	padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
	if (knitr::is_latex_output()) {
		sprintf("\\texttt{%s}", as.character(x))
	} else if (knitr::is_html_output()) {
		sprintf("<tt>%s</tt>", as.character(x))
	} else {
		x
	}
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(tibble)
library(dplyr)
library(purrr)
library(tidyr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2021/day/", params$task_nr)
puzzle <- GET(base_url,
				  session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	xml_find_all("///article") %>% 
	lapply(as.character)

sort_pattern <- function(a) {
	str_split(a, "") %>% 
		map_chr(~ .x %>% 
				  	sort() %>% 
				  	paste(collapse = ""))
}

puzzle_data <- GET(paste0(base_url, "/input"),
						 session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	strsplit("\n") %>% 
	`[[`(1) %>% 
	strsplit("\\|") %>% 
	map_dfr(~strsplit(trimws(.x), "\\s+") %>% 
			  	unlist() %>% 
			  	set_names(c(paste0("I", 1:10),
			  					paste0("O", 1:4))) %>% 
			  	as.list() %>% 
			  	as_tibble()) %>% 
	mutate(across(.fns = sort_pattern))
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

This task is as easy as simply counting all the outputs whose string length is two, three,
four or seven.

```{r solution-a}
puzzle_data %>% 
	select(starts_with("O")) %>% 
	pivot_longer(everything()) %>% 
	filter(nchar(value) %in% c(2:4, 7)) %>% 
	nrow()

```



## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

Kudos to my brother with whom I discussed the problem and who helped me to get to the same result more elegantly. The final idea is as follows:

1. Digits `1`, `4`, `7` and `8` are trivially identified by the length of their code.
1. Digit `3` is the only one among the length 5 codes, where `1` is a complete subset.
1. Counting the occurrence of each letter in the codes, we will identify the lower right
segment as it is the only one which occurs in 9 out of 10 digits.
1. The one digit which does not have it is the `2`.
1. The remaining 5 letter code must be `5`.
1. `6` is the only digit among the 6 letter codes which does not fully cover `1`.
1. Likewise is `0` the only digit among the 6 letter codes which does not fully cover `5`.
1. The last digit must be `9`.

In order to solve the puzzle the input codes were sorted, such that we can easily compare
inputs and code irrespective of the order.

```{r solution-b, eval = length(puzzle) > 1}
diff_patterns <- function(...) {
	f <- function(a, b) {
		rx <- str_replace_all(b, "(.)(?!$)", "\\1|")
		str_remove_all(a, rx)
	}
	
	reduce(list(...), f)
}

solve_riddle <- function(I1, I2, I3, I4, I5, I6, I7, I8, I9, I10,
								 O1, O2, O3, O4, ...) {
	row <- c(I1, I2, I3, I4, I5, I6, I7, I8, I9, I10)
	len <- nchar(row)
	S1 <- row[len == 2]
	S4 <- row[len == 4]
	S7 <- row[len == 3]
	S8 <- row[len == 7]
	S235 <- row[len == 5]
	S069 <- row[len == 6]
	S3 <- S235[!nzchar(diff_patterns(S1, S235))]
	el_e <- row %>% 
		str_split("") %>% 
		unlist() %>% 
		table() %>% 
		which.max() %>% 
		names()
	S2 <- S235[nzchar(diff_patterns(el_e, S235))]
	S5 <- setdiff(S235, c(S2, S3))
	S6 <- S069[nzchar(diff_patterns(S1, S069))]
	S0 <- S069[nzchar(diff_patterns(S5, S069))]
	S9 <- setdiff(S069, c(S6, S0))
	sol <- c(S0 = S0, S1 = S1, S2 = S2, S3 = S3, S4 = S4, S5 = S5, S6 = S6,
				S7 = S7, S8 = S8, S9 = S9)
	stopifnot(all(c(O1, O2, O3, O4) %in% sol))
	names(sol[match(c(O1, O2, O3, O4), sol)]) %>% 
		str_remove(fixed("S")) %>% 
		paste(collapse = "") %>% 
		as.integer()
}

sum(pmap_int(puzzle_data, solve_riddle))
```


