---
params:
   task_nr: 11
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(knitr)
library(cli)
library(bit64)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  text_block %>% 
    unlist() %>% 
    str_split(" ") %>% 
    extract2(1L) %>% 
    keep(nzchar) %>% 
    as.integer()
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```
```{r get_example_data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

Instead of storing *all* the stone ids, we store only their distribution (that is how many
of eahc stone id there are). Then, in the next iteration we simply add the number of 
parents to the number of their offspring. If, for instance, we had 10 stones with id 1234, 
we will have 10 stones with id 12 and 10 stones with id 34. A samll corner case needed to 
be accounted for, namely if a stone id splits into the same ids (e.g. stone ide 77 will 
split into id 7 *twice*). In this case we need also to add the number of the parent 
*twice*.

In the end we simply have to sum up all numbers and we get the final count.

```{r get-solution-1}
dance <- function(cur_seq, n) {
  dance_step <- function(d) {
    n <- nchar(as.character(d))
    if (d == 0L) {
      as.integer64(1L)
    } else if (n %% 2 == 0) {
      c(as.integer64(substr(as.character(d), 1L, n / 2L)),
        as.integer64(substr(as.character(d), n / 2L + 1L, n)))
    } else {
      as.integer64(d * 2024L)
    }
  }
  res <- rep(as.integer64(1L), length(cur_seq)) %>% 
    setNames(as.character(cur_seq))
  for (i in 1:n) {
    cur_res <- as.integer64(NULL)
    for (d in names(res)) {
      dnc <- dance_step(as.integer64(d))
      is_same_coef <- if_else(length(dnc) == 2L &&
                                length(unique(dnc)) == 1L, as.integer64(2L), 
                              as.integer64(1L))
      cur_res[as.character(dnc)] <- coalesce(cur_res[as.character(dnc)], 
                                             as.integer64(0L)) + is_same_coef * res[d]
    }
    res <- cur_res[cur_res > 0L]
  }
  sum(res)
}

dance(puzzle_data, 25L)
```


## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We simply need to increase the iteration count to 75.

```{r get-solution-2, cache = TRUE}
dance(puzzle_data, 75L)
```
