---
params:
   task_nr: 11
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
	if (knitr::is_latex_output()) {
		sprintf("\\texttt{%s}", as.character(x))
	} else if (knitr::is_html_output()) {
		sprintf("<tt>%s</tt>", as.character(x))
	} else {
		x
	}
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(tibble)
library(dplyr)
library(purrr)
library(tidyr)
library(glue)
library(stringr)
library(R6)
library(cli)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2021/day/", params$task_nr)
puzzle <- GET(base_url,
				  session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	xml_find_all("///article") %>% 
	lapply(as.character)

## use curly braces in the last pipe to avoid that . is additionally added
## as the first argument
puzzle_data <- GET(paste0(base_url, "/input"),
						 session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	str_split("\n") %>% 
	`[[`(1) %>% 
	str_split("") %>% 
	do.call(rbind, .) %>% 
	{matrix(as.numeric(.), nrow(.), ncol(.))}

```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We use again an [R6 Class](https://r6.r-lib.org/articles/Introduction.html) for 
its beauty rather than its necessity. To avoid infinite loops, we set octopuses
which flashed already to `Inf`. Adding one to `Inf` let the value remain at 
`Inf` and we can conveniently count only new flashes. At the end of a step, 
we set those cells back to `0`. Some syntactic sugar like the `print` member, or
allowing to set the flash threshold to an number other than `10` was added, for 
the sake of testing.

```{r solution-a}
Octopus <- R6Class(
	"Octopus",
	public = list(
		initialize = function(start, flash_signal = 10L) {
			stopifnot(is.matrix(start))
			private$.state <- start
			private$.flashes_at <- flash_signal
		},
		print = function() {
			str <- private$.state
			str[str >= private$.flashes_at] <- col_red("#")
			str <- apply(str, 1, paste, collapse = "") %>% 
				paste(collapse = "\n")
			cat(cli::col_grey(strrep("-", ncol(private$.state)), "\n"))
			cat(str, "\n")
			cat(cli::col_grey(strrep("-", ncol(private$.state)), "\n"))
			invisible(self)
		}
	),
	private = list(
		.state = NULL,
		.round = 0L,
		.nr_flashes = 0L,
		.flashes_at = 10L
	)
)

Octopus$set("private",
				".increase_neighbors",
				function() {
					flash <- is.finite(private$.state) &
						private$.state >= private$.flashes_at
					idx <- which(flash, arr.ind = TRUE) 
					amt <- expand.grid(row = -1:1,
											 col = -1:1) %>% 
						apply(1, `+`, t(idx), simplify = FALSE)  %>% 
						do.call(cbind, .) %>%
						t() %>% 
						as_tibble() %>% 
						filter(between(row, 1L, nrow(private$.state)),
								 between(col, 1L, ncol(private$.state))) %>% 
						count(row, col) %>% 
						as.matrix()
					private$.state[amt[, 1:2]] <- private$.state[amt[, 1:2]] + 
						amt[, 3]
					private$.state[flash] <- Inf
					sum(flash)
				})

Octopus$set("private",
				".add",
				function(verbose = FALSE) {
					if (verbose) {
						print(self)
					}
					private$.state <- private$.state + 1
					while ((flashes <- private$.increase_neighbors()) > 0) {
						if (verbose) {
							print(self)
						}
						private$.nr_flashes <- private$.nr_flashes + 
							flashes
					}
					private$.state[is.infinite(private$.state)] <- 0L
					private$.round <- private$.round + 1
					invisible(self)
				})

Octopus$set("public",
				"simulate",
				function(n, verbose = FALSE) {
					for (i in seq_len(n)) {
						private$.add(verbose)
					}
					private$.nr_flashes
				})
octo <- Octopus$new(puzzle_data)
octo$simulate(100)

```



## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

To solve this task, we simply add another method to our `Octopus` class, which
calls `.add` until we reach a state, where all the octopuses flashes. *N.B.* 
This is the case when all animals have an energy level of **0**, b/c this means
they flashed all in this round.

```{r solution-b, eval = length(puzzle) > 1}
Octopus$set("public",
				"find_climax",
				function(verbose = FALSE) {
					while (sum(private$.state == 0L) < 
							 prod(dim(private$.state))) {
						private$.add(verbose)
					}
					private$.round
				})
new_octo <- Octopus$new(puzzle_data)
new_octo$find_climax()
```


