---
params:
   task_nr: 17
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2024"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(knitr)
library(cli)
library(bit64)
library(igraph)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
     str_split("\n") %>% 
      extract2(1L)
  } 
  list(
    reg = c(A = str_subset(text_block, "A:") %>% 
              str_extract("\\d+") %>% 
              as.integer64(),
            B = str_subset(text_block, "B:") %>% 
              str_extract("\\d+") %>% 
              as.integer64(),
            C = str_subset(text_block, "C:") %>% 
              str_extract("\\d+") %>% 
              as.integer64()
    ),
    code = str_subset(text_block, "Program:") %>% 
      str_extract("(\\d+,?)+") %>% 
      str_split(",") %>% 
      extract2(1L) %>% 
      as.integer()
  )
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r get_example_data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

```{r get-solution-1}
bXor64 <- function(a, b) {
  tobs <- . %>% 
    as.integer64() %>% 
    as.bitstring() %>% 
    str_split("") %>% 
    extract2(1L) %>% 
    as.integer()
  a <- tobs(a)
  b <- tobs(b)
  bs <- paste((a + b) %% 2, collapse = "")
  class(bs) <- "bitstring"
  as.integer64(bs)
}

calculator <- function(data) {
  instructions <- data$code
  reg <- data$reg
  combo <- function(op) {
    stopifnot(between(op, 0L, 6L))
    if (op <= 3L) {
      as.integer64(op)
    } else {
      as.integer64(reg[op - 3L])
    }
  }
  
  div <- function(op, reg_sel) {
    reg_sel <- match.arg(reg_sel, names(reg))
    num <- reg["A"]
    den <- as.integer64(2 ^ combo(op))
    reg[reg_sel] <<- num %/% den 
  }
  
  adv <- function(op) {
    div(op, "A")
  }
  
  bxl <- function(op) {
    reg["B"] <<- bXor64(reg["B"], op)
  }
  
  bst <- function(op) {
    reg["B"]<<- combo(op) %% 8L
  }
  
  jnz <- function(op) {
    if (reg["A"] > 0L) {
      ptr <<- op
    }
  }
  
  bxc <- function(op) {
    reg["B"] <<- bXor64(reg["B"], reg["C"])
  }
  
  out <- function(op) {
    std_out <<- c(std_out, 
                  as.character(combo(op) %% 8) %>% 
                    str_split("") %>% 
                    extract2(1L) %>% 
                    paste(collapse = ","))
  }
  
  bdv <- function(op) {
    div(op, "B")
  }
  
  cdv <- function(op) {
    div(op, "C")
  }
  
  process <- function() {
    
    opcode <- instructions[ptr + 1L]
    operand <- instructions[ptr + 2L]
    switch(opcode + 1L,
           adv(operand),
           bxl(operand),
           bst(operand),
           jnz(operand),
           bxc(operand),
           out(operand),
           bdv(operand),
           cdv(operand))
    if (opcode != 3 || reg["A"] == as.integer64(0)) {
      ptr <<- ptr + 2L
    }
    invisible(ptr)
  }
  
  ptr <- 0L
  std_out <- NULL
  while(ptr < length(instructions)) {
    process()  
  }
  std_out %>% 
    as.integer()
}

sol1 <- calculator(puzzle_data)

sol1 %>% 
  paste(collapse = ",")
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

First we analyze the program. What it does is:

1. Calculating `A` modulo 8 and store it in `B`: `B <- A %% 8L`
1. Calculating `B` XOR 5 and store it in `B`: `B <- bXor64(B, 5L)`
1. Calculating `A` divided by 2 to the power of `B` and store it in `C`: 
`C <- A %/% (2 ^ B)`. This is effectively the same as bit shifting `A` by `B` bits to the
right (`B` is an integer between 0 and 7 at this point which in turn depends on `A`).
1. Calculating `B` XOR 6 and store it in `B`: `B <- bXor64(B, 6L)`
1. Calculating `B` XOR `C` and store it in `B`: `B <- bXor64(B, 6L)`. As `B` has only 3 
bits at this stage, this effectively changes only the last 3 bits of `C` and let the rest
remain unchanged.
1. Output the last 3 bits of `B`.
1. Divide `A` by 8: `A <- A %/% 8` (i.e. bit shift it 3 bits to the right).
1. Repeat until `A == 0`.

We can now implement this algorithm straight away:

```{r retro-algo}
get_next_digit <- function(A) {
  A <- as.integer64(A)
  B <- bXor64(A %% 8L, 5L)
  C <- A %/% (2 ^ B)
  B <- bXor64(bXor64(B, 6L), C)
  as.integer(B %% 8)
}

retro <- function(A) {
  A <- as.integer64(A)
  res <- integer(0L)
  repeat {
    res <- c(res, get_next_digit(A))
    A <- A %/% 8
    if (A == 0) {
      break
    }
  }
  res
}

all.equal(retro(puzzle_data$reg["A"]), sol1)
```

With this decomposed algorithm, we can generate a backtracking algorithm as follows:

1. Let $I$ denote the original instructions. Set $k$ to the length of $I$: $k=|I|$.
1. Calculate $A_k=I_k\oplus 6 \oplus 5$.
1. Generate candidates $A_{k-1}^{l} = 8\cdot A_k + l,0\leq l \leq 7$. These are all values
which lead to $A_k$ after one iteration.
1. For all $0\leq l \leq 7$ check, if $A_{k-1}^l$ produces $I_{k-1}$. Prune invalid 
candidates, set $k = k - 1$ and repeat with the valid candidates $A_{k-1}^{l^*}$.
1. Continue until we got $A_1^{l^*}$ and eventually return $\min_{l^*}A_1^{l^*}$.

```{r get-solution-2}
find_starting_value <- function(data, limit_n = length(data$code)){
  goal <- tail(data$code, limit_n)
  start <- tail(goal, 1L) %>% 
    bXor64(6L) %>% 
    bXor64(5L)
  backtrack <- function(a, k) {
    res <- NULL
    if (k == 0L) {
      res <-  as.character(a)
    } else {
      target <- goal[k]
      candidates <- (8L * a + 0:7) %>% 
        keep(~ get_next_digit(.x) == target)
      res <- map(candidates, ~ backtrack(.x, k - 1L)) %>% 
        unlist()
    }
    res
  }
  backtrack(start, length(goal) - 1L) %>% 
    as.integer64(res) %>%
    min()
}

find_starting_value(puzzle_data)
```