---
params:
   task_nr: 12
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
	width: 100%;
}

.back {
	font-size: 2rem;
}

.day-desc {
	border: 2px dotted;
	border-radius: 6px;
	border-color: var(--blockquote-border-color);
	padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
	if (knitr::is_latex_output()) {
		sprintf("\\texttt{%s}", as.character(x))
	} else if (knitr::is_html_output()) {
		sprintf("<tt>%s</tt>", as.character(x))
	} else {
		x
	}
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(tibble)
library(dplyr)
library(purrr)
library(igraph)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2021/day/", params$task_nr)
puzzle <- GET(base_url,
				  session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	xml_find_all("///article") %>% 
	lapply(as.character)

## use curly braces in the last pipe to avoid that . is additionally added
## as the first argument
puzzle_data <- GET(paste0(base_url, "/input"),
						 session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	str_split("\n") %>% 
	`[[`(1) %>% 
	head(-1L) %>% 
	lapply(str2lang) %>% 
	do.call(graph_from_literal, .)

```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We use the library `igraph` to deal with the network. First of all, lets plot the graph:

```{r plot-graph, fig.cap = "Cave Network"}
V(puzzle_data)$single <- str_detect(V(puzzle_data)$name, "^[:lower:]+$")
V(puzzle_data)$color <- "#F9DB6D"
V(puzzle_data)[c("start", "end")]$color <- c("#464D77", "#36827F")
V(puzzle_data)[c("start", "end")]$label.color <- "white"

set.seed(1) ## to keep plot the same
op <- par(bg = "transparent")
plot(puzzle_data, layout = layout_with_lgl(puzzle_data, root = "start"),
	  vertex.size = 25)
```

Then we can define an easy recursive function which walks the graph according 
to the following rules:

1. In each iteration we maintain a list of *visitable* neighbors. Those are the neighbors
which are not yet visited in the current walk, unless it is a *big* cave, which can - by
definition - visited nore than once.
1. If we reach `end` we increase the counter by 1 and the function ends.
1. Otherwise we call the function recursively again with the new neighbor as next node 
and where we add the current node to the path of visited nodes.

```{r solution-a, cache = FALSE}
walk_graph <- function(G = puzzle_data) {
	do_walk <- function(node, path) {
		path <- c(node, path)
		nbs <- neighbors(G, node)
		free_nbs <- difference(nbs, path[single])
		if (node$name == "end") {
			## we reached "end"
			res <- 1
		} else {
			res <- 0
			for (idx in free_nbs) {
				res <- res + Recall(V(G)[idx], path)
			}
		}
		res
	}
	do_walk(V(G)["start"], NULL)
}

walk_graph()

```



## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For the second part we need to relax the condition in so far that one small caves can 
be visited at most twice. To do so, we check if there is already a small cave, which
was visited twice and only in this condition we remove small caves from the list of
visitable caves.

*N.B.* The following code includes 2 versions of the same (and part one is largely a clone
of solution 1): a version with the `igraph` framework and one with a much simpler data
structure. It turned out that the `igraph` algo, though correct runs by far too long. My
guess is that this due to copy on change semantics and the fact that an `igraph` object is
rather heavy. Thus, I include a simpler - yet less verbose - version relying on basic data
structures.

```{r solution-b, eval = length(puzzle) > 1, cache = FALSE}
walk_graph2 <- function(G = puzzle_data, use_igraph = FALSE) {
	do_walk_igraph <- function(node, path) {
		path <- c(node, path)
		nbs <- difference(neighbors(G, node), V(G)["start"])
		if (any(duplicated(path[single]))) {
			free_nbs <- difference(nbs, path[single])
		} else {
			free_nbs <- nbs
		}
		
		if (node$name == "end") {
			## we reached "end"
			res <- 1
		} else {
			res <- 0
			for (idx in free_nbs) {
				res <- res + Recall(V(G)[idx], path)
			}
		}
		res
	}
	
	do_walk_basic <- function(node, path) {
		path <- c(node, path)
		nbs <- setdiff(G[[node]], "start")
		small_caves <- path[str_detect(path, "^[:lower:]+$")]
		if (any(duplicated(small_caves))) {
			free_nbs <- setdiff(nbs, small_caves)
		} else {
			free_nbs <- nbs
		}
		
		if (node == "end") {
			## we reached "end"
			res <- 1
		} else {
			res <- 0
			for (nb in free_nbs) {
				res <- res + Recall(nb, path)
			}
		}
		res
	}
	if (use_igraph) {
		do_walk_igraph(V(G)["start"], NULL)
	} else {
		G <- as_data_frame(G) %>% 
			### add edges in both directions
			bind_rows((.) %>% rename(from = to, to = from)) %>% 
			group_by(from) %>% 
			group_map(~ list(.x$to) %>% 
						 	set_names(.y %>% pull(from))) %>% 
			flatten()
		do_walk_basic("start", NULL)
	}
}

walk_graph2()
```



