---
params:
   task_nr: 12
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(knitr)
library(cli)
library(bit64)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) > 1L) {
    text_block <- paste(text_block, collapse = "\n")
  }
  text_block %>% 
    unlist() %>% 
    str_split("\n") %>% 
    extract2(1L) %>% 
    str_split("") %>% 
    do.call(rbind, .)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```
```{r get_example_data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We use again a backtracking algorithn to create all regions. In the backtrackng algorithm, 
we assign the number of non similar neighbors treating nodes sitting on the edge of
the matrix as having non matching neighbors outside the matrix boundaries.

Then the solution is to multiply the number of nodes in each region with the sum of non
matching neighbors to get the final price.

```{r get-solution-1}
get_regions <- function(garden_map, skip_border = FALSE) {
  dd <- dim(garden_map)
  visited <- matrix(FALSE, dd[1L], dd[2L])
  visited[is.na(garden_map)] <- TRUE
  ids <- unique(na.omit(c(garden_map)))
  regions <- vector("list", length(ids))
  names(regions) <- ids
  dirs <- rbind(
    cbind(-1L, 0L),
    cbind(0L, 1L),
    cbind(1L, 0L),
    cbind(0L, -1L)
  ) %>% 
    set_rownames(c("up", "right", "down", "left"))
  
  parse_region <- function(current_pos, current_region_id) {
    if (any(current_pos < c(1L, 1L)) ||
        any(current_pos > dd)) {
      ## position out of bounds
      current_region <- NULL
    } else {
      ## position in bounds
      if (visited[current_pos] || is.na(garden_map[current_pos]) ||
          garden_map[current_pos] != current_region_id) {
        ## wrong region
        current_region <- NULL
      } else {
        visited[current_pos] <<- TRUE
        neighbors <- c(current_pos) + t(dirs)
        border_touchpoints <- rowSums(t(neighbors < c(1L, 1L) | neighbors > dd))
        neighbors_idx <- t(neighbors)[border_touchpoints == 0, , drop = FALSE]
        foreign_touchpoints <- garden_map[neighbors_idx] != current_region_id 
        current_region <- cbind(current_pos, 
                                border = sum(border_touchpoints) + 
                                  sum(foreign_touchpoints))
        for (dir_idx in seq(1L, nrow(dirs))) {
          dir <- dirs[dir_idx, , drop = FALSE]
          current_region <- rbind(current_region, 
                                  Recall(current_pos + dir, current_region_id))
        }
      }
    }
    current_region
  }
  edge_id <- 0
  while (!all(visited)) {
    start_pos <- which(!visited, arr.ind = TRUE)[1, , drop = FALSE]
    current_id <- garden_map[start_pos]
    current_region <- parse_region(start_pos, current_id)
    if (skip_border) {
      current_region <- current_region[, -3L, drop = FALSE]
    }
    regions[[current_id]] <- c(regions[[current_id]], list(current_region))
  }
  regions
}

get_fence_price <- function(regions) {
  regions %>% 
    flatten() %>% 
    map_int(~ nrow(.x) * sum(.x[, "border"])) %>% 
    sum()
}

garden_regions <- get_regions(puzzle_data)

get_fence_price(garden_regions)
```


## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

To compute the side length of each region, we implement the following algorithm:
1. Get the **convex hull** of the region of interest (that is the rectangular using min 
and max of row and ocl indices).
1. This rectangular has 4 corners, hence 4 sides. Hence add 4 to the overall result.
1. Substract the original region from the convey hull. If this yields the empty set, we 
are done.
1. Otherwise, get all sub regions of this difference and calulcate again the convex hull 
of those.
1. For each convex hull, if we have at least one corner in common, we add 2 to the result 
otherwise 4, to avoid counting the same corner twice.
1. Repeat until each region is iterated over.



*Notes.* The algorithm is not overly performant, which is due to the fact that the complex 
hull is explicitely calculated i.e. each field is generated) multiple of times. The region
finding algorithm was re-implemented using matrix indices this time instead of the matrix 
itself, as we could then re-use the regiosn generated from part 1 (which returned matrix
indices).

```{r get-solution-2, cache = TRUE}
calculate_discount <- function(regions) {
  get_convex_hull <- function(region) {
    apply(region, 2, \(x) seq(min(x), max(x), 1L), simplify = FALSE) %>% 
      do.call(expand.grid, .) %>% 
      as.matrix()
  }
  
  is_convex <- function(region, convex_hull = get_convex_hull(region)) {
    region <- region[order(region[, 2L], region[, 1L]), , drop = FALSE]
    identical(c(region), c(convex_hull))
  }
  
  get_corners <- function(region) {
    apply(region, 2, \(x) range(x), simplify = FALSE) %>% 
      do.call(expand.grid, .) %>% 
      as.matrix()
  }
  
  contains <- function(region1, region2) {
    apply(region2, 1, function(coord) {
      any(colSums(coord == t(region1)) == 2L)
    })
  }
  
  substract <- function(region1, region2) {
    r1 <- as_tibble(region1)
    r2 <- as_tibble(region2)
    r1 %>% 
      anti_join(r2, c("row", "col")) %>% 
      as.matrix()
  }
  
  get_sub_regions <- function(region) {
    dirs <- rbind(
      cbind(-1L, 0L),
      cbind(0L, 1L),
      cbind(1L, 0L),
      cbind(0L, -1L)
    ) %>% 
      set_rownames(c("up", "right", "down", "left"))
    parse_region <- function(current_pos) {
      res <- NULL
      current_pos <- current_pos[, 1:2, drop = FALSE]
      idx <- contains(current_pos, current_region[, 1:2, drop = FALSE]) 
      if (!any(idx)) {
        res <- NULL
      } else {
        idx <- which(idx)
        if (current_region[idx, 3]) {
          ## node already visited
          res <- NULL
        } else {
          current_region[idx, 3] <<- TRUE
          res <- current_pos
          for (dir_idx in seq(1L, nrow(dirs))) {
            dir <- dirs[dir_idx, , drop = FALSE]
            res <- rbind(res,
                         Recall(current_pos + dir))
          }
        }
      }
      res
    }
    current_region <- cbind(region[, 1:2, drop = FALSE], visited = FALSE)
    regions <- list()
    while (any(!current_region[, "visited"])) {
      start_pos <- current_region[!current_region[, "visited"], , drop = FALSE][1, , drop = FALSE]
      regions <- c(regions, list(parse_region(start_pos)))
    }
    regions
  }
  
  calculate_borders <- function(region, prev_region) {
    region <- region[, 1:2, drop = FALSE] 
    if (nrow(region) == 0L) {
      res <- 0L
    } else {
      res <- 4L
      convex_hull <- get_convex_hull(region)
      if (!is.null(prev_region)) {
        corners <- get_corners(convex_hull)
        prev_corners <- get_corners(prev_region)
        if (any(contains(prev_corners, corners))) {
          res <- 2L
        }
      }
      if (!is_convex(region, convex_hull)) {
        diff_regions <- substract(convex_hull, region)
        sub_regions <- get_sub_regions(diff_regions)
        sub_border_regions <- vapply(sub_regions, \(r) calculate_borders(r, convex_hull), 
                                     integer(1L)) %>% 
          sum()
        res <- res + sub_border_regions
      }
    }
    res
  }
  map_int(flatten(regions), \(region) calculate_borders(region, NULL) * nrow(region)) %>% 
    sum()
}

calculate_discount(garden_regions)

```
