---
params:
   task_nr: 13
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
	width: 100%;
}

.back {
	font-size: 2rem;
}

.day-desc {
	border: 2px dotted;
	border-radius: 6px;
	border-color: var(--blockquote-border-color);
	padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
	if (knitr::is_latex_output()) {
		sprintf("\\texttt{%s}", as.character(x))
	} else if (knitr::is_html_output()) {
		sprintf("<tt>%s</tt>", as.character(x))
	} else {
		x
	}
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(tibble)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2021/day/", params$task_nr)
puzzle <- GET(base_url,
				  session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	xml_find_all("///article") %>% 
	lapply(as.character)

## use curly braces in the last pipe to avoid that . is additionally added
## as the first argument
input <- GET(paste0(base_url, "/input"),
				 session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	str_split("\n")

puzzle_data <- local({
	mat <- input[[1L]] %>% 
		str_detect("^[:digit:]+,[:digit:]+$") %>% 
		`[`(input[[1L]], .) %>% 
		str_split(",") %>% 
		do.call(rbind, .) %>% 
		{matrix(as.integer(.), ncol = 2L)} + 1L
	dots <- matrix(FALSE, max(mat[, 2L]), max(mat[, 1L]))
	dots[mat[, 2:1]] <- TRUE
	
	folds <- input[[1]] %>% 
		str_detect("^fold") %>% 
		`[`(input[[1L]], .) %>% 
		str_extract_all("[:digit:]+|[xy]")
	
	folds_df <- folds %>% 
		do.call(rbind, .) %>% 
		`colnames<-`(c("dir", "pos")) %>% 
		as_tibble() %>% 
		mutate(pos = as.integer(pos) + 1L)
	
	list(dots = dots, 
		  folds = folds %>% 
		  	map(~ list(dir = .x[[1L]],
		  				  pos = as.integer(.x[[2L]]) + 1L)), 
		  folds_df = folds_df)
})

```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

To solve this puzzle we work with logicals. Then the task boils down to splitting the 
matrix into pieces as indicated by the fold positions and mirroring the lower (right) 
part. The we can use an *OR*  operation on the two parts.

The folds are anyways always at the middle position as the following check shows:

```{r check-folds}
puzzle_data$folds_df %>% 
	count(dir) %>% 
	group_by(dir) %>% 
	group_map(function(dat, grp) {
		dd <- set_names(dim(puzzle_data$dots), c("y", "x"))
		xy <- grp %>% pull(dir)
		res <- Reduce(\(.x, .y) (.x - 1L) / 2L,
							seq_len(dat %>% pull(n)),
							init = dd[xy],
							accumulate = TRUE)
		puzzle_data$folds_df %>% 
			filter(dir == xy) %>% 
			mutate(calc = tail(res, -1L))
	})
```


```{r solution-a, cache = FALSE}
fold_paper <- function(dots, fold) {
	dir <- fold$dir
	pos <- fold$pos
	if (dir == "x") {
		p1 <- dots[, seq(1L, pos - 1L), drop = FALSE]
		p2 <- dots[, seq(ncol(dots), pos + 1L, -1L), drop = FALSE]
	} else if (dir == "y") {
		p1 <- dots[seq(1, pos - 1L), , drop = FALSE]
		p2 <- dots[seq(nrow(dots), pos + 1L, -1L), , drop = FALSE]
	}
	stopifnot(identical(dim(p1), dim(p2)))
	res <- p1 | p2
	class(res) <- "code_paper"
	res
}

print.code_paper <- function(x, ...) {
	res <- matrix(".", nrow(x), ncol(x))
	res[x] <- "#"
	res <- apply(res, 1, paste, collapse = "") %>% 
		paste(collapse = "\n")
	cat(res, "\n")
	invisible(x)
}

sum(fold_paper(puzzle_data$dots, puzzle_data$folds[[1]]))
```



## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

```{r solution-b, eval = length(puzzle) > 1, comment = NA}
print(reduce(puzzle_data$folds, fold_paper, .init = puzzle_data$dots))
```



