---
params:
   task_nr: 24
   year: 2018
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}

```

```{js cpp-highlighter, file = "Prism_CPP.js", echo = FALSE}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(purrr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  parse_army <- function(block) {
    stats <- str_extract_all(block, "\\d+") %>% 
      do.call(rbind, .) %>% 
      set_colnames(c("units", "hp", "dmg", "initiative"))
    storage.mode(stats) <- "integer"
    attack <- str_extract(block, "\\S+ damage") %>% 
      str_remove(" damage")
    immune <- str_extract(block, "immune to [^;)]+") %>% 
      str_remove("immune to ") %>% 
      str_split(", ")
    weak <- str_extract(block, "weak to [^;)]+") %>% 
      str_remove("weak to ") %>% 
      str_split(", ")
    as_tibble(stats) %>% 
      mutate(attack = attack,
             weak = weak,
             immune = immune)
  }
  army_blocks_start <- str_which(text_block, "Immune|Infection") + 1L
  army_blocks_end <- c(army_blocks_start[[2L]] - 2L, length(text_block))
  army_label <- c("immune", "infection")
  pmap(list(army_blocks_start, army_blocks_end, army_label),
        \(i, j, army) parse_army(text_block[i:j]) %>% 
         mutate(army = army, .before = 1L)) %>% 
    list_rbind() %>% 
    mutate(id = paste0(str_sub(army, 1L, 2L), 1:n()), .by = army,
           .before = 1L) %>% 
    mutate(army_id = if_else(army == "infection", 0L, 1L),
           .before = army)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- tibble(
  id = c("im1", "im2", "in1", "in2"), 
  army_id = rep(1:0, each = 2L),
  army = c("immune", "immune", "infection", "infection"), 
  units = c(17L, 989L, 801L, 4485L), 
  hp = c(5390L, 1274L, 4706L, 2961L), 
  dmg = c(4507L, 25L, 116L, 12L), 
  initiative = c(2L, 3L, 1L, 4L), 
  attack = c("fire", "slashing", "bludgeoning", "slashing"),
  weak = list(c("radiation", "bludgeoning"), 
              c("bludgeoning", "slashing"), 
              "radiation", 
              c("fire", "cold")), 
  immune = list(NA_character_, "fire", NA_character_, "radiation")
)
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

For the first part, we start with a (slowish) un-optimized R version, which uses
inner joins and sorting to find the relevant opponents. Event though the number of
armies is small, the inherent copy on modify semantic of R slows the algorithm down.

```{r get-solution-1, cache = TRUE}
choose_opponents <- function(armies) {
  armies <- armies %>% 
    mutate(power = units * dmg) %>% 
    arrange(desc(power), desc(initiative)) %>% 
    mutate(opponent_id = (army_id + 1L) %% 2L,
           id = factor(id, id)) 
  armies %>% 
    inner_join(armies,
               join_by(army_id == opponent_id),
               suffix = c("_attacker", "_defender"),
               relationship = "many-to-many") %>% 
    mutate(
      is_weak = map2_lgl(attack_attacker, weak_defender, ~ .x %in% .y),
      is_immune = map2_lgl(attack_attacker, immune_defender, ~.x %in% .y),
      damage_dealt = power_attacker * case_when(is_weak ~ 2L,
                                                is_immune ~ 0L,
                                                .default = 1L)) %>% 
    group_by(id_attacker) %>%
    arrange(desc(damage_dealt), desc(power_defender), desc(initiative_defender),
            .by_group = TRUE) %>%
    group_split() %>%
    reduce(.init = tibble(id_defender = character(0L)),
           .f = function(acc, group) {
             available <- group %>%
               filter(!id_defender %in% acc$id_defender)
             if (nrow(available) > 0L) {
               acc <- bind_rows(acc, available %>%
                                  slice(1L))
             }
             acc
           }) %>%
    arrange(desc(initiative_attacker)) %>% 
    mutate(
      id_attacker = as.character(id_attacker),
      id_defender,
      .keep = "none"
    )
}

attack_opponents <- function(attack_order, armies) {
  for (i in seq_len(nrow(attack_order))) {
    battle <- attack_order %>% 
      slice(i) %>% 
      as.list()
    attacker <- armies %>% 
        filter(id == battle$id_attacker)
    defender <- armies %>% 
      filter(id == battle$id_defender)
    if (attacker$units > 0L) {
      power <- attacker$units * attacker$dmg * 
        case_when(attacker$attack %in% defender$weak[[1L]] ~ 2L,
                  attacker$attack %in% defender$immune[[1L]] ~ 0L, ## should never happen
                  .default = 1L)
      armies <- armies %>% 
        mutate(units = if_else(id == battle$id_defender, 
                               units - power %/% hp, 
                               units))
    }
  }
  armies %>% 
    filter(units > 0L)
}

fight <- function(armies = puzzle_data) {
  
  is_done <- function(armies) {
    armies %>% 
      pull(army) %>% 
      unique() %>% 
      length() %>% 
      equals(1L)
  }
  while (!is_done(armies)) {
    attack_order <- choose_opponents(armies)
    armies <- attack_opponents(attack_order, armies)
  }
  armies %>% 
    summarize(remaining_force = sum(units)) %>% 
    pull(remaining_force)
}
fight(puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For the second part, we re-implment the solution in C++ for a substantial speed gain.
Then, we can use a binary search to find the minimal boost. We also included a stalemate
check in this version, as it could happen that the remaining groups are still alive but to
weak to kill a single opponent.

```{Rcpp cpp-algorithm, file = "immune.cpp", cache = TRUE}
```

```{r get-solution-2, cache = TRUE}
armies <- puzzle_data %>% 
  rowwise() %>% 
  group_split() %>% 
  map(function(row) {
    row <- as.list(row)
    if (is.na(row$weak)) {
      row$weak <- ""
    } else {
      row$weak <- paste(unlist(row$weak), collapse = ",")
    }
    if (is.na(row$immune)) {
      row$immune <- ""
    } else {
      row$immune <- paste(unlist(row$immune), collapse = ",")
    }
    row
  })

find_minimum_booster(armies[1:10], armies[11:20])
```
