---
params:
   task_nr: 7
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
	width: 100%;
}

.back {
	font-size: 2rem;
}

.day-desc {
	border: 2px dotted;
	border-radius: 6px;
	border-color: var(--blockquote-border-color);
	padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
	if (knitr::is_latex_output()) {
		sprintf("\\texttt{%s}", as.character(x))
	} else if (knitr::is_html_output()) {
		sprintf("<tt>%s</tt>", as.character(x))
	} else {
		x
	}
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(tibble)
library(dplyr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2021/day/", params$task_nr)
puzzle <- GET(base_url,
				  session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	xml_find_all("///article") %>% 
	lapply(as.character)

puzzle_data <- GET(paste0(base_url, "/input"),
						 session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	strsplit(",") %>% 
	`[[`(1) %>% 
	as.integer()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We simply need to calculate the 
[Manhattan distance](https://en.wikipedia.org/wiki/Taxicab_geometry) between all the 
points and each potential point in `range(puzzle_data)`. Points outside this range can be 
always improved by setting them to the nearest maximum. We will use `outer` to get all
distances and then apply `rowSum` to get the distance for each potential candidate. The 
point with minimum `rowSum` is the best option.

```{r solution-a}
cand <- puzzle_data %>% 
	range() %>% 
	set_names(c("from", "to")) %>% 
	as.list() %>% 
	do.call(seq, .)

mh_dist <- outer(cand, puzzle_data, `-`) %>% 
	abs()
min(rowSums(mh_dist))
```



## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

In this variant we have to calculate the distance sligthly different. Instead of taking 
the absolute value of the difference between the two points, we calculate the sum of all 
values from 1 to this distance:

```{r solution-b, eval = length(puzzle) > 1}
adj_dist <- outer(cand, puzzle_data, 
						FUN = Vectorize(function(x, y) sum(seq(0, abs(y - x)))))
min(rowSums(adj_dist))
```


