---
params:
   task_nr: 10
   year: 2016
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2016"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(igraph)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  values <- text_block %>% 
    str_subset(fixed("value")) %>% 
    str_extract_all("\\d+") %>% 
    do.call(rbind, .) %>% 
    set_colnames(c("value", "bot")) %>% 
    as_tibble() %>% 
    mutate(value = as.integer(value),
           bot = paste0("B", bot)) %>% 
    select(bot, value)
  bots <- text_block %>% 
    str_subset("^bot") %>% 
    str_extract_all("bot \\d+|output \\d+") %>% 
    do.call(rbind, .) %>% 
    set_colnames(c("from", "low", "high")) %>% 
    as_tibble() %>% 
    mutate(across(everything(), 
                  ~ str_replace_all(.x, c("bot " = "B", "output " = "O"))
    ))
  list(values = values, bots = bots)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

We construct a graph, where each node represents a bot or an output. Nodes are connected, 
iif bot `A` passes a chip to bot / output `B`. We assign values according to the original 
set up and then iterate through the graph until one bot received the 2 coins of interest. 

### Solution

```{r get-solution-1}
construct_bot_network <- function(bots, values) {
  nodes <- values %>% 
    pull(bot) %>% 
    c(bots %>% 
        unlist()) %>% 
    unique() %>% 
    sort()
  G <- make_empty_graph(length(nodes))
  vals <- values %>% 
    summarize(values = list(sort(value)), .by = bot)
  vals <- vals %>% 
    pull(values) %>% 
    set_names(vals %>% 
                pull(bot))
  V(G)$name <- nodes
  V(G)$type <- if_else(str_detect(V(G)$name, "^B"), "bot", "output")
  V(G)$value <- rep(list(NULL), vcount(G))
  V(G)$value[V(G)$name %in% names(vals)] <- vals[V(G)$name[V(G)$name %in% names(vals)]]
  edges <- bots %>% 
    rowwise() %>% 
    mutate(path = list(c(from, low, from, high))) %>% 
    pull(path) %>% 
    unlist()
  G <- add_edges(G, edges, dir = rep(c("low", "high"), nrow(bots)))
  G
}

pass_microchips <- function(G, cond) {
  while (!cond$is_done(G)) {
    has_pairs <- lengths(V(G)$value) == 2L & V(G)$type == "bot"
    srcs <- V(G)$name[has_pairs]
    for (src in srcs) {
      vals <- V(G)[src]$value[[1L]]
      V(G)[src]$value <- list(NULL)
      edges <- incident(G, src, "out")
      ep <- ends(G, edges)[, 2]
      low <- ep[edges$dir == "low"]
      high <- ep[edges$dir == "high"]
      V(G)[low]$value <- list(sort(c(V(G)[low]$value[[1L]], vals[1L])))
      V(G)[high]$value <- list(sort(c(V(G)[high]$value[[1L]], vals[2L])))
    }
  }
  cond$extractor(G)
}

bot_has_chips <- function(target) {
  list(
    is_done = function(G) {
      list(target) %in% V(G)$value
    },
    extractor = function(G) {
      V(G)$name[V(G)$value %in% list(target)] %>% 
        str_remove("^.") %>% 
        as.integer()
    }
  )
}

G <- construct_bot_network(puzzle_data$bots,
                           puzzle_data$values)
pass_microchips(G, bot_has_chips(c(17L, 61L)))
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We use the same algorithm, but this time we do not stop before the output nodes have
received their coins.

```{r get-solution-2}
output_has_chips <- function(bots) {
  list(
    is_done = function(G) {
      all(lengths(V(G)[bots]$value) > 0L)
    },
    extractor = function(G) {
      V(G)[bots]$value %>% 
        unlist() %>% 
        prod()
    }
  )
}

pass_microchips(G, output_has_chips(paste0("O", 0:2)))
```
