---
params:
   task_nr: 24
   year: 2022
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(numbers)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    str_split("") %>% 
    do.call(rbind, .)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We implement the solution using a BFS search, where the current state corresponds to the 
position and the current time. We can pre-compute the blizzard positions as they are 
cyclic. We include also multiple goals for part 2 of the puzzle.

```{r get-solution-1, cache = TRUE}
simulate_blizzards <- function(grid, trip = c("start", "goal")) {
  qs <- 1000L
  nrow <- nrow(grid)
  ncol <- ncol(grid)
  start_col <- which(grid[1, ] == ".")
  goal_col  <- which(grid[nrow, ] == ".")
  inner <- grid[2:(nrow - 1L), 2:(ncol - 1L)]
  H <- nrow(inner)
  W <- ncol(inner)
  start <- cbind(0L, start_col - 1L)
  goal <- cbind(H + 1L, goal_col - 1L)
  
  dirs <- rbind(
    "^" = c(-1L, 0L),
    ">" = c(0L, 1L),
    "v" = c(1L, 0L),
    "<" = c(0L, -1L),
    "x" = c(0L, 0L)
  )
  blizzards <- which(inner != ".", arr.ind = TRUE)
  blizzards <- cbind(blizzards, dir = match(inner[blizzards], rownames(dirs)))
  period <- LCM(H,W)
  
  # pre-compute blizzard positions
  blizz_lkp <- vector("list", period)
  for (t in 1:period) {
    mat <- matrix(FALSE, H, W)
    bi <- blizzards[, 1:2] - 1L + dirs[blizzards[, 3L], ] * t
    bi[, 1L] <- bi[, 1L] %% H + 1L
    bi[, 2L] <- bi[, 2L] %% W + 1L
    mat[bi] <- TRUE
    blizz_lkp[[t]] <- mat
  }
  
  bfs <- function(start_pos, goal_pos, start_minute = 0L) {
    queue <- vector("list", qs)
    qp <- 1L
    ql <- 1L
    queue[[qp]] <- list(pos = start_pos, t = 0L)
    visited <- new.env(parent = emptyenv())
    
    key <- function(pos, t) {
      paste(pos[, 1L], pos[, 2L], t %% period, sep = ",")
    }
    
    visited[[key(start_pos, 0L)]] <- TRUE
    while (qp <= ql) {
      state <- queue[[qp]]
      qp <- qp + 1L
      pos <- state$pos
      t <- state$t
      
      if (all(pos == goal_pos)) {
        return(t)
      }
      
      blizz <- blizz_lkp[[((t + start_minute ) %% period) + 1L]]
      for (k in 1:nrow(dirs)) {
        new_pos <- pos + dirs[k, ]
        if (!all(new_pos == start_pos) && !all(new_pos == goal_pos)) {
          if (new_pos[, 1L] < 1L || new_pos[, 1L] > H || 
              new_pos[, 2L] < 1L || new_pos[, 2L] > W) {
            next
          }
          if (blizz[new_pos]) {
            next
          } 
        }
        
        # Memo
        kkey <- key(new_pos, t + 1L)
        if (exists(kkey, visited, inherits = FALSE)) {
          next
        }
        visited[[kkey]] <- TRUE
        ql <- ql + 1L
        if (ql > length(queue)) {
          queue <- c(queue, vector("list", qs))
        }
        queue[[ql]] <- list(pos = new_pos, t = t + 1L)
      }
    }
  }
  poi <- list(start = start, goal = goal)
  n <- length(trip)
  res <- vector("integer", n - 1L)
  t <- 0L
  for (i in seq_len(n - 1L)) {
    res[i] <- bfs(poi[[trip[i]]], poi[[trip[i + 1L]]], t)
    t <- t + res[i]
  }
  sum(res)
}

simulate_blizzards(puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

As we implemented already a trip with multiple goals, we just have to provide a longer 
trip this time.


```{r get-solution-2, cache = TRUE}
simulate_blizzards(puzzle_data, c("start", "goal", "start", "goal")) 
```

