---
params:
   task_nr: 19
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2024"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(stringi)
library(knitr)
library(cli)
library(bit64)
library(igraph)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  list(
    towels = str_split(text_block[1L], ", ?") %>% 
      extract2(1L),
    patterns = text_block[-1L] %>% 
      keep(nzchar)
  )
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

Looking at the available towels we observe a couple of things:

1. All letters except `b` are available as **single letters**:
   ```{r single-letters}
   str_subset(puzzle_data$towels, "^.$") %>% 
     sort()
   ```
1. All digrams with `b` are evailable except `bg`:
   ```{r digrams-with-b}
   str_subset(puzzle_data$towels, ".b|b.") %>% 
     str_subset("^..$") %>% 
     sort()
   ```
1. A `gb` pattern not occuring at the end of the string is **never** a problem, as it can
be split up, e.g. `ugbbbu => u g bb bu` (single letters are all available and so are all 
`b`  digrams except `bg`).
1. Thus, only `gb` patterns at the end pose a potential problem, **unless** there is a 
special (longer) towel with `gb` in the end:
   ```{r life-savers}
   puzzle_data$towels %>% 
     str_subset("gb$")
   ```
1. However, if there is a special token fitting the end of a pattern, we **must not** end
with `gb` without a matching token for the same reasons.

This allows to construct an algorithm which iteratively remove patterns which end with 
`gb` without a matching towel.


```{r get-solution-1}
get_solvable_patterns <- function(data) {
  life_savers <- data$towels %>% 
    str_subset("gb$")
  has_token <- paste0("(", paste(life_savers, collapse = "|"), ")$")
  keep(data$patterns, function(pattern) {
    res <- TRUE
    while (res && str_detect(pattern, "gb$")) {
      if (str_detect(pattern, has_token)) {
        # Remove the matched token from the end
        pattern <- str_remove(pattern, has_token)
      } else {
        # No token can save this pattern
        res <- FALSE
      }
    }
    res
  }) 
}

valid_patterns <- get_solvable_patterns(puzzle_data)
length(valid_patterns)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We solve this part by means of dynamic programming.

1. Inititalize a vector (`ways_to_arrange`) with the same length as the pattern. Each
entry $x_i$ in this vector counts the number of arrangements for a substring of length 
$i$. The final reuslt will eventually be found at position $x_n$.
1. Iterate over all character positions $i$ in the pattern and form all substrings up to 
the position $i$. For instance `rrbg => [r], [rr, r], [rrb, rb, b], [rrbg, rbg, bg, g]`.
1. At each iteration check if the sub string represents an existing towel.
1. If this is the case, there are 2 possibilities:
   1. It is the complete substring from the beginning. That is the first part of the 
   pattern. In this case set $x_i$ to 1. This represents the fact that we can match this
   pattern in one way simply by using the fitting towel.
   1. If we look at a substring **not** starting at the beginning, we simply add the 
   $x_{i-1}$ to $x_i$. The idea is that with this new towel we add new possibilities to 
   the arrangement which correspond to the number of arrangements which we had before 
   adding this very towel.
   
For instance consider the example towels `r, wr, b, g, bwu, rb, gb, br` and the pattern 
`rrbgbr`. The algorithm would work like this:

1. We initialize `x` to `[0,0,0,0,0,0]`
1. $i=1$: 
   * The first (and only) substring is `r`. There is a towel for `r`: 
   `x = [1,0,0,0,0,0]`
1. $i=2$:
   * First substring is `rr`. There is no matching towel.
   * Second substring is `r`. There is a matching towel. We add the number of ways to
   arrange a pattern of length 1 to $x_2$: x = [1,1,0,0,0,0,0]
1. $i=3$: 
   * First substring is `rrb`. There is no matching towel.
   * Second substring is `rb`. There is a matching towel, use the number of arrangement
   for a string of length 1: `x = [1,1,1,0,0,0,0]`.
   * Third substring is `b`. Again there is a matching towel, add the number of 
   arrangements for a string of length 2: `x = [1,1,2,0,0,0,0]`.
1. $i=4$ (*N.B.* I skip irrelevant iterations, especially looking further back than 
$\max_i|t_i|$ positions, where $t_i$ is towel $i$ does not make sense, as there cannot be 
a matching towel of this length):
   * `rbg` does not match.
   * `bg` does not match.
   * `g` matches: `x = [1,1,2,2,0,0]`.
1. $i=5$:
   * `bgb` does not match.
   * `gb` matches: `x = [1,1,2,2,2,0]`.
   * `b` matches: `x = [1,1,2,2,4,0]`.
1. $i=6$:
   * `gbr` does not match.
   * `br` matches: `x = [1,1,2,2,4,2]`.
   * `r` matches: `x = [1,1,2,2,4,6]`.
1. Final result is 6.

```{r get-solution-2}
count_arrangements <- function(data) {
  max_len <- data$towels %>% 
    nchar() %>% 
    max()
  count <- function(pattern, towels) {
    idx <- seq(1L, nchar(pattern))
    ways_to_arrange <- rep(as.integer64(0L), nchar(pattern))
    for (end_pos in seq(1L, nchar(pattern))) {
      for (start_pos in seq(max(1L, end_pos - max_len + 1L), end_pos)) {
        sub_string <- substr(pattern, start_pos, end_pos)
        if (sub_string %in% towels) {
          if (start_pos == 1L) {
            ways_to_arrange[end_pos] <- 1L
          } else {
            ways_to_arrange[end_pos] <- ways_to_arrange[end_pos] + 
              ways_to_arrange[start_pos - 1L]
          }
        }
      }
    }
    tail(ways_to_arrange, 1L)
  }
  map(data$patterns, count, data$towels) %>% 
    do.call(c, .) %>% 
    sum()
}
count_arrangements(puzzle_data)
```

