---
params:
   task_nr: 6
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(knitr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    str_split("\n") %>% 
    extract2(1L) %>% 
    str_split("") %>% 
    do.call(rbind, .)
})
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We solve this puzzle in the following way:

1. From the current position, find the nearest obstacle in the same row or column, 
depending on the current orientation of the guard.
1. If there is no obstacle fill all the fields and stop the loop.
1. From the current position up to the field before the obstacle fill all empty spots.
1. Turn around and repeat.

*N.B.* The code below includes already parts (`extended_path`) which are needed for part 2
of the puzzle and will be explained there

```{r helper-funcs}
get_relative_position <- function(p1, p2) {
  x <- -1:1
  lkp <- expand.grid(s1 = x, s2 = x) %>% 
    mutate(direction = c("left_top", "left", "left_bottom",
                         "top", "same", "bottom",
                         "right_top", "right", "right_bottom"
    )) %>% 
    as_tibble()
  pos <- c(sign(p2 - c(p1))) %>% 
    matrix(ncol = 2) %>% 
    set_colnames(c("s1", "s2")) %>% 
    as_tibble()
  lkp %>% 
    inner_join(pos, c("s1", "s2")) %>% 
    pull(direction)
}

get_connection_node <- function(p1, p2) {
  ## p1 must be a single point
  stopifnot("`p1` must be single point" = length(c(p1)) == 2L)
  ## make sure ü2 has dimensions
  p2 <- matrix(as.matrix(p2), ncol = 2L)
  apply(p2, 1, function(p2) {
    pos <- get_relative_position(p1, p2)
    if (pos == "right_bottom") {
      if (p1[1L] + 1L == p2[1L]) {
        res <- p2
      } else {
        res <- cbind(p1[1L] + 1L, p2[2L] + 1L)
      }
    } else if (pos == "right_top") {
      if (p1[2L] + 1L == p2[2L]) {
        res <- p2
      } else {
        res <- cbind(p2[1L] - 1L, p1[2L] + 1L)
      }
    } else if (pos == "left_bottom") {
      if (p1[2L] - 1L == p2[2L]) {
        res <- p2
      } else {
        res <- cbind(p2[1L] + 1L, p1[2L] - 1L)
      }
    } else if (pos == "left_top") {
      if (p1[1L] - 1L == p2[1L]) {
        res <- p2
      } else {
        res <- cbind(p1[1L] - 1L, p2[2L] - 1L)
      }
    } else {
      res <- cbind(NA_integer_, NA_integer_)
    }
    res
  }) %>% 
    t()
}

o <- as.matrix(obs)
all_con_nodes <- lapply(seq(1L, nrow(o)), 
                        \(i) get_connection_node(o[i, ], o[-i, ])) %>% 
  do.call(rbind, .) %>% 
  as_tibble() %>% 
  distinct() %>% 
  setNames(c("row", "col")) %>% 
  inner_join(path)

```


```{r get-solution-1}
find_obstacles <- function(line, current_pos) {
  all_obstacles <- which(line == "#")
  distance <- all_obstacles - current_pos
  left_top <- distance[distance < 0L]
  right_bottom <- distance[distance > 0L]
  if (length(left_top) > 0L) {
    res <- c(left_top = all_obstacles[distance == max(left_top)] + 1L)
  } else {
    res <- c(left_top = 1L)
  }
  if (length(right_bottom) > 0L) {
    res <- c(res, right_bottom = all_obstacles[distance == min(right_bottom)] - 1L)
  } else {
    res <- c(res, right_bottom = length(line))
  }
  res
}

set_marks <- function(marks, new_mark) {
  marks %>% 
    str_split(fixed("|")) %>% 
    map_chr(
      function(mark) {
        mark <- discard(mark, ~ .x == ".")
        c(mark, new_mark)  %>% 
          unique() %>% 
          sort() %>% 
          paste(collapse = "|")
      }
    )
}

simulate_walk <- function(current_map, current_position, direction, paths = list()) {
  dd <- dim(current_map)
  if (direction %in% c("up", "down")) {
    idx <- cbind(1:dd[1L], current_position[2L])
    guard <- current_position[1L]
  } else if (direction %in% c("left", "right")) {
    idx <- cbind(current_position[1L], 1:dd[2L])
    guard <- current_position[2L]
  }
  closest_obstacles <- find_obstacles(current_map[idx], guard)
  if (direction %in% c("left", "up")) {
    path <- idx[seq(closest_obstacles[["left_top"]], guard), ]
    if (guard < nrow(idx)) {
      extended_path <- idx[seq(guard + 1L, closest_obstacles[["right_bottom"]]), ,
                           drop = FALSE]
    } else {
      extended_path <- idx[FALSE, , drop = FALSE]
    }
    new_position <- idx[closest_obstacles[["left_top"]], ]
    done <- closest_obstacles[["left_top"]] == 1L
  } else if (direction %in% c("right", "down")) {
    path <- idx[seq(guard, closest_obstacles[["right_bottom"]]), ]
    if (guard > 1L) {
      extended_path <- idx[seq(closest_obstacles[["left_top"]], guard - 1L), , 
                           drop = FALSE]
    } else {
      extended_path <- idx[FALSE, , drop = FALSE]
    }
    new_position <- idx[closest_obstacles[["right_bottom"]], ]
    done <- closest_obstacles[["right_bottom"]] == nrow(idx)
  }
  new_map <- current_map
  new_map[path] <- set_marks(new_map[path], c("X", direction))
  new_map[extended_path] <- set_marks(new_map[extended_path], direction)
  if (list(path) %in% paths) {
    attr(new_map, "loop") <- TRUE
    done <- TRUE
  } else {
    paths <- c(paths, list(path))
    attr(new_map, "loop") <- FALSE
  }
  if (!done) {
    new_direction <- case_when(
      direction == "up" ~ "right",
      direction == "right" ~ "down",
      direction == "down" ~ "left",
      direction == "left" ~ "up"
    )
    new_map <- Recall(new_map, new_position, new_direction, paths)
  }
  new_map 
}

simulate_walk3 <- function(current_map, current_position, direction) {
  dd <- dim(current_map)
  if (direction %in% c("up", "down")) {
    idx <- cbind(1:dd[1L], current_position[2L])
    guard <- current_position[1L]
  } else if (direction %in% c("left", "right")) {
    idx <- cbind(current_position[1L], 1:dd[2L])
    guard <- current_position[2L]
  }
  closest_obstacles <- find_obstacles(current_map[idx], guard)
  if (direction %in% c("left", "up")) {
    path <- idx[seq(closest_obstacles[["left_top"]], guard), ]
    if (guard < nrow(idx)) {
      extended_path <- idx[seq(guard + 1L, closest_obstacles[["right_bottom"]]), ,
                           drop = FALSE]
    } else {
      extended_path <- idx[FALSE, , drop = FALSE]
    }
    new_position <- idx[closest_obstacles[["left_top"]], ]
    done <- closest_obstacles[["left_top"]] == 1L
  } else if (direction %in% c("right", "down")) {
    path <- idx[seq(guard, closest_obstacles[["right_bottom"]]), ]
    if (guard > 1L) {
      extended_path <- idx[seq(closest_obstacles[["left_top"]], guard - 1L), , 
                           drop = FALSE]
    } else {
      extended_path <- idx[FALSE, , drop = FALSE]
    }
    new_position <- idx[closest_obstacles[["right_bottom"]], ]
    done <- closest_obstacles[["right_bottom"]] == nrow(idx)
  }
  new_map <- current_map
  new_map[path] <- set_marks(new_map[path], c("X", direction))
  new_map[extended_path] <- set_marks(new_map[extended_path], direction)
  if (!done) {
    new_direction <- case_when(
      direction == "up" ~ "right",
      direction == "right" ~ "down",
      direction == "down" ~ "left",
      direction == "left" ~ "up"
    )
    new_map <- Recall(new_map, new_position, new_direction)
  }
  new_map 
}

simulate_walk2 <- function(current_map, current_position, direction) {
  dd <- dim(current_map)
  done <- FALSE
  while (!done) {
    if (direction %in% c("up", "down")) {
      idx <- cbind(1:dd[1L], current_position[2L])
      guard <- current_position[1L]
    } else if (direction %in% c("left", "right")) {
      idx <- cbind(current_position[1L], 1:dd[2L])
      guard <- current_position[2L]
    }
    closest_obstacles <- find_obstacles(current_map[idx], guard)
    if (direction %in% c("left", "up")) {
      path <- idx[seq(closest_obstacles[["left_top"]], guard), ]
      if (guard < nrow(idx)) {
        extended_path <- idx[seq(guard + 1L, closest_obstacles[["right_bottom"]]), ,
                             drop = FALSE]
      } else {
        extended_path <- idx[FALSE, , drop = FALSE]
      }
      current_position <- idx[closest_obstacles[["left_top"]], ]
      done <- closest_obstacles[["left_top"]] == 1L
    } else if (direction %in% c("right", "down")) {
      path <- idx[seq(guard, closest_obstacles[["right_bottom"]]), ]
      if (guard > 1L) {
        extended_path <- idx[seq(closest_obstacles[["left_top"]], guard - 1L), , 
                             drop = FALSE]
      } else {
        extended_path <- idx[FALSE, , drop = FALSE]
      }
      current_position <- idx[closest_obstacles[["right_bottom"]], ]
      done <- closest_obstacles[["right_bottom"]] == nrow(idx)
    }
    current_map[path] <- set_marks(current_map[path], c("X", direction))
    current_map[extended_path] <- set_marks(current_map[extended_path], direction)
    direction <- case_when(
      direction == "up" ~ "right",
      direction == "right" ~ "down",
      direction == "down" ~ "left",
      direction == "left" ~ "up"
    )
  }
  current_map 
}

```

```{r get-solution-1}
maze <- puzzle_data
current_pos <- which(maze == "^", arr.ind = TRUE)
maze[current_pos] <- "."
walked_maze <- simulate_walk2(maze, current_pos, "up")
walked_maze %>% 
  str_count("X") %>% 
  sum()
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

A couple of observations for the placment of any new obstacles which should form a loop:

1. It must be on one of the fields which a guard passed in his original walk.
1. A loop is created if the new path induced by the new obstacle will lead into a path the
guard took already. There are 2 options for this:
1. The first field after the obstacle (i.e. after turning) was visited by the guard
previously **and** he passed this evry field in the same direction.
1. The **extension** of the new path eventually leads into a path already walked by
the guard (again in the same direction).

To translate this into an algorithm, we need to
1. Mark each field on the path with all the directions the guard took walking over this
field.
1. Extend each path from the turning point until the next obstacle (to make it easier to
detect whether a new path connects with a trodden one).
1. With this steps done, all we have to do is to loop through all fields on the path and
check whether placing an obstacle on this field would result in a path connecting with an
old path.
1. With this appraoch it could however happen, that we also create shortcuts to freedom.
That is we connect to the path which leads to the border. To avoid this we would need to
keep track of the timepoint when we visited a field. This seems to be too cumbersome for
the moment, hence we take a brute force approach: Get all reconnecting paths and check for each if we still could escape eventually.

```{r find_loops}
is_trodden <- function(walked_map, current_field) {
  all_directions <- walked_map[current_field] %>% 
    str_split(fixed("|")) %>% 
    extract2(1L) %>% 
    str_subset(fixed("X"), negate = TRUE)
  check_field <- function(current_field, direction) {
    new_direction <- case_when(
      direction == "up" ~ "right",
      direction == "right" ~ "down",
      direction == "down" ~ "left",
      direction == "left" ~ "up"
    )
    prev_field <- case_when(
      direction == "up" ~ cbind(current_field[1L] + 1L, current_field[2L]),
      direction == "right" ~ cbind(current_field[1L], current_field[2L] - 1L),
      direction == "down" ~ cbind(current_field[1L] - 1L, current_field[2L]),
      direction == "left" ~ cbind(current_field[1L], current_field[2L] + 1L)
    )
    new_field <- case_when(
      new_direction == "up" ~ cbind(prev_field[1L] - 1L, prev_field[2L]),
      new_direction == "right" ~ cbind(prev_field[1L], prev_field[2L] + 1L),
      new_direction == "down" ~ cbind(prev_field[1L] + 1L, prev_field[2L]),
      new_direction == "left" ~ cbind(prev_field[1L], prev_field[2L] - 1L)
    )
    str_detect(walked_map[new_field], new_direction) &&
      str_detect(walked_map[prev_field], fixed("#"), negate = TRUE)
  }
  map_lgl(all_directions, ~ check_field(current_field, .x)) %>% 
    any()
}

find_reconnecting_obstacles <- function(walked_map, starting_pos) {
  visited_fields <- str_detect(walked_map, "X") %>% 
    `dim<-`(dim(walked_map)) %>% 
    which(arr.ind = TRUE)
  visited_fields <- visited_fields[!(visited_fields[, 1L] == starting_pos[[1L]] &
                                       visited_fields[, 2L] == starting_pos[[2L]]), ,
                                   drop = FALSE]
  candidates <- apply(visited_fields, 1, function(ii)
    is_trodden(walked_map, cbind(ii[[1L]], ii[[2L]])))
  visited_fields[candidates, , drop = FALSE]
}

find_loop_obstacles <- function(walked_map, starting_pos) {
  orig_map <- walked_map
  orig_map[orig_map != "#"] <- "."
  candidates <- find_reconnecting_obstacles(walked_map, starting_pos)
  print(paste(nrow(candidates), "candidates found"))
  i <- 1 
  is_loop <- apply(candidates, 1, function(ii) {
    print(paste("Checking candidate", i))
    idx <- cbind(ii[1L], ii[2L])
    new_map <- orig_map
    new_map[idx] <- "#"
    i <<- i  + 1
    attr(simulate_walk(new_map, starting_pos, "up"), "loop")
  })
  candidates[is_loop, , drop = FALSE]
}

brute_force <- function(walked_map, starting_pos) {
  start <- Sys.time()
  print(start)
  orig_map <- walked_map
  orig_map[orig_map != "#"] <- "."
  visited_fields <- str_detect(walked_map, "X") %>% 
    `dim<-`(dim(walked_map)) %>% 
    which(arr.ind = TRUE)
  visited_fields <- visited_fields[!(visited_fields[, 1L] == starting_pos[[1L]] &
                                       visited_fields[, 2L] == starting_pos[[2L]]), ,
                                   drop = FALSE]
  #visited_fields <- visited_fields[1:10, ]
  print(paste("Checking", nrow(visited_fields), "Candidates"))
  i <- 1
  is_loop <- apply(visited_fields, 1, function(ii) {
    print(paste("Checking candidate", i))
    idx <- cbind(ii[1L], ii[2L])
    new_map <- orig_map
    new_map[idx] <- "#"
    i <<- i  + 1
    attr(simulate_walk(new_map, starting_pos, "up"), "loop")
  })
  duration <- Sys.time() - start
  print(duration)
  visited_fields[is_loop, , drop = FALSE]
}

sol <- brute_force(walked_maze, current_pos)

find_loop_obstacles(walked_maze, current_pos) %>% 
  nrow()

find_loop_obstacles(orig_maze_walked, cbind(7, 5))
```


```{r get-solution-2}
mark_fields <- function(current_map, current_position, direction) {
  dd <- dim(current_map)
  if (direction == "up") {
    idx <- cbind(1:current_position[1L], current_position[2L])
    sel <- max
    step_back <- c(1L, 0L)
    turn_to <- "right"
  } else if (direction == "left") {
    idx <- cbind(current_position[1L], 1:current_position[2L])
    sel <- max
    step_back <- c(0L, 1L)
    turn_to <- "up"
  } else if (direction == "down") {
    idx <- cbind(current_position[1L]:dd[1L], current_position[2L])
    sel <- min
    step_back <- c(-1L, 0L)
    turn_to <- "left"
  } else if (direction == "right") {
    idx <- cbind(current_position[1L], current_position[2L]:dd[2L])
    sel <- min
    step_back <- c(0L, -1L)
    turn_to <- "down"
  }
  obstacles <- which(current_map[idx] == "#")
  if (length(obstacles) == 0L) {
    marks <- current_map[idx] %>% 
      str_split(fixed("|"))
    
    new_marks <-  map_chr(marks, ~ c(.x, direction) %>% 
                            unique() %>% 
                            sort() %>% 
                            paste(collapse = "|")) %>% 
      str_remove("^\\|")
    current_map[idx] <- new_marks
    res <- current_map
  } else {
    nearest_obstacle <- idx[sel(obstacles), , drop = FALSE]
    new_position <- nearest_obstacle + step_back
    distance <- max(c(abs(new_position - current_position)))
    idx <- cbind(seq(current_position[, 1L], new_position[, 1L],
                     length.out = distance + 1L),
                 seq(current_position[, 2L], new_position[, 2L],
                     length.out = distance + 1L))
    marks <- current_map[idx] %>% 
      str_split(fixed("|")) 
    new_marks <- map_chr(marks, ~ c(.x, direction) %>% 
                           unique() %>% 
                           sort() %>% 
                           paste(collapse = "|")) %>% 
      str_remove("^\\|")
    current_map[idx] <- new_marks
    res <- Recall(current_map, new_position, turn_to)
  }
  res
}

```
