---
params:
   task_nr: 6
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(knitr)
library(cli)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    str_split("\n") %>% 
    extract2(1L) %>% 
    str_split("") %>% 
    do.call(rbind, .)
})
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We solve this puzzle by simulating the alk field by field. An earlier version of this code
used more sophisticated indexing, but was for some reasons not clear to me, slower.

Anyways, the algorithm is rather straight forward:

1. If we are not seeing a `#` move one field in the current direction and store the
step number and current direction in the path (this is done to recognize loops.)
1. If we hit a `#` first step back (reducing the step counter to undo the step) and then 
turn.
1. Continue until we would be outside the grid.

*N.B.* Most of the data stored (`step_number` or `direction`) is not needed for the first 
puzzle, where we simply want to get the number of fields visited. Also the part where we 
can resume an earlier path. However, we will need (and explain) these parts in the section 
of puzzle 2.



```{r get-solution-1}
get_field_id <- function(idx) {
  paste(idx, collapse = "/")
}

walk_maze <- function(maze, trodden_path = NULL) {
  pos <- which(maze == "^", arr.ind = TRUE)
  continue <- loop <- done <- FALSE
  dir <- "up"
  walk <- function(pos, dir, backward = FALSE) {
    offset <- switch(
      dir,
      up = cbind(-1L, 0L),
      right = cbind(0L, 1L),
      down = cbind(1L, 0L),
      left = cbind(0L, -1L)
    )
    pos + (1L - 2L * backward) * offset
  }
  path <- vector("list", prod(dim(maze)))
  i <- 1L
  step_counter <- 1L
  if (!is.null(trodden_path)) {
    idx <- seq_along(trodden_path)
    path[idx] <- trodden_path
    names(path)[idx] <- names(trodden_path)
    last_field_idx <- map_dbl(
      trodden_path,
      max
    ) %>% 
      which.max()
    last_field <- trodden_path[last_field_idx]
    step_counter <- max(last_field[[1L]])
    pos <- last_field %>% 
      names() %>% 
      str_split(fixed("/")) %>% 
      extract2(1L) %>% 
      as.integer() %>% 
      matrix(ncol = 2L)
    dir <- last_field[[1L]] %>% 
      tail(1L) %>% 
      names()
    i <- length(trodden_path) + 1L
    continue <- TRUE
  }
  while (!done) {
    cur_field <- maze[pos]
    if (cur_field == "#") {
      pos <- walk(pos, dir, TRUE)
      dir <- switch(
        dir,
        up = "right",
        right = "down",
        down = "left",
        left = "up"
      )
      step_counter <- step_counter - 1L
    } else {
      id <- get_field_id(pos)
      new_step <- step_counter %>% 
        setNames(dir)
      if (id %in% names(path)) {
        if (continue) {
          ## if we continue a path the first step is already recorded so skip
          continue <- FALSE
        } else {
          if (dir %in% names(path[[id]])) {
            ## loop condition
            done <- loop <- TRUE
          }
          path[[id]] <- c(path[[id]], new_step)
        }
      } else {
        path[[i]] <- new_step
        names(path)[[i]] <- id
        i <- i + 1L
      }
      pos <- walk(pos, dir)
      step_counter <- step_counter + 1L
    }
    done <- done || pos[1L] %in% c(0L, nrow(maze) + 1L) ||
      pos[2L] %in% c(0L, ncol(maze) + 1L)
  }
  res <- discard(path, is.null)
  attr(res, "loop") <- loop
  res
}

path <- walk_maze(puzzle_data)
length(path)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

Any obstacle creating a loop must be placed on the original path. An obstacle placed
somewhere else would simply be never hit. We loop through each field on the path, place
an obstacle there and re-simulate the walk. To detect whether we ended up in a loop, all 
we have to check whether the field we are entering was entered already before in the same
direction. To safe some time, we do not regenerate the walk from the beginning at each 
iteration but reuse the original path up to the field on which we are placing the new 
obstacle. For this reason, we stored the `step_counter` in the first place, because to
trim a path, we simply discard and step larger than a given step number.

```{r get-solution-2, cache = TRUE}
trim_path <- function(path, idx) {
  if (idx > 0L) {
    keep(path, ~ any(.x <= idx)) %>% 
      map(~ .x[.x <= idx])
  } else {
    NULL
  }
}

find_loops <- function(maze, path) {
  start_pos <- which(maze == "^", arr.ind = TRUE)
  path_fields_nm <- names(path) %>% 
    setdiff(get_field_id(start_pos))
  path_fields <- path_fields_nm %>% 
    str_split(fixed("/")) %>% 
    map(~ as.integer(.x) %>% matrix(ncol = 2))
  if (interactive()) {
    pb <- list(
      name = "Checking Candidates",
      type = "iterator",
      ## need to convert to double beforehand otherwise we get an error
      total = as.numeric(length(path_fields)) 
    )
  } else {
    pb <- FALSE
  }
  idx <- map_lgl(
    seq_along(path_fields), 
    function(i) {
      ## use global maze to avoid copying large objects in each iteration
      cand <- path_fields[[i]]
      maze[cand] <- "#"
      current_path <- trim_path(path, i - 1L)
      new_path <- walk_maze(maze, current_path)
      ## need to reset the maze
      maze[cand] <- "."
      attr(new_path, "loop")
    },
    .progress = pb
  )
  path_fields_nm[idx]
}
res <- find_loops(puzzle_data, path)
length(res)
```
