---
params:
   task_nr: 17
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
  })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}

code em {
font-weight:bold
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
	if (knitr::is_latex_output()) {
		sprintf("\\texttt{%s}", as.character(x))
	} else if (knitr::is_html_output()) {
		sprintf("<tt>%s</tt>", as.character(x))
	} else {
		x
	}
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(tibble)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2021/day/", params$task_nr)
puzzle <- GET(base_url,
				  session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	xml_find_all("///article") %>% 
	lapply(as.character)

puzzle_data <- GET(paste0(base_url, "/input"),
						 session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	str_extract_all("-?\\d+") %>% 
	`[[`(1L) %>% 
	as.integer() %>% 
	set_names(c("xmin", "xmax", "ymin", "ymax"))
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

This was the first puzzle, which I solved by logic rather than by coding. With
the given kinetics we can observe the following:

1. To reach a maximum height, we require `y > 0` and the higher `y` the higher
the maximum height.
1. Every trajectory with `y_0 > 0` will eventually return to the horizontal
baseline at  `(x / 0)`. This is because the `y` velocity will first decrease 
from `y_0` to `0` (reaching a maximum height at `sum(1:y_0)`) and then further 
decrease until it reaches `-y_0` which means it decreases exactly 
`sum(-1:-y_0) = -sum(1:y_0)` units.
1. In order to not overshoot the last decrease can be at most 
`r puzzle_data["ymin"]`.
1. That is `y_0` can be at most `r abs(puzzle_data["ymin"]) - 1` and the 
absolute height can be solved by good old Gauss.

```{r solution-a, cache = FALSE}
y0 <- abs(puzzle_data["ymin"]) - 1
(y0 + 1) * y0 / 2
```


## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

To get all feasible trajectories, we observe the following points:

1. The *minimal* `x` velocity is the minimal `x_0` which satisfies 
`sum(1:x_0) >= x_min`. Using Gauss  we conclude 
``(1 + x) * x / 2 >= `r puzzle_data["xmin"]` `` this can be solved by the 
quadratic formula:
   ```{r quadratic-formula-min}
   q_solve <- function(a, b, c) {
   	(-b + c(-1, 1) * sqrt(b ^ 2 - 4 * a * c)) / (2 * a)
   }
   sol <- q_solve(1 / 2, 1 / 2, -puzzle_data["xmin"])
   ceiling(sol[sol > 0])
   ```
1. We can reach each point `(x/y)` of the trench exactly by using the trajectory
`(x/-y)`.
1. Besides these direct trajectories, the maximum `x` velocity is 
`ceil(xmax / 2)`. If it were `ceil(xmax / 2) + 1` then the next step would be
already at `ceil(xmax / 2) + 1 + ceil(xmax / 2) + 1 - 1 >= xmax` and thus 
overshoot the area. Thus, 
the maximal `x` velocity is `r (x0max <- ceiling(puzzle_data["xmax"] / 2))`.
1. Likewise the minimum `y` velocity is `ceil(ymin / 2)`, i.e. 
`r (y0min <- ceiling(puzzle_data["ymin"] / 2))`.
1. As per part 1, the maximum `y` velocity is `r (y0max <- y0)`.
1. For each pair `(x0 / y0)` we can calculate the minimal steps which are needed 
to reach the trench in both directions, `kx` and `ky`, say. Then we take the 
maximum of those two values (`k`) and check whether we reached the trench after
`k` steps. It is not necessary to check any other value, because if we overshoot
after `k` steps, we will overshoot after `k + 1` steps. As we did not reach the
trench before `k` steps in at least one direction it also not necessary to check
any value smaller than `k`.
1. Overall, a semi-intelligent brute-force algorithm, iterates over each pair in
``(`r ceiling(sol[sol > 0])`...`r x0max`/`r y0min`...`r y0max`)`` calculates `k`
and counts how many trajectories reach the trench after `k` steps and adds 
`prod(dim(puzzle) + 1)` to the result accounting for the direct trajectories.^[The 
algo is rather stupid, because if for any given `(x / yk)` we cannot reach the 
trench, we do not need to check any `(x / yk + i)`. However, due to the small
sample size we do not really need to fine tune here.]

```{r solve}
get_k <- Vectorize(function(x0, y0, dims = puzzle_data) {
	## solve x0 + x0 - 1 + ... + x0 - k >= dims["xmin"] w.r.t. k
	## (x0 + x0 - k) * (x0 - (x0 - k) + 1) / 2 = dims["xmin"]
	## -1 / 2 * k² + (2x0 - 1) / 2 * k + x0 - d
	kx <- q_solve(-1 / 2, x0 - 1 / 2, x0 - dims["xmin"]) %>% 
		ceiling()
	kx <- kx[kx <= x0][[1L]]
	ky <- q_solve(-1 / 2, y0 - 1 / 2, y0 - dims["ymax"]) %>% 
		ceiling()
	ky <- ky[ky >= 0][[1L]]
	max(kx, ky)
}, c("x0", "y0"))

get_x <- Vectorize(function(x0, k) {
	stopifnot(x0 >= 0)
	if (k > x0) {
		(x0 + 1) * x0 / 2
	} else {
		(2 * x0 - k) * (k + 1) / 2
	}
})

get_y <- Vectorize(function(y0, k) {
	if (k == 0) {
		0
	} else {
		(2 * y0 - k) * (k + 1) / 2
	}
})

is_within <- function(x, y, k, dims = puzzle_data) {
	between(get_x(x, k), dims["xmin"], dims["xmax"]) &
		between(get_y(y, k), dims["ymin"], dims["ymax"])
}

solve <- function(dims = puzzle_data) {
	## total movement in x direction =
	## x0 + x0 - 1 + x0 - 2 + ... + 1 = (x0 + 1) * X0 / 2
	## To reach the trench we need at least dims["xmin"] steps in total
	## => solve (x0 + 1) * x0 / 2 = dims["xmin"] w.r.t to x0
	x0min <- q_solve(1 / 2, 1 / 2, -dims["xmin"])
	x0min <- ceiling(x0min[x0min > 0])
	x0max <- ceiling(dims["xmax"] / 2)
	y0min <- ceiling(dims["ymin"] / 2)
	y0max <- abs(dims["ymin"]) - 1
	sols <- expand.grid(x = x0min:x0max,
							  y = y0min:y0max) %>% 
		as_tibble() %>% 
		mutate(k = get_k(x, y, dims),
				 hits = is_within(x, y, k, dims)) %>% 
		filter(hits) %>% 
		nrow()
	sols + ((diff(range(dims[c("xmin", "xmax")])) + 1) *
			  	(diff(range(abs(dims[c("ymin", "ymax")]))) + 1))
}

solve(puzzle_data)
```

