---
params:
   task_nr: 15
   year: 2016
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2016"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  res <- str_extract_all(text_block, "\\d+") %>% 
    do.call(rbind, .)
  storage.mode(res) <- "integer"
  colnames(res) <- c("nr", "nr_pos", "time", "cur_pos")
  as_tibble(res)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

For each disc we want to solve the equation

$$(\text{start}_i + \text{offset}_i + t) \mod \text{size}_i \equiv 0$$
where $\text{offset}_i = i$ describes the time the capsule needs to reach the disc. The
minimal $t$ which solves all equations simultaneously is our solution. Thus, we look for 
the minimal $t$ such that

$$t \equiv (-\text{start}_i - i) \mod \text{size}_i$$

This problem can be solved with the [Chinese Remainder Theorem](https://en.wikipedia.org/wiki/Chinese_remainder_theorem)

```{r get-solution-1}
crt <- function(remainder, modulo) {
  r <- remainder[1L]
  m <- modulo[1]
  for (i in 2:length(remainder)) {
    k <- 0
    while((r + k * m) %% modulo[i] != remainder[i]) {
      k <- k + 1
    }
    r <- r + k * m
    m <- m * modulo[i]
  }
  r
}

remainder <- puzzle_data %>% 
  mutate(remainder = - (cur_pos + 1:n()) %% nr_pos) %>% 
  pull(remainder)

modulo <- puzzle_data %>% 
  pull(nr_pos)

crt(remainder, modulo)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For the second part we simply add another disc to the end.

```{r get-solution-2}
crt(c(remainder, 4), c(modulo, 11))
```
