---
params:
   task_nr: 21
   year: 2019
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

mark_code <- function(code, pos) {
  code <- str_replace(code, "\\.\\.\\.", "&#46;&#46;&#46;")
  left <- str_sub(code, 1L, pos - 1L)
  right <- str_sub(code, pos + 1L)
  char <- str_sub(code, pos, pos)
  sprintf("<code>%s<span style=\"color:red;\">%s</span>%s</code>",
          left, char, right)
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(knitr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
 text_block %>% 
    str_split(",") %>% 
    do.call(c, .) %>% 
    as.numeric()
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We build upon the code from [Day 19 @ 2019](2019_task19.html). First we need to try some
springscripts to get an idea of the holes. We start with the easy rules to jump if there
is a hole in 1, 2 or 4 fields:


```{r map-surface-1, results = "asis"}
parse_op_codes_fast <- function(op_codes = puzzle_data, input = numeric(0L), 
                                verbose = FALSE) {
  ip <- 1L
  rel_base <- 0L
  
  out_buffer <- numeric(0)
  out_len <- 0L
  out_block <- 1024L
  block_size <- 1024L
  
  grow_memory <- function(idx) {
    if (idx > length(op_codes)) {
      new_size <- ceiling(idx / block_size) * block_size
      op_codes <<- c(op_codes, rep(0, new_size - length(op_codes)))
    }
  }
  
  append_out <- function(val) {
    if (out_len + 1L > length(out_buffer)) {
      new_size <- length(out_buffer) + out_block
      out_buffer <<- c(out_buffer, rep(0, new_size - length(out_buffer)))
    }
    out_len <<- out_len + 1L
    out_buffer[out_len] <<- val
  }
  
  get_val <- function(param, mode) {
    if (mode == 0L) {
      grow_memory(param + 1L)
      return(op_codes[param + 1L])
    } else if (mode == 1L) {
      return(param)
    } else if (mode == 2L) {
      idx <- rel_base + param + 1L
      grow_memory(idx)
      return(op_codes[idx])
    } else {
      stop("unknown mode")
    }
  }
  
  get_addr <- function(param, mode) {
    if (mode == 0L) {
      return(param + 1L)
    } else if (mode == 2L) {
      return(rel_base + param + 1L)
    } else {
      stop("invalid write mode")
    }
  }
  
  halt <- FALSE
  while (!halt) {
    instr <- op_codes[ip]
    op <- instr %% 100L
    modes <- c(
      (instr %/% 100L) %% 10L, 
      (instr %/% 1000L) %% 10L, 
      (instr %/% 10000L) %% 10L
    )
    
    if (op == 1L) { # add
      a <- get_val(op_codes[ip + 1L], modes[1L])
      b <- get_val(op_codes[ip + 2L], modes[2L])
      addr <- get_addr(op_codes[ip + 3L], modes[3L])
      grow_memory(addr)
      op_codes[addr] <- a + b
      ip <- ip + 4L
    } else if (op == 2L) { # mul
      a <- get_val(op_codes[ip + 1L], modes[1L])
      b <- get_val(op_codes[ip + 2L], modes[2L])
      addr <- get_addr(op_codes[ip + 3L], modes[3L])
      grow_memory(addr)
      op_codes[addr] <- a * b
      ip <- ip + 4L
    } else if (op == 3L) { # input
      if (length(input) == 0L) {
        stop("No input available")
      }
      addr <- get_addr(op_codes[ip + 1L], modes[1])
      grow_memory(addr)
      op_codes[addr] <- input[1L]
      input <- input[-1]
      ip <- ip + 2L
    } else if (op == 4L) { # output
      val <- get_val(op_codes[ip + 1L], modes[1])
      append_out(val)
      if (verbose) cat(val, "\n")
      ip <- ip + 2L
    } else if (op == 5L) { # jump-if-true
      a <- get_val(op_codes[ip + 1L], modes[1L])
      b <- get_val(op_codes[ip + 2L], modes[2L])
      if (a != 0L) {
        ip <- b + 1L
      } else {
        ip <- ip + 3L
      }
    } else if (op == 6L) { # jump-if-false
      a <- get_val(op_codes[ip + 1L], modes[1L])
      b <- get_val(op_codes[ip + 2L], modes[2L])
      if (a == 0L) {
        ip <- b + 1L
      } else {
        ip <- ip + 3L
      }
    } else if (op == 7L) { # less than
      a <- get_val(op_codes[ip + 1L], modes[1L])
      b <- get_val(op_codes[ip + 2L], modes[2L])
      addr <- get_addr(op_codes[ip + 3L], modes[3L])
      grow_memory(addr)
      op_codes[addr] <- as.numeric(a < b)
      ip <- ip + 4L
    } else if (op == 8L) { # equals
      a <- get_val(op_codes[ip + 1L], modes[1L])
      b <- get_val(op_codes[ip + 2L], modes[2L])
      addr <- get_addr(op_codes[ip + 3L], modes[3L])
      grow_memory(addr)
      op_codes[addr] <- as.numeric(a == b)
      ip <- ip + 4L
    } else if (op == 9L) { # adjust relative base
      a <- get_val(op_codes[ip + 1L], modes[1L])
      rel_base <- rel_base + a
      ip <- ip + 2L
    } else if (op == 99L) { # halt
      halt <- TRUE
    } else {
      stop("Unknown opcode ", op)
    }
  }
  out_buffer[1:out_len]
}

encode_input <- function(springscript, walk) {
  cmd <- if_else(walk, "WALK", "RUN")
  paste(springscript, collapse = "\n") %>% 
    paste0("\n", cmd, "\n") %>% 
    charToRaw() %>% 
    as.integer()
}

decode_output <- function(out) {
  if (all(out <= 255)) {
    intToUtf8(out) %>% 
      str_split("\n") %>% 
      extract2(1L)
  } else {
    tail(out, 1L)
  }
}

run_script <- function(script, program, walk = TRUE) {
  parse_op_codes_fast(program, encode_input(script, walk)) %>% 
    decode_output()
}

show_holes <- function(chk_script, program = puzzle_data, walk = TRUE, numbered = TRUE) {
  res <- map_chr(chk_script, 
                 ~ run_script(.x, program, walk) %>% 
                   extract(11L)) %>% 
    paste0("`",. , "`")
  if (numbered) {
    iwalk(res, ~ cat(.y, ". ", .x, "\n", sep = ""))
  } else {
    walk(res, ~ cat(.x, "\n"))
  }
}

test_scripts <- list(
  "NOT A J", 
  "NOT B J", 
  "NOT D J"
)

show_holes(test_scripts, puzzle_data, TRUE)
```

Now we want to define a sequence which jumps over these patterns:

1. We observe that jumping if there is a hole one field away will jump over patterns 2 
and 3.
1. For pattern 1 we need to jump if there is solid ground 1 and 4 spaces away, while 
there is a hole 2 and 3 spaces away.
1. Eventually, we need to combine this 2 rules using the temporary register.

We feed that to our program and get yet another pattern:

```{r map-surface-2, results = "asis"}
test_scripts <- c(
  test_scripts,
  list(c(
    "NOT C J",
    "AND A J",
    "NOT B T",
    "AND T J",
    "AND D J",
    "NOT A T",
    "OR T J"
  )
  )
)
show_holes(test_scripts, puzzle_data, TRUE)
```


With these patterns we can define now the following Boolean logic (jump off point is
marked in red):

1. $\lnot A \land D$: jumps over `r mark_code("#####.###########", 5L)`.
1. $\lnot A \land \lnot B \land \lnot C \land D$: jumps over 
`r mark_code("#####...#########", 5L)`.
1. $A \land B \land \lnot C \land D$: Jumps over the first gap in 
`r mark_code("#####.#..########", 3L)`.
1. $\lnot A \land \lnot B \land C \land D$: Jumps over the second gap in 
`r mark_code("#####.#..########", 7L)`.
1. $A \land \lnot B \land \lnot C \land D$: Jumps over the first gap in 
`r mark_code("#####..#.########", 4L)` ; the second gap is taken care of pattern 1.

We see that all patterns contain the term `D` which we can factor out, which leads us to 
the form 
$$
J = D \land \left(\lnot A \lor (\lnot A \land \lnot B \land \lnot C) \lor 
   (A \land B \land \lnot C) \lor (\lnot A \land \lnot B \land C) \lor
   (A \land \lnot B \land \lnot C)\right)
$$
The term $\lnot A$ makes the terms $(\lnot A \land \lnot B \land \lnot C)$ and
$(\lnot A \land \lnot B \land C)$ redundant, which leaves us with

$$
J = D \land (\lnot A \lor \underbrace{(A \land B \land \lnot C) \lor 
(A \land \lnot B \land \lnot C)}_{=:F})
$$
We can further simplify the inner term:
$$
\begin{align}
F &= 
\underbrace{(A \lor A)}_{=1} \land (A\lor \lnot B) \land (A\lor \lnot C) \land (A\lor B) 
\land \underbrace{(B \lor \lnot B)}_{=1}\land (B \lor \lnot C) \land(A \lor \lnot C) \land
(\lnot B \lor \lnot C) \land \underbrace{(\lnot C \lor \lnot C)}_{=\lnot C}\\
&=\underbrace{(A\lor \lnot B) \land (A\lor B)}_{=A} \land 
\underbrace{(A\lor \lnot C) \land(A \lor \lnot C)}_{=(A \lor \lnot C)} \land 
\underbrace{(B \lor \lnot C) \land (\lnot B \lor \lnot C)}_{=\lnot C} \land \lnot C\\
&= A \land (A\lor \lnot C) \land \underbrace{(\lnot C \land \lnot C)}_{=\lnot C}\\
&=(A \land \lnot C) \land (A \land \lnot C)\\
&= A \land \lnot C
\end{align}
$$
With this simplification, we finally obtain the formula
$$
J = D \land (\lnot A \lor (A \land \lnot C))
$$
Which we translate to springscript as follows:

```{r get-solution-1}
springscript <- c(
  "NOT C J",
  "AND A J",
  "NOT A T",
  "OR T J",
  "AND D J"
)

run_script(springscript, puzzle_data)
```


## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We start again first by screening the area and see what happens if we run the previous 
solution in `RUN` mode:

```{r map-surface-large, results = "asis"}
test_scripts_wide <- c(
  test_scripts,
  list(
    springscript
  )
)

show_holes(test_scripts_wide, puzzle_data, FALSE)
```

We observe a new pattern `r mark_code("#####.#.#...#####", 3L)`. With the existing rules 
the droid would jump too early and land on the first plateau. We can avoid this by joining 
the term $(G \land E) \lor H$ by `AND` to the previous solution:

$$
J' = D \land (\lnot A \lor (A \land \lnot C)) \land ((G \land E) \lor H)
$$
However, this reveals yet another pattern:

```{r find-next-hole, results = "asis"}
show_holes(list(c(springscript,
        "OR G T",
        "AND E T",
        "OR H T",
        "AND T J"
      )), 
  puzzle_data,
  FALSE,
  FALSE
)
```

To find a valid solution we must ensure that

1. The droid does not jump unnecessarily (to avoid landing on a piece where a
consequent jump is too late). $\lnot (A \land B \land C \land D) ensures this.
1. The jump must be safe, i.e. $D$.
1. If a second jump right after the first jump is needed, it must be safe, i.e.
$E\lor H$.

Hence, we finally come up with the following solution:

$$
J_W = D \land (E \lor H) \land \lnot(A \land B \land C \land D)
$$
(we could remove $D$ from the first term because after expanding we will 
have $. \lor(D \land \lnot D)$)

```{r get-solution-2}
springscript_wide <- c(
      "OR E J",
      "OR H J",
      "AND D J",
      "OR A T",
      "AND B T",
      "AND C T",
      "NOT T T",
      "AND T J"
    )

run_script(springscript_wide, puzzle_data, FALSE)
```

