---
params:
   task_nr: 9
   year: 2019
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{js cpp-highlighter, file = "Prism_CPP.js", echo = FALSE}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(rlang)
library(bit64)
library(igraph)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    str_split(",") %>% 
    do.call(c, .) %>% 
    as.integer64()
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We further improve our parser from [Day 7 @ 2019](2019_task7.html) and adapt it to use 
*relative parameter* mode and implement opcode *9*.

```{r get-solution-1}
split_digits <- function(n) {
  i <- floor(log10(n)) + 1L
  digits <- integer64(i)
  while (n > 0) {
    digits[i] <- n %% 10L
    n <- n %/% 10L
    i <- i - 1L
  }
  digits
}

get_op <- function(op_code) {
  digits <- split_digits(op_code)
  n <- length(digits)
  digits <- tail(c(rep(as.integer64(0L), 5L), digits), 5L)
  list(op = as.integer(sum(digits[5:4] * 10L ^ (0:1))),
       modes = digits[3:1])
}

parse_op_codes <- function(input, op_codes = puzzle_data, 
                           ip = 1L, 
                           rel_base = as.integer64(0L),
                           verbose = FALSE) {
  out_buffer <- integer64(0L)
  get_val <- function(..., modes) {
    args <- enquos(...)
    nms <- vapply(args, quo_name, character(1L))
    args <- list(...)
    modes <- modes[seq_along(args)]
    res <- map2(args, modes, \(var, mod) {
      if (mod == 0L) {
        if ((var + 1L) > length(op_codes)) {
          enlarge(var + 1L)
        }
        op_codes[as.integer(var + 1L)]
      } else if (mod == 1L) {
        var
      } else if (mod == 2L) {
        if ((rel_base + var + 1L) > length(op_codes)) {
          enlarge(rel_base + var + 1L)
        }
        op_codes[as.integer(rel_base + var + 1L)]
      }
    }) %>% 
      set_names(nms)
    res
  }
  
  get_addr <- function(res, mode) {
    if (mode == 0L) {
      if (res == 65) browser()
      as.integer(res)
    } else if (mode == 2L) {
      if ((rel_base + res) == 65) browser()
      as.integer(rel_base + res)
    } else {
      stop("unsopported mode for output")
    }
  }
  
  enlarge <- function(idx) {
    gap <- as.integer(idx) - length(op_codes)
    if (gap > 0L) {
      op_codes <<- c(op_codes, rep(as.integer64(0L), gap))
    }
  }
  
  add <- function(x, y, res, modes) {
    args <- get_val(x, y, modes = modes)
    x <- args$x
    y <- args$y
    res <- get_addr(res, modes[3L])
    enlarge(res + 1L)
    op_codes[[res + 1L]] <<- x + y
    ip <<- ip + 4L
    TRUE
  }
  mul <- function(x, y, res, modes) {
    args <- get_val(x, y, modes = modes)
    x <- args$x
    y <- args$y
    res <- get_addr(res, modes[3L])
    enlarge(res + 1L)
    op_codes[[res + 1L]] <<- x * y
    ip <<- ip + 4L
    TRUE
  }
  set <- function(res, ..., modes) {
    if (length(input) == 0L) {
      return (FALSE)
    }
    res <- get_addr(res, modes[1L])
    enlarge(res + 1L)
    op_codes[[res + 1L]] <<- head(input, 1L)
    ip <<- ip + 2L
    input <<- tail(input, -1L)
    TRUE
  }
  out <- function(x, ..., modes) {
    args <- get_val(x, modes = modes)
    x <- args$x
    out_buffer <<- c(out_buffer, x)
    if (verbose) {
      cat(x, "\n")
    }
    ip <<- ip + 2L
    TRUE
  }
  jnz <- function(x, y, ..., modes) {
    args <- get_val(x, y, modes = modes)
    x <- args$x
    y <- args$y
    if (x != 0L) {
      ip <<- as.integer(y + 1L)  
    } else {
      ip <<- ip + 3L
    }
    TRUE
  }
  jeqz <- function(x, y, ..., modes) {
    args <- get_val(x, y, modes = modes)
    x <- args$x
    y <- args$y
    if (x == 0L) {
      ip <<- as.integer(y + 1L)  
    } else {
      ip <<- ip + 3L
    }
    TRUE
  }
  lt <- function(x, y, res, modes) {
    args <- get_val(x, y, modes = modes)
    x <- args$x
    y <- args$y
    res <- get_addr(res, modes[3L])
    enlarge(res + 1L)
    op_codes[[res + 1L]] <<- (x < y)
    ip <<- ip + 4L
    TRUE
  }
  eq <- function(x, y, res, modes) {
    args <- get_val(x, y, modes = modes)
    x <- args$x
    y <- args$y
    res <- get_addr(res, modes[3L])
    enlarge(res + 1L)
    op_codes[[res + 1L]] <<- (x == y)
    ip <<- ip + 4L
    TRUE
  }
  rbo <- function(x, ..., modes) {
    args <- get_val(x, modes = modes)
    x <- args$x
    rel_base <<- rel_base + x
    ip <<- ip + 2L
    TRUE
  }
  
  ops <- list(add, mul, set, out, jnz, jeqz, lt, eq, rbo)
  n <- length(op_codes)
  op <- get_op(op_codes[ip])
  done <- TRUE
  while(op$op %in% c(seq_along(ops), 99L)) {
    args <- as.list(op_codes[seq(ip + 1L, length.out = 3L)])
    cont <- do.call(ops[[op$op]], c(args, list(modes = op$modes)))
    if (!cont) {
      done <- FALSE
      break
    }
    op <- get_op(op_codes[ip])
    if (op$op == 99L) {
      break
    } else if (!op$op %in% seq_along(ops)) {
      cat("Unknown op code", op$op, "\n")
    }
  }
  out_buffer
}

parse_op_codes(1L, puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

While the original parser works and is very R idomatic, it is awefully slow. We refactor 
the code to gain a significant speed gain. Additionally, we implemented the code also in
C++ (in the appendix) for possible future use.

```{r get-solution-2}
parse_op_codes_fast <- function(op_codes = as.numeric(puzzle_data), input = numeric(), 
                                verbose = FALSE) {
  ip <- 1L
  rel_base <- 0L
  out_buffer <- numeric(0)
  out_len <- 0L
  out_block <- 1024L
  block_size <- 1024L  # Speicherblockgröße zum Wachsen
  
  grow_memory <- function(idx) {
    if (idx > length(op_codes)) {
      new_size <- ceiling(idx / block_size) * block_size
      op_codes <<- c(op_codes, rep(0, new_size - length(op_codes)))
    }
  }
  
  append_out <- function(val) {
    if (out_len + 1L > length(out_buffer)) {
      new_size <- length(out_buffer) + out_block
      out_buffer <<- c(out_buffer, rep(0, new_size - length(out_buffer)))
    }
    out_len <<- out_len + 1L
    out_buffer[out_len] <<- val
  }
  
  get_val <- function(param, mode) {
    if (mode == 0L) {
      grow_memory(param + 1L)
      return(op_codes[param + 1L])
    } else if (mode == 1L) {
      return(param)
    } else if (mode == 2L) {
      idx <- rel_base + param + 1L
      grow_memory(idx)
      return(op_codes[idx])
    } else {
      stop("unknown mode")
    }
  }
  
  get_addr <- function(param, mode) {
    if (mode == 0L) {
      return(param + 1L)
    } else if (mode == 2L) {
      return(rel_base + param + 1L)
    } else {
      stop("invalid write mode")
    }
  }
  
  halt <- FALSE
  while (!halt) {
    instr <- op_codes[ip]
    op <- instr %% 100L
    modes <- c(
      (instr %/% 100L) %% 10L, 
      (instr %/% 1000L) %% 10L, 
      (instr %/% 10000L) %% 10L
    )
    
    if (op == 1L) { # add
      a <- get_val(op_codes[ip + 1L], modes[1])
      b <- get_val(op_codes[ip + 2L], modes[2])
      addr <- get_addr(op_codes[ip + 3L], modes[3])
      grow_memory(addr)
      op_codes[addr] <- a + b
      ip <- ip + 4L
    } else if (op == 2L) { # mul
      a <- get_val(op_codes[ip + 1L], modes[1])
      b <- get_val(op_codes[ip + 2L], modes[2])
      addr <- get_addr(op_codes[ip + 3L], modes[3])
      grow_memory(addr)
      op_codes[addr] <- a * b
      ip <- ip + 4L
    } else if (op == 3L) { # input
      if (length(input) == 0L) {
        stop("No input available")
      }
      addr <- get_addr(op_codes[ip + 1L], modes[1])
      grow_memory(addr)
      op_codes[addr] <- input[1L]
      input <- input[-1]
      ip <- ip + 2L
    } else if (op == 4L) { # output
      val <- get_val(op_codes[ip + 1L], modes[1])
      append_out(val)
      if (verbose) cat(val, "\n")
      ip <- ip + 2L
    } else if (op == 5L) { # jump-if-true
      a <- get_val(op_codes[ip + 1L], modes[1])
      b <- get_val(op_codes[ip + 2L], modes[2])
      if (a != 0L) {
        ip <- b + 1L
      } else {
        ip <- ip + 3L
      }
    } else if (op == 6L) { # jump-if-false
      a <- get_val(op_codes[ip + 1L], modes[1])
      b <- get_val(op_codes[ip + 2L], modes[2])
      if (a == 0L) {
        ip <- b + 1L
      } else {
        ip <- ip + 3L
      }
    } else if (op == 7L) { # less than
      a <- get_val(op_codes[ip + 1L], modes[1])
      b <- get_val(op_codes[ip + 2L], modes[2])
      addr <- get_addr(op_codes[ip + 3L], modes[3])
      grow_memory(addr)
      op_codes[addr] <- as.numeric(a < b)
      ip <- ip + 4L
    } else if (op == 8L) { # equals
      a <- get_val(op_codes[ip + 1L], modes[1])
      b <- get_val(op_codes[ip + 2L], modes[2])
      addr <- get_addr(op_codes[ip + 3L], modes[3])
      grow_memory(addr)
      op_codes[addr] <- as.numeric(a == b)
      ip <- ip + 4L
    } else if (op == 9L) { # adjust relative base
      a <- get_val(op_codes[ip + 1L], modes[1])
      rel_base <- rel_base + a
      ip <- ip + 2L
    } else if (op == 99L) { # halt
      halt <- TRUE
    } else {
      stop("Unknown opcode ", op)
    }
  }
  return(out_buffer[1:out_len])
}

parse_op_codes_fast(as.numeric(puzzle_data), 2)
```


# C++ Version

```{Rcpp cpp-algorithm, file = "parser.cpp", cache = TRUE}
```
