---
params:
   task_nr: 8
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(knitr)
library(cli)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  text_block %>% 
    unlist() %>% 
    keep(nzchar) %>% 
    str_split("") %>% 
    do.call(rbind, .)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    str_split("\n") %>% 
    parse_puzzle_data()
})

example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

The approach is rather straight forward, while the only interesting part is that we encode
coordinates using complex numbers (where the real and imaginary part corresponds to the x
and y coordinates respectively). This allows for quite concise code as we do not have to 
deal with 2 dimensional data directly but can treat the coordinates as a single number 
(which makes filtering and mathematical operations a breeze - though in all fairness R is
rather good at vectorization anyways):

1. Get the antennae coordinates and split them according to their frequency symbol.
1. For each pair of antennae of the same frequency calculate the antinodes storing the
position using complex numbers as highlighted above.
1. Filter the new positions to be inside the map.

```{r get-solution-1}
get_antenna_locations <- function(antenna_map) {
  dd <- dim(antenna_map)
  locs <- !grepl(".", antenna_map, fixed = TRUE)
  dim(locs) <- dd
  idx <-which(locs, arr.ind = TRUE)
  antenna_list <- idx %>% 
    split(antenna_map[idx]) %>% 
    map(~ matrix(.x, ncol = 2))
  antenna_list
}

get_antinodes <- function(antenna_map) {
  dd <- dim(antenna_map)
  antenna_list <- get_antenna_locations(antenna_map)
  get_new_pos <- function(antenna_pos) {
    nr <- nrow(antenna_pos)
    dist_matrix <- matrix(NA_complex_, nr, nr)
    off_diag <- which(lower.tri(dist_matrix) | upper.tri(dist_matrix), 
                      arr.ind = TRUE)
    new_points <- apply(off_diag, 1, function(idx) {
      p1 <- antenna_pos[idx[1L], , drop = FALSE]
      p2 <- antenna_pos[idx[2L], , drop = FALSE]
      np <- 2L * p1 - p2
      ## store new point as complex number
      ## Re part is the x coordinate and the Im Part the y coordinate
      complex(real = np[, 1], imaginary = np[, 2])
    }) %>% 
      unique()
    
    new_points[between(Re(new_points), 1L, dd[1L]) &
                 between(Im(new_points), 1L, dd[2L])]
  }
  map(antenna_list, get_new_pos) %>% 
    unlist() %>% 
    unique()
}
get_antinodes(puzzle_data) %>% 
  length()
```


## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

With the selected parametrization, the task is as easy as adding the distance of the 
points mutiplied by a vector of scalars. We could optimize the solution by avoiding to
calculate out of bounds points, but given the cheap function here, we simply use 
corresponding filters to avoid those points.

```{r get-solution-2}
get_resonant_antinodes <- function(antenna_map) {
  dd <- dim(antenna_map)
  antenna_list <- get_antenna_locations(antenna_map)
  get_new_pos <- function(antenna_pos) {
    nr <- nrow(antenna_pos)
    dist_matrix <- matrix(NA_complex_, nr, nr)
    off_diag <- which(lower.tri(dist_matrix) | upper.tri(dist_matrix), 
                      arr.ind = TRUE)
    new_points <- apply(off_diag, 1, function(idx) {
      p1 <- antenna_pos[idx[1L], , drop = FALSE]
      p2 <- antenna_pos[idx[2L], , drop = FALSE]
      d <- p2 - p1
      np <- map(1:max(dd), ~ p1 + .x * d) %>% 
        do.call(rbind, .)
      complex(real = np[, 1], imaginary = np[, 2])
    }, simplify = FALSE) %>% 
      do.call(c, .) %>% 
      unique()
    new_points[between(Re(new_points), 1L, dd[1L]) &
                 between(Im(new_points), 1L, dd[2L])]
  }
  map(antenna_list, get_new_pos) %>% 
    unlist() %>% 
    unique()
}
get_resonant_antinodes(puzzle_data) %>% 
  length()
```
