---
params:
   task_nr: 2
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(tibble)
library(dplyr)
library(purrr)
library(stringr)
library(igraph)
library(kableExtra)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    str_split("\n+") %>% 
    extract2(1L) %>% 
    str_split("\\s+") %>% 
    lapply(as.integer) %>% 
    discard(~ any(is.na(.x)))
})
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

To investigate whether a sequence is considered "safe" we simply look at the differences
between consecutive numbers. If they all bear the same sign and are between `-3` and `3` 
the report can be considered safe. To get the solution we count how many reports were 
safe.

```{r get-solution-1}
is_safe <- function(report) {
  diffs <- diff(report)
  is_ordered <- abs(sum(sign(diffs))) == length(diffs)
  is_right_size <- all(abs(diffs) <= 3L)
  is_ordered && is_right_size
}

sum(vapply(puzzle_data, is_safe, logical(1L)))
```


## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

To solve the second part we simply apply `is_safe` to the whole report $r$, as well as 
each sub-report $r_i$, which is generated by removing the *i*th element from $r$.

```{r get-solution-2}
is_safe_robust <- function(report) {
  is_safe(report) ||
    any(vapply(seq_along(report), \(i) is_safe(report[-i]), logical(1L)))
}

sum(vapply(puzzle_data, is_safe_robust, logical(1L)))
```
