---
params:
   task_nr: 16
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
  })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}

code em {
font-weight:bold
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
	if (knitr::is_latex_output()) {
		sprintf("\\texttt{%s}", as.character(x))
	} else if (knitr::is_html_output()) {
		sprintf("<tt>%s</tt>", as.character(x))
	} else {
		x
	}
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(tibble)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2021/day/", params$task_nr)
puzzle <- GET(base_url,
				  session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	xml_find_all("///article") %>% 
	lapply(as.character)

puzzle_data <- GET(paste0(base_url, "/input"),
						 session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	str_remove("\n")
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We first decode the HEX string into binary. Then we use a recursive function, 
which takes the starting position for the packet as an argument. If the packet 
is an operator packet, we recurse into the sub packets and store the intersting
parts in a list of liust structure, where operator packets have a `child` slot, 
which holds all the child packets.

```{r solution-a, cache = FALSE}
hex2bin <- function(code = puzzle_data) {
	res <- code %>% 
		str_split("") %>%
		`[[`(1L) %>% 
		str_replace_all("([0-9A-F])", function(hc) {
			strtoi(hc, 16L) %>% 
				intToBits() %>% 
				rev() %>% 
				as.integer() %>% 
				tail(4) %>% 
				str_c(collapse = "")
		}) %>% 
		str_split("") %>% 
		unlist() %>% 
		as.integer()
	class(res) <- "bitstream"
	res
}

bin2dec <- function(bitstream) {
	stopifnot(all(bitstream %in% 0:1))
	sum(bitstream * rev(2 ^ (seq_along(bitstream) - 1)))
}

parse_packets <- function(bitstream) {
	do_parse <- function(start) {
		version <- bitstream[seq(start, length.out = 3L)] %>% 
			bin2dec() %>% 
			as.integer()
		type_id <- bitstream[seq(start + 3L, length.out = 3L)] %>% 
			bin2dec()
		if (type_id == 4L) {
			## literal value
			idx <- seq(start + 6L, length(bitstream), by = 5L)
			value_end <- idx[which.max(bitstream[idx] == 0L)] + 4L
			value <- bitstream[(start + 6L):value_end]
			## remove leading bits
			idx <- seq(1L, length(value), 5L)
			value <- list(value = value[-idx] %>% 
							  	bin2dec(),
							  version = version,
							  type = type_id,
							  packet_length = value_end - start + 1L)
			class(value) <- c("packet", "literal")
		} else {
			## operator
			length_type_id <- bitstream[start + 6L]
			if (length_type_id == 0L) {
				## total length
				len <- bitstream[seq(start + 7L, length.out = 15L)] %>% 
					bin2dec()
				iter <- start + 22L
				total_len <- 22L
				cond <- quote(iter < start + len + 22L)
			} else {
				## packet length
				total_cnt <- bitstream[seq(start + 7L, length.out = 11L)] %>%
					bin2dec()
				iter <- start + 18L
				total_len <- 18L
				cond <- quote(cnt <= total_cnt)
			}
			childs <- list()
			cnt <- 1L
			while (eval(cond)) {
				next_pack <- Recall(iter)
				pack_len <- next_pack$packet_length
				iter <- iter + pack_len
				total_len <- total_len + pack_len
				childs <- c(childs, list(next_pack))
				cnt <- cnt + 1L
			}
			value <- list(childs = childs,
							  version = version,
							  type = type_id,
							  packet_length = total_len,
							  length_type = length_type_id)
			class(value) <- c("packet", "operator", 
									if (length_type_id == 0L) 
										"total_length" else "total_packets")
		}
		value
	}
	do_parse(1L)
}


hier <- puzzle_data %>% 
	hex2bin() %>% 
	parse_packets()

sum_versions <- function(hier) {
	if (is.null(hier$childs)) {
		hier$version
	} else {
		hier$version + sum(vapply(hier$childs, sum_versions, integer(1L)))
	}
}

sum_versions(hier)
```


## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For the second part, we have to walk the parsed string recursively and execute 
the indicated operations.

```{r solution-b, eval = length(puzzle) > 1, comment = NA}
eval_hierachy <- function(hier = hier) {
	if (hier$type == 4L) {
		hier$value
	} else {
		fns <- list(
			sum,
			prod,
			min,
			max,
			## this should never be called in this branch, we add it to 
			## avoid additional if clasue
			stop, 
			\(x, y) as.integer(x > y),
			\(x, y) as.integer(x < y),
			\(x, y) as.integer(x == y)
		)
		fn <- fns[[hier$type + 1]]
		args <- lapply(hier$childs, eval_hierachy)
		do.call(fn, unname(args))
	}
}
eval_hierachy(hier)
```