---
params:
   task_nr: 9
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(knitr)
library(cli)
library(bit64)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  text_block %>% 
    unlist() %>% 
    str_remove_all("\n") %>% 
    str_split("") %>% 
    extract2(1L) %>% 
    as.integer()
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```
```{r get_example_data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

The idea is as follows:

1. We know that in the end there is a consecutive block with files, followed by a block
with empty space. The length of the files block equals the sum of all file blocks.
1. Thus, we first generate the position of the file blocks in the original disk image. 
1. Then, we compare the original file block positions with the file block size:
1. If the position is smaller than the final block size, the block keeps its index.
1. If it is greater than the final block size, we assign empty block indices to it. We
follow the rule that we assign the smallest indices to the latest files.
1. Eventually, we have the final block position together with the file id and can 
multiply those numbers.

```{r get-solution-1}
get_layout <- function(disk_map) {
  if (length(disk_map) %% 2 == 0) {
    fill_me <- NULL
  } else {
    fill_me <- NA_integer_
  }
  tibble(
    files = disk_map[seq(1L, length(disk_map), 2L)],
    free_space = c(disk_map[seq(2L, length(disk_map), 2L)], fill_me)
  ) %>% 
    mutate(file_id = 0:(n() - 1L), .before = 1L) %>% 
    mutate(start = lag(cumsum(files + free_space), default = 0L),
           indices = map2(start, files, ~ seq(.x, length.out = .y)))
}

checksum <- function(disk_map) {
  layout <- get_layout(disk_map) 
  indices <- layout %>% 
    reframe(file_id = rep(file_id, lengths(indices)),
            original_index = unlist(indices),
            has_space = original_index <= sum(lengths(indices)))
  empty_indices <- seq(0, max(indices %>% pull(original_index))) %>% 
    setdiff(indices %>% pull(original_index)) %>% 
    extract(seq(1, sum(!(indices %>% pull(has_space))))) %>% 
    rev()
  indices %>% 
    mutate(final_index = c(original_index[has_space],
                           empty_indices)) %>% 
    summarize(chk_sum = sum(as.integer64(file_id) *
                              as.integer64(final_index))) %>% 
    pull(chk_sum)
}

checksum(puzzle_data)
```


## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

This time we use an iterative process, where we try to move each file to the left, if 
there are some free indices which can house the file blocks. 


```{r get-solution-2}
checksum_defragmented <- function(disk_map) {
  layout <- get_layout(disk_map) %>% 
    mutate(free_indices = map2(start + files, 
                               free_space,
                               ~ seq(.x, length.out = coalesce(.y, 0L))),
           final_indices = rep(list(integer64(0L)), n()))
  for (row_index in seq(nrow(layout), 1L)) {
    record <- layout %>% 
      slice(row_index)
    size_needed <- record %>% 
      pull(files)
    size_available <- layout %>% 
      pull(free_space)
    slot_indices <- size_available >= size_needed & 
      seq_along(size_available) < row_index
    if (any(slot_indices)) {
      slot_index <- which.max(slot_indices)
      slot <- layout %>% 
        slice(slot_index)
      avail_indices <- slot %>% 
        pull(free_indices) %>% 
        unlist()
      used_indices <- head(avail_indices, size_needed)
      remaining_indices <- setdiff(avail_indices, used_indices)
      ## store the found indices for the current record
      layout[row_index, "final_indices"] <- list(list(as.integer64(used_indices)))
      ## remove the used indices from the found slot
      layout[slot_index, "free_indices"] <- list(list(remaining_indices))
      ## update the free_space colum
      layout[slot_index, "free_space"] <- length(remaining_indices)
    } else {
      layout[row_index, "final_indices"] <- map(layout[row_index, "indices"], 
                                                ~ map(.x, as.integer64))
    }
  }
  ## need this do.call(c, .) construct because an unlist would destroy the integer64s
  layout %>%
    summarize(chk_sum = sum(rep(file_id, files) * do.call(c, final_indices))) %>%
    pull(chk_sum)
}
checksum_defragmented(puzzle_data)
```
