---
params:
   task_nr: 19
   year: 2020
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{js cpp-highlighter, file = "Prism_CPP.js", echo = FALSE}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  rules_marker <- str_which(text_block, ":") %>% 
    max()
  rules <- text_block[1:rules_marker]
  lhs <- str_extract(rules, "^\\d+") %>% 
    as.integer() %>% 
    add(1L)
  rhs <- str_remove_all(rules, "^\\d+: |\"") %>% 
    str_split(fixed(" | ")) %>% 
    map(function(rhs) {
      parts <- str_split(rhs, "\\s")
      map(parts, function(x) {
        xv <- suppressWarnings(as.integer(x)) 
        if (any(is.na(xv))) {
          x
        } else {
          xv + 1L
        }
      })
    })
  list(rules = rhs[order(lhs)], msg = text_block[(rules_marker + 1L):length(text_block)])
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We use use the 
[Cocke–Younger–Kasami algorithm (CYK)](https://en.wikipedia.org/wiki/CYK_algorithm) on 
this grammar to determine whether a word can be generated by it. Thus, the grammar needs 
to be in [Chomsky normal form (CNF)](https://en.wikipedia.org/wiki/Chomsky_normal_form).

As it happens the grammar is *almost* in CNF, but we need to remove some unit production
rules first. We also take care of chain rules in case they occur (will be needed for part
2).

We have a working implementation in R which is rather slow (it can be found in the 
appendix). Thus we re-implemented the `cyk` part in C++ for a substantial speed gain.

```{Rcpp cpp-algorithm, file = "cyk.cpp", cache = TRUE}
```


```{r get-solution-1}
create_cnf <- function(rules) {
  names(rules) <- as.character(seq_along(rules))
  unit_rules <- keep(rules, ~ any(lengths(.x) == 1L)) 
  for (lhs in names(unit_rules)) {
    rule <- rules[[lhs]]
    new_rhs <- NULL
    for (i in seq_along(rule)) {
      rhs <- rule[[i]]
      if (length(rhs) == 1L && is.integer(rhs)) {
        new_rhs <- c(new_rhs, rules[[as.character(rhs)]])
      } else {
        new_rhs <- c(new_rhs, rule[i])
      }
    }
    rules[[lhs]] <- new_rhs
  } 
  chain_rules <- keep(rules, ~ any(lengths(.x) > 2L))
  if (length(chain_rules) > 0L) {
    for (lhs in names(chain_rules)) {
      rule <- rules[[lhs]]
      new_rhs <- NULL
      add_rules <- list()
      n <- length(rules)
      for (i in seq_along(rule)) {
        rhs <- rule[[i]]
        if (length(rhs) == 3L) {
          new_rhs <- c(new_rhs, list(c(rhs[1L], n + 1L)))
          add_rules <- c(add_rules, 
                         list(
                           list(rhs[-1L]))
          )
          n <- n + 1L
        } else {
          new_rhs <- c(new_rhs, rule[i])
        }
      }
      rules[[lhs]] <- new_rhs
    }
    rules <- c(rules, add_rules)
  }
  unname(rules)
}

make_key <- function(...) {
  paste(c(...), collapse = "|")
}

invert_rules <- function(rules) {
  rls <- new.env(parent = emptyenv())
  for (i in seq_along(rules)) {
    rule <- rules[[i]]
    for (r in rule) {
      stopifnot(length(r) %in% 1:2)
      rls[[make_key(r)]] <- c(i, rls[[make_key(r)]])
    }
  }
  rls
}

cyk <- function(words, rules, start = 1L) {
  cnf_rules <- create_cnf(rules)
  inv_rules <- invert_rules(cnf_rules)
  is_member(inv_rules, words, start) %>% 
    sum()
}

cyk(puzzle_data$msg, puzzle_data$rules) 
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]] %>% str_replace_all(fixed("*"), "\\*"))
```

### Solution

For the second part, we need to adapt the rules. Because we followed R's 1-based paradigm 
all indices need to be shifted by 1. Now we also have to take care about the rule with 
**three** elements on the right hand side and need to split them accordingly to maintain 
a CNF.

```{r get-solution-2}
new_rules <- puzzle_data$rules
new_rules[[9L]] <- list(
  43L,
  c(43L, 9L)
)
new_rules[[12L]] <- list(
  c(43L, 32L),
  c(43L, 12L, 32L)
)

cyk(puzzle_data$msg, new_rules)
```

# R Legacy Solution

```{r legacy-code, eval = FALSE}
cyk <- function(words, rules, start = 1L) {
  cnf_rules <- create_cnf(rules)
  inv_rules <- invert_rules(cnf_rules)
  is_member <- function(word) {
    chars <- str_split(word, "")[[1L]]
    n <- length(chars)
    V <- vector("list", n * n)
    dim(V) <- c(n, n)
    for (i in 1:n) {
      V[[i, 1]] <- inv_rules[[chars[i]]]
    }
    for (j in 2:n) {
      for (i in 1:(n + 1L - j)) {
        cell <- integer(0L)
        for (k in 1:(j - 1)) {
          B <- V[[i, k]]
          C <- V[[i + k, j - k]]
          if (length(B) == 0 || length(C) == 0) {
            next
          }
          
          for (b in B) {
            for (c in C) {
              rhs <- inv_rules[[make_key(b, c)]]
              if (!is.null(rhs)) {
                cell <- c(cell, rhs)
              }
            }
          }
        }
        cell <- unique(cell)
        if (length(cell) > 0L) {
          V[[i, j]] <- cell
        }
      }
    }
    start %in% V[[1L, n]]
  }
  res <- imap_lgl(words, ~ {cat("Testing word ", .y, "...\r", sep =""); is_member(.x)}) 
  cat("\n")
  sum(res)
}
```
