---
params:
   task_nr: 3
   year: 2025
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L)
  }
  res <- text_block[nzchar(text_block)] %>% 
    str_extract_all("\\d") %>% 
    do.call(rbind, .)
  storage.mode(res) <- "integer"
  res
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We use a recursive function where we loop through the positions of the digits from `9` to 
`1`. If a position is in increasing order and there is still enuogh space for the 
remaining digits we use it and repeat this process until all required batteries are found.

```{r get-solution-1}
find_max_joltage <- function(banks, nr_bat) {
  find_joltage <- function(bank) {
    digits <- 9:1
    n <- length(bank)
    positions_by_digit <- map(digits, ~ which(bank == .x))
    backtrack <- function(selected_pos) {
      remaining <- nr_bat - length(selected_pos)
      if (remaining == 0L) {
        return (selected_pos)
      }
      for (digit_positions in positions_by_digit) {
        for (pos in digit_positions) {
          enough_space <- (n - pos) >= (remaining - 1L)
          increasing <- pos > max(c(selected_pos, -Inf))
          if (enough_space && increasing) {
            return(Recall(c(selected_pos, pos)))
          }
        }
      }
    }
    res <- backtrack(integer(0L))
    sum(bank[res] * 10 ^ ((nr_bat - 1L):0))
  }
  apply(banks, 1L, find_joltage) %>% 
    sum()
}
find_max_joltage(puzzle_data, 2L)


```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We simply need to call the function with a larger number of required batteries.

```{r get-solution-2}
find_max_joltage(puzzle_data, 12L) %>% 
  print(digit = 12)
```
