---
params:
   task_nr: 14
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(knitr)
library(cli)
library(bit64)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- str_split(text_block, "\n")
  }
  text_block <- text_block %>% 
    unlist()
  res <- text_block %>% 
    str_extract_all("-?\\d+") %>% 
    do.call(rbind, .) %>% 
    set_colnames(c("x", "y", "delta_x", "delta_y"))
  storage.mode(res) <- "integer"
  res
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
attr(puzzle_data, "size") <- c(101L, 103L)
```

```{r get_example_data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
attr(example_data, "size") <- c(11L, 7L)
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

The solution is straight forward. Add the velocity multiplied by the seconds to the 
starting position and take the results modulo the grid size and you get the final 
position.

```{r get-solution-1}
visualize_puzzle <- function(pos, char = "#") {
  size <- attr(pos, "size")
  m <- matrix(".", size[2L], size[1L])
  pos <- pos %>% 
    select("y", "x") %>% 
    as.matrix()
  m[pos[, c("y", "x"), drop = FALSE] + 1L] <- char
  apply(m, 1, paste, collapse = "") %>% 
    paste(collapse = "\n") %>% 
    cat("\n")
}

move_robots <- function(all_robots, secs) {
  get_quadrant <- function(pos, size) {
    if (pos[1L] == ((size[1L] - 1L) / 2) || 
        pos[2L] == ((size[2L] - 1L) / 2L)) {
      0L
    } else {
      if (pos[1L] <= size[1L] / 2L) {
        if (pos[2L] <= size[2L] / 2) {
          1L
        } else {
          2L
        }
      } else {
        if (pos[2L] <= size[2L] / 2) {
          3L
        } else {
          4L
        }
      }
    }
  }
  
  move_robot <- function(a_robot, secs, size) {
    pos <- a_robot[c("x", "y")]
    vel <- a_robot[c("delta_x", "delta_y")]
    res <- (pos + secs * vel) %% size %>% 
      matrix(nrow = 1L) %>% 
      set_colnames(c("x", "y"))
    res <- cbind(res,
                 quadrant = get_quadrant(res, size))
    attr(res, "size") <- size
    res
  }
  
  size <- attr(all_robots, "size")
  
  res <- apply(all_robots, 1L, move_robot, secs, size) %>% 
    t() %>% 
    set_colnames(c("x", "y", "quadrant")) %>% 
    as_tibble()
  attr(res, "size") <- size
  res
}


res <- move_robots(puzzle_data, 100L) 

res %>% 
  filter(quadrant != 0) %>% 
  group_by(quadrant) %>% 
  summarize(n = n()) %>% 
  pull(n) %>% 
  prod()

```


## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We have no idea how the tree should look like, but it is safe to assume that in any 
picture robots are somehow connected to form a picture. Thus, robots cluster in the same
area. We can thus measure the variance in both directions and summing them up. The index
with the lowest sum could be a candidate where there is some structure in the picture.

```{r get-solution-2, cache = TRUE}
vars <- map_dbl(
  1:10000,
  \(i) move_robots(puzzle_data, i) %>% 
    summarize(v = var(x) + var(y)) %>% 
    pull(v)
)

(sol <- which.min(vars))
```

```{r visualize-solution}
visualize_puzzle(move_robots(puzzle_data, sol))
```

