---
params:
   task_nr: 17
   year: 2015
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2015"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  as.integer(text_block)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- c(20L, 15L, 10L, 5L, 5L)
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We solve this puzzle via dynamic programming. We create a vector where each entry 
represents the amount of ways to reach a sum equal to the index. That is $z_0$ indicates
the amount of ways to build a sum equal 0. We set $z_0$ to 1 (there is only one way to
build the sum 0, namely by adding no numbers as all numbers are greater than 0).

We then use two loops, the first iterates over each number, the second loops from the 
maximum capacity down to the current number. Let's assume that $x_i$ denotes the current
number from the list. Now to count the number of ways to reach a sum of $y$, we simply
have to add the number of ways how we reach  $y-x_i$ to get the proper count.

```{r get-solution-1}
count_sums <- function(numbers, capacity) {
  nr_ways <- vector("integer", capacity + 1L)
  nr_ways[1L] <- 1L
  walk(numbers, function(nr) {
    for (i in capacity:nr) {
      ## idea: 
      ## 1. res[i + 1L] counts the number of ways we can build the sum `i` (the plus one
      ##    is there because R is 1-based)
      ## 2. Likewise res[i - nr + 1L] counts the number of ways we can build sum `i - nr`
      ## 3. We are currently examining summand `nr`, thus we can add `nr` to `i - nr` to
      ##    end up at `i`. Hence, we simply add the number of ways to reach `i - nr` to
      ##    the number of ways we reached `i` (in earlier iterations) 
      nr_ways[i + 1L] <<- nr_ways[i + 1L] + nr_ways[i - nr + 1L] 
    }
  })
  tail(nr_ways, 1L)
}
count_sums(puzzle_data, 150)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

This time we need also to keep track about the amount of summands. The algorithm follows
largely the same ideas as before, but this time we check if the previous sum was reached
with less then current number of operands plus one, if so, we take this number of ways as 
new minimum and update teh values accordingly. If we are already on the track of smallest
number of summands, we update the number of ways as before by adding the value of the 
partial sum.

```{r get-solution-2}
count_min_sums <- function(numbers, capacity) {
  nr_operands <- rep(Inf, capacity + 1L)
  nr_ways <- rep(0L, capacity + 1L) 
  
  nr_operands[1L] <- 0L
  nr_ways[1L] <- 1L
  walk(numbers, function(nr) {
    for (i in capacity:nr) {
      if (nr_operands[i - nr + 1L] + 1L < nr_operands[i + 1L]) {
        ## we got `i - nr` with less summands minus one (the one we would be adding now)
        ## update teh numebr of operands (one more) and overwrite the number of ways 
        ## (we want to count only the minumum, hence not adding but simply overwriting)
        nr_operands[i + 1L] <<- nr_operands[i - nr + 1L] + 1L
        nr_ways[i + 1L] <<- nr_ways[i - nr + 1L]
      } else if (nr_operands[i - nr + 1L] + 1L == nr_operands[i + 1L]) {
        ## same as before if we are on the track of smallest number of operands
        nr_ways[i + 1L] <<- nr_ways[i + 1L] + nr_ways[i - nr + 1L]
      }
    }
  }
  )
  list(min_count = nr_operands[capacity + 1L], min_ways = nr_ways[capacity + 1L])
}
count_min_sums(puzzle_data, 150L)
```
