---
params:
   task_nr: 23
   year: 2015
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2015"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(tidyr)
library(igraph)
library(glue)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block 
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

To understand the algorithm, we draw it first as a graph.

```{r make-graph, fig.height = 12}
make_flow_chart <- function(ops) {
  parse_nodes <- function(op) {
    str_match_all(op, "(...) ([^ ,]+)(?:, )?([-+]\\d+)?") %>% 
      do.call(rbind, .) %>% 
      set_colnames(c("string", "op", "reg", "offset")) %>% 
      as_tibble() %>% 
      mutate(id = 1:n(), .before = 1L) %>% 
      mutate(offset = if_else(op == "jmp",  reg, offset),
             reg = if_else(op == "jmp", NA_character_, reg),
             offset = as.integer(offset),
             label = case_when(
               op == "jio" ~ glue("{reg} == 1?"),
               op == "jie" ~ glue("{reg} %% 2 == 0?"),
               op == "inc" ~ glue("{reg}++"),
               op == "tpl" ~ glue("{reg} = 3\U00B7{reg}"),
               op == "hlf" ~ glue("{reg} = {reg} / 2")
             ),
             target = if_else(!is.na(offset), id + offset, NA_integer_))

  }
  ops_data <- parse_nodes(ops)
  ops_data_no_jmp <- ops_data %>% 
    filter(op != "jmp")
  n <-  nrow(ops_data_no_jmp) + 1L ## for exit
  G <- make_empty_graph(n)
  V(G)$name <- ops_data_no_jmp %>% 
    pull(id) %>% 
    paste0("E", .) %>% 
    c(glue("E{nrow(ops_data) + 1L}"))
  V(G)$type <- ops_data_no_jmp %>% 
    pull(op) %>% 
    c("out")
  V(G)$label <- ops_data_no_jmp %>% 
    pull(label) %>% 
    c("Ouput b")
  V(G)$shape <- "rectangle"
  V(G)$size <- case_match(
    V(G)$type,
    "inc" ~ 40L,
    c("tpl", "hlf", "jio") ~ 60L,
    .default = 100L)
  V(G)$size2 <- 65L
  V(G)$color <- case_when(
    V(G)$label == "b++" ~ "firebrick",
    V(G)$type %in% c("jio", "jie") ~ "steelblue",
    V(G)$type %in% c("inc", "tpl", "hlf") ~ "gray80",
    V(G)$type == "out" ~ "forestgreen"
    
  ) 
  V(G)$label.color <- if_else(
    V(G)$color == "gray80", "black", "white"
  )
  double_jmp <- ops_data %>% 
    inner_join(ops_data, c(target = "id")) %>%
    filter(op.y == "jmp") %>% 
    mutate(id, target = target.y, .keep = "none")
  ops_data <- ops_data %>% 
    rows_update(double_jmp, "id")
  edges <- ops_data %>% 
    mutate(
      edge_str = if_else(lead(op) == "jmp", 
                         glue("{id},{lead(target)}"),
                         glue("{id},{if_else(is.na(target), id + 1L, target)}")
      ),
      edge_str = if_else(op %in% c("jio", "jie"),
                         glue("{edge_str},{id},{id + 1L}"),
                         edge_str)
      ) %>% 
    filter(op != "jmp") %>%
    pull(edge_str) %>%
    str_extract_all("\\d+") %>%
    unlist() %>%
    paste0("E", .)
  
  G <- G %>% 
     add_edges(edges)
  E(G)$arrow.size <- .75
  E(G)$arrow.width <- .75
  lay <- matrix(c(0L, -2L, -2L, -2L, -2L, -2L, -2L, -2L, -2L, -2L, 
                  -2L, -2L, -2L, -2L, -2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
                  2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 0L, -2L, 
                  -2L, -3L, -3L, -1L, 0L, 28L, 27L, 26L, 25L, 24L, 23L, 22L, 21L, 
                  20L, 19L, 18L, 17L, 16L, 15L, 14L, 27L, 26L, 25L, 24L, 23L, 22L, 
                  21L, 20L, 19L, 18L, 17L, 16L, 15L, 14L, 13L, 12L, 11L, 10L, 9L, 
                  8L, 7L, 6L, 5L, 4L, 3L, 2L, 1L, 2L, 0L), ncol = 2L)
  G <- set_graph_attr(G, "layout", lay)
  G <- set_graph_attr(G, "rescale", FALSE)
  G
}

G <- make_flow_chart(puzzle_data) 
plot(G, ylim = c(0, 28), xlim = c(-3.5, 2.5), margin = 0, asp = 0)
```

We can see that the algorithm calculates a starting value for `a` (Depending on whether
we started with 1 or any other value (0 in our case)) and then does a loop:

1. If `a` equals 1 stop.
1. Else if it is even divide `a` by 2 otherwise multiply by 3 and add 1.
1. Repeat.
1. Return the number of iterations.

This is known as the **Collatz conjecture**. 

```{r get-starting-value-0, echo = FALSE}
a0 <- (((0L + 2L) * 27L + 2L) * 3L + 2L) * 27L + 1L
```


For `a == 0L` the starting value is `r format(a0, big.mark = ",")`.


```{r get-solution-1}
<<get-starting-value-0>>
collatz_conjecture <- function(n) {
  b <- 0L
  while (n != 1L) {
    b <- b + 1L
    if (n %% 2L == 0L) {
      n <- n / 2L
    } else {
      n <- 3L * n + 1L
    }
  }
  b
}

collatz_conjecture(a0)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

```{r get-starting-value-1, echo = FALSE}
a1 <- ((((((((1L * 3L) + 2L) * 3L + 2L) * 9L + 2L) * 3L + 1L) * 3L + 1L) * 3L + 2L) * 
         3L + 1L) * 9L +1L
```

This time we simply walk the other branch which yields another starting value 
(`r format(a1, big.mark = ",")`).

```{r get-solution-2}
<<get-starting-value-1>>
collatz_conjecture(a1)
```
