---
params:
   task_nr: 22
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2024"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(tibble)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(stringi)
library(knitr)
library(cli)
library(bit64)
library(igraph)
library(Rcpp)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    as.integer()
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

Looking at the instructions we observe that these boil down to:

1. Bit shift the input to the left by 6 bits, XOR it with the original result and keep
only the first 24 bits.
1. Bit shift to the right b< 5 bits, XOR and truncate as before.
1. Ecentually bit shift to the left again, this time by 11 bits, XOR and truncate.

While we could do so in R directly, it proves useful to write a samell C++ wrapper, as
this will speed up the calculation tremendously. The implementation is straight forward
and with respect to the second task, we return all (pseudo-) random numbers generated, as
we will need them for the second part of this puzzle.

```{Rcpp bit-shift, cache = TRUE}
#include <Rcpp.h>

// [[Rcpp::export]]
Rcpp::NumericVector bit_shift(int number, int iterations) {
  const long long MASK = 1LL << 24; 
  Rcpp::NumericVector results(iterations);
  
  long long current = number; 
  
  for (int i = 0; i < iterations; i++) {
    current = ((current << 6) ^ current) % MASK; 
    current = ((current >> 5) ^ current) % MASK;          
    current = ((current << 11) ^ current) % MASK;
    
    results[i] = current;
  }
  
  return results;
}
```


```{r get-solution-1}

pseudos <- map(puzzle_data, ~ bit_shift(.x, 2000L))

map(pseudos, ~ .x[2000L] %>% as.integer64()) %>% 
  do.call(c, .) %>% 
  sum()
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We use agai C++ to implement this solution with the following idea:

1. For each starting pseudo random number create a vector of differences between the one
digits among all 2000 pseudo random numbers.
1. For each 4 consecutive differences add the final value to a hash table, iif the element
key (concatenated differences) does not exist. That is fro each random number we maintain
a hash table of unique differences, keeping their selling value at the first occurence.
1. Once we have a hash table of all length 4 difference sequences, we create yet another
hash table, where we add the sellign bids for each monkey.
1. In a last loop we go through each ofd those entries and return the maximum price.

```{Rcpp hash-map, cache = TRUE}
#include <Rcpp.h>
#include <unordered_map>
#include <iostream>
#include <string>
#include <sstream>

// [[Rcpp::export]]
int get_max_revenue(Rcpp::List pseudo_random) {
  std::vector<std::unordered_map<std::string, int>> hash_maps;
  std::unordered_map<std::string, int> max_map;
  int result = 0;
  // create all hash maps
  for (std::size_t i = 0; i < pseudo_random.size(); i++) {
    std::unordered_map<std::string, int> hash_map;
    Rcpp::IntegerVector vec = pseudo_random[i];
    std::vector<int> diffs(vec.size());
    diffs[0] = 0;
    for (std::size_t j = 1; j < vec.size(); j++) {
      diffs[j] = (vec[j] % 10) - (vec[j - 1] % 10);
      if (j >= 3) {
        std::ostringstream key_stream;
        std::string key;
        key_stream << diffs[j - 3] << "," << diffs[j - 2] << "," << 
          diffs[j - 1] << "," << diffs[j];
        key = key_stream.str();
        if (hash_map.find(key) == hash_map.end()) {
          hash_map[key] = vec[j] % 10;
        }
      }
    } 
    hash_maps.push_back(hash_map);
  }
  // loop through all hash maps and sum up the corresponding elements
  for (const auto& hash : hash_maps) {
    for (const auto& element: hash) {
      if (max_map.find(element.first) == max_map.end()) {
        max_map[element.first] = element.second;
      } else {
        max_map[element.first] += element.second;
      }
    }
  }
  for (const auto& max_element: max_map) {
    if (max_element.second > result) {
      result = max_element.second;
    }
  }
  return result;
}
```


```{r get-solution-2}
get_max_revenue(pseudos)
```

