---
params:
   task_nr: 16
   year: 2020
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(tidyr)
library(bit64)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  markers <- str_which(text_block, "your ticket|nearby tickets")
  validity <- text_block[1:(markers[1L] - 2L)] %>% 
    str_split(": | or ") %>% 
    do.call(rbind, .) %>% 
    set_colnames(c("field", "range1", "range2")) %>% 
    as_tibble() %>% 
    separate_wider_delim(range1:range2, "-", names = c("from", "to"), names_sep = "_") %>% 
    mutate(across(starts_with("range"), as.integer))
  my_ticket <- text_block[markers[1L] + 1L] %>% 
    str_split(",") %>% 
    extract2(1L) %>% 
    as.integer()
  other_tickets <- text_block[(markers[2L] + 1):length(text_block)] %>% 
    str_split(",") %>% 
    map(as.integer)
  list(validity = validity,
       my_ticket = my_ticket,
       other_tickets = other_tickets)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We construct a vector with all valid values and extract invalid fields for each ticket.


```{r get-solution-1}
check_tickets <- function(tickets) {
  valid_numbers <- tickets$validity %>% 
    rowwise() %>% 
    mutate(valid = list(c(range1_from:range1_to, range2_from:range2_to))) %>% 
    pull(valid) %>% 
    unlist()
  bad_tickets <- puzzle_data$other_tickets %>% 
    map_lgl(~ !all(.x %in% valid_numbers))
  error_rate <- puzzle_data$other_tickets[bad_tickets] %>% 
    map_int(~ sum(.x[!.x %in% valid_numbers])) %>% 
    sum()
  list(valid_tickets =  puzzle_data$other_tickets[!bad_tickets], error_rate = error_rate)
}

tickets <- check_tickets(puzzle_data)
tickets$error_rate
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]] %>% str_replace_all(fixed("*"), "\\*"))
```

### Solution

For each ticket we check which fields are theoretically valid. Then, we sort the fields 
according to the number of theoretically valid fields. We will find a field
for which there is only one possibility. We remove this possibility from the next field, 
which will yield again a unique field name. Eventually we can assign each field to its
position and use these positions to get the checksum based on our own ticket.

```{r get-solution-2}
get_checksum <- function(valid_tickets, validity, my_ticket) {
  tickets <- do.call(rbind, tickets$valid_tickets)
  valid_fields <- vector("list", ncol(tickets)) %>% 
    set_names(seq_len(ncol(tickets)))
  for (i in seq_len(ncol(tickets))) {
    vals <- tickets[, i]
    valid_fields[[i]] <- validity %>% 
      rowwise() %>% 
      mutate(is_valid = all(between(vals, range1_from, range1_to) |
                              between(vals, range2_from, range2_to))) %>% 
      filter(is_valid) %>% 
      pull(field)
  }
  valid_fields <- valid_fields[order(lengths(valid_fields))]
  field_ids <- accumulate(valid_fields, 
                          function(l, el) {
                            keep <- setdiff(el, l$drop)
                            list(drop = c(l$drop, keep), keep = keep)
                          }, .init = list(drop = character(0L), keep = character(0L))) %>% 
    map("keep") %>% 
    unlist()
  ids <- field_ids[str_detect(field_ids, "departure")] %>% 
    names() %>% 
    as.integer()
  prod(my_ticket[ids]) %>% 
    as.integer64()
}

get_checksum(tickets, puzzle_data$validity, puzzle_data$my_ticket)
```
