---
params:
   task_nr: 7
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(knitr)
library(cli)
library(bit64)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  formulas <- text_block %>% 
    unlist() %>% 
    keep(nzchar)
  res <- str_extract(formulas, "^\\d+") %>% 
    as.integer64() %>% 
    as.list()
  ops <- str_remove(formulas, "^\\d+: *") %>% 
    str_split(fixed(" ")) %>% 
    map(as.integer64)
  map2(ops, res, ~ list(ops = .x, res = .y))
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    str_split("\n") %>% 
    parse_puzzle_data()
})

test_input <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

```{r get-solution-1-implementation, echo = FALSE}
nr_calls <- 0
is_valid1 <- function(res, operands, i, division = FALSE) {
  nr_calls <<- nr_calls + 1L
  ret <- FALSE
  if (i == 0L) {
    ret <- (division && res == 1L) || (!division && res == 0L)
  } else {
    operand <- operands[i]
    if ((res - operand) >= 0L) {
      if (Recall(res - operand, operands, i - 1L, FALSE)) {
        ret <- TRUE
      }
    }
    if (res %% operand == 0L) {
      if (Recall(res / operand, operands, i - 1L, TRUE)) {
        ret <- TRUE
      }
    }
  }
  ret
}

get_valid <- function(calibrations, validator) {
  nr_calls <<- 0
  map_lgl(calibrations, function(calibration) {
    ops <- calibration$ops
    res <- calibration$res
    validator(res, ops, length(ops))
  })
}

idx1 <- get_valid(puzzle_data, is_valid1)

res1 <- puzzle_data[idx1] %>% 
    map("res") %>% 
    do.call(c, .) %>% 
    sum()
```

We use a backtracking algorithm with all the operators from the **right**. If the current 
result is divisible without remainder by the current operator, we recurse into a division 
branch with the current result divided by the current operator. If the current result is
greater than the current operator we recurse into a subtraction branch with the
current result minus the current operator. If no numbers are left, we check whether we 
end up at zero for a subtraction branch or at one for a division branch. We stop early if
at any recursion neither of the 2 pre-conditions (no remainder and greater 
result) holds true.

The benefit from doing it from the right is that we can branch a lot of cases where the
division would not work. A brute force method would need the following amount of 
recursions:

```{r nr-recursions-1}
puzzle_data %>% 
  map_dbl(~ 2 ^ (.x$ops %>% length())) %>% 
  sum()
```

while the branching brought that number down to `r format(nr_calls, big.mark = ",")`.

```{r get-solution-1}
<<get-solution-1-implementation>>
res1
```


## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

```{r get-solution-2-implemebtation, echo = FALSE}
is_valid2 <- function(res, operands, i, division = FALSE) {
  nr_calls <<- nr_calls + 1L
  ret <- FALSE
  if (i == 0L) {
    ret <- (division && res == 1L) || (!division && res == 0L)
  } else {
    operand <- operands[i]
    if ((res - operand) >= 0L) {
      if (Recall(res - operand, operands, i - 1L, FALSE)) {
        ret <- TRUE
      }
    }
    if (res %% operand == 0L) {
      if (Recall(res / operand, operands, i - 1L, TRUE)) {
        ret <- TRUE
      }
    }
    str_operand <- as.character(operand)
    str_res <- as.character(res)
    n <- nchar(str_res)
    n_operand <- nchar(str_operand)
    if (substr(str_res, n - n_operand + 1L, n) == str_operand) {
      ## concatenate could be a viable solution
      ## remove last character and recurse
      res <- substr(str_res, 1L, n - n_operand) %>% 
        as.integer64()
      if (Recall(res, operands, i - 1L, FALSE)) {
        ret <- TRUE
      }
    }
  }
  ret
}

idx2 <- get_valid(puzzle_data[!idx1], is_valid2)

res2 <- puzzle_data[!idx1][idx2] %>% 
  map("res") %>% 
  do.call(c, .) %>% 
  sum()
```

We simply adapt our `is_valid1` function by adding the concat operator. We run the 
backtracking algorithm only on cases where we did not find a match in part one, saving 
also some decent amount of runs.

For the concat operator the (even more direly needed) branching condition is, whether the 
current result ends in the current number. If so, we recurse simply by removing those
numbers from the end.

Without branching we would need a whooping number of

```{r nr-recursions-2}
puzzle_data %>% 
  map_dbl(~ 3 ^ (.x$ops %>% length())) %>% 
  sum()
```

runs, while in our implementation we brought that number down to 
`r format(nr_calls, big.mark = ",")`.

```{r get-solution-2}
<<get-solution-2-implemebtation>>
res1 + res2
```
