---
params:
   task_nr: 13
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(knitr)
library(cli)
library(bit64)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- str_split(text_block, "\n")
  }
  text_block <- text_block %>% 
    unlist()
  if (nzchar(tail(text_block, 1L))) {
    text_block <- c(text_block, "")
  }
  pos <- c(0L, which(!nzchar(text_block)))
  chunks <- map(seq_along(head(pos, -1L)), 
                ~ text_block[seq(pos[.x] + 1L, pos[.x + 1] - 1L)])
  map(chunks, function(block) {
    buttons <- str_extract_all(block[1:2], "\\d+") %>% 
      unlist() %>% 
      as.integer() %>% 
      matrix(nrow = 2L) %>% 
      set_colnames(c("A", "B"))
    goal <- str_extract_all(block[3L], "\\d+") %>% 
      unlist() %>% 
      as.integer() %>% 
      matrix(nrow = 2L) %>% 
      set_colnames("goal")
    cbind(buttons, goal)
  })
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r get_example_data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We could solve the problem using *Linear Programming*, but as it turns out each problem
is uniquely solvable and has exactly one solution. Henc, we simply can rely on simple
Linear Algebra by inverting the coefficient matrix to solve `A %*% x = b` and then compute
the objective function `3 * sol[1] + sol[2]`. A simple check is needed to guarantee 
integer solutions.

```{r get-solution-1}
solve_machine <- function(all_coefs) {
  map_dbl(all_coefs, function(coefs) {
    sol <- solve(coefs[, 1:2]) %*% (coefs[, 3, drop = FALSE])
    if (sum((sol - round(sol, 0L)) ^ 2) < 1e-7) {
      sum(c(3, 1) * c(sol))
    } else {
      0
    }
  }) %>% 
    sum() %>% 
    as.integer64()
}
solve_machine(puzzle_data)
```


## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We simply add the offset to the RHS of each problem and get the new solution.

```{r get-solution-2}
new_machines <- map(puzzle_data, function(machine) {
  machine[, 3] <- machine[, 3, drop = FALSE] + 10000000000000
  machine
})
solve_machine(new_machines)
```
