---
params:
   task_nr: 25
   year: 2017
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}

.page-content table tr pre {
  border-radius: revert;
  background-color: revert;
  color: revert;
  padding: revert;
  border: revert;
}
```

```{js cpp-highlighter, file = "Prism_CPP.js", echo = FALSE}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  n <- text_block %>% 
    extract2(2L) %>% 
    str_extract("\\d+") %>% 
    extract2(1L) %>% 
    as.integer()
  start <- text_block %>% 
     extract2(1L) %>% 
    str_remove_all("^Begin in state |\\.$") %>% 
    extract2(1L)
  blk_idx <- text_block %>% 
    str_which("^In state")
  parse_block <- function(block) {
    block_id <- str_remove_all(block[[1L]], "In state |:$") %>% 
      extract2(1L)
    ops <- str_extract_all(block[c(3:5, 7:9)], 
                           "\\d|right|left|state .") %>% 
      str_remove(fixed("state "))
    res <- list(
      list(
      c(
        list(as.integer(ops[1L])),
        if_else(ops[2L] == "right", 1L, -1L),
        ops[3]
      ),
      c(
        list(as.integer(ops[4L])),
        if_else(ops[5L] == "right", 1L, -1L),
        ops[6]  
      )
    )
    )
    names(res) <- block_id
    res
  }
  ops <- map(blk_idx, ~ parse_block(text_block[.x:(.x + 8L)])) %>% 
    flatten()
  list(ops = ops, n = n, start = start)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

Given the large repetition count, we go for a `C++` solution. We ust a `std::list` as it
allows for conveniently moving forward and backward. New items are anyways just added at 
either side of the list. The trickiest part was thus, just to find a proper data structure
for encoding this puzzle's data.

```{Rcpp cpp-algorithm, file = "turing.cpp", cache = TRUE}
```

```{r get-solution-1, cache = TRUE}
turing(puzzle_data$n, puzzle_data$start, puzzle_data$ops)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

Hohoho, that's it for 2017!
