---
params:
   task_nr: 18
   year: 2023
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L)
  }
  text_block[nzchar(text_block)] %>% 
    str_remove_all("[()]") %>% 
    str_split(" ") %>% 
    do.call(rbind, .) %>% 
    set_colnames(c("dir", "dist", "col")) %>% 
    as_tibble() %>% 
    mutate(dist = as.integer(dist),
           col = str_remove(col, "^#"))
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}

cat(puzzle[[1]])
```

### Solution

We solve the first part by a brute-force algorithm, where we explicitly generate the 
trench. Then, we flood the trench from the outside and count all non flooded elements
to get the area of the lagoon.

```{r get-solution-1}
flood_cave <- function(cave) {
  class(cave) <- "cave"
  nr <- nrow(cave)
  nc <- ncol(cave)
  dirs <- rbind(
    U = c(-1L, 0L),
    R = c(0L, 1L),
    D = c(1L, 0L),
    L = c(0L, -1L)
  )
  pos <- rbind(
    cbind(1L, 1:nc),
    cbind(nr, 1:nc),
    cbind(1:nr, 1L),
    cbind(1:nr, nc)
  )
  visited <- matrix(FALSE, nr, nc)
  pos <- pos[cave[pos] == ".", , drop = FALSE]
  while (nrow(pos) > 0L) {
    visited[pos] <- TRUE
    cave[pos] <- "~"
    new_pos <- map(1:4, function(r) {
      t(t(pos) + dirs[r, ])
    }) %>% 
      do.call(rbind, .)
    ok <- new_pos[, 1L] >= 1L & new_pos[, 1L] <= nr &
      new_pos[, 2L] >= 1L & new_pos[, 2L] <= nc
    new_pos <- new_pos[ok, , drop = FALSE]
    pos <- new_pos[!visited[new_pos] & cave[new_pos] == ".", , drop = FALSE]
    pos <- pos[!duplicated(pos), , drop = FALSE]
      
  }
  cave
}

create_trench <- function(ops) {
  n <- nrow(ops)
  pos <- matrix(NA_integer_, n + 1L, 3L, dimnames = list(NULL, c("row", "col", "color")))
  pos[1L, ] <- c(0L, 0L, NA_integer_)
  dirs <- rbind(
    U = c(-1L, 0L),
    R = c(0L, 1L),
    D = c(1L, 0L),
    L = c(0L, -1L)
  )
  for (i in seq_len(n)) {
    last_pos <- pos[i, , drop = FALSE]
    op <- ops %>% 
      slice(i) %>% 
      as.list()
    pos[i + 1L, 1:2] <- last_pos[, 1:2] + op$dist * dirs[op$dir, ]
    pos[i + 1L, 3L] <- strtoi(op$col, 16L)
  }
  pos
}

calculate_lagoon_area <- function(ops) {
  trench <- create_trench(ops)
  mins <- apply(trench, 2L, min) - 1L
  mins[3L] <- 0L
  coords <- sweep(trench, 2, mins)
  start <- coords[1L, ]
  cave <- array(".", apply(coords[, 1:2], 2, max))
  for (i in 2:nrow(coords)) {
    end <- coords[i, ]
    idx <- expand.grid(row = start[1L]:end[1L], col = start[2L]:end[2L]) %>% 
      as.matrix()
    cave[idx] <- "#"
    start <- end
  }
  res <- flood_cave(cave)
  sum(res != "~")
}
calculate_lagoon_area(puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

Given the large number for part 2 brute force is not feasible anymore. Instead we use the
[shoelace formula](https://en.wikipedia.org/wiki/Shoelace_formula) to calculate its area.
The perimeter can also be easily derived from the coordinates.
Then we use [Pick's theorem](https://en.wikipedia.org/wiki/Pick%27s_theorem) to count the 
number of integer points within and on its boundary.


```{r get-solution-2}
calculate_lagoon_area_large <- function(ops) {
  dirs <- c("0" = "R", "1" = "D", "2" = "L", "3" = "U")
  ops <- ops %>% 
    mutate(
      dir = dirs[str_sub(col, -1L)],
      dist = strtoi(str_sub(col, 1L, -2L), 16L)
      )
  trench <- create_trench(ops) 
  n <- nrow(trench)
  area <- perimeter <- 0L
  storage.mode(trench) <- "numeric"
  x1 <- trench[1L, 1L]
  y1 <- trench[1L, 2L]
  for (i in 2:n) {
    x2 <- trench[i, 1L]
    y2 <- trench[i, 2L]
    area <- area + y1 * (x2 - x1)
    perimeter <- perimeter + abs(x2 - x1) + abs(y2 - y1)
    x1 <- x2
    y1 <- y2
  }
  unname(abs(area) + perimeter / 2 + 1) ## pick's theorem
}
calculate_lagoon_area_large(puzzle_data) %>% 
  print(digit = 16L)
```
