---
params:
   task_nr: 24
   year: 2021
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2021"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```

```{js sql-highlighter, file = "Prism_CPP.js", echo = FALSE}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(tibble)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(stringi)
library(knitr)
library(cli)
library(rlang)
library(DT)
library(bit64)
library(rlang)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  block_idx <- str_which(text_block, "^inp")
  map2(block_idx, c(tail(block_idx, -1L) - 1L, length(text_block)), ~
         text_block[.x:.y])
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

# Puzzle Day `r params$task_nr`

## Part 1


### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

Our first observation is that the instructions are largely the same for each digit, except
at positions 
`r do.call(rbind, puzzle_data) %>% apply(2L, \(c) length(unique(c)) != 1L) %>% which()`.
In particular the usage of the internal registers is the same for each instruction and 
only some constants change.

```{r show-instructions}
codes <- do.call(rbind, puzzle_data) %>% 
  set_colnames(paste("Step", 1:ncol(.))) %>% 
  set_rownames(paste("Instruction", 1:nrow(.)))

diffs <- do.call(rbind, puzzle_data) %>% 
  apply(2L, \(c) length(unique(c)) != 1L) %>% 
  which()

datatable(
  codes, 
  class = c("compact", "nowrap", "hover", "row-border"),
  options = list(
    pageLength = nrow(codes),
    dom = "t",
    ordering = FALSE,
    columnDefs = list(
      list(
        className = "dt-center", targets = "_all"
      )
    ))
)%>% 
  formatStyle(diffs, 
              backgroundColor = "firebrick", 
              color = "white", 
              fontStyle = "italic")

```

Now we can translate each instruction set to native R code:

* `r tt(puzzle_data[[1]][1])`: Read input (`d`, say) and store it in `w`: `w <- d`.
* `r tt(puzzle_data[[1]][2:6])`: Copy the value of `z` to `x`, take it modulo 26 and add a
constant `c1` (this was one of the changing instructions). Replace `z` by `z` divided by 
`c2` (again a differnt constant for each instruction): `x <- z %% 26 + c1` and
`z <- z %/% c2`.
* `r tt(puzzle_data[[1]][7:8])`: Store 1 in `x`, if `x`does not equal `w`, otherwise store 
0: `x <- as.integer(x != w)`.
* `r tt(puzzle_data[[1]][9:12])`: Multiply `x`by 25 and add 1 and sore the reuslt in `y`:
`y <- 25L * x + 1L`.
* `r tt(puzzle_data[[1]][13])`: Multiply `z` by `y` and store it in `z`: 
`z <- z * y`.
* `r tt(puzzle_data[[1]][14:17])`: Add constant `c3` to to `w`, multiply by `x` and store
the result in `y`: `y <- (w + c3) * x`.
* `r tt(puzzle_data[[1]][18])`: Add `z` to `y` and store the result in `z`: 
`z <- z + y`.

To find valid codes, we start at the end where we know that `z` must equal 0 in order to
result in a valid code. From there, with a given digit we simply brute-force all possible 
`z` of the *previous* digit and recurse with all valid digits.

Since this is a rather heavy brute-force algorithm, we fall back to C++ to gain a 
considerable amount of speed.

A smarter algorithm would be to limit the list of candidates `z` to make the search space
smaller and to introduce some hash table to avoid recalculating known values.

As a matter of fact the first version used a hash map which reduced the amount of runs by
a factor of 11, but overall the algorithm was slower due to the overhead involved in 
checking feasible values.

```{Rcpp cpp-algorithm, file = "nomad.cpp", cache = TRUE}
```


```{r get-solution-1, cache = TRUE}
consts <- map(puzzle_data, 
              ~ str_extract(.x[c(5:6, 16)], "-?\\d+") %>% 
                as.integer() %>% 
                set_names(paste0("c", 1:3))) %>% 
  do.call(rbind, .) %>% 
  as_tibble()

all_solutions <- solve_nomad(consts) %>% 
  as.integer64()

max(all_solutions)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

As we generated already all valid codes for part 1, we simply need to return the smallest
code for this part.

```{r get-solution-2}
min(all_solutions)
```
