---
params:
   task_nr: 22
   year: 2022
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  maze_marker <- text_block %>% 
    str_detect("[.#]")
  lines <- text_block[maze_marker] %>% 
    str_split("")
  fill <- max(lengths(lines))
  maze <- map(lines, ~ c(.x, rep(" ", fill))[1:fill]) %>% 
    do.call(rbind, .)
  maze[maze == " "] <- NA_character_
  instructions <- text_block[!maze_marker] %>% 
    str_extract_all("\\d+|[LR]") %>% 
    extract2(1L)
  list(maze = maze, instructions = instructions)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We use a straight forward implementation of the algorithm. If we hit an `NA` field we have
to wrap around to the first non `NA` file on the other side. 

```{r get-solution-1}
wrap_flat <- function(dir, maze, pos) {
  if (dir %% 2L == 0L) {
    slice <- maze[, pos[2L]]
  } else {
    slice <- maze[pos[1L], ]
  }
  ground <- which(!is.na(slice))
  if (dir <= 2L) {
    wrap <- ground[1L]
  } else {
    wrap <- ground[length(ground)]
  }
  if (dir %% 2L == 0L) {
    list(pos = cbind(wrap, pos[2L]), dir = dir)
  } else {
    list(pos = cbind(pos[1L], wrap), dir = dir)
  }
}

walk_maze <- function(maze, steps, wrap_fn) {
  dirs <- rbind(
    ">" = c(0L, 1L),
    "v" = c(1L, 0L),
    "<" = c(0L, -1L),
    "^" = c(-1L, 0L)
  )
  rot <- c(L = -1L, R = 1L)
  dir <- 1L
  pos <- cbind(1, which(!is.na(maze[1, ]))[[1L]])
  dd <- dim(maze)
  for (step in steps) {
    n <- as.integer(step) %>% 
      suppressWarnings()
    if (is.na(n)) {
      ## rotation
      dir <- (dir - 1L + rot[step]) %% 4L + 1L
    } else {
      cur_dir <- dirs[dir, ] 
      for (i in seq_len(n)) {
        new_pos <- pos + cur_dir
        new_pos[, 1L] <- (new_pos[, 1L] - 1L) %% dd[1L] + 1L
        new_pos[, 2L] <- (new_pos[, 2L] - 1L) %% dd[2L] + 1L
        if (is.na(maze[new_pos])) {
          wrap <- wrap_fn(dir, maze, pos)
          new_pos <- wrap$pos
          if (maze[new_pos] != "#") {
            dir <- wrap$dir
          }
        }
        if (maze[new_pos] == "#") {
          break
        }
        pos <- new_pos
      }
    }
  }
  1000L * pos[1L] + 4L * pos[2L] + dir - 1
}
walk_maze(puzzle_data$maze, puzzle_data$instructions, wrap_flat) 

```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We need to define a more sophisticated wrap function and can otherwise use the same 
algorithm as before.

```{r get-solution-2}
wrap_cube <- function(dir, maze, pos) {
  surface_id <- case_when(
    pos[1L] <=  50L & pos[2L] <= 100L ~ 1L,
    pos[1L] <=  50L & pos[2L] >  100L ~ 2L,
    pos[1L] >   50L & pos[1L] <= 100L ~ 3L,
    pos[1L] >  100L & pos[1L] <= 150L ~ 4L,
    pos[1L] >  150L & pos[2L] <=  50L ~ 5L,
    pos[1L] >  150L & pos[2L] >   50L ~ 6L
  )
}

walk_maze(puzzle_data$maze, puzzle_data$instructions, wrap_flat) 
```

