---
params:
   task_nr: 4
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(knitr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    str_split("\n") %>% 
    extract2(1L) %>% 
    head(-1L) %>% 
    str_split("") %>% 
    do.call(rbind, .)
})
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

First, we define a function that, from a given starting point extracts all 8 words:

```{r word-extractor}
get_word <- function(row, col, .data = puzzle_data, .length = 4L) {
  pos <- expand.grid(
    row_offset = -1:1,
    column_offset = -1:1
  ) %>% 
    as_tibble() %>% 
    filter(row_offset !=0 | column_offset != 0) %>% 
    pmap(function(row_offset, column_offset) {
      t(outer(c(row_offset, column_offset), seq(0, .length - 1L)) + c(row, col))
    })
  dd <- dim(.data)
  valid_pos <- pos %>% 
    discard(\(m) {
      any(m <= 0L | m[, 1] > dd[1L] | m[, 2L] > dd[2L])
    })
  valid_pos %>% 
    map_chr(\(idx) paste(c(.data[idx]), collapse = ""))
}
```

For instance, the 3 words from position `(1 / 1)` (to the left, down and down left, the 
other directions would be out of bounds) are `r combine_words(get_word(1L, 1L))`.

What is left is to get all positions starting with an `X` and get all the words from 
there. Then, we count how often we get the word `XMAS`:

```{r get-solution-1}
xpos <- which(puzzle_data == "X", arr.ind = TRUE) %>% 
  as_tibble()

pmap(xpos, get_word) %>% 
  flatten() %>% 
  str_count(fixed("XMAS")) %>% 
  sum()
```


## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We apply the same idea, first for each position we get the x-shaped words (note that the 
position is the middle point):

```{r x-word-extractor}
get_xword <- function(row, col, .data = puzzle_data) {
  idx <- rbind(
    c(-1, 0, 1, -1, 1),
    c(-1, 0, -1, 1, 1)
  ) + c(row, col)
  paste(c(.data[t(idx)]), collapse = "")
}
```

Each `A` is the middle point of a potential x-word (exclude `As` in first / last 
column row as they cannot be a middle point). For each of these positions check whether
the x-word is any of the permutations of `MAS` (we pasted the letters top-left / middle /
bottom-left / top-right / bottom-right), those are `MAMSS`, `MASMS`, `SAMSM` and
`SASMM`.

```{r get-solution-2}
apos <- which(puzzle_data == "A", arr.ind = TRUE) %>% 
  as_tibble() %>% 
  filter(between(row, 2, nrow(puzzle_data) - 1),
         between(col, 2, ncol(puzzle_data) - 1))

pmap(apos, get_xword) %>% 
  flatten() %>% 
  str_count("MAMSS|MASMS|SAMSM|SASMM") %>% 
  sum()
```
