---
params:
   task_nr: 19
   year: 2023
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(rlang)
library(igraph)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L)
  }
  sep <- which(!nzchar(text_block))
  parse_rule <- function(line) {
    rules <- str_remove_all(line, "^[a-z]+\\{|\\}")  %>% 
      str_extract_all("[^,]+") %>% 
      extract2(1L)
    is_dflt <- str_which(rules, ":", TRUE)
    rules <- str_replace_all(rules, ":(.*)", " ~ \"\\1\"")
    rules[is_dflt] <- paste0("TRUE ~ \"", rules[is_dflt], "\"")
    call2("case_when", !!!parse_exprs(rules))
  }
  parse_rules_lims <- function(line) {
    nm <- str_extract(line, "^\\w+")
    rule <- str_remove_all(line, "^\\w+\\{|\\}")
    dflt <- str_remove_all(line, "^.+,|\\}$")
    mtch <- str_match_all(rule, "([xmas])([<>])(\\d+):(\\w+)") %>% 
      extract2(1L)
    tibble(
      id = nm,
      val = mtch[, 2],
      lwr = if_else(mtch[, 3L] == ">", as.integer(mtch[, 4]), NA_integer_),
      upr = if_else(mtch[, 3L] == "<", as.integer(mtch[, 4]), NA_integer_),
      target = mtch[, 5L],
      dflt = dflt
    )
  }
  rules <- text_block[1:(sep[1L] - 1L)]
  vals <- text_block[(sep[1L] + 1L):(length(text_block) - 1)] %>% 
    str_extract_all("\\d+") %>% 
    do.call(rbind, .) %>% 
    set_colnames(c("x", "m", "a", "s")) %>% 
    as_tibble() %>% 
    mutate(across(everything(), as.integer))
  nms <- str_extract_all(rules, "^[a-z]+")
  rules_lims <- map(rules, parse_rules_lims) %>% 
    list_rbind()
  rules <- map(rules, parse_rule) %>% 
    set_names(nms)
  
  list(vals = vals, rules = rules, rules_lims = rules_lims)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}

cat(puzzle[[1]])
```

### Solution

```{r get-solution-1}
sort_pieces <- function(factory) {
  pieces <- factory$vals
  rules <- factory$rules
  pieces <- pieces %>% 
    mutate(rule_target = "in")
  while (!all(pieces$rule_target %in% c("R", "A"))) {
    pieces <- pieces %>% 
      mutate(rule = rules[rule_target]) %>% 
      rowwise() %>% 
      mutate(rule_target = if (rule_target %in% c("R", "A")) rule_target 
                               else eval_tidy(rule))
  }
  pieces %>% 
    filter(rule_target == "A") %>% 
    mutate(rating = x + m + a + s) %>% 
    summarize(rating = sum(rating)) %>% 
    pull(rating) %>% 
    sum()
}

sort_pieces(puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

Given the large number for part 2 brute force is not feasible anymore. Instead we use the
[shoelace formula](https://en.wikipedia.org/wiki/Shoelace_formula) to calculate its area.
The perimeter can also be easily derived from the coordinates.
Then we use [Pick's theorem](https://en.wikipedia.org/wiki/Pick%27s_theorem) to count the 
number of integer points within and on its boundary.


```{r get-solution-2}
count_combs <- function(rules) {
  pos <- "in"
  nodes <- rules %>% 
    distinct(id) %>% 
    pull(id)
  lims <- vector("list", length(nodes)) %>% 
    set_names(nodes)
  while (length(pos) > 0L) {
    cur_pos <- pos[1L]
    pos <- pos[-1L]
    cur_rule <- rules %>% 
      filter(id == cur_pos)
    for (i in seq_len(nrow(cur_rule))) {
      
    }
  }
}

count_combs(ex$rules_lims)
```
