---
params:
   task_nr: 3
   year: 2017
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    as.integer()
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We observe the following things:

1. We label the first square (just the file with 1) as square $s_0$ and each surrounding
square with $s_i$. Fields 2 - 9 are for example on square $s_1$, fields 10 - 25 on square
$s_2$ and so on.
1. The right lower corner of each square $s_i$ ends with the number $(2i + 1)^2$.
1. Either the horizontal or the vertical distance to each field on square $s_i$ is $i$.
1. The other direction is the distance to the middle point of this square to the field. 


Thus, to get the distance for field $j$ the algorithm finds first $i$ by solving 
$j \leq (2i + 1)^2$ and then gets the distance to the nearest middle point.


```{r get-solution-1}
get_distance <- function(j) {
  if (j == 1L) {
    0L
  } else {
    i <- ceiling((sqrt(j) - 1L) / 2L)
    if (sqrt(j) == floor(sqrt(j)) && (sqrt(j) %% 2L == 1L)) { 
      i <- i - 1
    }
    rc <- (2L * i + 1L) ^ 2 ## right corner
    mp <- rc - (2L * 0:3 + 1L) * i ## middle points
    min(abs(mp - j)) + i
  }
}

get_distance(puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We solve the second part by constructing the matrix iteratively.

```{r get-solution-2}
embed_matrix <- function(m, n = 2) {
  for (i in 1:n) {
    m <- cbind(NA, rbind(NA, m , NA), NA)
  }
  m
}

fill_ulam_spiral <- function(max_val) {
  dir <- rbind(
    ">"  = c(0L, 1L),
    "^"  = c(-1L, 0L),
    "<"  = c(0L, -1L),
    "v"  = c(1L, 0L),
    "^>" = c(-1L, 1L),
    "^<" = c(-1L, -1L),
    "v>" = c(1L, 1L),
    "v<" = c(1L, -1L)
  )
  n <- 9L
  m <- matrix(NA, n, n)
  val <- 1L
  pos <- cbind((n + 1L) / 2L, (n + 1L) / 2L)
  m[pos] <- val
  j <- 2L
  while (val < max_val) {
    i <- ceiling((sqrt(j) - 1L) / 2L)
    l <- 2L * i + 1L
    if (pos[1L] >  n - 2L) {
      k <- 2L
      m <- embed_matrix(m, k)
      pos <- t(t(pos) + c(k, k))
      n <- n + 2L * k
    }
    movements <- rep(c(">", "^", "<", "v", ">"),
                     c(1L, l - 2L, l - 1L, l - 1L, l - 1L))
    for (next_dir in movements) {
      pos <- pos + dir[next_dir, ]
      val <- sum(m[t(t(dir) + c(pos))], na.rm = TRUE)
      m[pos] <- val
      j <- j + 1L
      if (val >= max_val) {
        break
      }
    }
  }
  val
}
fill_ulam_spiral(puzzle_data)
```
