---
params:
   task_nr: 14
   year: 2020
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{js cpp-highlighter, file = "Prism_CPP.js", echo = FALSE}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(bit64)
library(tidyr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  bm <- str_which(text_block, "mask")
  end <- c(tail(bm, -1L), length(text_block) + 1L)
  map2(bm, end, function(s, e) {
    bl <- text_block[s:(e - 1L)]
    mask <- str_extract(bl[1L], "[01X]+")[[1L]]
    set_op <- str_extract_all(bl[-1L], "\\d+") %>% 
      do.call(rbind, .) %>% 
      set_colnames(c("addr", "val"))
    storage.mode(set_op) <- "integer"
    list(mask = mask, mem = set_op)
  })
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

To apply a bitmask we use `(x & ~clear_mask) | set_mask` where `clear_mask` contains `1`
at each position where we want to set a bit to `0` and a `0` at the positions which
should not be changed. Likewise `set_mask` contains a `1` at each position which we want 
to set to set to `1` and `0` otherwise.

*N.B.* As we need bit 64 bitwise operations, we include a simple C++ code to avoid the 
need to work on bit vectors in R for speed reasons.

```{Rcpp cpp-algorithm, file = "bit64.cpp", cache = TRUE}
```

```{r get-solution-1}
bit2num <- function(x) {
  bits <- str_split(x, "") %>% 
    extract2(1L) %>% 
    as.integer()
  n <- length(bits)
  sum(2 ^ ((n - 1L):0) * bits) %>% 
    as.integer64
}

get_clear_mask <- function(mask, which = c("value", "addr")) {
  which <- match.arg(which)
  if (which == "value") {
    str_replace_all(mask, "1", "X") %>% 
      extract2(1L) %>% 
      str_replace_all("0", "1") %>% 
      extract2(1L) %>% 
      str_replace_all("X", "0") %>% 
      extract2(1L) %>% 
      bit2num()
  } else {
    str_replace_all(mask, "X", "1") %>% 
      extract2(1L) %>% 
      bit2num()
  }
}

get_set_mask <- function(mask, which = c("value", "addr")) {
  which <- match.arg(which)
  if (which == "value") {
    str_replace_all(mask, "X", "0") %>% 
      extract2(1L) %>% 
      bit2num()
  } else {
    digits <- str_split(mask, "") %>% 
      extract2(1L) %>% 
      as.list()
    digits[digits == "X"] <- list(c("0", "1"))
    do.call(expand_grid, digits) %>% 
      unite("mask", sep = "") %>% 
      rowwise() %>% 
      mutate(mask = bit2num(mask)) %>% 
      pull(mask)
  }
}

apply_mask <- function(vals, masks) {
  bitwOr64(bitwAnd64(vals, bitwNot64(masks$clear, 36L)), masks$set)
}

set_memory <- function(ops) {
  mem_size <- vapply(ops, \(.x) max(.x$mem[, "addr"]), integer(1L)) %>% 
    max()
  mem <- integer64(mem_size)
  for (op in ops) {
    masks <- list(set = get_set_mask(op$mask, "value"),
                  clear = get_clear_mask(op$mask, "value"))
    mem[op$mem[, "addr"]] <- apply_mask(as.integer64(op$mem[, "val"]), masks) 
  }
  sum(mem)
}

set_memory(puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]] %>% str_replace_all(fixed("*"), "\\*"))
```

### Solution

To deal with the floating bits for the address mask we use the following approach:

1. We set all bits to `0` where the bitmask contains a `1` or a `X`. We store this 
operation in `clear_mask`
1. Then, we generate all possible bitmasks by replacing all `X` by `0` and `1` and looking
at all combination (`set_mask`)
1. To get the final address we look at `x & ~clear_mask | set_mask` as before.


```{r get-solution-2}

set_memory_floating <- function(ops) {
  mem <- NULL
  for (op in ops) {
    masks <- list(set = get_set_mask(op$mask, "addr"),
                  clear = get_clear_mask(op$mask, "addr"))
    for (i in seq_len(nrow(op$mem))) {
      addr <- apply_mask(as.integer64(op$mem[i, "addr"]), masks) %>% 
        as.character()
      mem[addr] <- as.list(as.integer64(op$mem[i, "val"]))
    }
  }
  do.call(c, mem) %>% 
    sum()
}

set_memory_floating(puzzle_data)
```
