---
params:
   task_nr: 16
   year: 2015
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2015"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(tibble)
library(tidyr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  aunties <- str_remove(text_block, "^Sue \\d+: ") %>% 
    str_extract_all("[[:alpha:]]+: \\d+") %>% 
    imap(function(notes, idx) {
      vals <- str_split(notes, ": ") %>% 
        do.call(rbind, .) %>% 
        set_colnames(c("item", "amount")) %>% 
        as_tibble() %>% 
        mutate(amount = as.integer(amount)) %>% 
        mutate(aunt = idx, .before = 1L)
      }) %>% 
    list_rbind()
  note <- tibble(
    item = c("children", "cats", "samoyeds", "pomeranians", "akitas", "vizslas", 
             "goldfish", "trees", "cars", "perfumes"),
    amount = c(3L, 7L, 2L, 3L, 0L, 0L, 5L, 3L, 2L, 1L)
  ) 
  list(aunties = aunties, note = note)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We solve this puzzle by transforming the data into wide format (one column per item), 
and doing an `inner_join` first to get all aunties which have at most items as indicated
in the note. Then, we filter these matched down to those, which have an exact match or a
missing value.

```{r get-solution-1}
match_evidence <- function(aunties, note) {
  aunties_wide <- aunties %>% 
    pivot_wider(
      id_cols = aunt,
      names_from = item,
      values_from = amount,
      values_fill = -1L
    )
  note_wide <- note %>% 
    pivot_wider(
      names_from = item,
      values_from = amount
    )
  inner_join(
    aunties_wide,
    note_wide,
    join_by(
      children <= children,
      cats <= cats,
      samoyeds <= samoyeds,
      pomeranians <= pomeranians,
      akitas <= akitas,
      vizslas <= vizslas,
      goldfish <= goldfish,
      trees <= trees,
      cars <= cars,
      perfumes <= perfumes
    ),
    suffix = c(".auntie", ".note")
  ) %>% 
    pivot_longer(
      -aunt
    ) %>% 
    separate_wider_delim(name, ".", names = c("item", "src")) %>% 
    pivot_wider(id_cols = aunt:item, names_from = src) %>% 
    mutate(match = auntie == -1 | (auntie == note)) %>% 
    summarize(match = all(match), .by = aunt) %>% 
    filter(match)
}

match_evidence(puzzle_data$aunties, puzzle_data$note)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We use a similar approach as before, but modifying the join and filter criteria to the new
instructions.

```{r get-solution-2}
match_evidence2 <- function(aunties, note) {
  aunties_wide <- aunties %>% 
    pivot_wider(
      id_cols = aunt,
      names_from = item,
      values_from = amount,
      values_fill = -1L
    ) %>% 
    mutate(
      across(c(cats, trees), ~ if_else(.x == -1L, Inf, .x))
    )
  note_wide <- note %>% 
    pivot_wider(
      names_from = item,
      values_from = amount
    )
  inner_join(
    aunties_wide,
    note_wide,
    join_by(
      children <= children,
      cats > cats,
      samoyeds <= samoyeds,
      pomeranians < pomeranians,
      akitas <= akitas,
      vizslas <= vizslas,
      goldfish < goldfish,
      trees > trees,
      cars <= cars,
      perfumes <= perfumes
    ),
    suffix = c(".auntie", ".note")
  ) %>% 
    pivot_longer(
      -aunt
    ) %>% 
    separate_wider_delim(name, ".", names = c("item", "src")) %>% 
    pivot_wider(id_cols = aunt:item, names_from = src) %>% 
    mutate(
      match = case_when(
        item %in% c("cats", "trees") ~ is.infinite(auntie) | auntie > note,
        item %in% c("pomeranians", "goldfish") ~ auntie == -1L | auntie < note,
        TRUE ~ auntie == -1L | auntie == note
      )
    ) %>% 
    summarize(match = all(match), .by = aunt) %>% 
    filter(match)
}

match_evidence2(puzzle_data$aunties, puzzle_data$note)
```
