---
params:
   task_nr: 7
   year: 2015
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2015"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(tibble)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(stringi)
library(knitr)
library(cli)
library(igraph)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  ops <- text_block %>% 
    str_split("\\s") %>% 
    map(function(ops) {
      c(rep(NA_character_, 5L - length(ops)), ops)
    }) %>% 
    do.call(rbind, .)
  ops <- ops[, -4]
  colnames(ops) <- c("lhs", "op", "rhs", "res")
  as_tibble(ops)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We solve this puzzle by means of graph theory.

1. For each operation `lhs op rhs -> res` we create the nodes `lhs -> op`, `rhs -> op` and
`op -> res` (note that the  `NOT` operator has no `lhs` and a pure assignment has neither 
`op` nor `lhs`).
1. We sort the graph then topologically, that is nodes with no incoming nodes (this will
be the number literals) come first, followed by their dependent nodes and so on.
1. Then all what is left, is to iterate through the graph in topological order:
   1. If we hit a node with a number literal, assign this value to the node.
   1. If we hit an operator node, perform the corresponding operation with the value from 
   its predecessors (some care has to be taken for the non-commutative `SHIFT` operators 
   to make sure their operands are used in the right order)
   1. If we hit a output node, simply copy the value from its predecessor.
1. The final result will then be stored in node `a`.

```{r get-solution-1}
construct_graph <- function(ops) {
  ops <- ops %>% 
    mutate(
      op_label = case_when(
      is.na(op) ~ op,
      is.na(lhs) ~ paste(op, rhs, sep = "_"),
      TRUE ~ paste(lhs, op, rhs, sep = "_")
    )
    )
  el <- matrix(character(0), 0, 2) %>% 
    set_colnames(c("from", "to"))
  for (i in seq_len(nrow(ops))) {
    gate <- ops %>% 
      slice(i) %>% 
      as.list()
    if (is.na(gate$op)) {
      ## assignment operator
      el <- el %>% 
        rbind(
          cbind(gate$rhs, gate$res)
        )
    } else if (gate$op == "NOT") {
      el <- el %>% 
        rbind(
          cbind(gate$rhs, gate$op_label),
          cbind(gate$op_label, gate$res)
          )
    } else {
      el <- el %>% 
        rbind(
          cbind(gate$lhs, gate$op_label),
          cbind(gate$rhs, gate$op_label),
          cbind(gate$op_label, gate$res)
        )
    }
  }
  G <- graph_from_edgelist(
    el
  )
  V(G)$type <- case_when(
    !is.na(strtoi(V(G)$name)) ~ "literal",
    str_detect(V(G)$name, "^NOT") ~ "not",
    str_detect(V(G)$name, "_LSHIFT_") ~ "lshift",
    str_detect(V(G)$name, "_RSHIFT_") ~ "rshift",
    str_detect(V(G)$name, "_AND_") ~ "and",
    str_detect(V(G)$name, "_OR_") ~ "or",
    TRUE ~ "var"
  )
  G
}

calculate_wires <- function(G) {
  V(G)$value <- NA_integer_
  v_sorted <- topo_sort(G)
  for (vi in v_sorted) {
    me <- V(G)[vi]$name
    nbs <- neighbors(G, vi, "in")
    type <- V(G)[vi]$type
    if (type == "literal") {
      stopifnot(length(nbs) == 0L)
      ## it is a literal node with a value
      V(G)[vi]$value <- strtoi(V(G)[vi]$name)
    } else if (type == "var") {
      stopifnot(length(nbs) == 1L)
      V(G)[vi]$value <- nbs$value
    } else if (type == "not") {
      stopifnot(length(nbs) == 1L)
      V(G)[vi]$value <- bitwNot(nbs$value) %% 2 ^ 16
    } else if (type == "and") {
      stopifnot(length(nbs) == 2L)
      V(G)[vi]$value <- bitwAnd(nbs[1L]$value, nbs[2L]$value) %% 2L ^ 16L
    } else if (type == "or") {
      stopifnot(length(nbs) == 2L)
      V(G)[vi]$value <- bitwOr(nbs[1L]$value, nbs[2L]$value) %% 2L ^ 16L
    } else if (type == "lshift") {
      stopifnot(length(nbs) == 2L)
      parts <- str_split(me, "_[^_]+_") %>% 
        extract2(1L)
      nbs <- nbs[match(nbs$name, parts)]
      V(G)[vi]$value <- bitwShiftL(nbs[1L]$value, nbs[2L]$value) %% 2L ^ 16L
    } else if (type == "rshift") {
      stopifnot(length(nbs) == 2L)
      parts <- str_split(me, "_[^_]+_") %>% 
        extract2(1L)
      nbs <- nbs[match(nbs$name, parts)]
      V(G)[vi]$value <- bitwShiftR(nbs[1L]$value, nbs[2L]$value) %% 2L ^ 16L
    }
  }
  G
}
  
G <- construct_graph(puzzle_data) %>% 
  calculate_wires()

(res <- V(G)["a"]$value)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We have to simply rename the incoming node to `b` to hold the result from part 1 and 
recalculate the graph.

```{r get-solution-2}
rewire_graph <- function(G, new_val) {
  b_in <- neighbors(G, "b", "in")
  V(G)[b_in]$name <- new_val
  G
}

G2 <- rewire_graph(G, res) %>% 
  calculate_wires()
V(G2)["a"]$value
```