---
params:
   task_nr: 4
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
   
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
   width: 100%;
}

.back {
   font-size: 2rem;
}

.day-desc {
   border: 2px dotted;
   border-radius: 6px;
   border-color: var(--blockquote-border-color);
   padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
   if (knitr::is_latex_output()) {
      sprintf("\\texttt{%s}", as.character(x))
   } else if (knitr::is_html_output()) {
      sprintf("<tt>%s</tt>", as.character(x))
   } else {
      x
   }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(tibble)
library(dplyr)
library(stringr)
library(purrr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = TRUE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2021/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
   content(encoding = "UTF-8") %>% 
   xml_find_all("///article") %>% 
   lapply(as.character)

puzzle_data <- GET(paste0(base_url, "/input"),
                   session_cookie) %>% 
   content(encoding = "UTF-8") %>% 
   strsplit("\n") %>% 
   `[[`(1)

read_boards <- function(stream) {
   markers <- vapply(stream, nchar, integer(1))
   idx <- which(markers == 0L)
   idx <- matrix(c(head(idx, -1), tail(idx, -1)), ncol = 2) %>% 
      `colnames<-`(c("from", "to"))
   apply(idx, 1, function(ii) {
      tab <- stream[seq(ii[["from"]] + 1L, ii[["to"]] - 1L)] %>% 
         paste(collapse = "\n") %>% 
         trimws() %>% 
         read.table(text = .) %>% 
         as.matrix() %>% 
         `colnames<-`(NULL)
   }, simplify = FALSE)
}

puzzle_data <- list(draw = puzzle_data[1] %>% 
                       strsplit(",") %>% 
                       `[[`(1) %>% 
                       as.numeric(),
                    boards = read_boards(puzzle_data))
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

The most challenging part about this puzzle was to bring the data into the right format.
Once we have a vector of draws and a list of boards we can loop over the draws, mark all
matches as `TRUE` and calculate `[row|col]Sums`. If either matches the number of rows 
(cols) we have a BINGO. We use again `purrr:reduce` instead of an old-fashioned `for` 
loop just for the fun of it.

The binary function takes the following relevant arguments^[The original boards are passed
in as an argument as well though we could simply use it without an explicit argument. But
I prefer expliciteness overimpliciteness.]:
* `matches` a list of length 5 lists. The first element corresponds to the board.
The second element is a logical matrix showing the matches. The third element is a logical
flagging a bingo. The forth element is the last number drawn. The last number is the round
whihc we are in.
* `draw` a numeric given the next number.

```{r solution-a}
play_bingo <- function(rounds, draw, 
                       boards = puzzle_data$boards, 
                       stop_on_first = TRUE) {
   bingo_found <- FALSE
   res <- map(rounds, function(round) {
      bingo <- round$bingo
      if (!bingo) {
         res <- round$matches | round$board == draw 
         if (any(rowSums(res) == nrow(round$board),
                 colSums(res) == ncol(round$board))) {
            bingo <- TRUE
            bingo_found <<- TRUE
         }
         list(board = round$board, matches = res, bingo = bingo, draw = draw,
              round = round$round + 1)
      } else {
         round
      }
   })
   all_bingos <- all(map_lgl(res, ~ .x$bingo))
   if (bingo_found & stop_on_first | all_bingos) {
      done(res)
   } else {
      res
   }
}

init <- map(puzzle_data$boards, ~ list(board = .x,
                                       matches = matrix(FALSE, nrow(.x), ncol(.x)),
                                       bingo = FALSE,
                                       draw = NA, round = 0))

bingo_game <- reduce(puzzle_data$draw, play_bingo, 
                     .init = init)


winner <- Filter(\(l) l$bingo, bingo_game)
map_dbl(winner, ~ sum(.x$board[!.x$matches]) * .x$draw)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We can reuse the previous code, but instead of exiting pre-maturely on first BINGO, we
play until all boards have a bingo.

```{r solution-b, eval = length(puzzle) > 1}
bingo_game <- reduce(puzzle_data$draw, play_bingo, 
                     stop_on_first = FALSE, .init = init)

max_round <- map_dbl(bingo_game, ~ .x$round)
winner <- bingo_game[which.max(max_round)]
map_dbl(winner, ~ sum(.x$board[!.x$matches]) * .x$draw)

```


