---
params:
   task_nr: 10
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
	width: 100%;
}

.back {
	font-size: 2rem;
}

.day-desc {
	border: 2px dotted;
	border-radius: 6px;
	border-color: var(--blockquote-border-color);
	padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
	if (knitr::is_latex_output()) {
		sprintf("\\texttt{%s}", as.character(x))
	} else if (knitr::is_html_output()) {
		sprintf("<tt>%s</tt>", as.character(x))
	} else {
		x
	}
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(tibble)
library(dplyr)
library(purrr)
library(tidyr)
library(glue)
library(stringr)
library(R6)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2021/day/", params$task_nr)
puzzle <- GET(base_url,
				  session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	xml_find_all("///article") %>% 
	lapply(as.character)

## use curly braces in the last pipe to avoid that . is additionally added
## as the first argument
puzzle_data <- GET(paste0(base_url, "/input"),
						 session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	str_split("\n") %>% 
	`[[`(1) %>% 
	str_split("") %>% 
	head(n = -1L) ## remove empty string at end
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

My first intuition was to use RegEx, but after reading some [SO Answers](https://stackoverflow.com/questions/546433/regular-expression-to-match-balanced-parentheses) I am not sure whether RegEx is the right tool. The answer shows some funky posisbilities to solve it via RegEx, but hey, why going through all this hassle. Instead I
will use a good old loop (you may have guessed, `purrr::reduce` to the rescue, simply 
because of its nifty `rlang::done` support).

The idea is that we loop through the string and maintain a LIFO stack of the opened 
brackets. When we encounter a closing bracket, we check whether it has a corresponding 
open bracket on the top of the stack. If this is not the case, we indicate an error.


```{r solution-a}
## a naive implementation of a stack
Stack <- R6Class(
	"qstack",
	public = list(
		initialize = function(init_size = 1000L) {
			private$.stack <- rep(list(NULL), init_size)
			private$.head <- 0L
			private$.init_size <- private$.size <- as.integer(init_size)
		},
		push = function(elem) {
			if (private$.head == private$.size) {
				message(
					sprintf(paste("stack size is too small -",
									  "trying to reallociate %d more slots"),
							  private$.init_size))
				private$.stack <- c(private$.stack, 
										  rep(list(NULL), private$.init_size))
				private$.size <- private$.size + private$.init_size
			}
			private$.head <- private$.head + 1L
			private$.stack[[private$.head]] <- elem
			invisible(self)
		},
		pop = function() {
			if (private$.head == 0) {
				NULL 
			} else {
				el <- private$.stack[[private$.head]]
				private$.stack[private$.head] <- list(NULL)
				private$.head <- private$.head -1
				el
			}
		},
		size = function() {
			private$.head
		},
		max_size = function() {
			private$.size
		},
		print = function() {
			if (private$.head == 0L) {
				cat("# Empty stack\n")
			} else {
				indent <- "    "
				cat(glue("# Stack with {private$.head} element(s)"), "\n")
				str <- capture.output(private$.stack[seq(1, private$.head, 1L)])
				str <- paste0(indent, str)
				str <- str_replace_all(str, glue("^{indent}$"), 
											  glue("\n{indent}^^^\n", .trim = FALSE))
				str[length(str)] <- ""
				str <- str_replace(
					str, 
					glue("{indent}(\\[\\[{private$.head}\\]\\])"),
					"--> \\1")
				cat(str, sep = "\n")
			}
		}
	),
	private = list(
		.stack = NULL,
		.head = NA_integer_,
		.init_size = NA_integer_,
		.size = NA_integer_
	)
)

brackets_data <- tribble(
	~ open, ~ close, ~value_missing, ~value_adding,
	"(",    ")",     3L,             1L,
	"[",    "]",     57L,            2L,
	"{",    "}",     1197L,          3L,
	"<",    ">",     25137L,         4L)

is_opening <- function(chr) {
	chr %in% brackets_data$open
}

match_brackets <- function(a, b) {
	cls <- brackets_data %>% 
		filter(close == b)
	does_match <- cls %>% 
		filter(open == a) %>% 
		nrow() == 1L
	if (does_match) {
		0L
	} else {
		cls %>% 
			pull(value_missing)
	}
}

get_closing_value <- function(a) {
	brackets_data %>% 
		filter(open == a) %>% 
		pull(value_adding)
}

check_line <- function(line) {
	stack <- Stack$new(length(line))
	
	check <- function(value, chr) {
		if (is_opening(chr)) {
			stack$push(chr)
			res <- 0L
		} else {
			el <- stack$pop()
			res <- match_brackets(el, chr)
		}
		if (res > 0) {
			done(list(result = res, stack = stack))
		} else {
			list(result = res, stack = stack)
		}
	}
	
	reduce(line, check, .init = list(result = 0L, stack = stack))
}

res <- map(puzzle_data, check_line)
sum(map_int(res, pluck, "result"))

```



## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We filter the results from part 1 by dropping corrupt and valid codes, which leaves us
with incomplete codes (valid codes have an empty stack as all opening brackets were
dutifully closed). For those, we just close each bracket still on the stack and update
the penailty formula as indicated.


```{r solution-b, eval = length(puzzle) > 1}
## Please note the R 4.1. shortcut for anonymous function
incomplete <- Filter(\(r) r$result == 0L & r$stack$size() > 0, res)
get_score <- function(res) {
	sum <- 0
	while(res$stack$size() > 0) {
		sum <- sum * 5 + get_closing_value(res$stack$pop())	
	}
	sum
}

map_dbl(incomplete, get_score) %>% 
	median()
```


