---
params:
   task_nr: 9
   year: 2017
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- c(0L, 2L, 7L, 0L)
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We write a parser, which first remove `!.` sequences, then removes garbage sequences and
finally parses the groups. This is done by replacing each `{x}` by `list(x)`. The we need 
toparse the string and evaluate it. This will give us a nested list. The we recurse 
through this list and whenever we encounter a pure `list()` element we return the depth 
increased by 1. We also add the increased depth at the üarent to count this element and 
eventually we sum up the values to get the final score.

```{r get-solution-1}
parse_stream <- function(stream = puzzle_data) {
  cleaned_stream <- stream %>% 
    str_remove_all("!.") %>% 
    str_remove_all("<[^>]*>") %>% 
    str_replace_all(",{2,}", ",") %>% 
    str_replace_all(",\\}", "}") %>% 
    str_replace_all("\\{,", "{")
  parsed_stream <- cleaned_stream
  while (str_detect(parsed_stream, "\\{")) {
    parsed_stream <- str_replace_all(parsed_stream, "\\{([^{]*)\\}", "list(\\1)")
  }
  lists <- parse(text = parsed_stream) %>% 
    eval()
  count_groups <- function(l, list_index) {
    if (identical(l, list())) {
      result <- list_index + 1L
    } else {
      result <- map(l, count_groups, list_index + 1) %>% 
        c(list_index + 1L) %>% 
        unlist() %>% 
        sum()
    }
    result
  }
  count_groups(lists, 0L)
} 
parse_stream(puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We solve this with some regex. First we remove all canceled characters, then extract all 
garbage groups, remove the the garbage delimiters and fnally count the characters.

```{r get-solution-2}
count_garbage <- function(stream) {
  stream %>% 
    str_remove_all("!.") %>% 
    str_extract_all("<[^>]*>") %>% 
    extract2(1L) %>% 
    str_remove_all("^<|>$") %>% 
    nchar() %>% 
    sum()
}
count_garbage(puzzle_data)
```
