---
params:
   task_nr: 12
   year: 2019
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(bit64)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  res <- text_block %>% 
    str_extract_all("-?\\d+") %>% 
    do.call(rbind, .) %>% 
    set_colnames(c("x", "y", "z"))
  storage.mode(res) <- "integer"
  res
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

The first part can be solved literally, iterate over n, calculate the velocity and 
the new position.

```{r get-solution-1}
move_moons <- function(pos = puzzle_data, n = 1000L) {
  nr_moons <- nrow(pos)
  vel <- matrix(0L, 4L, 3L) 
  for (steps in seq_len(n)) {
    for (i in seq_len(nr_moons - 1L)) {
      for (j in (i + 1L):nr_moons) {
        delta <- sign(pos[i, ] - pos[j, ])
        vel[i, ] <- vel[i, ] - delta
        vel[j, ] <- vel[j, ] + delta
      }
    }
    pos <- pos + vel
  }
  sum(rowSums(abs(pos)) * rowSums(abs(vel)) )
}
move_moons(puzzle_data, 1000)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For the second part, we observe that the three dimensions evolve independently. Thus, 
instead for finding the period for all dimensions **simultaneously**, we find the period 
for each dimension **seperately**. Then, all we need to do is to find the lowest common 
multiple of these periods, which corresponds to the overall period.

```{r get-solution-2, cache = TRUE}
gcd <- function(a, b) {
  while (b != 0) {
    t <- b
    b <- a %% b
    a <- t
  }
  abs(a)
}

lcm <- function(...) {
  x <- list(...) %>% 
    lapply(as.integer64)
  Reduce(\(a, b) abs(a * b) %/% gcd(a, b), x)
}

find_period <- function(pos = puzzle_data) {
  solve_period <- function(dim) {
    nr_moons <- nrow(pos)
    pos <- start_pos <- pos[, dim]
    vel <- start_vel <- rep(0L, nr_moons) 
    found <- FALSE
    period <- as.integer64(0L)
    while (!found) {
      for (i in seq_len(nr_moons - 1L)) {
        for (j in (i + 1L):nr_moons) {
          delta <- sign(pos[i] - pos[j])
          vel[i] <- vel[i] - delta
          vel[j] <- vel[j] + delta
        }
      }
      pos <- pos + vel
      period <- period + 1L
      found <- all(pos == start_pos) && all(vel == start_vel)
    }
    period
  }
  lapply(1:3, solve_period) %>% 
    do.call(lcm, .)
}
find_period(puzzle_data)

```
