---
params:
   task_nr: 1
   year: 2023
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L)
  }
  text_block[nzchar(text_block)]
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

Hohoho! Welcome to this year's AoC. Let the games begin (again). `regex` will solve this
puzzle in no time.

```{r get-solution-1}
decode_calibrations <- function(calibration) {
  rex <- paste0("(?m)^.*?(?<first>\\d)(?:.*(?<last>\\d))?.*$")
  nrs <- str_match_all(paste(calibration, collapse = "\n"), rex)[[1L]]
  ## if there is just one number duplicate the first number
  paste0(nrs[, "first"], coalesce(nrs[, "last"], nrs[, "first"])) %>% 
    as.integer() %>% 
    sum()
}

decode_calibrations(puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

Pretty much the same as part 1, but this time we also take into account spelled out 
digits. Care has to be taken of tokens like `oneight`, b/c in the beginning of the string 
this should translate to `1` while at the end it should translate to `8`.

```{r get-solution-2}
decode_calibrations_advanced <- function(calibration) {
  digits <- tibble(
    n1 = 1:9,
    d1 = c("one", "two", "three", "four", "five",
           "six", "seven", "eight", "nine")
  )
  lkp <- expand.grid(d1 = digits$d1, d2 = digits$d1, stringsAsFactors = FALSE) %>% 
    mutate(combo = if_else(str_sub(d1, -1L) == str_sub(d2, 1L, 1L),
                           paste0(d1, str_sub(d2, 2L)),
                           NA_character_),
           n1 = match(d1, digits$d1),
           n2 = match(d2, digits$d1)) %>% 
    filter(!is.na(combo)) %>% 
    bind_rows(
      tibble(key = digits$d1, n1 = digits$n1, n2 = digits$n1)
    ) %>%
    mutate(key = coalesce(combo, key)) %>%
    select(key, n1, n2)

  parse_number <- function(x, take_first, lkp) {
    xn <- suppressWarnings(as.integer(x))
    nok <- is.na(xn)
    idx <- match(x[nok], lkp$key)
    xn[nok] <- if (take_first) lkp$n1[idx] else lkp$n2[idx]
    xn
  }

  alt <- paste(c(lkp$key, "\\d"), collapse = "|")
  rex <- paste0("(?m)^.*?(?<first>", alt, ")(?:.*(?<last>",
                alt, "))?.*$")

  nrs <- str_match_all(paste(calibration, collapse = "\n"), rex)[[1L]]
  ## if there is just one number duplicate the first number
  nrs[, "last"] <- coalesce(nrs[, "last"], nrs[, "first"])
  sum(parse_number(nrs[, "first"], TRUE, lkp) * 10L + 
        parse_number(nrs[, "last"], FALSE, lkp))
}

decode_calibrations_advanced(puzzle_data)
```
