---
params:
   task_nr: 23
   year: 2020
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(bit64)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    str_split("") %>% 
    extract2(1L) %>% 
    as.integer()
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We implement the swap using a linked list. `circle[i]` stores which cup is **after** cup
`i`. A swap needs then to change `circle[current]` to the position the third cup (`p3`) 
was pointing to and `p3` must point where the destination cup was originally pointing to.
Eventually, the destination cup must point to to the start of the 3 removed cups.

```{r get-solution-1}
move_cups <- function(start_pos, times, n = length(start_pos)) {
  circle <- integer(n)
  k <- length(start_pos)
  
  for (i in seq_len(k - 1L)) {
    circle[start_pos[i]] <- start_pos[i + 1L]
  }
  
  if (n > k) {
    circle[start_pos[k]] <- k + 1L
    for (i in (k + 1L):(n - 1L)) {
      circle[i] <- i + 1L
    }
    circle[n] <- start_pos[1L]
  } else {
    circle[start_pos[k]] <- start_pos[1L]
  }
  current <- start_pos[1]

  for (i in seq_len(times)) {

    # pick up
    p1 <- circle[current]
    p2 <- circle[p1]
    p3 <- circle[p2]

    after_pick <- circle[p3]
    circle[current] <- after_pick

    # destination
    dest <- current - 1L
    if (dest == 0L) {
      dest <- n
    }
    while (dest == p1 || dest == p2 || dest == p3) {
      dest <- dest - 1L
      if (dest == 0L) {
        dest <- n
      }
    }

    # insert
    circle[p3] <- circle[dest]
    circle[dest] <- p1

    # advance
    current <- circle[current]
  }
  circle
}
find_position <- function(start_pos, times) {
  cups <- move_cups(puzzle_data, times)
  res <- integer()
  x <- cups[1]
  while (x != 1) {
    res <- c(res, x)
    x <- cups[x]
  }
  paste(res, collapse = "") %>% 
    as.integer()
}

find_position(puzzle_data, 100L)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]] %>% str_replace_all(fixed("*"), "\\*"))
```

### Solution

With the pointer logic in place increasing the numbers is still a feasible operation.

```{r get-solution-2}
find_stars <- function(start_pos, times, n) {
  cups <- move_cups(puzzle_data, times, n)
  as.integer64(cups[1L]) * as.integer64(cups[cups[1L]])
}

find_stars(puzzle_data, 10000000L, 1000000L)

```
