---
params:
   task_nr: 15
   year: 2019
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
 text_block %>% 
    str_split(",") %>% 
    do.call(c, .) %>% 
    as.numeric()
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We further finetune the code from [Day 13 @ 2019](2019_task13.html) (and get rid of some 
legacy code). The basic idea is to first build the complete maze BFS (as we need the full 
maze for part 2 anyways). Then, we run a BFS to find the shortest path form start to end.

```{r get-solution-1}
print.maze <- function(x, ...) {
  margins <- apply(x[, 1:2], 2L, min)
  x_norm <- sweep(x[, 1:2], 2, margins, \(x, y) x - y + 1L)
  m <- matrix(".", max(x_norm[, 2L]), max(x_norm[, 1L]))
  m[x_norm[, 2:1]] <- case_when(
    x[, 3L] == -1L ~ "S",
    x[, 3L] == 0L  ~ "#",
    x[, 3L] == 1L  ~ " ",
    x[, 3L] == 2L  ~ "Z"
  )
  apply(m, 1, paste, collapse = "") %>% 
    paste(collapse = "\n") %>% 
    cat()
  invisible(x)
}

parse_op_codes_fast <- function(op_codes = puzzle_data, input = numeric(0L), 
                                verbose = FALSE) {
  ip <- 1L
  rel_base <- 0L
  
  out_buffer <- numeric(0L)
  out_len <- 0L
  out_block <- integer(1L)
  block_size <- 1L  # size of memory
  
  maze <- matrix(c(0L, 0L, -1L), 1L, 3L, 
                 dimnames = list(NULL, c("x", "y", "type")))
  
  grow_memory <- function(idx) {
    if (idx > length(op_codes)) {
      new_size <- ceiling(idx / block_size) * block_size
      op_codes <<- c(op_codes, rep(0, new_size - length(op_codes)))
    }
  }
  
  append_out <- function(val) {
    if (out_len + 1L > length(out_buffer)) {
      new_size <- length(out_buffer) + out_block
      out_buffer <<- c(out_buffer, rep(0, new_size - length(out_buffer)))
    }
    out_len <<- out_len + 1L
    out_buffer[out_len] <<- val
  }
  
  get_val <- function(param, mode) {
    if (mode == 0L) {
      grow_memory(param + 1L)
      return(op_codes[param + 1L])
    } else if (mode == 1L) {
      return(param)
    } else if (mode == 2L) {
      idx <- rel_base + param + 1L
      grow_memory(idx)
      return(op_codes[idx])
    } else {
      stop("unknown mode")
    }
  }
  
  get_addr <- function(param, mode) {
    if (mode == 0L) {
      return(param + 1L)
    } else if (mode == 2L) {
      return(rel_base + param + 1L)
    } else {
      stop("invalid write mode")
    }
  }
  
  get_key <- function(...) {
    paste(c(...), collapse = ",")
  }
  
  pick_dir <- function(pos) {
    for (d in seq_along(dirs)) {
      dir <- dirs[[d]]
      np <- pos + dir
      k <- get_key(np)
      if (!exists(k, visited)) {
        return (d)
      }
    }
    ## no unseen possibility at this stage => go back
    if (length(stack) > 1L) {
      return(stack[[length(stack)]]$back)
    }
    return (NA_integer_)
  }
  
  dirs <- list(c(0L, -1L), 
               c(0L, 1L), 
               c(-1L, 0L), 
               c(1L, 0L))
  opp_dir <- c(2L, 1L, 4L, 3L)
  
  
 visited <- new.env(parent = emptyenv())

  pos <- c(0L, 0L)
  
  visited[[get_key(pos)]] <- TRUE
  
  stack <- list(
    list(
      pos = pos,
      back = NA_integer_
    )
  )
  
  halt <- done <- FALSE
  while (!halt) {
    instr <- op_codes[ip]
    op <- instr %% 100L
    modes <- c(
      (instr %/% 100L) %% 10L, 
      (instr %/% 1000L) %% 10L, 
      (instr %/% 10000L) %% 10L
    )
    
    if (op == 1L) { # add
      a <- get_val(op_codes[ip + 1L], modes[1L])
      b <- get_val(op_codes[ip + 2L], modes[2L])
      addr <- get_addr(op_codes[ip + 3L], modes[3L])
      grow_memory(addr)
      op_codes[addr] <- a + b
      ip <- ip + 4L
    } else if (op == 2L) { # mul
      a <- get_val(op_codes[ip + 1L], modes[1L])
      b <- get_val(op_codes[ip + 2L], modes[2L])
      addr <- get_addr(op_codes[ip + 3L], modes[3L])
      grow_memory(addr)
      op_codes[addr] <- a * b
      ip <- ip + 4L
    } else if (op == 3L) { # input
      if (done) {
        halt <- TRUE
        next
      }
      dir <- pick_dir(pos)
      if (is.na(dir)) {
        done <- TRUE
        next
      }
      addr <- get_addr(op_codes[ip + 1L], modes[1L])
      grow_memory(addr)
      op_codes[addr] <- dir
      ip <- ip + 2L
    } else if (op == 4L) { # output
      val <- get_val(op_codes[ip + 1L], modes[1L])
      step <- dirs[[dir]]
      
      prev_pos <- pos
      next_pos <- pos + step
      k <- get_key(next_pos)
      if (!exists(k, visited)) {
        maze <- rbind(maze, c(next_pos, val))
      }
      visited[[k]] <- TRUE
      if (val %in% 1:2) {
        pos <- next_pos
        back_dir <- opp_dir[dir]
        if (!is.na(stack[[length(stack)]]$back) &&
            dir == stack[[length(stack)]]$back) {
          ## backtrack
          stack <- stack[-length(stack)]
        } else {
          ## new depth save state
          stack <- c(stack, list(list(pos = pos, back = back_dir)))
        }

      }

      if (verbose) cat(val, "\n")
      ip <- ip + 2L
    } else if (op == 5L) { # jump-if-true
      a <- get_val(op_codes[ip + 1L], modes[1L])
      b <- get_val(op_codes[ip + 2L], modes[2L])
      if (a != 0L) {
        ip <- b + 1L
      } else {
        ip <- ip + 3L
      }
    } else if (op == 6L) { # jump-if-false
      a <- get_val(op_codes[ip + 1L], modes[1L])
      b <- get_val(op_codes[ip + 2L], modes[2L])
      if (a == 0L) {
        ip <- b + 1L
      } else {
        ip <- ip + 3L
      }
    } else if (op == 7L) { # less than
      a <- get_val(op_codes[ip + 1L], modes[1L])
      b <- get_val(op_codes[ip + 2L], modes[2L])
      addr <- get_addr(op_codes[ip + 3L], modes[3L])
      grow_memory(addr)
      op_codes[addr] <- as.numeric(a < b)
      ip <- ip + 4L
    } else if (op == 8L) { # equals
      a <- get_val(op_codes[ip + 1L], modes[1L])
      b <- get_val(op_codes[ip + 2L], modes[2L])
      addr <- get_addr(op_codes[ip + 3L], modes[3L])
      grow_memory(addr)
      op_codes[addr] <- as.numeric(a == b)
      ip <- ip + 4L
    } else if (op == 9L) { # adjust relative base
      a <- get_val(op_codes[ip + 1L], modes[1L])
      rel_base <- rel_base + a
      ip <- ip + 2L
    } else if (op == 99L) { # halt
      halt <- TRUE
    } else {
      stop("Unknown opcode ", op)
    }
  }
  class(maze) <- "maze"
  maze
}
shortest_path <- function(maze) {
  start <- maze[maze[, "type"] == -1L, c("x", "y")]
  goal <- maze[maze[, "type"] == 2L, c("x", "y")]
  
  free_cells <- maze[maze[, "type"] != 0L, c("x", "y")]
  free_keys <- paste(free_cells[, 1L], free_cells[, 2L], sep = ",")
  free_set <- new.env(parent = emptyenv())
  visited <- new.env(parent = emptyenv())
  for (k in free_keys) {
    free_set[[k]] <- TRUE
  }

  get_key <- function(p) {
    paste(p, collapse = ",")
  }
  
  dirs <- list(
    c(0L, -1L), 
    c(0L, 1L), 
    c(-1L, 0L), 
    c(1L, 0L)
  )
  
  # BFS-Queue
  queue <- list(
    list(
      pos = start, 
      path = list()
      )
  )
  visited[[get_key(start)]] <- TRUE

  while (length(queue) > 0L) {
    current <- queue[[1L]]
    queue <- queue[-1L]
    
    if (identical(current$pos, goal)) {
      return(unlist(current$path))
    }
    for (d in seq_along(dirs)) {
      np <- current$pos + dirs[[d]]
      k <- get_key(np)
      if (exists(k, free_set) && !exists(k, visited)) {
        visited[[k]] <- TRUE
        
        queue <- c(
          queue, 
          list(list(pos = np, 
                    path = c(current$path, d))
          )
        )
      }
    }
  }
  return(NULL)
}

maze <- parse_op_codes_fast(puzzle_data)
shortest_path(maze) %>% 
  length()
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For the second part, we start from the goal and flood the maze with oxygen.

```{r get-solution-2}
flood <- function(maze) {
  get_neighbors <- function(pos) {
    dirs <- rbind(
      c(-1L, 0L),
      c(1L, 0L),
      c(0L, -1L),
      c(0L, 1L)
    )
    nbs <- apply(pos, 1L, function(r) t(t(dirs) + r), simplify = FALSE) %>% 
      do.call(rbind, .)
    nbs <- nbs[!duplicated(nbs), , drop = FALSE]
    nbs <- nbs[between(nbs[, 1L], 1L, nrow(m)) &
                 between(nbs[, 2L], 1L, ncol(m)), , drop = FALSE]
    if (nrow(nbs) == 0L) {
      return(matrix(NA_integer_, 0L, 2L))
    }
    free <- is.na(m[nbs])
    return (nbs[free, , drop = FALSE])
  }
  margins <- apply(maze[, 1:2], 2L, min)
  x_norm <- sweep(maze[, 1:2], 2, margins, \(x, y) x - y + 1L)
  m <- matrix("#", max(x_norm[, 2L]), max(x_norm[, 1L]))
  m[x_norm[, 2:1]] <- case_when(
    maze[, 3L] %in% c(-1L, 1L, 2L) ~ NA_character_,
    maze[, 3L] == 0L  ~ "#"
  )
  queue <- x_norm[maze[, 3L] == 2L, 2:1, drop = FALSE]
  mins <- -1L
  while(nrow(queue) > 0L) {
    mins <- mins + 1L
    pos <- queue
    queue <- get_neighbors(pos)
    m[pos] <- "O"
  }
  mins
}
flood(maze)
```
