---
params:
   task_nr: 17
   year: 2023
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(collections)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L)
  }
  res <- text_block[nzchar(text_block)] %>% 
    str_split("") %>% 
    do.call(rbind, .)
  storage.mode(res) <- "integer"
  res
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}

cat(puzzle[[1]] %>% 
      str_replace_all("\\\\", "\\\\\\\\"))
```

### Solution

We solve this using a Dijkstra algorithm with an extended state, including the direction
and the umber of steps taken so far in this directions to honor the movement restricions.

```{r get-solution-1, cache = TRUE}
find_path <- function(map) {
  nr <- nrow(map)
  nc <- ncol(map)
  goal <- matrix(dim(map), 1L, dimnames = list(NULL, c("row", "col")))
  start <- cbind(row = 1L, col = 1L)
  dirs <- rbind(
    "^" = c(-1L, 0L),
    ">" = c(0L, 1L),
    "v" = c(1L, 0L),
    "<" = c(0L, -1L)
  )
  dist <- array(Inf, dim = c(dim(map), 4L, 3L))
  pq <- priority_queue()
  for (dir in 1:4) {
    state <- cbind(start, dir = dir, steps = 0L, cost =  0L)
    dist[cbind(start, dir, 1L)] <- 0L
    pq$push(state, 0L)
  }
  while (pq$size() > 0L) {
    state <- pq$pop()
    if (all(state[, 1:2] == goal)) {
      return(state[, "cost"])
    }
    if (state[, "steps"] > 0L && state[, "cost"] > dist[state[, -5L, drop = FALSE]]) {
      ## we reached this field already at lower costs => skip
      next
    }
    if (state[, "steps"] < 3L && state[, "steps"] > 0L) {
      ## walk straight
      new_pos <- state[, 1:2, drop = FALSE] + dirs[state[, "dir"], ]
      if (new_pos[, 1L] >= 1L && new_pos[, 1L] <= nr &&
          new_pos[, 2L] >= 1L && new_pos[, 2L] <= nc) {
        new_cost <- state[, "cost"] + map[new_pos]
        idx <- cbind(new_pos, dir = state[, "dir"], steps = state[, "steps"] + 1L)
        if (new_cost < dist[idx]) {
          dist[idx] <- new_cost
          pq$push(cbind(idx, cost = new_cost), -new_cost)
        }
      }
    }
    for (new_dir in 1:4) {
      if (new_dir %% 2L == state[, "dir"] %% 2L) {
        ## no straight or 180° turns
        next
      }
      new_pos <- state[, 1:2, drop = FALSE] + dirs[new_dir, ]
      if (new_pos[, 1L] >= 1L && new_pos[, 1L] <= nr &&
          new_pos[, 2L] >= 1L && new_pos[, 2L] <= nc) {
        new_cost <- state[, "cost"] + map[new_pos]
        idx <- cbind(new_pos, dir = new_dir, steps = 1L)
        if (new_cost < dist[idx]) {
          dist[idx] <- new_cost
          pq$push(cbind(idx, cost = new_cost), -new_cost)
        }
      }
    }
  }
  NA_integer_
}

find_path(puzzle_data)

```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We use the same algorithmic idea, but implement the new moving restrictions.

```{r get-solution-2, cache = TRUE}
find_path_ultra <- function(map) {
  nr <- nrow(map)
  nc <- ncol(map)
  goal <- matrix(dim(map), 1L, dimnames = list(NULL, c("row", "col")))
  start <- cbind(row = 1L, col = 1L)
  dirs <- rbind(
    "^" = c(-1L, 0L),
    ">" = c(0L, 1L),
    "v" = c(1L, 0L),
    "<" = c(0L, -1L)
  )
  dist <- array(Inf, dim = c(dim(map), 4L, 10L))
  pq <- priority_queue()
  for (dir in 1:4) {
    state <- cbind(start, dir = dir, steps = 0L, cost =  0L)
    dist[cbind(start, dir, 1L)] <- 0L
    pq$push(state, 0L)
  }
  r <- 0L
  while (pq$size() > 0L) {
    r <- r + 1
    state <- pq$pop()
    if (all(state[, 1:2] == goal)) {
      return(state[, "cost"])
    }
    if (state[, "steps"] > 0L && state[, "cost"] > dist[state[, -5L, drop = FALSE]]) {
      ## we reached this field already at lower costs => skip
      next
    }
    if (state[, "steps"] < 10L && state[, "steps"] > 0L) {
      ## walk straight
      new_pos <- state[, 1:2, drop = FALSE] + dirs[state[, "dir"], ]
      if (new_pos[, 1L] >= 1L && new_pos[, 1L] <= nr &&
          new_pos[, 2L] >= 1L && new_pos[, 2L] <= nc) {
        new_cost <- state[, "cost"] + map[new_pos]
        idx <- cbind(new_pos, dir = state[, "dir"], steps = state[, "steps"] + 1L)
        if (new_cost < dist[idx]) {
          dist[idx] <- new_cost
          pq$push(cbind(idx, cost = new_cost), -new_cost)
        }
      }
    }
    if (state[, "steps"] == 0L || state[, "steps"] >= 4L) {
      for (new_dir in 1:4) {
        if (new_dir %% 2L == state[, "dir"] %% 2L) {
          ## no straight or 180° turns
          next
        }
        new_pos <- state[, 1:2, drop = FALSE] + dirs[new_dir, ]
        if (new_pos[, 1L] >= 1L && new_pos[, 1L] <= nr &&
            new_pos[, 2L] >= 1L && new_pos[, 2L] <= nc) {
          new_cost <- state[, "cost"] + map[new_pos]
          idx <- cbind(new_pos, dir = new_dir, steps = 1L)
          if (new_cost < dist[idx]) {
            dist[idx] <- new_cost
            pq$push(cbind(idx, cost = new_cost), -new_cost)
          }
        }
      }
    }
  }
  NA_integer_
}

find_path_ultra(puzzle_data)
```
