---
params:
   task_nr: 11
   year: 2022
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(glue)
library(bit64)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  parse_monkey <- function(monkey_block) {
    items <- str_remove_all(monkey_block[2L], "Starting items: |\\s+") %>% 
      str_split(",") %>% 
      unlist() %>% 
      as.integer()
    fn_boilerplate <- paste("function(old) {{",
                            "   {code}",
                            "   return (new)",
                            "}}", sep = "\n")
    code <- str_remove(monkey_block[3L], "\\s*Operation: ")
    op <- glue(fn_boilerplate) %>% 
      parse(text = .) %>% 
      eval()
    args <- str_extract_all(monkey_block[4:6], "\\d+") %>% 
      unlist() %>% 
      as.integer()
    fn_boilerplate <- paste(
      "function(wl) {{",
      "   if_else(wl %% {args[[1L]]} == 0L, {args[[2L]]}, {args[[3L]]})",
      "}}", sep = "\n")
    test <- glue(fn_boilerplate) %>% 
      parse(text = .) %>% 
      eval()
    list(
      items = items,
      op = op,
      test = test,
      mod = args[1L]
    )
  }
  monkey_markers <- str_which(text_block, "Monkey")
  map(monkey_markers, ~ parse_monkey(text_block[.x:(.x + 5L)]))
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

The first part is rather straight forward, pass the items as indicated and count how often
a monkey inspects an element to find the largest numbers. We enlarge the item vector and
assign rather than copy in an attempt to be reasonable fast.


```{r get-solution-1}
play_keep_away <- function(monkeys, rounds, relief = function(x) floor(x / 3)) {
  nr_items <- map_int(monkeys, ~ length(.x$items)) %>% 
    sum()
  for (m in seq_along(monkeys)) {
    ## add required fields and enlarge item vectors to avoid copying
    k <- length(monkeys[[m]]$items)
    monkeys[[m]]$nr_inspections <- as.integer64(0L)
    monkeys[[m]]$nr_items <- k
    monkeys[[m]]$items <- c(monkeys[[m]]$items, rep(NA_integer_, nr_items - k))
  }
  for (rnd in seq_len(rounds)) {
    for (m in seq_along(monkeys)) {
      monkey <- monkeys[[m]]
      if (monkey$nr_items == 0L) {
        next
      }
      
      items <- relief(monkey$op(monkey$items[seq_len(monkey$nr_items)]))
      targets <- monkey$test(items)
      for (nr in unique(targets)) {
        mv <- items[targets == nr]
        k <- length(mv)
        nr <- as.integer(nr) + 1L
        cur <- monkeys[[nr]]$nr_items
        monkeys[[nr]]$items[(cur + 1L):(cur + k)] <- mv
        monkeys[[nr]]$nr_items <- cur + k
      }
      monkeys[[m]]$items[] <- NA_integer_
      monkeys[[m]]$nr_items <- 0L
      monkeys[[m]]$nr_inspections <- monkeys[[m]]$nr_inspections + length(items)
    }
  }
  map(monkeys, "nr_inspections") %>% 
    do.call(c, .) %>% 
    sort(decreasing = TRUE) %>% 
    extract(1:2) %>% 
    prod()
}

play_keep_away(puzzle_data, 20L)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We cannot use the same algorithm because the worry level would explode. However, we can 
use a relief function which forms the LCM from all modulo operands, and uses this as a new
modulo operator for a global relief function. As all operands are prime numbers, we do not
lose any information by restricting the worry level to be in this range.

```{r get-solution-2}
lcm <- map_int(puzzle_data, "mod") %>% 
  prod()
play_keep_away(puzzle_data, 10000, \(x) x %% lcm)
```