---
params:
   task_nr: 22
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("2021_%s",
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
	width: 100%;
}

.back {
	font-size: 2rem;
}

.day-desc {
	border: 2px dotted;
	border-radius: 6px;
	border-color: var(--blockquote-border-color);
	padding: 0 20px;
}

code em {
	font-weight:bold
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
	if (knitr::is_latex_output()) {
		sprintf("\\texttt{%s}", as.character(x))
	} else if (knitr::is_html_output()) {
		sprintf("<tt>%s</tt>", as.character(x))
	} else {
		x
	}
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(tibble)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(R6)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2021/day/", params$task_nr)
puzzle <- GET(base_url,
				  session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	xml_find_all("///article") %>% 
	lapply(as.character)

puzzle_data <- GET(paste0(base_url, "/input"),
						 session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	str_split("\n") %>% 
	`[[`(1L) %>% 
	Filter(nzchar, .) %>% 
	str_extract_all("on|off|-?\\d+") %>% 
	do.call(rbind, .) %>% 
	`colnames<-`(c("switch", "x0", "x1", "y0", "y1", "z0", "z1")) %>% 
	as_tibble() %>% 
	mutate(across(x0:z1, as.integer))
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1L]])
```


### Solution

The first observation is that an intersection between 2 cubes is another cube, whose
dimensions are defined by the `max` and the `min` of the respective interval limits. For
instance the intersection between `[1,4] x [2,5] x [3,6]` and `[0,1] x [1,3] x [4,5]` is
determined by `[max(1,0),min(4,1)] x [max(2,1),min(5,3)] x [max(3,4),min(6,5)]` = 
`[1,1] x [2,3] x [4,5]`.

The second observation is that counting all blocks, which are switched on by a single cube
corresponds to simply getting the volume of the cube.

The third observation leads us to the [Inclusion-Exclusion Principle](https://en.wikipedia.org/wiki/Inclusion%E2%80%93exclusion_principle). That is,
to get all blocks which are switched on, we have to sum up all volumes and add and remove
the intersecting cubes according to the inclusion exclusion principle. That is, we have to 
maintain a list of all intersecting cubes and when we add (or remove, see next point) a
new cube we simply have to intersect the new cube wiht all previous intersections.

The final observation is that __substracting__ a cube differs from __adding__ a cube only 
by the fact for the latter we add the cube itself as well, while for the former we simply
intersect the deleting cube with all cubes already in the list.

This suggests the following algorithm:

1. Add the first cube to the list `I`.
1. Loop through all remaining cubes and for each new cube `I_j` intersect it with all
cubes and intersections already in the list `I` resulting in `I'`.
1. If we are adding, add the cube itself to `I'`.
1. Keep a count of how many cubes took part in an intersections.
1. Add the new intersections `I'` to `I`.
1. Once done, sum the volumes of each intersection multiplied by `(-1)` if there is
an even number of cubes in the respective intersection.

To solve the puzzle eventually, all which is left to do is to remove the cubes which are 
outside the dedicated area.

```{r solution-a}
solve <- function(data = puzzle_data, filter_inner = TRUE) {
	volume <- function(x0, x1, y0, y1, z0, z1, ...) {
		(x1 - x0 + 1) * (y1 - y0 + 1) * (z1 - z0 + 1) 
	}
	intersect <- function(intersections, new_cube) {
		left <- grep("0$", names(intersections), value = TRUE)
		right <- grep("1$", names(intersections), value = TRUE)
		right_cols <- map_dfc(right, ~ list(pmin(intersections[[.x]], 
															  new_cube[[.x]])) %>% 
									 	set_names(.x))
		left_cols <- map_dfc(left, ~ list(pmax(intersections[[.x]], 
															new_cube[[.x]])) %>% 
										set_names(.x))
		res <- bind_cols(left_cols, right_cols)
		res %>% 
			mutate(level = (intersections[["level"]] %||% -1) + 1) %>% 
			filter(x0 <= x1 & y0 <= y1 & z0 <= z1) %>% 
			mutate(volume = do.call(volume, .))
		
	}
	if (filter_inner) {
		data <- data %>% 
			filter(x0 >= -50 & y0 >= -50 & z0 >= -50 &
					 	x1 <= 50 & y1 <= 50 & z1 <= 50)
	}
	
	all_intersections <- data %>% 
		slice(1L) %>% 
		select(-switch) %>% 
		mutate(volume = do.call(volume, .),
				 level = 0L)
	for (r_idx in seq(2L, nrow(data))) {
		new_cube <- data %>% 
			slice(r_idx) %>% 
			mutate(volume = do.call(volume, .),
					 level = 0L)
		new_intersections <- intersect(all_intersections, new_cube)
		if (new_cube$switch == "on") {
			new_intersections <- bind_rows(new_intersections, new_cube %>% 
													 	select(-switch))
		}
		all_intersections <- bind_rows(all_intersections, new_intersections)
	}
	all_intersections %>% 
		summarise(volume = sum((-1) ^ level * volume)) %>% 
		pull(volume) %>% 
		sprintf("%.0f", .)
}
solve()
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

```{r solution-b}
solve(filter_inner = FALSE)
```

## Original Solution

My original solution based on the same idea, but was utterly slow due to overengineered 
`S4` class and growing lists. I list it here for the sake of completeness, but the running 
time suggests not to try it with the data set.

```{r solution-a-original, eval = FALSE}
Cube <- setClass("Cube", slots = c(.x0 = "integer", .x1 = "integer",
											  .y0 = "integer", .y1 = "integer",
											  .z0 = "integer", .z1 = "integer"))

setGeneric("is_empty", function(object) {
	standardGeneric("is_empty")
})

setGeneric("volume", function(object) {
	standardGeneric("volume")
})

setValidity("Cube", function(object) {
	all_slots <- c(outer(c(".x", ".y", ".z"), 0:1, paste0))
	empty_slots <-  all_slots %>% 
		map_lgl(~ length(slot(object, .x)) == 0L)
	if (all(empty_slots)) {
		TRUE
	} else if (any(empty_slots)) {
		## only some are empty but __not__ all
		glue("slot `{all_slots[empty_slots]}` is empty but at least ",
			  "one other lsot is not empty")
	} else {
		## all slots are filled
		OK <- c(object@.x1 >= object@.x0,
				  object@.y1 >= object@.y0,
				  object@.z1 >= object@.z0)
		if (any(!OK)) {
			faulty_slots <- c("x", "y", "z")[!OK]
			start_points <- map_int(faulty_slots, 
											~ slot(object, paste0(".", .x, "0")))
			end_points <- map_int(faulty_slots, 
										 ~ slot(object, paste0(".", .x, "1")))
			glue("left interval limit `{faulty_slots}0` ",
				  "[{start_points}] is bigger than ",
				  "right interval limit `{faulty_slots}1` [{end_points}]")
		} else {
			TRUE
		}
	}
})

setMethod("initialize", "Cube",
			 function(.Object, x0 = NULL, x1 = NULL,
			 			y0 = NULL, y1 = NULL,
			 			z0 = NULL, z1 = NULL, ...) {
			 	.Object <- callNextMethod(.Object, ...)
			 	.Object@.x0 <- as.integer(x0)
			 	.Object@.x1 <- as.integer(x1)
			 	.Object@.y0 <- as.integer(y0)
			 	.Object@.y1 <- as.integer(y1)
			 	.Object@.z0 <- as.integer(z0)
			 	.Object@.z1 <- as.integer(z1)
			 	validObject(.Object)
			 	.Object
			 })

setMethod("is_empty", "Cube",
			 function(object) {
			 	## it suffices to test .x0 for zero length b/c the validator assures
			 	## that either all are empty or none
			 	length(object@.x0) == 0L
			 })

setMethod("volume", "Cube",
			 function(object) {
			 	if (is_empty(object)) {
			 		0L
			 	} else {
			 		(object@.x1 - object@.x0 + 1L) *
			 			(object@.y1 - object@.y0 + 1L) *
			 			(object@.z1 - object@.z0 + 1L)
			 	}
			 })

setMethod("show", "Cube",
			 function(object) {
			 	if (is_empty(object)) {
			 		cat("<Empty Cube>\n")	
			 	} else {
			 		cat(glue("[{object@.x0}, {object@.x1}] x ",
			 					"[{object@.y0}, {object@.y1}] x ",
			 					"[{object@.z0}, {object@.z1}]")
			 		)
			 	}
			 })

setMethod("-", c("Cube", "Cube"),
			 function(e1, e2) {
			 	if (is_empty(e1) | is_empty(e2)) {
			 		Cube()
			 	} else {
			 		x0 <- max(e1@.x0, e2@.x0)
			 		x1 <- min(e1@.x1, e2@.x1)
			 		y0 <- max(e1@.y0, e2@.y0)
			 		y1 <- min(e1@.y1, e2@.y1)
			 		z0 <- max(e1@.z0, e2@.z0)
			 		z1 <- min(e1@.z1, e2@.z1)
			 		if (x1 < x0 || y1 < y0 || z1 < z0) {
			 			Cube()
			 		} else {
			 			Cube(x0, x1, y0, y1, z0, z1)
			 		}
			 	}
			 })

solve <- function(data = puzzle_data, filter_inner = TRUE, verbose = FALSE) {
	cutter <- Cube(-50, 50, -50, 50, -50, 50)
	pos <- neg <- vector("list", 32000)
	pos_i <- neg_i <- 1
	for (i in 1L:nrow(data)) {
		row <- data[i, ]
		new_cube <- do.call(Cube, as.list(row[-1L]))
		if (verbose) {
			cat("[", str_pad(as.character(i), 3), "/", nrow(data), "]: ", sep = "")
			cat(if (row[1L] == "on") "Adding" else "Removing")
			cat(" <")
			show(new_cube)
			cat(">...\n")
		}
		if (filter_inner) {
			new_cube <- cutter - new_cube
		}
		if (!is_empty(new_cube)) {
			subst <- map(pos[seq(1, length.out = pos_i - 1)],
							 ~ .x - new_cube) %>% 
				Filter(Negate(is_empty), .)
			add <- map(neg[seq(1, length.out = neg_i - 1)], 
						  ~ .x - new_cube) %>% 
				Filter(Negate(is_empty), .)
			if (row[1L] == "on") {
				add <- c(add, new_cube)
			}
			if (pos_i + length(add) > length(pos)) {
				pos <- c(pos, vector("list", length(pos)))
			}
			if (neg_i + length(subst) > length(neg)) {
				neg <- c(neg, vector("list", length(neg)))
			}
			pos[seq(pos_i, length.out = length(add))] <- add
			neg[seq(neg_i, length.out = length(subst))] <- subst
			pos_i <- pos_i + length(add)
			neg_i <- neg_i + length(subst)
		}
	}
	pos <- Filter(Negate(is.null), pos)
	neg <- Filter(Negate(is.null), neg)
	sum(map_int(pos, volume)) -
		sum(map_int(neg, volume))
}
```

