---
params:
   task_nr: 25
   year: 2015
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2015"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(tidyr)
library(bit64)
library(Rcpp)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  list(start_value = 20151125L,
       target = 
         text_block %>% 
         str_extract_all("\\d+") %>% 
         unlist() %>% 
         as.integer() %>% 
         set_names(c("row", "col")))
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

The algorithm is rather straight forward:

1. Find out the index `n` of the element.
1. Iterate `n - 1` times and apply the algorithm.

However, as loops are rather slow in `R` (as usual), we use a C++ function (the R function
remains for reference).

```{r get-solution-1}
get_index <- function(row, col) {
  diagonal <- row + col - 1L
  (diagonal * (diagonal - 1L)) / 2 + col
}

get_code <- function(row, col, start) {
  i <- 1
  n <- get_index(row, col)
  x <- as.integer64(start)
  while(i < n) {
    x <- (x * 252533L) %% 33554393L 
    i <- i + 1L
    if (i %% 100000L == 0) {
      cat("Iteration:", i, "\n")
    }
  }
  x
}

cppFunction(
  '
  long long get_code_cpp(const int n, const int start) {
    long long x = start;
    for (int j = 1; j < n; ++j) {
       x = (x * 252533) % 33554393;
    }
    return x;
  } 
  '
)

get_code_cpp(get_index(puzzle_data$target["row"], puzzle_data$target["col"]), 
             puzzle_data$start_value)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We succesfully solved all puzzles in this year. CU next year! Ho-Ho-Ho!
```{r get-solution-2}
```
