---
params:
   task_nr: 18
   year: 2015
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2015"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```

```{js cpp-highlighter, file = "Prism_CPP.js", echo = FALSE}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    str_split("") %>% 
    do.call(rbind, .) %>% 
    set_class("light_array")
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

The solution is rather straight forward. However, since R is rather slow (while featuring 
very nice matrix indexing options), we re-implemented the solution in C++, which gave us a 
huge speed gain. The original code can be found in the appendix.

```{Rcpp cpp-algorithm, file = "light_array.cpp", cache = TRUE}
```


```{r get-solution-1}
count_lights(puzzle_data, 100L)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We just need to leave out the corners in our calculation and off we go.

```{r get-solution-2}
count_lights(puzzle_data, 100L, FALSE)
```

# R Legacy Solution

```{r legacy-code, eval = FALSE}
print.light_array <- function(x, ..., raw = FALSE) {
  if (raw) {
    NextMethod(x)
  } else {
    apply(x, 1, paste, collapse = "") %>% 
      paste(collapse = "\n") %>% 
      cat()
  }
}

get_neighbors <- function(dims) {
  rows <- 1:dims[1L]
  cols <- 1:dims[2L]
  idx <- expand.grid(row = rows, col = cols) %>% 
    as.matrix()
  nbs <- lapply(seq_len(nrow(idx)), function(i) {
    node <- c(idx[i, ])
    dirs <- -1:1
    nbs <- expand.grid(row = dirs, col = dirs)
    nbs <- nbs[!(nbs[, "row"] == 0L & nbs[, "col"] == 0L), ]
    nbs <- t(t(nbs) + node)
    cbind(row_orig = node[1L],
          col_orig = node[2L],
          nbs[between(nbs[, "row"], 1L, dims[1L]) &
                between(nbs[, "col"], 1L, dims[2L]), ]
    )
  })
  do.call(rbind, nbs) %>% 
    as_tibble() %>% 
    group_by(row_orig, col_orig) %>% 
    summarize(nbs = list(cbind(row, col)),
              .groups = "rowwise") %>% 
    rename(row = row_orig, col = col_orig) %>% 
    mutate(is_corner = (row == 1L & col == 1L) |
                 (row == 1L & col == dims[2L]) |
                 (row == dims[1L] & col == 1L) |
                 (row == dims[1L] & col == dims[2L]))
}

get_new_status <- function(row, col, nbs, light_array) {
  nb_status <- sum(light_array[nbs] == "#")
  if (light_array[row, col] == "#") {
    new_status <- if_else(between(nb_status, 2L, 3L), "#", ".")
  } else {
    new_status <- if_else(nb_status == 3L, "#", ".")
  }
  new_status
}

switch_lights <- function(light_array, iterations, consider_corners = TRUE) {
  dd <- dim(light_array)
  nbs <- get_neighbors(dd)
  for (i in 1:iterations) {
    cat("Iteration ", i, "\n")
    new_lights <- nbs %>% 
      mutate(status = get_new_status(row, col, nbs, light_array),
             status = case_when(
               consider_corners | !is_corner ~ status,
               TRUE ~ "#"
             )
      ) %>% 
      pull(status)
    light_array[] <- new_lights
  }
  light_array
}

count_lights <- function(light_array, iterations, consider_corners = TRUE) {
  final_state <- switch_lights(light_array, iterations, consider_corners)
  sum(final_state == "#")
}

count_lights(puzzle_data, 100L)
count_lights(puzzle_data, 100L, FALSE)
```
