---
params:
   task_nr: 9
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
	width: 100%;
}

.back {
	font-size: 2rem;
}

.day-desc {
	border: 2px dotted;
	border-radius: 6px;
	border-color: var(--blockquote-border-color);
	padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
	if (knitr::is_latex_output()) {
		sprintf("\\texttt{%s}", as.character(x))
	} else if (knitr::is_html_output()) {
		sprintf("<tt>%s</tt>", as.character(x))
	} else {
		x
	}
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(tibble)
library(dplyr)
library(purrr)
library(tidyr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2021/day/", params$task_nr)
puzzle <- GET(base_url,
				  session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	xml_find_all("///article") %>% 
	lapply(as.character)

## use curly braces in the last pipe to avoid that . is additionally added
## as the first argument
puzzle_data <- GET(paste0(base_url, "/input"),
						 session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	str_split("\n") %>% 
	`[[`(1) %>% 
	str_split("") %>% 
	do.call(rbind, .) %>% 
	{matrix(as.numeric(.), nrow(.), ncol(.))}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

To solve this puzzle, we simply create the nighbors by dropping the first / last 
row / column and pad with an `Inf` row / column and compare the original matrix to 
those neighbors.

```{r solution-a}
nc <- ncol(puzzle_data)
nr <- nrow(puzzle_data)
left_neighbors  <- cbind(Inf, puzzle_data[, -nc])
right_neighbors <- cbind(puzzle_data[, -1], Inf)
upper_neighbors <- rbind(Inf, puzzle_data[-nr, ])
lower_neighbors <- rbind(puzzle_data[-1, ], Inf)
low_points <- puzzle_data < left_neighbors &
	puzzle_data < right_neighbors &
	puzzle_data < upper_neighbors &
	puzzle_data < lower_neighbors
sum(puzzle_data[low_points] + 1)
```



## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For this puzzle we use a recursive function, where we start in a basin (as determined in
part 1) and walk up, down, left, and right counting all cells until we either

1. Reach an already visited field.
1. Reach the edge

*N.B.* The algo could be improved, if we moved the logic for checking whether we are on
an invalid field *before* recalling the function. This approach, however, makes it easier, 
as we do not have to take about bad indices 4 times.

```{r solution-b, eval = length(puzzle) > 1}
walk_basin <- function(row, col,
							  map = puzzle_data) {
	visited <- matrix(FALSE, nrow(map), ncol(map))
	do_walk <- function(row, col) {
		nr <- nrow(map)
		nc <- ncol(map)
		if (!between(row, 1, nr) || !between(col, 1, nc)) {
			0L
		} else if (map[row, col] == 9 || visited[row, col]) {
			visited[row, col] <<- TRUE
			0L
		} else {
			visited[row, col] <<- TRUE
			1L + Recall(row - 1, col) + 
				Recall(row + 1, col) +
				Recall(row, col - 1) + 
				Recall(row, col + 1)
		}
	}
	map2_int(row, col, do_walk)
}

which(low_points, arr.ind = TRUE) %>% 
	as_tibble() %>% 
	mutate(basin_size = walk_basin(row, col, puzzle_data)) %>% 
	arrange(desc(basin_size)) %>% 
	slice(1:3) %>% 
	pull(basin_size) %>% 
	prod

```


