---
params:
   task_nr: 23
   year: 2021
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2021"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{js sql-highlighter, file = "Prism_CPP.js", echo = FALSE}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(tibble)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(stringi)
library(knitr)
library(cli)
library(bit64)
library(igraph)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    str_subset("[ABCD]") %>% 
    str_extract_all("[ABCD]") %>% 
    list_transpose()
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

We use a brute-force appraoch, where we simply try out all potential moevments. We could
improve the code by carefully reflecting on smart moving strategies to avoid exploring
the while space. However, the problem seems to be small enough that such improvements
seem unnecessary.

However, we had to fall back to `C++` code for efficiency reasons, as in `R` we have
copy on modifcation semantics, which is, in eahc recursion we are copying lists, which
significantly slows down the code.

The idea of the `C++` is as follows:

1. We define a  `State` class, which maintains the bots on the `hallway` and the bots
in the `rooms`. The former is implemented as an `std::array<char, 11>` and the latter as 
a `std::array<std::stack<char>, 4>`.
1. The class defines functions for moving bots in and out, checking if a path is free and
returning all movable bots in both the rooms and the hallway.
1. It provides a `get_map_id()` function, which gives a condensed textual representation 
of the current state, which later will serve as a key for the has map, where we maintain 
the costs of any given mpa already calculated (to avoid re-calculating known costs).
1. The main function `solve` is a recursive function, which
  1. Checks if we have a final state, and returns cost zero (no more movements needed)
  1. Checks if a certain satet was already calculated in the past and returns the cached 
  value in this case.
  1. Move all bots from the hallway home and recalls itself wiht this new state and sums
  the costs.
  1. Tries each possible space for a bot currently sitting in its false room and keeps
  the cheapest solution in each recursion.

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

```{Rcpp cpp-algorithm, file = "bots.cpp", cache = TRUE}
```

```{r get-solution-1}
bring_bots_home(puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We simply call `bring_bots_home` with the extended rooms.

```{r get-solution-2}
new_rooms<- list(
  c("A", "D", "D", "B"), 
  c("D", "C", "B", "C"), 
  c("B", "B", "A", "D"), 
  c("C", "A", "C", "A")
  ) 
bring_bots_home(new_rooms)
```

# R Legacy Solution

For the sake of completeness, here's the previous `R` codes, which is, however, due to the
copy on change semantic not really performing well:

```{r r-legacy-solution, eval = FALSE}
encode_map <- function(data) {
  hall <- data[2L, seq(2L, ncol(data) - 1L)]
  rooms <- map(4L + 2L * 0:3, ~ data[seq(3L, nrow(data) - 1L), .x]) %>% 
    set_names(LETTERS[1:4])
  list(hall = hall, rooms = rooms)
}

get_map_id <- function(state) {
  unlist(state) %>% 
    paste(collapse = "")
}

is_done <- function(state) {
  rooms <- state$rooms
  reduce(names(rooms), function(res, room_id) {
    res && all(rooms[[room_id]] == room_id)
  }, .init = TRUE)
}

is_path_blocked <- function(bot, hall_pos, room_pos, hall) {
  !all(hall[seq(hall_pos, room_pos)][-1L] %in% c(".", "x"))
}

get_cost <- function(bot, in_room_pos, room_pos, hall_pos) {
  room_idx <- match(bot, LETTERS[1:4]) - 1L
  vertical <- in_room_pos
  horizontal <- abs(hall_pos - room_pos)
  cost <- 10 ^ room_idx
  (vertical + horizontal) * cost
}

get_new_position <- function(hall_pos, state, room_id = NA, dir = c("home", "out")) {
  dir <- match.arg(dir)
  hall <- state$hall
  if (dir == "home") {
    bot <- hall[hall_pos]
    room_id <- bot
    room <- state$rooms[[room_id]]
    room_pos <- (match(bot, LETTERS) - 1L) * 2L + 3L
  } else if (dir == "out") {
    ## invariant: we do not call this function in the out direction with an empty room
    stopifnot(!is.na(room_id))
    room <- state$rooms[[room_id]]
    bot <- room[which(room != ".")[1L]]
    room_pos <- (match(room_id, LETTERS) - 1L) * 2L + 3L
  }
  room_size <- length(room)
  if (is_path_blocked(bot, hall_pos, room_pos, hall)) {
    in_room_pos <- NA
  } else if (dir == "home") {
    ## bot wants to move from hall to room
    new_roomie <- bot
    new_idler <- "."
    cand <- which(room == ".")
    if (length(cand) == 0L) {
      ## no empty spaces
      in_room_pos <- NA
    } else if (length(cand) == room_size) {
      ## all empty spaces
      in_room_pos <- room_size
    } else {
      ## room has already some inhabitants
      neighbors <- tail(room, room_size - max(cand))
      if (all(neighbors == bot)) {
        ## all inhabitants are of same type
        in_room_pos <- max(cand)
      } else {
        in_room_pos <- NA
      }
    }
  } else if (dir == "out") {
    ## bot wants to move from room to hall
    new_roomie <- "."
    new_idler <- bot
    if (hall[hall_pos] != ".") {
      in_room_pos <- NA
    } else {
      in_room_pos <- which(room == bot)[1L]
    }
  }
  if (!is.na(in_room_pos)) {
    room[in_room_pos] <- new_roomie
    hall[hall_pos] <- new_idler
    state$rooms[[room_id]] <- room
    state$hall <- hall
    res <- list(
      is_free = TRUE,
      cost = get_cost(bot, in_room_pos, room_pos, hall_pos),
      state = state
    )
  } else {
    res <- list(
      is_free = FALSE,
      cost = NA_integer_,
      state = list()
    )
  }
  res
}

get_top_bot <- function(room) {
  room_id <- names(room)
  room <- room[[1L]]
  if (all(room %in% c(".", room_id))) {
    res <- NA_integer_
  } else {
    ## there is at least one wrong bot in here
    res <- which(room != ".")[1L]
  }
  res
}

bring_bots_home <- function(initial_state) {
  hash_list <- list()
  i <- 1
  move_bots <- function(state) {
    if (i %% 1000 == 0) print(i)
    i <<- i + 1
    id <- get_map_id(state)
    if (is_done(state)) {
      return(0L)
    }
    if (!is.null(hash_list[[id]])) {
      return(hash_list[[id]])
    }
    ## first move all hall bots home
    hall <- state$hall
    rooms <- state$rooms
    for (hall_pos in seq_along(hall)) {
      bot <- hall[hall_pos]
      if (!bot %in% c(".", "x")) {
        move_attempt <- get_new_position(hall_pos, state, dir = "home")
        if (move_attempt$is_free) {
          return(move_attempt$cost + Recall(move_attempt$state))
        }
      }
    }
    min_costs <- Inf
    for (room_id in names(rooms)) {
      top_bot_idx <- get_top_bot(rooms[room_id])
      if (!is.na(top_bot_idx)) {
        ## we can move a bot from this room
         for (hall_pos in seq_along(hall)) {
           move_attempt <- get_new_position(hall_pos, state, room_id, "out")
           if (move_attempt$is_free) {
             min_costs <- min(min_costs, move_attempt$cost + Recall(move_attempt$state))
           }
         }
      }
    }
    hash_list[[id]] <<- min_costs
    min_costs
  }
  
  move_bots(initial_state)
}

initial_state <- encode_map(puzzle_data)
bring_bots_home(initial_state)

initial_state2 <- initial_state
initial_state2$rooms <- list(A = c("A", "D", "D", "B"), 
                             B = c("D", "C", "B", "C"),
                             C = c("B", "B", "A", "D"), 
                             D = c("C", "A", "C", "A"))
bring_bots_home(initial_state2)
```

