---
params:
   task_nr: 25
   year: 2021
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2021"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(tibble)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(stringi)
library(knitr)
library(cli)
library(rlang)
library(DT)
library(bit64)
library(rlang)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    str_split("") %>% 
    do.call(rbind, .)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1


### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

The solution is rather straight forward thanks to R's nice matrix indexing possibilities:
1. Get all positions of of `EAST` movers in matrix index form (i.e. column one give the 
row index, column 2 the column index).
1. Add 1 to eahc row position and reset to 1 in case we are out of bounds.
1. Check if the space is free, if so move all of the cucumbers.
1. Do the same for the `SOUTH` movers.
1. Count how many cucumbers we moved. Repeat this until nobody can move.


```{r get-solution-1, cache = TRUE}
maze_to_string <- function(maze) {
  apply(maze, 1, paste, collapse = "") %>% 
    paste(collapse = "\n")
}

move_cucumbers <- function(maze) {
  new_maze <- maze
  dd <- dim(maze)
  old_east <- new_east <- which(maze == ">", arr.ind = TRUE)
  new_east[, 2L] <- (new_east[, 2L] + 1L) 
  new_east[, 2L] <- if_else(new_east[, 2L] > dd[2L], 1L, new_east[, 2L])
  ok <- new_maze[new_east] == "."
  nr_movers <- sum(ok)
  new_maze[old_east[ok, , drop = FALSE]] <- "."
  new_maze[new_east[ok, , drop = FALSE]] <- ">"
  old_south <- new_south <- which(new_maze == "v", arr.ind = TRUE)
  new_south[, 1L] <- new_south[, 1L] + 1L
  new_south[, 1L] <- if_else(new_south[, 1L] > dd[1L], 1L, new_south[, 1L])
  ok <- new_maze[new_south] == "."
  nr_movers <- nr_movers + sum(ok)
  new_maze[old_south[ok, , drop = FALSE]] <- "."
  new_maze[new_south[ok, , drop = FALSE]] <- "v"
  list(new_maze = new_maze,
       could_move = nr_movers > 0L)
}

count_moves <- function(maze) {
  could_move <- TRUE
  count <- 0
  while (could_move) {
    count <- count + 1
    res <- move_cucumbers(maze)
    maze <- res$new_maze
    could_move <- res$could_move
  }
  count
}
count_moves(puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution


```{r get-solution-2}
```
