---
params:
   task_nr: 16
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(knitr)
library(cli)
library(bit64)
library(igraph)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      unlist() %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      head(-1L)
  }
  str_split(text_block,
            "") %>% 
    do.call(rbind, .)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r get_example_data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

The idea of the algorithm is as follows:

1. First, find node which has currently the cheapest path to it **and** has still some
neighbors which have no costs associated yet.
1. If we can walk in the current direction from this node, do so, otherwise make a turn
and move then. Keep track of the costs for this operation. If we changed direction, store
the new direction for this node.
1. Continue until we reach the final node.

As we are greedily moving along the cheapest path, we will eventually get to the end point
with lowest costs.

*N.B.* Updating the number of free neighbors of all nodes in each iteration is a bit of
an overkill and could be optimised to only update hte free neighbors in the local 
neighborhood.

```{r get-solution-1}
get_shortest_path <- function(reindeer_map) {
  dirs <- rbind(
    cbind(-1L, 0L),
    cbind(0L, 1L),
    cbind(1L, 0L),
    cbind(0L, -1L)
  ) %>% 
    set_rownames(c("^", ">", "v", "<"))
  dd <- dim(reindeer_map)
  costs <- matrix(Inf, dd[1L], dd[2L])
  direction <- matrix(NA_character_, dd[1L], dd[2L])
  
  turn <- function(current_dir, clockwise) {
    if (clockwise) {
      switch(current_dir,
             "^" = ">",
             ">" = "v",
             "v" = "<",
             "<" = "^"
      )
    } else {
      switch(current_dir,
             "^" = "<",
             "<" = "v",
             "v" = ">",
             ">" = "^"
      )
    }
  }
  move <- function(current_pos, current_dir) {
    current_pos + dirs[current_dir, , drop = FALSE]
  }
  
  get_free_nbs <- function() {
    is_walkable <- reindeer_map == "." | reindeer_map == "E" | reindeer_map == "S"
    is_unassigned <- is.infinite(costs)
    free <- is_walkable & is_unassigned
    rbind(free[-1L, ], FALSE) +
      rbind(FALSE, free[-dd[1L], ]) +
      cbind(free[, -1L], FALSE) +
      cbind(FALSE, free[, -dd[2L]])
  }
  n_nbs <- get_free_nbs()
  start_pos <- which(reindeer_map == "S", arr.ind = TRUE)
  end_pos <- which(reindeer_map == "E", arr.ind = TRUE)
  costs[start_pos] <- 0L
  direction[start_pos] <- ">"
  while(is.infinite(costs[end_pos])) {
    nodes_with_neighbors <- which(n_nbs > 0, arr.ind = TRUE)
    filtered_costs <- costs[nodes_with_neighbors]
    cheapest_node <- nodes_with_neighbors[filtered_costs == min(filtered_costs), ,                                          drop = FALSE][1, , drop = FALSE]
    current_dir <- direction[cheapest_node] 
    stopifnot(!is.na(current_dir))
    all_dirs <- c(current_dir, turn(current_dir, TRUE), turn(current_dir, FALSE))
    candidates <- rbind(
      move(cheapest_node, current_dir),
      move(cheapest_node, turn(current_dir, TRUE)),
      move(cheapest_node, turn(current_dir, FALSE))
    ) %>% 
      set_rownames(all_dirs)
    for (nb_dir in all_dirs) {
      candidate <- candidates[nb_dir, , drop = FALSE]
      if (reindeer_map[candidate] %in% c("E", ".") && is.infinite(costs[candidate])) {
        ## consider only neighbors whihc have unassigned costs and are walk-able
        new_costs <-  costs[cheapest_node] + 1L + 
          if_else(nb_dir == current_dir, 0L, 1000L)
        ## update costs
        costs[candidate] <- new_costs
        ## store direction
        direction[candidate] <- nb_dir
        ## update free neighbor count (could be optimized)
        n_nbs <- get_free_nbs()
        break
      }
    }
  }
  costs[end_pos]
}
get_shortest_path(puzzle_data)

```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

```{r get-solution-2}
construct_graph <- function(reindeer_map) {
  dd <- dim(reindeer_map)
  dirs <- rbind(
      cbind(-1L, 0L),
      cbind(0L, 1L),
      cbind(1L, 0L),
      cbind(0L, -1L)
    ) %>% 
      set_rownames(c("^", ">", "v", "<"))
  turn <- function(dir, clockwise) {
    if (clockwise) {
      switch(dir,
             "^" = ">",
             ">" = "v",
             "v" = "<",
             "<" = "^")
    } else {
      switch(dir,
             "^" = "<",
             ">" = "^",
             "v" = ">",
             "<" = "v")
    }
  }
  
  make_vertex_id <- function(current_pos, current_dir) {
    paste(current_pos, collapse = "/") %>% 
      paste(current_dir, sep = "/")
  }
  
  create_vertex <- function(current_pos, current_dir) {
    id <- make_vertex_id(current_pos, current_dir)
    if (!id %in% V(G)$name) {
      G <<- G %>% 
        add_vertices(1L, name = id)
    }
    id
  }
  
  is_valid_pos <- function(new_pos) {
    all(new_pos >= c(1L, 1L) & new_pos <= dd) &&
      reindeer_map[matrix(new_pos, nrow = 1)] != "#" 
  }
  
  get_neighbor_nodes <- function(current_pos) {
    nbs <- t(t(dirs) + c(current_pos))
    nbs[apply(nbs, 1, is_valid_pos), , drop = FALSE]
  }
  
  get_edges <- function() {
    edges <- tibble(from = character(0L), 
                    to = character(0L),
                    weight = integer(0L))
    no_walls <- which(reindeer_map != "#", arr.ind = TRUE)
    for (row in seq_len(nrow(no_walls))) {
      current_node <- no_walls[row, , drop = FALSE]
      nbs <- get_neighbor_nodes(current_node)
      nb_dirs <- rownames(nbs)
      for (dir in nb_dirs) {
        nb <- nbs[dir, , drop = FALSE]
        from <- make_vertex_id(current_node, dir)
        to <- make_vertex_id(nb, dir)
        edges <- edges %>% 
          bind_rows(
            tibble(from = from, to = to, weight = 1L)
          )
        if (turn(dir, TRUE) %in% nb_dirs) {
          edges <- edges %>% 
            bind_rows(
              tibble(from = from, to = make_vertex_id(current_node, turn(dir, TRUE)), 
                     weight = 1000L)
            )
        } else if (turn(dir, FALSE) %in% nb_dirs) {
          edges <- edges %>% 
            bind_rows(
              tibble(from = from, to = make_vertex_id(current_node, turn(dir, FALSE)), 
                     weight = 1000L)
            )
        }
      }
    }
  edges
  }
  edges <- get_edges()
  list(G = graph_from_data_frame(edges, directed = TRUE), edges = edges)
}
vn <- NULL
res <- construct_graph(example_data)

goal <- "8/12/v"
distances(res$G, "14/2/>", goal)
shortest_paths(res$G, "14/2/>", goal)$vpath[[1L]]


```
