---
params:
   task_nr: 16
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2024"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(knitr)
library(cli)
library(bit64)
library(igraph)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      unlist() %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      head(-1L)
  }
  str_split(text_block,
            "") %>% 
    do.call(rbind, .)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r get_example_data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

The idea of the algorithm is as follows:

1. First, find node which has currently the cheapest path to it **and** has still some
neighbors which have no costs associated yet.
1. If we can walk in the current direction from this node, do so, otherwise make a turn
and move then. Keep track of the costs for this operation. If we changed direction, store
the new direction for this node.
1. Continue until we reach the final node.

As we are greedily moving along the cheapest path, we will eventually get to the end point
with lowest costs.

*N.B.* Updating the number of free neighbors of all nodes in each iteration is a bit of
an overkill and could be optimised to only update hte free neighbors in the local 
neighborhood.

```{r get-solution-1}
get_shortest_path <- function(reindeer_map) {
  dirs <- rbind(
    cbind(-1L, 0L),
    cbind(0L, 1L),
    cbind(1L, 0L),
    cbind(0L, -1L)
  ) %>% 
    set_rownames(c("^", ">", "v", "<"))
  dd <- dim(reindeer_map)
  costs <- matrix(Inf, dd[1L], dd[2L])
  direction <- matrix(NA_character_, dd[1L], dd[2L])
  
  turn <- function(current_dir, clockwise) {
    if (clockwise) {
      switch(current_dir,
             "^" = ">",
             ">" = "v",
             "v" = "<",
             "<" = "^"
      )
    } else {
      switch(current_dir,
             "^" = "<",
             "<" = "v",
             "v" = ">",
             ">" = "^"
      )
    }
  }
  move <- function(current_pos, current_dir) {
    current_pos + dirs[current_dir, , drop = FALSE]
  }
  
  get_free_nbs <- function() {
    is_walkable <- reindeer_map == "." | reindeer_map == "E" | reindeer_map == "S"
    is_unassigned <- is.infinite(costs)
    free <- is_walkable & is_unassigned
    rbind(free[-1L, ], FALSE) +
      rbind(FALSE, free[-dd[1L], ]) +
      cbind(free[, -1L], FALSE) +
      cbind(FALSE, free[, -dd[2L]])
  }
  n_nbs <- get_free_nbs()
  start_pos <- which(reindeer_map == "S", arr.ind = TRUE)
  end_pos <- which(reindeer_map == "E", arr.ind = TRUE)
  costs[start_pos] <- 0L
  direction[start_pos] <- ">"
  while(is.infinite(costs[end_pos])) {
    nodes_with_neighbors <- which(n_nbs > 0, arr.ind = TRUE)
    filtered_costs <- costs[nodes_with_neighbors]
    cheapest_node <- nodes_with_neighbors[filtered_costs == min(filtered_costs), ,                                          drop = FALSE][1, , drop = FALSE]
    current_dir <- direction[cheapest_node] 
    stopifnot(!is.na(current_dir))
    all_dirs <- c(current_dir, turn(current_dir, TRUE), turn(current_dir, FALSE))
    candidates <- rbind(
      move(cheapest_node, current_dir),
      move(cheapest_node, turn(current_dir, TRUE)),
      move(cheapest_node, turn(current_dir, FALSE))
    ) %>% 
      set_rownames(all_dirs)
    for (nb_dir in all_dirs) {
      candidate <- candidates[nb_dir, , drop = FALSE]
      if (reindeer_map[candidate] %in% c("E", ".") && is.infinite(costs[candidate])) {
        ## consider only neighbors whihc have unassigned costs and are walk-able
        new_costs <-  costs[cheapest_node] + 1L + 
          if_else(nb_dir == current_dir, 0L, 1000L)
        ## update costs
        costs[candidate] <- new_costs
        ## store direction
        direction[candidate] <- nb_dir
        ## update free neighbor count (could be optimized)
        n_nbs <- get_free_nbs()
        break
      }
    }
  }
  costs[end_pos]
}
get_shortest_path(puzzle_data)

```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

To solve this puzzle, we fall back to *graph theory*. We construct the graph as follows:

1. For each field in the map we create **4** nodes representing the 4 directions in which 
we could look. For instance the field `14/2` will be represented by the 4 nodes `14/2/^`,
`14/2/>`, `14/2/v` and `14/2/<`.
1. We connect neighboring fields along their same direction and assign a weight of 1 to
this edge. This represents moving form one field to another in a straight line.
For instance, the connected fields `14/2` and `13/2` would yield an edge 
`14/2/^ --> 13/2/^`.
1. Next, we connect nodes at the same position but different directions if they can be 
reached via a turn and assign weight 1000. For instance we connect `14/2/>` with `14/2/^`
and `14/2/v` which represents a turn.
1. For convenience we define 2 artificial nodes `S` (start) and `E` (end). We connect `S` 
with weight 0 to the starting position with direction `>` (which is our starting 
direction) and `E` with all possible direction of the end point (we do not care how we
reach the end point, that is facing south or facing north would both yield a valid end).
1. Eventually, to get the solution, we simply construct all shortest paths from `S` to 
`E`, reduce the node list to fields (that is stripping the direction part of the 
identifier) and count unique fields.

The algorithm does not create all directions for the fields, but only those which are
relevant for the solution. For instance if a field has walls to its East and West 
(allowing moves only to the South and North) we do not create nodes with directions `>` or
`<`, as it does not make sense to turn into these directions.

*N.B.*  This graph construction allows a swell to solve part one of the puzzle simply by 
getting the weight of the shortest path.

```{r get-solution-2, cache = TRUE}
construct_graph <- function(reindeer_map) {
  dd <- dim(reindeer_map)
  dirs <- rbind(
      cbind(-1L, 0L),
      cbind(0L, 1L),
      cbind(1L, 0L),
      cbind(0L, -1L)
    ) %>% 
      set_rownames(c("^", ">", "v", "<"))
  turn <- function(dir, n) {
    if (n == 1L) {
      switch(dir,
             "^" = ">",
             ">" = "v",
             "v" = "<",
             "<" = "^")
    } else if (n == -1L) {
      switch(dir,
             "^" = "<",
             ">" = "^",
             "v" = ">",
             "<" = "v")
    } else if (n == 2L) {
      switch(dir,
             "^" = "v",
             ">" = "<",
             "v" = "^",
             "<" = ">")
    }
  }
  
  make_vertex_id <- function(current_pos, current_dir) {
    paste(current_pos, collapse = "/") %>% 
      paste(current_dir, sep = "/")
  }
  
  is_valid_pos <- function(new_pos) {
    all(new_pos >= c(1L, 1L) & new_pos <= dd) &&
      reindeer_map[matrix(new_pos, nrow = 1)] != "#" 
  }
  
  get_neighbor_nodes <- function(current_pos) {
    nbs <- t(t(dirs) + c(current_pos))
    nbs[apply(nbs, 1, is_valid_pos), , drop = FALSE]
  }
  
  get_edges <- function() {
    edges <- tibble(from = character(0L), 
                    to = character(0L),
                    weight = integer(0L))
    no_walls <- which(reindeer_map != "#", arr.ind = TRUE)
    for (row in seq_len(nrow(no_walls))) {
      current_node <- no_walls[row, , drop = FALSE]
      nbs <- get_neighbor_nodes(current_node)
      nb_dirs <- rownames(nbs)
      for (dir in nb_dirs) {
        nb <- nbs[dir, , drop = FALSE]
        from <- make_vertex_id(current_node, dir)
        to <- make_vertex_id(nb, dir)
        edges <- edges %>% 
          bind_rows(
            tibble(from = from, to = to, weight = 1L)
          )
        par_dir <- turn(dir, 2L)
        turn_dirs <- setdiff(nb_dirs, dir)
        for (turn_dir in turn_dirs) {
          if (turn_dir != par_dir) {
            edges <- edges %>% 
              bind_rows(
                tibble(from = make_vertex_id(current_node, par_dir), 
                       to = make_vertex_id(current_node, turn_dir), 
                       weight = 1000L)
              )
          }
        }
      }
    }
  edges
  }
  
  edges <- get_edges()
  start_pos <- which(reindeer_map == "S", arr.ind = TRUE)
  end_pos <- which(reindeer_map == "E", arr.ind = TRUE)
  
  edges <- edges %>% 
    bind_rows(
      tibble(from = rep(make_vertex_id(start_pos, ">"), 2L),
             to = c(make_vertex_id(start_pos, "^"),
                    make_vertex_id(start_pos, "v")),
             weight = rep(1000L, 2L)),
      tibble(from = map_chr(rownames(dirs), 
                            ~ make_vertex_id(end_pos, .x)),
             to = rep("E", 4L),
             weight = rep(0L, 4L)),
      tibble(from = "S",
             to = make_vertex_id(start_pos, ">"),
             weight = 0L),
    )
  
  graph_from_data_frame(edges, directed = TRUE)
}

get_all_spots <- function(G) {
  paths <- all_shortest_paths(G, "S", "E")$vpath
  paths %>% 
    do.call(c, .) %>% 
    unique() %>% 
    difference(V(G)[name %in% c("S", "E")]) %>% 
    names() %>% 
    str_remove("/.$") %>% 
    unique()
}


G <- construct_graph(puzzle_data)
get_all_spots(G) %>% 
  length()
```

*N.B.* This graph construction allows for solving part 1 also quite easily:

```{r alternative-solution-1}
distances(G, "S", "E")
```