---
params:
   task_nr: 12
   year: 2022
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  res <- text_block %>% 
    str_split("") %>% 
    do.call(rbind, .)
  start <- which(res == "S", arr.ind = TRUE)
  end <- which(res == "E", arr.ind = TRUE)
  res[start] <- "a"
  res[end] <- "z"
  map <- match(res, letters)
  dim(map) <- dim(res)
  attr(map, "start") <- start
  attr(map, "dest") <- end
  map
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

This is a simple application for the Breadth-First-Search algorithm.


```{r get-solution-1}
find_shortest_path <- function(map, start = attr(map, "start"), prune = Inf) {
  dest <- attr(map, "dest")
  visited <- array(FALSE, dim(map))
  queue <- matrix(NA_integer_, length(map), 3L)
  queue_ptr <- queue_len <- 1L
  queue[queue_ptr, ] <- c(start, 0L)
  visited[start] <- TRUE
  dirs <- rbind(
    "^" = c(-1L, 0L),
    ">" = c(0L, 1L),
    "v" = c(1L, 0L),
    "<" = c(0L, -1L)
  )
  nr <- nrow(map)
  nc <- ncol(map)
  while (queue_ptr <= queue_len) {
    node <- queue[queue_ptr, , drop = FALSE]
    queue_ptr <- queue_ptr + 1L
    pos <- node[, 1:2, drop = FALSE]
    cur_lev <- map[pos]
    if (all(pos == dest)) {
      return(node[, 3L])  
    }
    if (node[, 3L] >= prune) {
      return(Inf)
    }
    nbs <- t(t(dirs) + c(pos))
    nbs <- nbs[nbs[, 1L] >= 1L & nbs[, 1L] <= nr &
                 nbs[, 2L] >= 1L & nbs[, 2L] <= nc, , drop = FALSE]
    valid_nbs <- nbs[map[nbs] <= cur_lev + 1L & !visited[nbs], , drop = FALSE]
    n_nbs <- nrow(valid_nbs)
    if (n_nbs > 0L) {
      visited[valid_nbs] <- TRUE
      queue[(queue_len + 1L):(queue_len + n_nbs), ] <- cbind(valid_nbs, node[3L] + 1L)
      queue_len <- queue_len + n_nbs
    }
  }
  Inf
}

find_shortest_path(puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We apply the BFS to all possible starting points and find the shortest. To save some time
we use a pruning factor to skip explorations which are longer than the currently best 
solution.

*N.B.* Yet a faster algorithm would be to walk backwards from the destination and 
continue walking until we reached all `a` positions.

```{r get-solution-2}
find_most_scenic_path <- function(map) {
  starts <- which(map == 1L, arr.ind = TRUE)
  min_dist <- 504L
  for (i in seq_len(nrow(starts))) {
    dist <- find_shortest_path(map, starts[i, , drop = FALSE], min_dist)
    min_dist <- min(dist, min_dist)
  }
  min_dist
}

find_most_scenic_path(puzzle_data)
```