---
params:
   task_nr: 9
   year: 2016
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2016"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(glue)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

We use the following algorithm:

1. Get everything up to (but not including) the first (from the left) `(nxm)` marker.
1. Add the length of this tring to the result.
1. Extract the `(nxm)` pattern and add `n * m` to the result (we repeat a substring of
length `n` `m` times.)
1. Remove the marker plus `n` characters from the string.
1. Repeat until the string is fully consumed.

### Solution

```{r get-solution-1}
get_decompressed_length <- function(string) {
   compression_pattern <- "\\((\\d+)x(\\d+)\\)"
   ## everything from start to the next pattern
   chunk_pattern <- glue("^.*?(?={compression_pattern})") 
   len <- 0

   while (str_length(string) > 0L) {
     left <- str_extract(string, chunk_pattern)
     rest <- str_remove(string, chunk_pattern)
     len <- len + str_length(left)
     compression <- str_match_all(rest, glue("^{compression_pattern}")) %>% 
       extract2(1L) %>% 
       c() %>% 
       extract(-1L) %>% 
       as.integer()
     len <- len + prod(compression)
     string <- str_remove(rest, glue("^{compression_pattern}.{{{compression[1L]}}}"))
   }
   len
}

get_decompressed_length(puzzle_data)


```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

This time we us a recursive version of the same idea. Instead of simply discarding the
next `n` characters, we recurse with this substring and multiply it by the outer `m`.

```{r get-solution-2}
get_decompressed_length_recursive <- function(string) {
  len <- 0
  compression_pattern <- "\\((\\d+)x(\\d+)\\)"
  chunk_pattern <- glue("^.*?(?={compression_pattern})")
  if (!str_detect(string, compression_pattern)) {
    ## no more compression markers
    len <- str_length(string)
  } else {
    ## everything from start to the next pattern
    left <- str_extract(string, chunk_pattern)
    rest <- str_remove(string, chunk_pattern)
    len <- len + str_length(left)
    compression <- str_match_all(rest, glue("^{compression_pattern}")) %>% 
      extract2(1L) %>% 
      c() %>% 
      extract(-1L) %>% 
      as.integer()
    next_chunk <- str_remove(rest, glue("^{compression_pattern}")) %>% 
      str_extract(glue(".{{{compression[1L]}}}"))
    rest <- str_remove(rest, glue("^{compression_pattern}.{{{compression[1L]}}}"))
    len <- len + compression[2L] * Recall(next_chunk) + Recall(rest)
  }
  len
}

get_decompressed_length_recursive(puzzle_data)
```
