---
params:
   task_nr: 11
   year: 2018
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
   as.integer()
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We solve this puzzle by explicitely generating the grid first making use of `R's` 
vectorization possibilities. Then, we caluclate the crll which yields the highest power 
level.

```{r get-solution-1}
get_power_level <- function(coords, serial_nr) {
  rack_id <- coords[, 1L] + 10L
  ((rack_id * coords[, 2L] + serial_nr) * rack_id) %/% 100L %% 10L - 5L 
}

make_grid <- function(serial_nr = puzzle_data, nr = 300L, nc = 300L) {
  grid <- matrix(NA_integer_, nr, nc)
  coords <- cbind(c(col(grid)), c(row(grid)))
  grid[coords[, 2:1]] <- get_power_level(coords, serial_nr)
  grid
}

find_cell <- function(grid, nr = 3L, nc = 3L) {
  grid <- t(grid)
  n <- nrow(grid)
  m <- ncol(grid)
  starts <- expand.grid(row = 1:(n - nr + 1L), 
                        col = 1:(m - nc + 1L))
  sums <- apply(starts, 1L, function(pos) {
    r <- pos[1]
    c <- pos[2]
    sum(grid[r:(r + nr - 1L), c:(c + nc - 1L)])
  })
  idx <- which.max(sums)
  cbind(starts[idx, ], power_level = max(sums))
}

grid <- make_grid(puzzle_data)
find_cell(grid)[, 1:2] %>% 
  paste(collapse = ",")

```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

Now we just have to iterate over all possible sub grid sizes and return the one with the 
largest power level. To do so efficiently we pre-calculate a matrix `summed` where an 
entry at index `i,j` represents the sum of the grid from `(1, 1) - (i, j)`. Them to find
the sum starting at any field, we get first the sum starting at `(1, 1)` till the right 
bottom corner of the field of interest, then subtract the rows above the field, then the
columns left of the field and eventually - as we were subtracting the area to the top left 
of the field twice - eventually add the field to the top left again.

```{r get-solution-2, cache = TRUE}
find_cell_any_size <- function(grid) {
  n <- nrow(grid)
  m <- ncol(grid)
  
  summed <- matrix(0L, n, m)
  summed[1L, 1L] <- grid[1L, 1L]
  
  for (i in 2:n) {
    summed[i, 1L] <- summed[i - 1L, 1L] + grid[i, 1L]
  }
  for (j in 2:m) {
    summed[1L, j] <- summed[1L, j - 1L] + grid[1L, j]
  }
  
  for (i in 2:n) {
    for (j in 2:m) {
      summed[i, j] <- grid[i, j] + summed[i - 1, j] + 
        summed[i, j - 1] - summed[i - 1, j - 1]
    }
  }

  get_subgrid_sum <- function(x, y, size) {
    x2 <- x + size - 1L
    y2 <- y + size - 1L
    total <- summed[y2, x2]
    if (x > 1L) {
      total <- total - summed[y2, x - 1L]
    }
    if (y > 1L) {
      total <- total - summed[y - 1, x2]
    }
    if (x > 1 && y > 1) {
      total <- total + summed[y - 1, x - 1]
    }
    total
  }

  best <- NULL
  max_power <- -Inf

  for (size in seq_len(n)) {
    for (y in 1:(n - size + 1L)) {
      for (x in 1:(m - size + 1L)) {
        power <- get_subgrid_sum(x, y, size)
        if (power > max_power) {
          max_power <- power
          best <- c(x = x, y = y, size = size, max_power = power)
        }
      }
    }
  }
  best
}

find_cell_any_size(grid)[1:3] %>% 
  paste(collapse = ",")
```
