---
params:
   task_nr: 24
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2024"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(tibble)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(stringi)
library(knitr)
library(cli)
library(bit64)
library(igraph)
library(tidyr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  assignments <- text_block %>% 
    str_subset(":")
  assignments <- str_extract(assignments, "\\d$") %>% 
    as.integer() %>% 
    set_names(assignments %>% 
                str_remove(": \\d")) %>% 
    as.list() %>% 
    c(
      AND = function(x, y) as.integer(x & y),
      OR = function(x, y) as.integer(x | y),
      XOR = function(x, y) as.integer(x != y)
    ) %>% 
    list2env()
  rules <- text_block %>%
    str_subset(":", TRUE) %>% 
    str_match_all("([^ ]+) ([^ ]+) ([^ ]+) -> ([^ ]+)") %>% 
    do.call(rbind, .) %>% 
    set_colnames(c("_", "x", "op", "y", "res")) %>% 
    as_tibble() %>% 
    select(x:res)
  
  create_call <- function(x, op, y, res){
    call("<-", as.name(res), call(op, as.name(x), as.name(y)))
  }
  
  rules <- rules %>% 
    mutate(cmd = do.call(Map, c(create_call, .)) %>% unname(),
           is_atomic = str_detect(x, "^[xy]") &
             str_detect(y, "^[xy]"))
  list(assignments = assignments,
       rules = rules)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

The problem boils down to bringing the rows in the right direction, such that terms, which
depend on other terms are not evaluated before.

```{r get-solution-1}
rearrange_rules <- function(data) {
  rules <- data$rules
  ordered_rules <- rules %>% filter(is_atomic)
  rules <- rules %>% 
    filter(!is_atomic)
  while (nrow(rules) > 0) {
    atoms <- ordered_rules %>% 
      pivot_longer(c(x, y, res)) %>% 
      pull(value)
    for (i in seq_len(nrow(rules))) {
      row <- rules[i, ]
      if (row$x %in% atoms && row$y %in% atoms) {
        ordered_rules <- rbind(ordered_rules, row)
        rules <- rules[-i, ]
        break
      }
    }
  }
  ordered_rules
}

evaluate_rules <- function(data, assignments = data$assignments) {
  new_rules <- rearrange_rules(data)
  walk(new_rules$cmd, ~ eval(.x, assignments))
  res <- names(assignments) %>% 
    str_subset("^z") %>% 
    sort(decreasing = TRUE) %>% 
    mget(assignments) %>% 
    unlist() %>% 
    paste(collapse = "") %>% 
    str_pad(64L, pad = "0")
  class(res) <- "bitstring"
  list(res = as.integer64(res), new_rules = new_rules)
}

res <- evaluate_rules(puzzle_data)
res$res
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For teh seocnd part we transform the rules into a graph first.

```{r create_graph}
create_calculator <- function(rules) {
  G <- make_empty_graph(directed = TRUE)
  rules <- rules %>% 
    mutate(op = paste(x, op, y, sep = "_"))
  nodes <- rules %>% 
    select(x:res)  %>% 
    pivot_longer(everything()) %>% 
    pull(value) %>% 
    unique()
  G <- G %>% 
    add_vertices(length(nodes), name = nodes) 
  for (i in seq_len(nrow(rules))) {
    row <- rules %>% 
      slice(i)
    G <- G %>% 
      add_edges(c(row$x, row$op, row$y, row$op, row$op, row$res))
  }
  nms <- V(G)$name
  V(G)$type <- case_when(
    str_detect(nms, "AND|OR|XOR") ~ "operator",
    str_detect(nms, "^[xy]") ~ "input",
    str_detect(nms, "^z") ~ "output",
    TRUE ~ "intermediate"
  )
  V(G)$op <- case_when(
    str_detect(nms, "_AND_") ~ "AND",
    str_detect(nms, "_XOR_") ~ "XOR",
    str_detect(nms, "_OR_") ~ "OR",
    TRUE ~ NA_character_
  )
  V(G)$color <- case_when(
    V(G)$type == "input" ~ "#FF7F00",
    V(G)$type == "output" ~ "forestgreen",
    V(G)$type == "operator" ~ "steelblue",
    V(G)$type == "intermediate" ~ "beige"
  )
  G %>% 
    simplify()
}

use_calculator <- function(G, x, y) {
  out_bits <- V(G)[type == "output"]$name %>% 
    str_extract("\\d+") %>% 
    unlist() %>% 
    as.integer() %>% 
    max()
  in_bits <- V(G)[type == "input"]$name %>% 
    str_extract("\\d+") %>% 
    unlist() %>% 
    as.integer() %>% 
    max()
  bitify <- . %>% 
    as.integer64() %>% 
    as.bitstring() %>% 
    str_split("") %>% 
    extract2(1L) %>% 
    as.integer() %>% 
    tail(in_bits + 1L)
  
  values <- rep(NA_integer_, vcount(G)) %>% 
    set_names(V(G)$name)
  values[sprintf("x%02d", in_bits:0)] <- bitify(x)
  values[sprintf("y%02d", in_bits:0)] <- bitify(y)
  sorted_nodes <- topo_sort(G, mode = "out")$name
  calc_nodes <- sorted_nodes %>% 
    setdiff(c(sprintf("x%02d", in_bits:0), sprintf("y%02d", in_bits:0)))
  for (node in calc_nodes) {
    me <- V(G)[node]
    parents <- adjacent_vertices(G, node, "in")[[1L]]$name
    if (me$type %in% c("output", "intermediate")) {
      stopifnot(length(parents) == 1L)
      val <- values[parents]
    } else {
      if (is.na(me$op) ||
                length(parents) != 2L) browser()
      if (me$op == "AND") {
        val <-  as.integer(values[[parents[[1L]]]] & values[[parents[[2L]]]])
      } else if (me$op == "OR") {
        val <- as.integer(values[[parents[[1L]]]] | values[[parents[[2L]]]])
      } else if (me$op == "XOR") {
        val <- (values[[parents[[1L]]]] + values[[parents[[2L]]]]) %% 2L
      }
    }
    values[[node]] <- val
  }
  res <- values[sprintf("z%02d", out_bits:0)] %>% 
    paste(collapse = "")
  class(res) <- "bitstring"
  as.integer64(res)
}

G <- create_calculator(res$new_rules) 
```

The rules should form a calculator. Thus, we first check which bits are wrongly wired by 
simply using $2^n, 0\leq n\leq 45$ and $0$ as inputs and check at which bits the results 
mismatch:

```{r get-wrong-bits, cache = TRUE}
get_wrong_bits <- function(G) {
  stringify <- . %>% 
    as.integer64() %>% 
    as.bitstring() %>% 
    str_split("") %>% 
    extract2(1L) %>% 
    tail(45)
  
  map(
    0:44,
    function(i) {
      a <- as.integer64(2 ^ i)
      b <- as.integer64(0)
      wr <- use_calculator(G, a, b)
      tr <- a + b
      mismatches <- rev(stringify(wr)) != rev(stringify(tr))
      if (any(mismatches)) {
        which(mismatches)
      } 
    }
  ) %>% 
    discard(is.null)
}

(wrong_bits <- get_wrong_bits(G))
```

Now that we have identified the wrong bits we can take a closer look to the respective 
subgraph and cehck the wiring. Let's first look into a subgraph where the calculation works

```{r working-bits}
get_subgraph <- function(G, input_bit, label_bad = TRUE) {
  input_path <- neighborhood(G, 4L, paste0(c("x", "y"), 
                                          sprintf("%02d", input_bit)),
                             mode = "out")
  output_path <-  neighborhood(G, 4L, paste0(c("x", "y"), 
                                          sprintf("%02d", input_bit - 1L)),
                             mode = "out")
  G <- subgraph(G,
    V(G)[unlist(input_path), unlist(output_path)])
  V(G)$label <- case_when(
    str_detect(V(G)$name, "_OR_") ~ "||",
    str_detect(V(G)$name, "_AND_") ~ "&&",
    str_detect(V(G)$name, "_XOR_") ~ "^^",
    TRUE ~ V(G)$name
  )
  V(G)$label.color <- if_else(V(G)$type == "intermediate", "black", "white")
  carry_over <- neighborhood(G, nodes = V(G)[str_detect(name, "_OR_")], 
                             mindist = 1, mode = "out") %>% 
    unlist() %>% 
    setdiff(V(G)[sprintf("z%02d", input_bit)])
  V(G)[carry_over]$label <- "CO"
  V(G)[carry_over]$color <- "#984EA3"
  V(G)[carry_over]$label.color <- "white"
  V(G)$size <- 22
  E(G)$arrow.size <- .5
  if (label_bad) {
    V(G)[sprintf("z%02d", input_bit)]$color <- "firebrick"
    V(G)[sprintf("z%02d", input_bit)]$label.color <- "white"
  }
  G %>% 
    set_graph_attr("layout", layout_as_tree(G))
}

get_subgraph(G, 2, FALSE) %>% 
  plot()
```

Let's look at the 3rd bit ($z_{02}$):
1. First we calculate $x_{02} \oplus y_{02}$ (depicted by `^^` in the graph), which is 
already very close to a binary addition: if both inputs equal 0 or 1 the result is 0, 
otherwise it is one.
1. However, we need to make sure that we also consider any carr-over from teh bits before.
Thus, we have to XOR the result again with the previous carry-over. (`CO` in the graph), 
which results in the final bit.
1. The last thing to do is to calculate the new carry-over bit, which can be calculated by 
the formula 
$\text{CO}_n = \left(\text{CO}_{n-1} \land (x_n \oplus y_n)\right) \lor (x_n \land y_n)$,
which translates to: either both bits on were 1, or at least one bit was 1 **and** the 
previous carry-over was 1.

Now that we know how a normal bit wiring should look like, we can look into the faulty 
bits to check what is amiss:

#### First Faulty Bit 

```{r first-bit}
get_subgraph(G, wrong_bits[[1L]][1L] - 1L) %>% 
  plot()
```

We see that `z05` and `tst` are wrongly wired.

#### Second Faulty Bit 

```{r second-bit}
get_subgraph(G, wrong_bits[[2L]][1L] - 1L) %>% 
  plot()
```

We see that `z11` and `sps` are wrongly wired.


#### Third Faulty Bit 

```{r third-bit}
get_subgraph(G, wrong_bits[[3L]][1L] - 1L) %>% 
  plot()
```

We see that `z23` and `frt` are wrongly wired.


#### Fourth Faulty Bit 

```{r fourth-bit}
get_subgraph(G, wrong_bits[[4L]][1L] - 1L) %>% 
  plot()
```

We see that `cgh` and `pmd` are wrongly wired.

Overall, wires `cgh,frt,pmd,sps,tst,z05,z11` and `z23` were wrongly wired and we see that
by fixing those we have a working binary adder:

```{r get-solution-2, cache = TRUE}
switch_lines <- function(G, from, to) {
  
  from_sp <- adjacent_vertices(G, from, "in")[[1L]]$name
  to_sp <- adjacent_vertices(G, to, "in")[[1L]]$name
  ids <- G %>% 
    get_edge_ids(c(from_sp, from, to_sp, to))
  G %>% 
    delete_edges(ids) %>% 
    add_edges(c(from_sp, to, to_sp, from))
}

G2 <- G %>% 
  switch_lines("z05", "tst") %>% 
  switch_lines("z11", "sps") %>% 
  switch_lines("z23", "frt") %>% 
  switch_lines("cgh", "pmd")


get_wrong_bits(G2)

```

