---
params:
   task_nr: 22
   year: 2017
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{js cpp-highlighter, file = "Prism_CPP.js", echo = FALSE}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  res <- text_block %>% 
    str_split("") %>% 
    do.call(rbind, .)
  res[] <- res == "#"
  storage.mode(res) <- "logical"
  res
  
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

For the first part (with reasonably sized inputs) we simply use an iterative approach, 
bursting as indicated

```{r get-solution-1}
burst <- function(k, node_status = puzzle_data) {
  dirs <- rbind(">" = c(0L, 1L),
                "v" = c(1L, 0L),
                "<" = c(0L, -1L),
                "^" = c(-1L, 0L))
  n <- nrow(node_status)
  dir <- 4L
  pos <- cbind(row = (nrow(node_status) + 1L) / 2L, 
               col = (ncol(node_status) + 1L) / 2L)
  nr_infections <- 0L
  for (i in 1:k) {
    if (node_status[pos]) {
      dir <- dir + 1L
    } else {
      dir <- dir - 1L
    }
    dir <- (dir - 1L) %% 4L + 1L
    nr_infections <- nr_infections + !node_status[pos]
    node_status[pos] <- !node_status[pos]
    pos <- pos + dirs[dir, , drop = FALSE]
    if (any(!between(c(pos), 1L, n))) {
      new <- matrix(FALSE, 3 * n, 3 * n)
      new[(n + 1L):(2 * n), (n + 1L):(2 * n)] <- node_status
      node_status <- new
      pos <- pos + cbind(n, n)
      n <- 3 * n
    }
  }
  nr_infections
}
burst(10000L, puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We solve the second part in `C++` as the numbers became too large for `R` to be efficient.
As always we include an `R` reference implementation in the appendix. 

```{Rcpp cpp-algorithm, file = "infection.cpp", cache = TRUE}
```

```{r get-solution-2, cache = TRUE}
burst_improved(10000000L, puzzle_data)
```

## R - Version

```{r burst-improved-r-version, eval = FALSE}
burst_improved <- function(k, node_status = puzzle_data) {
  storage.mode(node_status) <- "integer"
  node_status <- 2L * node_status
  dirs <- rbind(">" = c(0L, 1L),
                "v" = c(1L, 0L),
                "<" = c(0L, -1L),
                "^" = c(-1L, 0L))
  n <- nrow(node_status)
  dir <- 4L
  pos <- cbind(row = (nrow(node_status) + 1L) / 2L, 
               col = (ncol(node_status) + 1L) / 2L)
  nr_infections <- 0L
  for (i in 1:k) {
    if (node_status[pos] == 0L) {
      dir <- dir - 1L
    } else if (node_status[pos] == 2L) {
      dir <- dir + 1L
    } else if (node_status[pos] == 3L) {
      dir <- dir + 2L
    }
    dir <- (dir - 1L) %% 4L + 1L
    nr_infections <- nr_infections + (node_status[pos] == 1L)
    node_status[pos] <- (node_status[pos]+ 1L) %% 4L
    pos <- pos + dirs[dir, , drop = FALSE]
    if (any(!between(c(pos), 1L, n))) {
      new <- matrix(FALSE, 3 * n, 3 * n)
      new[(n + 1L):(2 * n), (n + 1L):(2 * n)] <- node_status
      node_status <- new
      pos <- pos + cbind(n, n)
      n <- 3 * n
    }
  }
  nr_infections
}

```