---
params:
   task_nr: 24
   year: 2016
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2016"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(collections)
library(gtools)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    str_split("") %>% 
    do.call(rbind, .)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

This is another instance of a path finding algorithm. We will use an A* search. Since the
order of the targets is not determined, we will first solve this small 
Travelling Salesman Problem by brute force to determine the optimal order of targets.

```{r get-solution-1, cache = TRUE}
get_route <- function(maze, roundtrip = FALSE, start = "0") {
  targets <- targets <- maze %>% 
    str_extract_all("[^.#]") %>% 
    unlist()
  n <- length(targets)
  targets_pos <- maze %in% targets
  dim(targets_pos) <- dim(maze)
  targets_pos <- which(targets_pos, arr.ind = TRUE) %>% 
    set_rownames(., maze[.])
  dist <- outer(1:n, 1:n, \(i, j) {
    abs(targets_pos[i, "row"] - targets_pos[j ,"row"]) +
      abs(targets_pos[i, "col"] - targets_pos[j , "col"])
  })
  rownames(dist) <- colnames(dist) <- targets
  path <- setdiff(targets, start)
  all_routes <- cbind(start, 
                      permutations(length(path), length(path), path))
  if (roundtrip) {
    all_routes <- cbind(all_routes, start)
  }
  min_dist <- Inf
  for (route_idx in 1:nrow(all_routes)) {
    route <- all_routes[route_idx, ]
    cur_dist <- 0L
    for (i in seq(1, length(route) - 1L)) {
      cur_dist <- cur_dist + dist[route[i], route[i + 1L]]
    }
    if (cur_dist < min_dist) {
      best_route <- route
      min_dist <- cur_dist
    }
  }
  targets_pos[tail(best_route, -1L), , drop = FALSE]
}

get_neighbors <- function(maze, cur_pos, target) {
  cur_pos <- cur_pos[, c("row", "col"), drop = FALSE]
  dims <- dim(maze)
  dir <- rbind(">" = c(0L, 1L),
               "v" = c(1L, 0L),
               "<" = c(0L, -1L),
               "^" = c(-1L, 0L)) %>% 
    set_colnames(c("row", "col"))
  target_pos <- target[, c("row", "col"), drop = FALSE]
  cand <- t(t(dir) + c(cur_pos))
  cand <- cand[between(cand[, 1L], 1L, dims[1L]) & 
                 between(cand[, 2L], 1L, dims[2L]), , drop = FALSE]
  walkable <- maze[cand] != "#"
  nbs <- cand[walkable, , drop = FALSE]
  heuristic <- t(abs(t(nbs) - c(target_pos))) %>% 
    rowSums()
  cbind(nbs, h = heuristic)
}

add_to_queue <- function(queue, cost, nbs, next_target, visited, target_index) {
  if (nrow(nbs) > 0) {
    for (i in 1:nrow(nbs)) {
      new_pos <- nbs[i, , drop = FALSE]
      f <- cost + new_pos[, "h"]
      new_pos <- cbind(new_pos, "cost" = cost + 1L)
      load <- list(visited = visited,
                   next_target = next_target[, c("row", "col"), drop = FALSE],
                   cur_pos = new_pos[, c("row", "col", "cost"), drop = FALSE],
                   target_index = target_index)
      queue$push(load, -f)
    }
  }
}

find_shortest_path <- function(maze, roundtrip = FALSE, start = "0") {
  shortest_path <- Inf
  
  pq <- priority_queue()
  hash <- list()
  start_pos <- which(maze == start, arr.ind = TRUE) %>% 
    set_colnames(c("row", "col")) %>% 
    set_rownames(start) %>% 
    cbind(cost = 0L)
  route <- get_route(maze, roundtrip, start) 
  visited <- rep(FALSE, nrow(route)) %>% 
    set_names(rownames(route))
  target_index <- 1L
  next_target <- route[target_index, , drop = FALSE]
  nbs <- get_neighbors(maze, start_pos, next_target)
  add_to_queue(pq, start_pos[, "cost"], nbs, next_target, visited, target_index)
  
  target_ids <- paste(as.numeric(visited), collapse = "")
  hash[[target_ids]] <- rbind(hash[[target_ids]],
                              start_pos[, c("row", "col"), drop = FALSE])
  while (pq$size() > 0L) {
    next_node <- pq$pop()
    cur_pos <- next_node$cur_pos
    next_target <- next_node$next_target
    if (all(cur_pos[, c("row", "col")] == next_target[, c("row", "col")])) {
      ## we hit a target
      target <- rownames(next_target)
      if (!next_node$visited[target]) {
        next_node$visited[target] <- TRUE
        next_node$target_index <- next_node$target_index + 1
      }
      if (all(next_node$visited)) {
        shortest_path <- cur_pos[, "cost"]
        break
      } else {
        next_target <- route[next_node$target_index, , drop = FALSE]
      }
    }
    target_ids <- paste(as.numeric(next_node$visited), collapse = "")

    hash[[target_ids]] <- rbind(hash[[target_ids]],
                                cur_pos[, c("row", "col"), drop = FALSE])
    nbs <- get_neighbors(maze, cur_pos, next_target)
    dupes <- !duplicated(rbind(hash[[target_ids]], 
                               nbs[, c("row", "col"), drop = FALSE]))
    idx <- dupes[-(1:nrow(hash[[target_ids]]))]
    add_to_queue(pq, cur_pos[, "cost"], nbs[idx, , drop = FALSE], 
                 next_target, next_node$visited, next_node$target_index)
  }
  shortest_path
}
find_shortest_path(puzzle_data)

```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For the second part, we rquire the TSP to end in the start city (which leads to another
route).

```{r get-solution-2, cache = TRUE}
find_shortest_path(puzzle_data, TRUE)
```
