---
params:
   task_nr: 1
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Preamble

It has been 3 years since I last participated in the AoC, so I needed to revisit my old 
code base to see how I have done it the last time. Fortunatlye, things were rather 
automatized already and I also remembered that I needed to logon once by hand to retrieve 
the session cookie to automatize the data retrieval.

So for future references: visit the page, open up Chrome DevTools, go to *Network*, reload
the page and then click on *Cookies*, where you will find the cookie named `session`.

This is the one which we need to store in our `keyring`.


# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(tibble)
library(dplyr)
library(purrr)
library(stringr)
library(igraph)
library(kableExtra)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

puzzle_data <- local({
  res <- GET(paste0(base_url, "/input"),
             session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    str_split("\n+") %>% 
    extract2(1L) %>% 
    str_split("\\s+")
  do.call(rbind, res) %>% 
    set_colnames(c("start", "end")) %>% 
    as_tibble() %>% 
    filter(nzchar(start)) %>% 
    mutate(across(everything(), as.integer))
})
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

The first 10 rows of the data looks like this:

```{r show-data, echo = FALSE}
kbl(
  puzzle_data %>% 
    slice(1:10)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)
```


We simply sort both columns subtract them and sum the differences:

```{r get-solution-1}
puzzle_data %>% 
  summarize(solution = sum(abs(sort(start) - sort(end))))
```



## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

To solve the second part, we simply inner-join the `start` column with the `end` column, 
allowing for a "many-to-many" relationship. In this way we automatically create as many 
copies of a value in `start` as there are matching values in `end`. Eventually we have to 
sum up these values to get the answer.

```{r get-solution-2}
inner_join(
  puzzle_data %>% 
    select(start),
  puzzle_data %>% 
    select(end),
  c(start = "end"),
  relationship = "many-to-many"
) %>% 
  summarize(solution = sum(start))
```


