---
params:
   task_nr: 19
   year: 2016
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2016"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(DT)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    as.integer()
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

For the first part we implement the selection process straight forward:

```{r get-solution-1}
eliminate_elves <- function(elves) {
  n <- length(elves)
  if (n == 1) {
    return(elves)
  }
  if (n %% 2L == 0L) {
    remaining <- elves[seq(1L, by = 2L, length.out = n / 2L)]
  } else {
    remaining <- elves  [seq(3L, by = 2L, length.out = floor(n / 2L))]
  }
  Recall(remaining)
}
eliminate_elves(1L:puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For the second part, let's have a look at the winners for $n=1,\ldots,27$:

```{r show-winners}
winners <- tibble(n = 1:27,
                  Winner = c(
                    1, 1, 3, 1, 2, 3, 5, 7, 9, 1, 2, 3, 4, 5, 6, 7, 8, 9, 11, 
                    13, 15, 17, 19, 21, 23, 25, 27)) %>% 
  mutate(
    m = 3 ^ floor(log(n, base = 3)),
    type = case_when(
    n == m ~ "power of 3",
    m < n & n <= 2 * m ~ "inc 1",
    2 * m < n & n < 3 * m ~ "inc 2"
    )
  )
winners %>% 
  datatable(
    class = c("compact", "nowrap", "hover", "row-border"),
    rownames = FALSE,
    options = list(
      pageLength = nrow(winners),
      dom = "t",
      ordering = FALSE,
      columnDefs = list(
        list(
          className = "dt-center", targets = "_all"
        ),
        list(
          visible = FALSE,
          targets = 2:3
        )
      ))
  ) %>% 
  formatStyle(1:2,
              "type",
              target = "row",
              backgroundColor = styleEqual(c("power of 3", "inc 1", "inc 2"), 
                                           c("#440154FF", "#21908CFF", "#FDE725FF")),
              color = styleEqual(c("power of 3", "inc 1", "inc 2"), 
                                           c("white", "black", "black"))
              )

```

We observe:

1. If the numbers of elves is a power of 3 the last elf will win.
1. Between powers the id of the winner elf increases linearly by a rate of 1 for the 
first $3^m$ elves and then by a rate of 2.

With this we can get a closed formula for the winning elf:

```{r get-solution-2}
get_winner <- function(N) {
  m <- 3 ^ floor(log(N, base = 3L))
  if (m == N) {
    N
  } else if (m < N && N <= 2 * m) {
    N - m
  } else if (2 * m < N && N < 3 * m) {
    2 * N - 3 * m
  }
}

get_winner(puzzle_data)
```
