---
params:
   task_nr: 12
   year: 2016
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2016"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

col_text <- function(x, bg_col, fg_col = "black") {
  sprintf("<span style='background-color:%s; color: %s'>%s</span>",
          bg_col, fg_col, x)
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(DT)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  str_split(text_block, " ") %>% 
    map(~ c(., NA_character_)[1:3] %>% 
          set_names(c("op", "arg1", "arg2"))) %>% 
    bind_rows()
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

The instructions are the following:
```{r show-instructions}
codes <- puzzle_data  %>% 
  set_colnames(c("Operation", "Argument 1", "Argument 2")) %>% 
  mutate(outer_block = c(0L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 1L, 2L, 1L, 0L, 
                         1L, 0L, 0L, 2L, 0L, 3L, 1L, 0L, 1L, 0L, 3L))

datatable(
  codes, 
  class = c("compact", "nowrap", "hover", "row-border"),
  options = list(
    pageLength = nrow(codes),
    dom = "t",
    ordering = FALSE,
    columnDefs = list(
      list(
        className = "dt-center", targets = "_all"
      ),
      list(
        visible = FALSE,
        targets = 4
      )
    ))
) %>% 
  formatStyle(1:3,
              "outer_block",
              target = "cell",
              backgroundColor = styleEqual(0:3, c("transparent", "#FDE725FF", "#440154FF", 
                                                  "#21908CFF")), 
              color = styleEqual(0:3, rep(c("black", "white", "black"), c(2L, 1L, 1L))) 
  ) 


```

Analyzing the instructions one will notice certain patterns:

1. A `inc a | dec b | jnz b -2` block first increases `a` by one, the decreases `b` by one
and jumps back two instructions - that is repeats this sequence until `b` equals zero. 
This is the same as adding `b` to `a` (we will refer to this pattern as *addition loop*).
1. More generally, a `dec a | jnz a -n` sequence forms a loop, which repeats the previous
`n` lines `a` times.
1. Finally, a `jnz a n` with a positive `n` skips the next `n` lines conditionally on `a`.

With these observations we conclude that

1. Lines 6 to 9 are skipped.
1. We have 2 addition loops marked in `r col_text("yellow", "#FDE725FF")` (not counting 
the skipped loop in lines 6 to 9).
1. We have 2 outer loops marked in `r col_text("turquoise", "#21908CFF")` and 
`r col_text("purple", "#440154FF", "white")` respectively.
1. The `r col_text("turquoise loop", "#21908CFF")` simply adds the value of `d` `c` times.

```{r get-solution-1}
get_code <- function(a, b, c, d) {
  for (i in 1:d) {
    c <- a
    a <- a + b
    b <- c
  }
  a + 14L * 14L
}
get_code(1L, 1L, 0L, 26L)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

With a different starting value for `c` we now include lines 6 - 9, which simply set 
different values for `c` and `d`.

```{r get-solution-2}
get_code(1L, 1L, 7L, 33L)
```


## Brute Force Algorithm

If you simply follow the rules line by line you could use the following algorithm. 
However, since this would introduce a lot of loops where a simple addition would do the 
same job, it would be awefully slow (imagine calculating `a + 100` by adding 1 to `a` a
hundred times).

```{r brute-force, eval = FALSE}
automata <- function(ops = puzzle_data) {
  reg = c(a = 0L, b = 0L, c = 0L, d = 0L)
  line <- 1
  cpy <- function(arg1, arg2) {
    new <- suppressWarnings(as.integer(arg1))
    if (is.na(new)) {
      new <- reg[arg1]
    }
    reg[arg2] <<- new
    line <<- line + 1L
  }
  inc <- function(arg1, arg2) {
    reg[arg1] <<- reg[arg1] + 1L
    line <<- line + 1L
  }
  dec <- function(arg1, arg2) {
    reg[arg1] <<- reg[arg1] - 1L
    line <<- line + 1L
  }
  jnz <- function(arg1, arg2) {
    arg2 <- as.integer(arg2)
    chk <- suppressWarnings(as.integer(arg1))
    if (is.na(chk)) {
      chk <- reg[arg1]
    }
    if (chk != 0L) {
      line <<- line + arg2
    } else {
      line <<- line + 1
    }
  }
  n <- nrow(ops)
  while (line <= n) {
    ops %>% 
      slice(line) %>% 
      rename(name = op) %>% 
      as.list() %>% 
      do.call(call, .) %>% 
      eval()
  }
  reg["a"]
}
```