---
params:
   task_nr: 5
   year: 2022
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  moves_idx <- str_which(text_block, "move")[1L]
  moves <- str_extract_all(text_block[moves_idx:length(text_block)], "\\d+") %>% 
    do.call(rbind, .) %>% 
    set_colnames(c("n", "from", "to")) %>% 
    as_tibble() %>% 
    mutate(across(everything(), as.integer))
  start <- text_block[1:(moves_idx - 2L)]
  l <- max(map_int(start, nchar))
  start_mat <- str_split_fixed(start, "", l)
  NOK <- apply(start_mat, 2L, \(c) all(c %in% c(" ", "[", "]")))
  start_mat <- start_mat[, !NOK]
  start_mat[start_mat == " "] <- NA_character_
  list(moves = moves, start_pos = apply(start_mat, 2L, \(c) list(c[!is.na(c)])) %>% 
         flatten())
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We apply one crane movement after another, using the list of 9 vectors as data storage. In
prepaation for part 2, we also include a parameter, which determiens wheterh the crates
need to be shifted.

```{r get-solution-1}
move_crane <- function(stacks, moves, reverse) {
  handler <- if (reverse) rev else identity
  for (i in seq_len(nrow(moves))) {
    move <- moves %>% 
      slice(i) %>% 
      as.list()
    load <- head(stacks[[move$from]], move$n)
    stacks[[move$from]] <- tail(stacks[[move$from]], -move$n)
    stacks[[move$to]] <- c(handler(load), stacks[[move$to]])
  }
  map_chr(stacks, 1L) %>% 
    paste(collapse = "")
}
move_crane(puzzle_data$start_pos, puzzle_data$moves, TRUE)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

All we have to do is, to add the crates in the same order instead of reverting their 
order.


```{r get-solution-2}
move_crane(puzzle_data$start_pos, puzzle_data$moves, FALSE)
```
