---
params:
   task_nr: 22
   year: 2015
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2015"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{js cpp-highlighter, file = "Prism_CPP.js", echo = FALSE}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(tidyr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  stats <- str_extract_all(text_block, "\\d+") %>% 
    flatten() %>% 
    as.integer() %>% 
    set_names(c("boss_HP", "dmg")) %>% 
    as.list()
  spells <- tibble(
    name = c("Magic Missile",
             "Drain",
             "Shield",
             "Poison",
             "Recharge"),
    damage = c(4L, 2L, 0L, 3L, 0L),
    heal = c(0L, 2L, 0L, 0L, 0L),
    armor = c(0L, 0L, 7L, 0L, 0L),
    mana = c(0L, 0L, 0L, 0L, 101L),
    duration = c(0L, 0L, 6L, 6L, 5L),
    costs = c(53L, 73L, 113L, 173L, 229L)
  )
  list(stats = c(stats, mana = 500L, hp = 50L), spells = spells)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

This was a difficult puzzle, where we overread a lot of details in hte first iterations 
(e.g. that buffs are applied not only before the player's turn but also before the
boss' turn). We started out with a recursive  R implementation, but as there are quite 
some combinations to check, R's copy on modify semantic became soon a deal-breaker. Thus, 
we switched to C++.

We learned an aweful lot about smart pointers (`unique_ptr` with its move semantics and
`shared_ptr`) just to realize that using a **unique** pointer, for something which 
effectively should be shared, is not the smartest move.

We needed a lot of iterations before the code worked, which was due to the self inflicted 
complications with moving unique pointers and even more importantly due to not following 
the instructions of the game.

Eventually We came up with the following solution (after switching from a recursive 
function to an iterative method). The idea is that we save the current state, cast an 
affordable spell (which is not yet casted) and add the new state to the stack. 
If we find a solution (i.e. the boss is dead), we update the mana count. During 
the process, if we hit a state which is more expensive than the best solution so far, 
we pruned to save some iterations.

```{Rcpp cpp-algorithm, file = "wizard_game.cpp", cache = TRUE}
```

```{r get-solution-1}
get_lowest_mana(puzzle_data$stats, puzzle_data$spells)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

While the first part of the solution took a considerable amount of time, the second one
was very easy, as we simply had to add another game step. This time we understood the
instructions properly and just added a flag to the algorithm to determine which version
we want to simulate.

```{r get-solution-2}
get_lowest_mana(puzzle_data$stats, puzzle_data$spells, TRUE)
```
