---
params:
   task_nr: 24
   year: 2015
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2015"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{js cpp-highlighter, file = "Prism_CPP.js", echo = FALSE}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(tidyr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    as.integer()
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We use a backtracking algorithm in C++ as always when we need to backtack a significant
amount of time with heavy data structures (`R` would make endless copies and slowing down
the process significantly).

The workhorse function is `can_partition` which checks whether the numbers can be split
into partition of the same sum. If we want to partition a vector into `k` groups, the
sum of all numbers must be dividable by k, and the sum divided by k must be the target 
sum. Then in the backtracking algorithm, we try to add a number to the partition and 
recurse. If the sum of selected numbers equals the target, we recurse again, decreasing 
the number of groups by one. Thta is, we found one partition and check whtehr the 
remaining numbers can also form partitions of the same sum, but we need one group less 
(the very one we just formed).

If there is only one number left, the algorithm returns `true`, as per definition this
number must equal the target. If at any stage the recursion returned `false`, we backtrack 
by deslecting this number and tring the next one. If we did not return with `true` from
the recursion we return `false` as we could not find a valid partition.

With this workhorse function what is left is to loop through all group sizes from 1 to
the number of elements in our vector. We create all combinations of this size, form the
first group with these combinations (only if the partial sum equals the target) 
and check whether we can partition the remaining
numbers into k - 1 groups. If this is possible, we store the result, keeping the best
solution (minimum group size, and minimum quantum energy).

As soon as we found the first minimum solution, we stop, as we do not need to look for
larger groups.

```{Rcpp cpp-algorithm, file = "quantum.cpp", cache = TRUE}
```

```{r get-solution-1}
find_min_quantum_entanglement(puzzle_data, 3L)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution


```{r get-solution-2}
find_min_quantum_entanglement(puzzle_data, 4L)
```
