---
params:
   task_nr: 12
   year: 2015
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2015"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(purrr)
library(stringr)
library(jsonlite)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block 
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We rely on exiting JSON parsers to parse the string, then we extract all numbers and sum
them up.

```{r get-solution-1}
sum_numbers <- function(json_data) {
 unlist(json_data) %>% 
   as.integer() %>% 
   suppressWarnings() %>% 
   sum(na.rm = TRUE)
}
json_data <- fromJSON(puzzle_data, simplifyVector = FALSE)
sum_numbers(json_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We use a recursive function to iterate through the elements. Whenever `red` is part of the
elements we do nothing, else we add the numbers to the final results. We jsut have to make
sure that we apply this logic only to JSON objects. To identify whether the original JSON
string was an object or an array, we check whether the slot has names (in this case it was
an object), otherwise it was an array.

```{r get-solution-2}
sum_filtered_numbers <- function(json_data) {
  sum_list <- function(branch) {
    is_object <- branch %>% 
      names() %>% 
      is.null() %>% 
      not()
    slot_sum <- children_sum <- 0L
    if (!is_object || !"red" %in% branch) {
      slot_sum <- keep(branch, is.numeric) %>% 
        unlist() %>% 
        sum()
      children_sum <- keep(branch, is.list) %>% 
        map_int(sum_list) %>% 
        sum()
    }
    slot_sum + children_sum
  }
  sum_list(json_data)
}

sum_filtered_numbers(json_data)
```
