---
params:
   task_nr: 21
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2024"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(tibble)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(stringi)
library(knitr)
library(cli)
library(bit64)
library(igraph)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    str_split("")
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

This puzzle took my way longer than it should have. Anyways, the idea is as follows:

1. Create graphs for both pads to facilitate calculating shorted paths between keys.
1. To significantly reduce the amount of paths between 2 keys we use several heuristics:
   1. As we always start at `A` we see that `<` is the most expensive key to reach, 
   followed by `v`.
   1. `^` and `<` have teh same costs, however, the former requires a `<` on the next
   robot, making it slightly more expansive. Thus the cheapest keys to reach are `>`, 
   `^`, `v` and `<` (in this order).
   1. Consecutives moves in the same directiona re cheaper than alternating on the next
   layer. `>>^` and `>^>` will end up on the same key, but the next robot needs more moves
   to type the latter sequence: `vAA^<A` vs `vA^<A>vA`.
1. We end up with a list of shortest moves. For some key combinations there are still 2
options, for instance we can move from `A -> v` wither by `v|<` or via `<|v`.
1. We maintain a list of all moves from each key to all the other keys. If an entry has
only a single entry it means that there is just one optimal move, if it has 2, we have to
calculate the length for both options. Becasue of the advantage of consecutive moves, 
there are at most 2 paths from any key on the numeric pad to any other key.
1. With this setting the algorithm is rather straight forward:
   1. Start with the (numeric) code.
   1. Tokenize the code, i.e. create the movments needed. For instance the code `540A`
   translates to `[A,5], [5,4], [4,0], [0,A]` representing the movements needed.
   1. Use the lookup table created before to get all key presses on the (next) directional
   pad.
   1. For each keypress sequence just generated recurse into the funciton 
   increasing the depth counter until you reach the final depth.
   1. If there was more than one option, calculate both options and keep track of the
   minimum.
   1. Add the result of the minimum path to the current string length.
   1. Once you are at the final depth simply count the number of keys and stop.
   1. We must keep a cache of lengths of certain key sequences at eahc depth, otherwise
   we would end up in way too many recursions. That is if we found out that at depth 4 a
   key seqeunce of `v|A` requires x keys at the final depth we store this information. The
   next time we need to calculate the number of keys needed for `v|A` at depth 4, we 
   **do not** recurse again, but ismply use the cached value.

```{r get-solution-1}
make_pad <- function(type = c("num", "dir")) {
  type <- match.arg(type)
  if (type == "num") {
    G <- make_lattice(c(3L, 4L), directed = TRUE, mutual = TRUE)
    V(G)$name <- c("--", "0", "A", 1:9)
    E(G)$name <- c(">", "^", ">", "<", "^", "<", "^",
                   rep(c(">", "^", "v", ">", "<", "^", "v", "<", "^", "v"), 2L),
                   ">", "v", ">", "<", "v", "<", "v")
  } else if (type == "dir") {
    G <- make_lattice(c(3L, 2L), directed = TRUE, mutual = TRUE)
    V(G)$name <- c("<", "v", ">", "--", "^", "A")
    E(G)$name <- c(">", "^", ">", "<", "^", "<", "^",
                   ">", "v", ">", "<", "v", "<", "v")
  }
  V(G)$shape <- "square"
  V(G)$color <- "lightgray"
  V(G)$size <- 50
  E(G)$arrow.size <- 0
  costs <- c(">" = 1, "^" = 2, "v" = 3, "<" = 4)
  E(G)$weight <- costs[E(G)$name]
  G %>% 
    delete_edges(incident(G, "--")) %>% 
    set_graph_attr("layout", layout_on_grid(G, width = 3L, height = 4L))
}

plot_graphs <- function(graphs) {
  op <- par(mfrow = 1:2)
  on.exit(par(op))
  for (i in graphs) {
    plot(i)
  }
}

get_dup_length <- function(path) {
  ## idea find paths which have many moves into the same direction
  nodes <- str_split(path, "\\|")[[1L]]
  len <- rle(nodes)$lengths
  length(len) * (length(nodes) - max(len))
}

get_shortest_paths <- function(G) {
  map(V(G)[name != "--"]$name, function(start) {
    path <- all_shortest_paths(G, start, V(G)[name != "--"], mode = "out")
    tibble(from = start,
           to = map_chr(path$vpaths, ~ tail(.x, 1L)$name),
           path = map_chr(path$epaths, 
                          ~ E(G)[.x]$name %>% 
                            c("A") %>% 
                            paste(collapse = "|")),
           dup_len = map_int(path, get_dup_length)
    )
  }) %>% 
    list_rbind() %>% 
    slice_min(dup_len, n = 1L, by = c(from, to))
}

tokenize_path <- function(path) {
  nodes <- str_split(path, "\\|")[[1L]]
  cbind(c("A", head(nodes, -1L)),
        nodes) %>% 
    apply(1, paste, collapse = "|")
}

graphs <- c("num", "dir") %>% 
  set_names() %>% 
  map(make_pad)

shortest_paths <- map(graphs, get_shortest_paths) %>% 
  list_rbind() %>% 
  mutate(start = paste(from, to, sep = "|"),
         end = path, .keep = "none") %>% 
  summarize(path = list(end), .by = start) %>% 
  deframe()

get_input_length <- function(start, max_depth, shortest_paths) {
  cache <- vector("list", max_depth + 2L)
  get_next_sequence <- function(old_path, depth) {
    if (old_path %in% names(cache[[depth + 1L]])) {
      res <- cache[[depth + 1L]][[old_path]]
    } else {
      if (depth == max_depth + 1L) {
        res <- str_remove_all(old_path, "\\|") %>% 
          str_length() %>% 
          as.integer64()
      } else {
        tokens <- tokenize_path(old_path)
        res <- as.integer64(0L)
        for (i in seq_along(tokens)) {
          new_path <- shortest_paths[[tokens[[i]]]]
          new_paths_lengths <- map(new_path, get_next_sequence, depth + 1L) %>% 
            do.call(c, .)
          res <- res + min(new_paths_lengths)
        }
      }
      res <- res %>% 
        set_names(old_path)
      ## need this slightly awkward `if` because I cannot pre fill a list with
      ## an empty int64 (if I try as.integer64(NULL) and `c` the results it gets 
      ## messed up)
      if (is.null(cache[[depth + 1L]])) {
        cache[[depth + 1L]] <<- res
      } else {
        cache[[depth + 1L]] <<- c(cache[[depth + 1L]], 
                                 res)
      }
    }
    res
  }
  res <- get_next_sequence(start, 0L)
  res
}

calculate_complexity <- function(codes, max_depth, shortest_paths) {
  map(codes, function(code) {
    int_code <- paste(code %>% head(-1L),
                      collapse = "") %>% 
      as.integer64()
    code_len <- get_input_length(paste(code, collapse = "|"),
                                 max_depth, shortest_paths)
    int_code * code_len
  }) %>% 
    do.call(c, .) %>% 
    as.integer64() %>% 
    sum()
}

calculate_complexity(puzzle_data, 2L, shortest_paths)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution


```{r get-solution-2}
calculate_complexity(puzzle_data, 25L, shortest_paths)
```

