---
params:
   task_nr: 20
   year: 2023
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(numbers)
library(collections)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L)
  }
  layout <- text_block[nzchar(text_block)] %>% 
    str_split(" -> |, ")
  nms <- map_chr(layout, 1L)
  type <- if_else(str_detect(nms, "^[%&]"), str_sub(nms, 1L, 1L), nms)
  nbs <- map2(layout, type, ~ list(nb = tail(.x, -1L), type = .y))
  names(nbs) <- if_else(str_detect(nms, "^[%&]"), str_sub(nms, 2L), nms)
  nbs
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We keep track about the state of each module. For flip-flops it is their power status (on
or off) and for conjunction modules it is the last seen pulse per input. Then, we maintain
a queue of next pulses to send and update the signal depending on the module type.

*N.B. 1* In an earlier version we also kept a memo hash to recognize cycles and to shortcut
them. However, neither for part 1 nor part 2 this s necessary, so we removed this bit.

*N.B. 2* We allow for some additional functionality, which is only needed and explained 
in part 2.


```{r get-solution-1}
count_pulses <- function(layout, nr_presses, stop_on_rx = FALSE) {
  state <- imap(layout, function(mod, nm) {
    if (mod$type == "broadcaster") {
      c(state = FALSE)
    } else if (mod$type == "%") {
      c(state = FALSE)
    } else if (mod$type == "&") {
      inp <- keep(layout, ~ nm %in% .x$nb) %>% 
        names()
      rep(FALSE, length(inp)) %>% 
        set_names(inp)
    }
  })
  j <- 1L
  pulse_cnt <- c(0L, 0L)
  rx_parent <- keep(layout, ~ "rx" %in% .x$nb) %>% 
    names()
  rx_inputs <- state[[rx_parent]] %>% 
    names()
  
  period <- last_seen <- rep(NA_integer_, length(rx_inputs)) %>% 
    set_names(rx_inputs)
  stable <- rep(FALSE, length(rx_inputs)) %>% 
    set_names(rx_inputs)
  
  note_high <- function(sender, j) {
    if (is.na(last_seen[[sender]])) {
      # first High
      last_seen[[sender]] <<- j
    } else if (is.na(period[[sender]])) {
      # second High -> first diff
      period[[sender]] <<- j - last_seen[[sender]]
      last_seen[[sender]] <<- j
    } else {
      # third or later high
      d <- j - last_seen[[sender]]
      if (d == period[[sender]]) {
        stable[[sender]] <<- TRUE
      } else {
        ## if distance is varying - replace old info
        period[[sender]] <<- d
      }
      last_seen[[sender]] <<- j
    }
  }
  
  while (j <= nr_presses) {
    queue <- queue()
    queue$push(c(broadcaster = FALSE))
    while (queue$size() > 0L) {
      signal <- queue$pop()
      pulse_cnt[signal + 1L] <- pulse_cnt[signal + 1L] + 1L
      nm <- names(signal)
      if (nm %in% names(layout)) {
        recv <- layout[[nm]]
        nb <- recv$nb
        n <- length(nb)
        is_conjunction <- map_lgl(layout[nb], ~ !is.null(.x) && .x$type == "&")
        cjs <- nb[is_conjunction]
        send <- TRUE
        if (recv$type == "%") {
          if (!signal) {
            ## flip-flop receives low pulse
            state[[nm]]["state"] <- signal <- !state[[nm]]["state"]
          } else {
            ## flip-flop get high pulse don't send anything
            send <- FALSE
          }
        } else if (recv$type == "&") {
          signal <- !all(state[[nm]])
        }
        if (stop_on_rx) {
          if (signal && send && length(nb) && nm %in% rx_inputs) {
            note_high(nm, j)
          }
          if (all(stable)) {
            return(reduce(period, LCM))
          }
        }
        if (send) {
          for (n in nb) {
            el <- signal %>% 
              set_names(n)
            queue$push(el)
          }
          walk(cjs, function(cj) {
            state[[cj]][nm] <<- signal
          })
        }
      }
    }
    j <- j + 1
  }
  prod(pulse_cnt)
}

count_pulses(puzzle_data, 1000L) %>% 
  print(digit = 16)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For the second part we note that the only module sending pulses to `rx` is a *conjunction*
module. It sends the required `low` pulse, whenever all of its inputs are `high`. Thus, we
look for the periodicity of those inputs and as soon as we identified the period (as soon 
as 2 time differences between `high` pulses is constant) of all inputs, we simply need to 
calculate the least common multiple of these periods to get the needed total length.

```{r get-solution-2, cache = TRUE}
count_pulses(puzzle_data, Inf, TRUE)
```
