---
params:
   task_nr: 24
   year: 2017
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}

.page-content table tr pre {
  border-radius: revert;
  background-color: revert;
  color: revert;
  padding: revert;
  border: revert;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(digest)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    str_split("/") %>% 
    map(as.integer)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We solve this puzzle by backtracking with memoization to avoid recalculating knwon paths. 

```{r get-solution-1, cache = TRUE}
get_strongest_bridge <- function(pieces) {
  start <- keep(seq_along(pieces), \(i) any(pieces[[i]] == 0))
  memo <- new.env()
  n <- length(pieces)
  best_weight <- 0
  best_path <- NULL
  dfs <- function(current_val, used, path) {
    key <- paste(current_val, digest(used), sep = "_")
    if (exists(key, memo)) {
      return(memo[[key]])
    }
    
    max_weight <- sum(unlist(pieces[path]))
    max_path <- path
    
    for (idx in which(!used)) {
      piece <- pieces[[idx]]
      if (any(piece == current_val)) {
        next_val <- setdiff(piece, current_val)
        if (length(next_val) == 0) {
          ## double piece
          next_val <- current_val
        }
        next_used <- used
        next_used[idx] <- TRUE
        next_path <- c(path, idx)
        res_weight <- dfs(next_val, next_used, next_path)
        if (res_weight > max_weight) {
          max_weight <- res_weight
          max_path <- next_path
        }
      }
    }
    
    if (max_weight > best_weight) {
      best_weight <<- max_weight
      best_path <<- max_path
    }
    memo[[key]] <- max_weight
    max_weight
  }
  
  used <- rep(FALSE, n)
  
  for(si in start) {
    piece <- pieces[[si]]
    current_val <- setdiff(piece, 0L)
    if (length(current_val) == 0L) {
      current_val <- 0L
    }
    start_used <- used
    start_used[si] <- TRUE
    dfs(current_val, start_used, si)
  }
  best_weight
}
get_strongest_bridge(puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We use the same approach as before replacing weight by length as the optimization 
criterion. In case there is a tie for the length, we use the weight as a tie-breaker.

```{r get-solution-2, cache = TRUE}
get_weight_of_longest_bridge <- function(pieces) {
  start <- keep(seq_along(pieces), \(i) any(pieces[[i]] == 0))
  memo <- new.env()
  n <- length(pieces)
  best_length <- 0
  best_path <- NULL
  dfs <- function(current_val, used, path) {
    key <- paste(current_val, digest(used), sep = "_")
    if (exists(key, memo)) {
      return(memo[[key]])
    }
    
    max_length <- length(path)
    max_path <- path
    
    for (idx in which(!used)) {
      piece <- pieces[[idx]]
      if (any(piece == current_val)) {
        next_val <- setdiff(piece, current_val)
        if (length(next_val) == 0) {
          ## double piece
          next_val <- current_val
        }
        next_used <- used
        next_used[idx] <- TRUE
        next_path <- c(path, idx)
        res_length <- dfs(next_val, next_used, next_path)
        if (res_length > max_length) {
          max_length <- res_length
          max_path <- next_path
        } else if (res_length == max_length) {
          old_weight <- sum(unlist(pieces[max_path]))
          new_weight <- sum(unlist(pieces[next_path]))
          if (new_weight > old_weight) {
            max_length <- res_length
            max_path <- next_path
          }
        }
      }
    }
    if (max_length > best_length) {
      best_length <<- max_length
      best_path <<- max_path
    } else if (max_length == best_length) {
      old_weight <- sum(unlist(pieces[best_path]))
      new_weight <- sum(unlist(pieces[max_path]))
      if (new_weight > old_weight) {
        best_length <<- max_length
        best_path <<- max_path
      }
    }
    memo[[key]] <- max_length
    max_length
  }
  
  used <- rep(FALSE, n)
  
  for(si in start) {
    piece <- pieces[[si]]
    current_val <- setdiff(piece, 0L)
    if (length(current_val) == 0L) {
      current_val <- 0L
    }
    start_used <- used
    start_used[si] <- TRUE
    dfs(current_val, start_used, si)
  }
  sum(unlist(pieces[best_path]))
}
get_weight_of_longest_bridge(puzzle_data)

```

