---
params:
   task_nr: 16
   year: 2023
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L)
  }
  text_block[nzchar(text_block)] %>% 
    str_split("") %>% 
    do.call(rbind, .)
  
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}

cat(puzzle[[1]] %>% 
      str_replace_all("\\\\", "\\\\\\\\"))
```

### Solution

We make use of matrix indexing. Whenever, we hit a mirror or a splitter we change the 
direction and add another beam respectively. We also keep track of visited fields 
(together with the direction) and revisit them only of we er coming with a different
direction. We use a bitmask to store this information.

```{r get-solution-1}
set_bit <- function(bitmask, i) {
  new_bit <- bitwShiftL(1L, i - 1L)
  bitwOr(bitmask, new_bit) 
}

is_visited <- function(bitmask, i) {
  new_bit <- bitwShiftL(1L, i - 1L)
  bitwAnd(bitmask, new_bit) != 0L
}

count_energized_tiles <- function(mirrors, start = cbind(1L, 0L, 2L)) {
  dirs <- rbind(
    "^" = c(-1L, 0L),
    ">" = c(0L, 1L),
    "v" = c(1L, 0L),
    "<" = c(0L, -1L)
  )
  nr <- nrow(mirrors)
  nc <- ncol(mirrors)
  visited <- matrix(0L, nr, nc)
  pos <- start
  turns <- matrix(c(2L, 1L, 4L, 3L,
                    4L, 3L, 2L, 1L,
                    -1L, 0L, -1L, 0L,
                    0L, -1L, 0L, -1L,
                    0L, 0L, 0L, 0L),
                  4L, 5L,
                  dimnames = list(rownames(dirs), c("/", "\\", "-", "|", ".")))
  while (nrow(pos) > 0L) {
    new_pos <- pos
    new_pos[, 1:2] <- new_pos[, 1:2, drop = FALSE] + dirs[pos[, 3L], , drop = FALSE]
    new_pos <- new_pos[new_pos[, 1L] >= 1L & new_pos[, 1L] <= nr &
      new_pos[, 2L] >= 1L & new_pos[, 2L] <= nc, , drop = FALSE]
    new <- !is_visited(visited[new_pos[, 1:2, drop = FALSE]], new_pos[, 3L])
    new_pos <- new_pos[new, , drop = FALSE]
    cur_dir <- new_pos[, 3L]
    visited[new_pos[, 1:2, drop = FALSE]] <- set_bit(visited[new_pos[, 1:2, drop = FALSE]],
                                                     cur_dir)
    tiles <- mirrors[new_pos[, 1:2, drop = FALSE]]
    pos <- pos[0L, , drop = FALSE]
    for (i in seq_along(tiles)) {
      turn <- turns[cur_dir[i], tiles[i]]
      if (turn == -1L) {
        if (cur_dir[i] %% 2L == 1L) {
          new_pos[i, 3L] <- 2L
          pos <- rbind(pos, new_pos[i , , drop = FALSE])
          new_pos[i, 3L] <- 4L
          pos <- rbind(pos, new_pos[i , , drop = FALSE])
        } else {
          new_pos[i, 3L] <- 1L
          pos <- rbind(pos, new_pos[i , , drop = FALSE])
          new_pos[i, 3L] <- 3L
          pos <- rbind(pos, new_pos[i , , drop = FALSE])
        }
      } else if (turn > 0L) {
        new_pos[i, 3L] <- turn
        pos <- rbind(pos, new_pos[i, , drop = FALSE])
      } else{
        pos <- rbind(pos, new_pos[i, , drop = FALSE])
      }
    }
  }
  sum(visited != 0L)
}
count_energized_tiles(puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We loop through all possible starts and find the maximum number of energized tiles using
the algorithm form part 1.

```{r get-solution-2, cache = TRUE}
find_best_starting_position <- function(mirrors) {
  nr <- nrow(mirrors)
  nc <- ncol(mirrors)
  starts <- rbind(
    cbind(1:nr, 0L, 2L),
    cbind(1:nr, nc + 1L, 4L),
    cbind(0L, 1:nc, 3L),
    cbind(nr + 1L, 1:nc, 1L)
  )
  n <- nrow(starts)
  res <- map_int(seq_len(n), ~ count_energized_tiles(mirrors, starts[.x, , drop = FALSE]))
  max(res)
}

find_best_starting_position(puzzle_data)
```
