---
params:
   task_nr: 6
   year: 2015
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2015"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(tibble)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(stringi)
library(knitr)
library(cli)
library(digest)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  coords <- text_block %>% 
    str_extract_all("\\d+") %>% 
    do.call(rbind, .) %>% 
    set_colnames(c("x0", "y0", "x1", "y1"))
  storage.mode(coords) <- "integer"
  coords %>% 
    add(1L) %>% 
    as_tibble() %>% 
    mutate(op = text_block %>% 
             str_extract("\\D+") %>% 
             str_trim(), .before = 1L)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We create a grid of booleans storing the current state and iterate through the 
instructions and execute each of them in turn. In the end we calculate the switched on 
lights.

```{r get-solution-1}
get_indices <- function(x0, y0, x1, y1, ...) {
  expand.grid(
    row = x0:x1,
    col = y0:y1
  ) %>% 
    as.matrix()
}

get_light_show <- function(ops, fns) {
  light_array <- matrix(FALSE, 1000L, 1000L)
  for (i in seq_len(nrow(ops))) {
    op <- ops %>% 
      slice(i) %>% 
      as.list()
    idx <- do.call(get_indices, op)
    light_array[idx] <- fns[[op$op]](light_array[idx])
  }
  light_array
}

fns_boolean <-   fns <- list(
    toggle = function(x) !x,
    "turn on" = function(x) TRUE,
    "turn off" = function(x) FALSE
  )

get_light_show(puzzle_data, fns_boolean) %>% 
  sum()
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

This time, instead of booleans we store integers reflecting their brightness.

```{r get-solution-2}
fns_integer <-   fns <- list(
    toggle = function(x) x + 2,
    "turn on" = function(x) x + 1,
    "turn off" = function(x) pmax(x - 1, 0)
  )

get_light_show(puzzle_data, fns_integer) %>% 
  sum()
```


For the fun of it, here's a picture of the lights:

```{r visualize-lights, cache = TRUE}
heatmap(
  get_light_show(puzzle_data, fns_integer),
  Rowv = NA,
  Colv = NA,
  col =  rgb(red = 0:255 / 255, blue = 0:255 / 255, green = 0:255 / 255),
  scale = "none",
  xaxt = "n",
  yaxt = "n",
  labRow = NA,
  labCol = NA
)
```

