---
params:
   task_nr: 21
   year: 2022
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(igraph)
library(bit64)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  tokens <- text_block %>% 
    str_remove_all(":") %>% 
    str_split (" ")
  ve <- map(tokens, function(line) {
    if (length(line) == 2L) {
      vals <- suppressWarnings(as.integer(line))
      vertices <- tibble(name = line,
                         value = vals,
                         type = if_else(is.na(vals), "intermediate", "input"),
                         operation = NA_character_)
      edges <- tibble(from = line[2L], to = line[1L])
    } else if (length(line) == 4L) {
      vertices <- tibble(name = line[1L],
                         value = NA_integer_,
                         type = "calculation",
                         operation = line[3L])
      edges <- tibble(from = line[c(2L, 4L)],
                      to = line[1L],
                      ops = c("left", "right"))
    }
    list(edges = edges, vertices = vertices)
  })
  list(
    vertices = map(ve, "vertices") %>% 
      list_rbind(),
    edges = map(ve, "edges") %>% 
      list_rbind()
  )
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We construct a graph from the monkey network, sort it topologically and propagate the 
values. 

```{r get-solution-1}
construct_graph <- function(monkeys) {
  graph_from_data_frame(
    monkeys$edges, TRUE, distinct(monkeys$vertices)
  )
}

calculate_value <- function(G) {
  nodes <- topo_sort(G)
  for (idx in seq_along(nodes)) {
    node <- nodes[idx]
    type <- node$type
    if (type == "intermediate") {
      father <- neighbors(G, node, "in")
      stopifnot(length(father) == 1L)
      V(G)[node]$value <- V(G)[father]$value
    } else if (type == "calculation") {
      incoming <- incident(G, node, "in")
      vals <- tail_of(G, incoming)$value
      x <- vals[incoming$ops == "left"]
      y <- vals[incoming$ops == "right"]
      V(G)[node]$value <- do.call(V(G)[node]$operation, list(x, y))
    }
  }
  G
}

G <- construct_graph(puzzle_data) %>% 
  calculate_value()
G2 <- construct_graph(ex) %>% 
  calculate_value()
V(G)["root"]$value %>% 
  as.integer64()
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For the second part we walk backwards from the root. We separate the graph into 2 
components, the one containing node `humn` and the rest. The rest will give us a fixed 
value, which we will use to calculate the desired value of the last node in the path from
`humn` to the current end.

```{r get-solution-2, cache = TRUE}
find_start_value <- function(G, source = "humn") {
  V(G)["root"]$operation <- "=="
  end <- V(G)["root"]
  done <- FALSE
  browser()
  while (!done) {
    path <- shortest_paths(G, source, end)$vpath[[1L]]
    new_end <- path[length(path) - 1L]
    incoming <- incident(G, end, "in")
    goal <- incoming %>% 
      tail_of(G, .) %>% 
      setdiff(new_end)
    goal <- goal$value
    op <- end$operation
    if (is.na(op)) {
      
    } else if (op == "==") {
      V(G)[new_end]$value <- goal
      end <- V(G)[new_end]
    } else if (op == "+") {
      
    } else if (op == "-") {
    } else if (op == "/") {
      
    } else if (op == "*") {
    }
  }
}

find_start_value(G2)
```

