---
params:
   task_nr: 14
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("%s_%s", format(Sys.Date(), "%Y"), 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
	width: 100%;
}

.back {
	font-size: 2rem;
}

.day-desc {
	border: 2px dotted;
	border-radius: 6px;
	border-color: var(--blockquote-border-color);
	padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
	if (knitr::is_latex_output()) {
		sprintf("\\texttt{%s}", as.character(x))
	} else if (knitr::is_html_output()) {
		sprintf("<tt>%s</tt>", as.character(x))
	} else {
		x
	}
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(tibble)
library(dplyr)
library(purrr)
library(stringr)
library(igraph)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2021/day/", params$task_nr)
puzzle <- GET(base_url,
				  session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	xml_find_all("///article") %>% 
	lapply(as.character)

puzzle_data <- local({
	res <- GET(paste0(base_url, "/input"),
				  session_cookie) %>% 
		content(encoding = "UTF-8") %>% 
		str_split("\n+") %>% 
		`[[`(1L)
	rules <- head(tail(res, -1L), -1L)
	lkp <- str_extract(rules, ".$") %>% 
		set_names(str_extract(rules, "^.."))
	list(start = res[[1L]],
		  rules = lkp)
})
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We split the string in pairs of two and replace matches according to the given
rules. Then we drop the first letter in each pair or triplet, as it is already
present in the pair or triplet before.

```{r solution-a, cache = FALSE}
format_rule <- function(rule) {
	sprintf("%s -> %s", names(rule), rule)
}

insert <- function(rule) {
	str_c(str_sub(names(rule), 1L, 1L),
			rule,
			str_sub(names(rule), 2L, 2L))
}

replace_line <- function(n, 
								 str = puzzle_data$start,
								 rules = puzzle_data$rules) {
	wrapper <- function(str, i) {
		idx <- seq(1L, str_length(str) - 1L)
		tokens <- str_sub(str, idx, idx + 1L)
		tokens <- insert(rules[tokens])
		tokens[-1L] <- str_sub(tokens[-1L], 2L)
		str_c(tokens, collapse = "")
	}
	all_ltrs <- str_c(rules, collapse = "") %>% 
		str_split("") %>% 
		`[[`(1) %>% 
		unique()
	reduce(seq(1L, n), wrapper, .init = str) %>% 
		str_count(fixed(all_ltrs)) %>% 
		range() %>% 
		diff()
}

replace_line(10)
```


## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

While the previous brute-force solution works for small `n` it quickly becomes not
manageable, because at each step the string can potentially grow by 
`r length(puzzle_data$rules)` (`=: k`) characters. That is, while the upper bound for the
string length after 10 iterations is `k ^ 10 = ` 
`` `r length(puzzle_data$rules) ^ 10` ``,
the upper bound for 40 iterations is already 
`` r length(puzzle_data$rules) ^ 40` ``. Of course not all `r length(puzzle_data$rules)` 
will hit in each of the iterations, hence we manage with 10 iterations but clearly for 40
iterations we have to find a smarter algorithm.

We will resort to a directed graph. Each node represents a tuple like 
`` `r names(puzzle_data$rules[1])` ``. By applying rule 
`` `r format_rule(puzzle_data$rules[1])` `` this tuple produces 
`` `r insert(puzzle_data$rules[1])` ``, which eventually results in the two *new* tuples
`` `r str_sub(insert(puzzle_data$rules[1]), 1:2, 2:3)` ``. In the graph we hence connect
`` `r names(puzzle_data$rules[1])` `` with 
`` `r str_sub(insert(puzzle_data$rules[1]), 1:2, 2:3)` `` via a directed edge. In this 
way we connect all tuples to its *2*  resulting tuples to form a graph. We connect a 
dedicated `start` node to the tuples found in the start word. We assign an initial `load` 
of `1` to the start node. If a tuple appears `k` times in the start string, there 
will be `k` edges from the start node to this very tuple.

Then the algorithm works as follows:

1. For each outgoing edge, add the full load to the child node. That is, if there are two
nodes `P` and `P'` say, which have an edge into `C`, then we add the load of `P` plus the 
load of `P'` to the load of `C`. Note, that the graph may have some loops (a rule like
`KN -> K` will lead to such a loop). This reflects that each tuple when polymerized will
create a new tuple. If there are for instance 3 `KN -> H` nodes and 2 `NN -> H`  nodes we
will - inter alia - have `3 + 2 = 5` new `HN` nodes. The amount is reflectd by the 
`load`.
1. We also have to reduce the number of "mother" polymers accordingly. In the above
example we will reduce the load of `KN` and `NN` by 3 and 2 respectively. Please note, 
that in one run we can produce __and__ destroy a certain tuple.
1. Once we transferred the loads to the childs, we recurse into all childs and continue
until we reached the desired depth.


```{r solution-b, eval = length(puzzle) > 1, comment = NA}
get_neighbor_node <- function(tuple) {
	nbs <- insert(tuple)
	from <- rep(1:2, length(tuple))
	str_sub(rep(nbs, each = 2), from, from + 1)
}

create_graph_from_rules <- function(start = puzzle_data$start,
												rules = puzzle_data$rules) {
	gr_dat <- data.frame(from    = rep(names(rules), each = 2L),
								to      = get_neighbor_node(rules),
								insert  = rep(rules, each = 2L),
								visited = 0L,
								load    = 0L)
	G <- gr_dat %>% 
		select(1:2) %>% 
		graph_from_data_frame(vertices = gr_dat %>% 
									 	select(-to) %>% unique())
	V(G)$color <- "orange"
	V(G)$shape <- "circle"
	idx <- seq_len(str_length(start) - 1L) 
	new_edges <- rep(str_sub(start, idx, idx + 1L), each = 2L)
	new_edges[2L * (0:(length(new_edges) / 2L - 1L)) + 1L] <- "start"
	G +
		vertices("start", 
					color = "steelblue", 
					shape = "square",
					load = 1L,
					visited = 1L) +
		edges(new_edges)
}

walk_graph <- function(max_depth,
							  start = puzzle_data$start,
							  rules = puzzle_data$rules) {
	op <- options(scipen = 99)
	on.exit(options(op))
	G <- create_graph_from_rules(start, rules)
	start_cnt <- tibble(ltr = unique(str_split(start, "")[[1L]]),
							  n   = str_count(start, ltr))
	do_walk <- function(nodes, depth) {
		if (depth == max_depth) {
			nodes <- difference(V(G), V(G)["start", visited == 0L])
			tibble(ltr = nodes$insert,
							  n   = nodes$visited) %>%
				group_by(ltr) %>% 
				summarize(n = sum(n)) %>% 
				full_join(start_cnt,
							 by = "ltr") %>% 
				transmute(ltr,
							 n = coalesce(n.x, 0L) + coalesce(n.y, 0L)) %>% 
				pull(n) %>% 
				range() %>% 
				diff()
		} else {
			incs <- incident_edges(G, nodes) %>% 
				unlist()
			nbs <- adjacent_vertices(G, nodes) %>% 
				unlist() %>% 
				unique()
			src_dest <- tibble(src  = tail_of(G, incs),
									 dest = head_of(G, incs),
									 load = V(G)[src]$load)
			dest <- src_dest %>% 
				group_by(dest) %>% 
				summarize(load = sum(load))
			V(G)[dest$dest]$load <<- V(G)[dest$dest]$load + dest$load
			src <- src_dest %>% 
				distinct(src, load)
			V(G)[src$src]$load <<- V(G)[src$src]$load - src$load
			V(G)[nbs]$visited <<- V(G)[nbs]$visited + V(G)[nbs]$load
			Recall(V(G)[nbs], depth + 1L)
		}
	}
	do_walk(V(G)["start"], 0L)
}

options(scipen = 1)
walk_graph(40)
```



