---
params:
   task_nr: 6
   year: 2023
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L)
  }
  res <- text_block %>% 
    str_extract_all("\\d+") %>% 
    do.call(rbind, .) %>% 
    t() %>% 
    set_colnames(c("time", "dist"))
  storage.mode(res) <- "integer"
  res
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

Let $t$ be the amount of time we hold down the button, $T$ the length of the race and $r$ 
the record. The distance our boat will glide is given by

$$
d = (T - t)\cdot t
$$
We want to go further than `r`, which  yields
$$
(T - t)\cdot t > r
$$
This is a qudaratic equation:
$$
t^2 - T\cdot t + r = 0
$$
which has the solutions
$$
t_{1,2} = \frac{T}{2} \pm \sqrt{\frac{T^2}{4} - r}
$$

Assume WLOG $t_1 = \min(t_1, t_2)$ and $t_2 = \max(t_1,t_2)$, then the desired number of
winning solutions is given by
$$
n = \lfloor t_2 \rfloor - \lceil t_1 \rceil
$$
*N.B.* If $t_{1}\in\mathbb{N}$ set $t_1 = t_1 + 1$ and if $t_{2}\in\mathbb{N}$ set 
$t_2 = t_2 - 1$ to avoid counting a draw as a win.

```{r get-solution-1}
solve_race <- function(races) {
  D <- sqrt(races[, "time"] ^ 2 / 4 - races[, "dist"])
  t1 <- races[, "time"] / 2 - D
  t2 <- races[, "time"] / 2 + D
  lwr <- pmin(t1, t2)
  upr <- pmax(t1, t2)
  exact <- abs(lwr - round(lwr, 0L)) <= sqrt(.Machine$double.eps)
  lwr[exact] <- lwr[exact] + 1
  exact <- abs(upr - round(upr, 0L)) <= sqrt(.Machine$double.eps)
  upr[exact] <- upr[exact] - 1
  prod(floor(upr) - ceiling(lwr) + 1L)
}

solve_race(puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

With the closed formula in place, we just need to plug in the much larger number to solve
the puzzle.

```{r get-solution-2}
solve_single_race <- function(races) {
  race <- apply(races, 2L, paste, collapse = "")
  storage.mode(race) <- "numeric"
  race <- matrix(race, nrow = 1L, dimnames = list(NULL, colnames(races)))
  solve_race(race)
}
solve_single_race(puzzle_data)
```
