---
params:
   task_nr: 14
   year: 2018
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{js cpp-highlighter, file = "Prism_CPP.js", echo = FALSE}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    as.integer()
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
puzzle[[1]] <- str_replace(puzzle[[1]], "<code>\\\\</code>", "<code>\\\\\\\\</code>")
cat(puzzle[[1]])
```

### Solution

For the first part, we create the vector as requested. The vector can grow by 2 fields in
each iteration. Thus to avoid physically growing the vector in each iteration, we create
a large enough vector in the beginning and keep track about its real length.


```{r get-solution-1}
brew_cacao <- function(n = puzzle_input) {
  cacao_quality <- rep(NA_integer_, n + 10L)
  cacao_quality[1:2] <- c(3L, 7L)
  k <- 2L
  i <- 1L
  j <- 2L
  while (k < n + 10L) {
    new_score <- sum(cacao_quality[c(i, j)])
    if (new_score >= 10L) {
      new_score <- c(1L, new_score - 10L)
    }
    cacao_quality[k + seq_along(new_score)] <- new_score
    k <- k + length(new_score)
    i <- (i + cacao_quality[i]) %% k + 1L
    j <- (j + cacao_quality[j]) %% k + 1L
  }
  cacao_quality[seq(n + 1L, length.out = 10L)] %>% 
    paste(collapse = "")
}
brew_cacao(puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We use again an iterative approach looking for the first match. But since `R` is a bit 
slow, we fall back to `C++` (the `R` version can be found in the appendix).

```{Rcpp cpp-algorithm, file = "cacao.cpp", cache = TRUE}
```

```{r get-solution-2, cache = TRUE}
find_cacao(puzzle_data)
```

## R - Version

```{r cacao-r-version, eval = FALSE}
find_cacao <- function(n = puzzle_data) {
  target <- as.character(n) 
  n_target <- nchar(target)
  cacao_quality <- rep(NA_integer_, 256L)
  cacao_quality[1:2] <- c(3L, 7L)
  k <- 2L
  i <- 1L
  j <- 2L
  found <- FALSE
  while (!found || k < n_target) {
    if (length(cacao_quality) < k + 2L) {
      cacao_quality <- c(cacao_quality, rep(NA_integer_, length(cacao_quality)))
    }
    new_score <- cacao_quality[i] + cacao_quality[j]
    if (new_score >= 10L) {
      new_score <- c(1L, new_score - 10L)
    }
    cacao_quality[k + seq_along(new_score)] <- new_score
    k <- k + length(new_score)
    i <- (i + cacao_quality[i]) %% k + 1L
    j <- (j + cacao_quality[j]) %% k + 1L
    if (k >= n_target) {
      idx <- seq(k - n_target + 1L, length.out = n_target)
      found <- paste(cacao_quality[idx], collapse = "") == target
      if (found) {
        return(k - n_target)
      } else if (length(new_score == 2L)) {
        found <- paste(cacao_quality[idx - 1L], collapse = "") == target
        if (found) {
          return(k - n_target - 1L)
        }
      }
    }
  }
}
```
