---
params:
   task_nr: 18
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2024"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(knitr)
library(cli)
library(bit64)
library(igraph)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  } 
  res <- text_block %>% 
    str_split(fixed(",")) %>% 
    lapply(as.integer) %>% 
    do.call(rbind, .) %>% 
    set_colnames(c("x", "y")) 
  res[, c(2, 1)] + 1L
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
attr(puzzle_data, "dims") <- c(71, 71)
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
attr(example_data, "dims") <- c(7, 7)
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We construct a graph to rely on a "shortest path" algorithm to solve this puzzle. We
use `make_lattice` to make a full lattice graph first and then delete all edges which are
incident to a wall. This approach is way fast rather than looping through all entries and
creating each edge separately.

```{r get-solution-1}
show_map <- function(memory_map) {
  apply(memory_map, 1, paste, collapse = "") %>% 
    paste(collapse = "\n") %>% 
    cat("\n")
}

make_vertex_id <- function(coord, end) {
  start <- c(1L, 1L)
  if (all(c(coord) == start)) {
    "S"
  } else if (all(c(coord) == end)) {
    "E"
  } else {
    paste(coord - 1L, collapse = "/")
  }
}

make_memory_map <- function(corrupted_bytes, 
                            subset = 1:nrow(corrupted_bytes)) {
  dd <- attr(corrupted_bytes, "dims")
  map <- matrix(".", dd[1L], dd[2L])
  map[corrupted_bytes[subset, ]] <- "#"
  walls <- which(map == "#", arr.ind = TRUE)
  G <- make_lattice(dd)
  V(G)$name <- expand.grid(
    col = seq(1L, dd[1L]),
    row = seq(dd[2L], 1L)
  ) %>%
    rowwise() %>%
    mutate(id = make_vertex_id(cbind(row, col), dd)) %>%
    pull(id)
  wall_ids <- apply(walls, 1, make_vertex_id, end = dd)
  wall_edges <- do.call(c, incident_edges(G, wall_ids))
  
  V(G)$shape <- if_else(V(G)$name %in% wall_ids, "square", "circle")
  V(G)$color <- case_when(
    V(G)$name %in% wall_ids ~ "darkgray", 
    V(G)$name %in% c("S", "E") ~ "firebrick",
    TRUE ~ "beige")
  V(G)$label.color <- if_else(V(G)$name %in% c("S", "E"), "white", "black")
  G <- G %>% 
    delete_edges(wall_edges)
  G %>% 
    set_graph_attr("layout", layout_on_grid(G)) %>% 
    set_graph_attr("dims", dd) %>% 
    set_graph_attr("data", corrupted_bytes)
}


find_path <- function(memory_map) {
  distances(memory_map, "S", "E")
}

kb_map <- make_memory_map(puzzle_data, 1:1024)
kb_map %>% 
  find_path()
```

An example graph (after 12 ananoseconds of falling bytes) is visualized below:

```{r visualize-example-graph}
example_data <- structure(
  c(5L, 3L, 6L, 1L, 2L, 4L, 5L, 6L, 7L, 4L, 7L, 2L, 3L,
    6L, 6L, 6L, 5L, 5L, 5L, 2L, 2L, 1L, 6L, 7L, 1L, 6L, 
    5L, 5L, 4L, 3L, 7L, 3L, 2L, 1L, 4L, 3L, 6L, 2L, 6L,
    3L, 7L, 2L, 1L, 7L, 2L, 7L, 2L, 1L, 2L, 3L),
  dim = c(25L, 2L), 
  dimnames = list(NULL, c("x", "y")), dims = c(7, 7))

ex_map <- make_memory_map(example_data, 1:12)

plot(ex_map)
```


## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

With the data stored as graph, the problem reduces to iteratively removing edges and to 
check at each removal, whether start and end are still connected.

```{r get-solution-2}
check_critical_byte <- function(map, from_index) {
  dd <- graph_attr(map, "dims")
  data <- graph_attr(map, "data")
  bytes_seq <- seq(from_index, nrow(data))
  res <- NULL
  for (byte_idx in bytes_seq) {
    byte <- data[byte_idx, , drop = FALSE] 
    byte_id <- make_vertex_id(byte, dd)
    byte_edge <- incident(map, byte_id)
    new_map <- map %>% 
      delete_edges(byte_edge)
    V(new_map)$shape[V(new_map)$name == byte_id] <- "square"
    V(new_map)$color[V(new_map)$name == byte_id] <- "darkgray"
    dis <- c(distances(new_map, "S", "E"))
    if (is.infinite(dis)) {
      res <- byte
      break
    }
    map <- new_map
  }
  list(res = res[, 2:1] - 1L, before = map, after = new_map)
}

check_critical_byte(kb_map, 1025L)$res
```

We can visualize the example before and after the critical byte fell:

```{r visualize-critical_path, fig.show = "hold", out.width = "50%"}
ex_cp <- check_critical_byte(ex_map, 13L)
plot(ex_cp$before)
plot(ex_cp$after)
```

