---
params:
   task_nr: 20
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   o_file <- sprintf("2021_%s",
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
	width: 100%;
}

.back {
	font-size: 2rem;
}

.day-desc {
	border: 2px dotted;
	border-radius: 6px;
	border-color: var(--blockquote-border-color);
	padding: 0 20px;
}

code em {
	font-weight:bold
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
	if (knitr::is_latex_output()) {
		sprintf("\\texttt{%s}", as.character(x))
	} else if (knitr::is_html_output()) {
		sprintf("<tt>%s</tt>", as.character(x))
	} else {
		x
	}
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(tibble)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2021/day/", params$task_nr)
puzzle <- GET(base_url,
				  session_cookie) %>% 
	content(encoding = "UTF-8") %>% 
	xml_find_all("///article") %>% 
	lapply(as.character)

puzzle_data <- local({
	lines <- GET(paste0(base_url, "/input"),
					 session_cookie) %>% 
		content(encoding = "UTF-8") %>% 
		str_split("\n") %>% 
		`[[`(1L) %>% 
		Filter(nzchar, .)
	decoder <- lines[[1L]] %>% 
		str_replace_all(fixed("#"), "1") %>% 
		str_replace_all(fixed("."), "0") %>% 
		str_split("") %>% 
		`[[`(1L) %>%
		as.integer()
	picture <- lines[-1L] %>% 
		str_replace_all(fixed("#"), "1") %>% 
		str_replace_all(fixed("."), "0") %>% 
		str_split("") %>% 
		do.call(rbind, .) %>% 
		{matrix(as.integer(.), nrow(.), ncol(.))} %>% 
		`class<-`(c(class(.), "picture"))
	list(decoder = decoder,
		  picture = picture)
})
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1L]])
```


### Solution

We observe the following points:

1. The infinite part of the picture which is at least one row (column) away from the 
picture will alternate. This is due to the fact that `0` (i.e. all 9 pixels around a pixel are dark `.`) decodes to `` `r puzzle_data$decoder[1]` ``, while `511` (i.e. all 9 pixels
around a pixel are bright `#`) decode to `` `r puzzle_data$decoder[512]` ``.^[For the
sample data that is not the case, as here `0` decodes to `0`. Hence, I parametrized this
fact in the function.]
1. That is in any odd iteration, the number of bright pixels is `Inf` while in any even
iteration we do not have to consider the outside at all.
1. At each iteration the picture grows by 1 row (column) at each side:

```
.....    #####    .....
.....    #####    .zzzz   
..... => ##yyy => .zzzz
...xx    ##yyy    .zzzz 
...xx    ##yyy    .zzzz
```

With these observations the algo is rather straight forward:
1. At each iteration allow for a picture with `dim(pic_{i + 1}) = dim(pic_i) + 2`.
1. Replace pixels via decoder and count (i.e. sum) all `1's`.


```{r solution-a, cache = TRUE}
print.picture <- function(x, top_left = c(1, 1), bottom_right = dim(x), ...) {
	rel <- x[top_left[1]:bottom_right[1], top_left[2]:bottom_right[2]]
	rel <- matrix(as.character(rel), nrow(rel), ncol(rel))
	rel[rel == "1"] <- "#"
	rel[rel == "0"] <- "."
	str <- apply(rel, 1, paste, collapse = "") %>% 
		paste(collapse = "\n")
	cat(str)
	invisible(x)
}

enhance <- function(picture, it, decoder) {
	outside <- 
		if (decoder[1L] == 0L) {
			0L
		} else {
			(it + 1L) %% 2L
		}
	
	new_pic <- rbind(outside,
						  cbind(outside, picture, outside),
						  outside)
	new_pixel <- function(row, col) {
		idx <- cbind(row, col)
		idx[!between(row, 1, nrow(new_pic)) |
			 	!between(col, 1, ncol(new_pic))] <- NA
		vals <- new_pic[idx]
		vals[is.na(vals)] <- outside
		decoder[sum(vals * 2 ^ (rev(seq_along(vals) - 1))) + 1]
	}
	
	idx <- tibble(col = c(col(new_pic)),
					  row = c(row(new_pic))) %>% 
		`[`(rep(seq_len(nrow(.)), each = 9), )
	offsets <- expand.grid(col = -1:1, row = -1:1) %>% 
		`[`(rep(1:9, length(new_pic)), )
	idx <- (idx + offsets) %>% 
		mutate(grp = rep(seq_along(new_pic), each = 9)) %>% 
		select(row, col, grp) %>% 
		as_tibble()
	
	pixels <- idx %>% 
		group_by(grp) %>% 
		summarise(pixel = new_pixel(row, col)) %>% 
		pull(pixel) %>% 
		`dim<-`(dim(new_pic)) %>% 
		`class<-`(class(picture))
	pixels
}

solve <- function(n, data = puzzle_data) {
	if (data$decoder[1L] == 1L && n %% 2) {
		Inf
	} else {
		reduce(seq_len(n),
				 enhance,
				 decode = data$decode,
				 .init = data$picture)
	}
}

sum(solve(2L))


```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We simply call out function with a higher iteration number.

```{r solution-b, cache = TRUE, eval = length(puzzle) > 1}
sum(solve(50L))
```

