---
params:
   task_nr: 13
   year: 2015
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2015"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(tibble)
library(igraph)
library(gtools)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  who <- str_extract(text_block, "^\\S+")
  neighbor <- str_extract(text_block, "\\S+$") %>% 
    str_remove("\\.$")
  pts <- str_extract(text_block, "-?\\d+")
  sign <- str_extract(text_block, "lose|gain")
  tibble(
    from = who,
    to = neighbor,
    pts = as.integer(pts) * if_else(sign == "lose", -1L, 1L)
  )
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We construct a complete graph with the sympathy as edge weight, then we need to find a 
cylce with maximum weight.

```{r get-solution-1}
find_cycles <- function(G) {
  n <- vcount(G)
  id <- 2:n
  ## we have a complete graph, thus to get all cycles we fix the first seat and
  ## permute all the other seats
  all_cycles <- cbind(
    1L, 
    permutations(n - 1L, n - 1L, id)
  )
  apply(all_cycles, 1, function(ids) {
    ## first we need to get an vertex sequence: 1 -> 2, 2 -> 3, ..., n -> 1
    es <- ids[c(1L, rep(2:length(ids), each = 2L), 1L)]
    ## next we want to get the opposing vertex sequence: 2 -> 1, 3 -> 2, ..., 1 -> n
    ## thus, we fist get the even indices of es
    idx <- seq(2L, length(es), by = 2L)
    ## c(rbind(x, y)) == c(x[1], y[1], x[2], y[2], ...)
    ## this results in the proper indices c(2, 1, 4, 3, 6, 5, ...)
    idx <- c(rbind(idx, idx - 1L))
    ## finally we get the reversed vertex sequence
    es_rev <- es[idx]
    ## get the edge ids
    left_cycle <- get_edge_ids(G, es)
    right_cycle <- get_edge_ids(G, es_rev)
    ## mix them together as before
    E(G)[c(rbind(left_cycle, right_cycle))]
  },
  simplify = FALSE)
}

find_maximum_cycle <- function(cycles) {
  sympathy <- map_int(cycles, ~ sum(.x$pts))
  max(sympathy)
}

G <- graph_from_data_frame(puzzle_data)
all_cycles <- find_cycles(G)
find_maximum_cycle(all_cycles)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For the second part, we need to find the minimum sum of sympathy between two neighbors. 
This will be the place, where we sit the new person (this will split up the worst sympathy 
rating and replace it by the value of 0). We subtract this minimum from the total sum of
sympathy and get the result.

```{r get-solution-2}
find_augmented_maximum_cycle <- function(cycles) {
  sympathy <- map_int(cycles, function(cycle) {
    weight <- cycle$pts
    idx <- seq(1L, length(weight), by = 2L)
    mutual_sympathy <- weight[idx] + weight[idx + 1L]
    best_place <- min(mutual_sympathy)
    sum(weight) - best_place
  })
  max(sympathy)
}

find_augmented_maximum_cycle(all_cycles)
```
