---
params:
   task_nr: 21
   year: 2023
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(bit64)
library(numbers)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L)
  }
  text_block[nzchar(text_block)] %>% 
    str_split("") %>% 
    do.call(rbind, .)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

The first part is rather straight forward. We use modulo arithmetic as a preparation for 
part 2.

```{r get-solution-1}
walk_garden <- function(garden, n_steps) {
  dirs <- rbind(
    "^" = c(-1L, 0L),
    ">" = c(0L, 1L),
    "v" = c(1L, 0L),
    "<" = c(0L, -1L)
  )
  pos <- which(garden == "S", arr.ind = TRUE)
  s <- 1L
  while (s <= n_steps) {
    new_pos <- map(1:4, ~ t(t(pos) + dirs[.x, ])) %>% 
      do.call(rbind, .)
    wrap <- sweep(new_pos, 2L, dim(garden), \(m, d) (m - 1L) %% d + 1L)
    new_pos <- new_pos[garden[wrap] != "#", , drop = FALSE]
    pos <- new_pos[!duplicated(new_pos), , drop = FALSE]
    s <- s + 1L
  }
  nrow(pos)
}

walk_garden(puzzle_data, 64)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For the second part we observe that

1. **Periodic world.** The map garden is a finite $W \times H$ tile that repeats 
infinitely in all directions. Walkability at absolute coordinates $(x,y)$ is decided by 
wrapping back to the base tile via $((y \bmod H), (x \bmod W))$.
1. **Residue class matters (not touching the first border).** We do **not** rely on when a 
wavefront “touches the first tile’s border”. What matters is the **residue class** of the 
step count modulo the tile’s **stride $L$**. For a square tile with odd side length 
(e.g., $131 \times 131$) and 4‑neighborhood moves, using $L=W=H$ works well. The start in 
the geometric center gives especially clean behavior for one residue class, but the 
method holds for every residue class.
1. **Stride and residue subsequences.** Fix a residue $r \in \{0,\dots,L-1\}$ and consider 
the subsequence of step counts 
$$n \;=\; r + L \cdot k \quad (k = 0,1,2,\dots).$$
Along this subsequence, the modulo‑tile “phase” (i.e., how moves align with the periodic 
pattern) remains constant.
1. **Quadratic law per residue class.** In 2D, the number of cells reachable after exactly 
$n$ steps grows quadratically in the “shell index” $k$. Consequently, for each residue 
$r$ there exists an exact quadratic polynomial
$$f_r(k) \;=\; A_r\,k^2 + B_r\,k + C_r,$$
such that $\#\text{reachable}(n) = f_r\big((n-r)/L\big)$ whenever 
$n \equiv r \pmod L$.
  > Practically: the leading term is quadratic (area‑like growth); lower‑order terms encode 
  periodic effects and become constants once the residue class is fixed.

1. **Determine coefficients with 3 supports.** To find $A_r,B_r,C_r$, simulate **three** 
step counts within the **same residue class**:
$$n_0 = r,\quad n_1 = r + L,\quad n_2 = r + 2L.$$
Collect the exact counts $f_r(0), f_r(1), f_r(2)$ and perform a quadratic fit 
(equivalently, check that the second finite difference $f(k+2) - 2f(k+1) + f(k)$
is constant). If a residue class stabilizes a bit later, just use more supports 
$k=0\ldots m$ and fit degree 2 once the second difference is constant.
1. **Evaluate for any target $n$.** For a given $n$, compute $r = n \bmod L$ and 
$k = (n - r)/L$. Evaluate $f_r(k)$ from the fitted polynomial and obtain the exact number 
of cells reachable after exactly $n$ steps—without simulating up to $n$.

```{r get-solution-2, cache = TRUE}
walk_garden_infinite <- function(garden, n_steps, supports = 0:2) {
  L <- LCM(nrow(garden), ncol(garden))
  if (any(dim(garden) %% 2L == 0L)) {
    L <- LCM(L, 2L)
  }

  r <- n_steps %% L
  k <- (n_steps - r) / L
  collect_y <- function(k_supports) {
    x <- r + k_supports * L
    y <- map_int(x, ~ walk_garden(garden, .x))
    list(k = as.integer(k_supports), x = as.integer(x), y = as.integer64(y))
  }
  
  is_stable <- function(k, y) {
    if (length(k) < 3L) {
      return(FALSE)
    }
    # Ensure k are consecutive integers (0,1,2,... or any contiguous range)
    kd <- sort(k) %>% 
      diff()
    if (!all(kd == 1L)) {
      return(FALSE)
    }
    y_num <- as.numeric(y)
    d1 <- diff(y_num)
    d2 <- diff(d1)
    length(d2) >= 1L && all(abs(d2 - d2[1L]) < 0.5) 
  }

  k_supports <- as.integer(supports) %>% 
    unique() %>% 
    sort()
  cur <- collect_y(k_supports)

  extra_used <- 0L
  while (!is_stable(cur$k, cur$y)) {
    if (length(k_supports) == 0L) {
      next_k <- 0L
    } else {
      next_k <- max(k_supports) + 1L
    }
    k_supports <- c(k_supports, next_k)
    cur <- collect_y(k_supports)
    extra_used <- extra_used + 1L
  }

  y <- tail(cur$y, 3L)
  x <- r + k_supports * L
  
  a <- (y[3L] - 2 * y[2L] + y[1L]) / 2
  b <- (y[2L] - y[1L]) - a
  c <- y[1L]
  as.integer64(a * k ^ 2L + b * k + c)
}

walk_garden_infinite(puzzle_data, 26501365)
```
