---
params:
   task_nr: 3
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- format(Sys.Date(), "%Y");
   o_file <- sprintf("%s_%s", year, tools::file_path_sans_ext(basename(inputFile)));
   day <- as.numeric(gsub(".*(\\d+)$", "\\1", o_file));
   rm <- readLines(here::here("README.md"));
   line <- grep(paste0("Day ", day, "([^0-9]?$|\\])"), rm);
   rm[line] <- sprintf("- [x] [Day %i](https://thothal.github.io/AoC/%s.html)", day, o_file);
   cat(rm, file = here::here("README.md"), sep = "\n");
   index <- readLines(here::here("index.Rmd"));
   
   if (!any(grepl(paste0(o_file, ".html"), index, fixed = TRUE))) {
      rx <- if (day > 1) sprintf("%s_task%i.html", year, day - 1) else sprintf("# %s", year);
      line <- grep(rx, index, fixed = TRUE);
      new_line <- sprintf("* [Task %i](%s.html)", day, o_file);
      
      if (day == 1) {
         new_line <- c("", new_line);
      }
      

      index <- c(index[seq(1, line)], new_line, if(line < length(index)) index[(line + 1):length(index)]);
      cat(index, file = here::here("index.Rmd"), sep = "\n");
      rmarkdown::render(here::here("index.Rmd"), encoding = encoding, output_file = here::here("docs", "index"));
      
   }
   
   rmarkdown::render(inputFile, encoding = encoding, output_file = o_file, output_dir = here::here("docs")) })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
width: 100%;
}

.back {
font-size: 2rem;
}

.day-desc {
border: 2px dotted;
border-radius: 6px;
border-color: var(--blockquote-border-color);
padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center")
## Define a function to write text in typewriter font
tt <- function(x) {
   if (knitr::is_latex_output()) {
      sprintf("\\texttt{%s}", as.character(x))
   } else if (knitr::is_html_output()) {
      sprintf("<tt>%s</tt>", as.character(x))
   } else {
      x
   }
}
```

# Setup

## Libraries

First we need to load the required libraries. 

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(tibble)
library(dplyr)
library(stringr)
library(purrr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache=FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2021/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
   content(encoding = "UTF-8") %>% 
   xml_find_all("///article") %>% 
   lapply(as.character)

puzzle_data <- GET(paste0(base_url, "/input"),
                   session_cookie) %>% 
   content(encoding = "UTF-8") %>% 
   read.table(text = ., colClasses = "character") %>% 
   as_tibble() %>% 
   set_names("status_code")
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We pad all binary numbers from left with `r tt("0")`'s and transform the data
into a matrix where each column represents a digit of the binary code.^[This 
step is strictly speaking not necessary as, if readed in als character, all
numbers are properly formtted. Only if readed in as numbers (which is the
default if `read.table` is not told otherwise) we see unequal length. However,
we keep this code to play it safe.] 
Since, we  padded all numbers to the maximum length, all items will be properly 
filled. 

Then we can simply calculate the column sum to determine the bit. We will code 
the bits with `logicals` which makes the negation easy (*N.B.* Integers would 
have been likewise easy to use). The last step is then to calculate the decimal
representation (with the highest bit on the left).

```{r solution-a}
digits <- max(nchar(puzzle_data$status_code))
n <- NROW(puzzle_data)
puzzle_data <- puzzle_data %>% 
   mutate(status_code = str_pad(as.character(status_code), digits, pad = "0"))
bin_matrix <- puzzle_data %>% 
   pull(status_code) %>% 
   str_split_fixed(boundary("character"), digits) %>% 
   as.numeric() %>% 
   matrix(n, digits) %>%
   `colnames<-`(paste0("V", seq_len(digits)))
ones <- bin_matrix %>% 
   colSums()

gamma <- ones > round(n / 2, 0)
epsilon <- !gamma

bin2dec <- function(x) {
   sum(x * 2 ^ seq(length(x) - 1, 0, -1))
}

bin2dec(gamma) * bin2dec(epsilon)

```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We will use `purrr::reduce` to apply the filtering function to the columns of 
the matrix (which is back-transformed to a `tibble` to make use of the list like 
structure). The filtering function is a binary function with parameters `cond` 
and `col`.^[Technically it also takes a third parameter `type`, indicating 
whether we filter for  majority or minority.]
* `cond` is a logical vector of length `NROW(data)` indicating whether 
a certain row is still in the race.
* `col` is simply the next column to investigate.

The function stops early (using `rlang::done`), if there is only one valid item
in the list. We make again use of the fact that `TRUE == 1` and `FALSE == 0`.

```{r solution-b, eval = length(puzzle) > 1}
col_filter <- function(cond, col, type = c("majority", "minority")) {
   if (sum(cond) > 1) {
      ## are there more than one cndidates
      type <- match.arg(type)
      ## how many 1s are there among the candidates
      k <- sum(col[cond])
      n <- length(col[cond])
      ## for which bit should we filter:
      ## - if there are more ones than zeros (k >= n - k) and we want to go 
      ##   for the majority than it should be 1
      ## - if there are less ones than zero (k < n - k) and we want to go 
      ##   for the majority than it should be 0
      goal <- switch(type, majority = (k >= n - k), minority = (k < n - k))
      new_cond <- rep(FALSE, length(col))
      ## candidates stay candidates if the have the right bit
      new_cond[cond] <- col[cond] == goal
      new_cond
   } else {
      done(cond)
   }
}

oxygen <- bin_matrix %>% 
   as_tibble() %>% 
   reduce(col_filter, .init = rep(TRUE, NROW(.))) %>% 
   `[`(bin_matrix, .) %>% 
   bin2dec()
co2 <- bin_matrix %>% 
   as_tibble() %>% 
   reduce(col_filter, type = "minority", .init = rep(TRUE, NROW(.))) %>% 
   `[`(bin_matrix, .) %>% 
   bin2dec()

oxygen * co2
```


