---
params:
   task_nr: 22
   year: 2020
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  player_markers <- str_which(text_block, "Player")
  list(player_1 = as.integer(text_block[2:(player_markers[2L] - 1L)]),
       player_2 = as.integer(text_block[(player_markers[2L] + 1L):length(text_block)]))
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We solve this game by using a pointer to the current card as well as a large enough vector
to store all cards. This avoids creating a lot of copies as just need to move single 
values.

```{r get-solution-1}
play_combat <- function(cards) {
  p1 <- cards$player_1
  p2 <- cards$player_2
  n <- length(p1)
  m <- length(p2)
  p1 <- c(p1, rep(NA_integer_, m))
  p2 <- c(p2, rep(NA_integer_, n))
  while (m > 0L && n > 0L) {
    c1 <- p1[1L]
    c2 <- p2[1L]
    if (c1 > c2) {
      p1 <- lead(p1)
      p1[n:(n + 1L)] <- c(c1, c2)
      n <- n + 1L
      p2 <- lead(p2)
      m <- m - 1L
    } else if (c1 < c2) {
      p2 <- lead(p2)
      p2[m:(m + 1L)] <- c(c2, c1)
      m <- m + 1L
      p1 <- lead(p1)
      n <- n - 1L
    }
  }
  if (m == 0L) {
    ## player 1 won
    sum(seq_along(p1) * rev(p1))
  } else {
    ## player 2 won
    sum(seq_along(p2) * rev(p2))
  }
}

play_combat(puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]] %>% str_replace_all(fixed("*"), "\\*"))
```

### Solution

We implement the recursive version of this game according to the rules

```{r get-solution-2, cache = TRUE}

play_recursive_combat <- function(cards) {
 lead_int <- function(x) {
    n <- length(x)
    if (n <= 1L) {
      return(rep(NA_integer_, n))
    }
    c(x[-1L], NA_integer_)
  }

  make_key <- function(p1, n, p2, m) {
    paste(
      paste(p1[seq_len(n)], collapse = "/"),
      paste(p2[seq_len(m)], collapse = "/"),
      sep = "_"
    )
  }
  seen <- new.env(parent = emptyenv())
  play_game <- function(p1, p2) {
    decks_hash <- new.env(parent = emptyenv())
    n <- length(p1)
    m <- length(p2)
    p1 <- c(p1, rep(NA_integer_, m))
    p2 <- c(p2, rep(NA_integer_, n))
    while (m > 0L && n > 0L) {
      key <- make_key(p1, n, p2, m)
      if (is.null(decks_hash[[key]])) {
        decks_hash[[key]] <- TRUE
      } else {
        return (list(winner = 1L))
      }
      c1 <- p1[1L]
      c2 <- p2[1L]
      if (n - 1L >= c1 && m - 1L >= c2) {
        ## players have enough cards
        p1_new <- p1[seq(2L, length.out = c1)]
        p2_new <- p2[seq(2L, length.out = c2)]
        key_sub <- make_key(p1_new, length(p1_new), p2_new, length(p2_new))
        res <- seen[[key_sub]]
        if (is.null(res)) {
          res <- Recall(p1_new, p2_new)
          seen[[key_sub]] <<- res
        }
      } else {
        res <- list(winner = (c1 < c2) + 1L)
      }
      if (res$winner == 1L) {
        p1 <- lead_int(p1)
        p1[n] <- c1
        p1[n + 1L] <- c2
        n <- n + 1L

        p2 <- lead_int(p2)
        m <- m - 1L
      } else if (res$winner == 2L) {
        p2 <- lead_int(p2)
        p2[m] <- c2
        p2[m + 1L] <- c1
        m <- m + 1L

        p1 <- lead_int(p1)
        n <- n - 1L
      }
    }
    if (m == 0L) {
      ## player 1 won
      list(winner = 1L, p1 = p1, p2 = p2)
    } else {
      ## player 2 won
      list(winner = 2L, p1 = p1, p2 = p2)
    }
  }
  res <- play_game(cards$player_1, cards$player_2)
  if (res$winner == 1L) {
    sum(seq_along(res$p1) * rev(res$p1))
  } else {
    sum(seq_along(res$p2) * rev(res$p2))
  }
}
play_recursive_combat(puzzle_data)
```
