---
params:
   task_nr: 7
   year: 2016
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2016"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    str_extract_all("[^\\[\\]]+|\\[[^\\[\\]]+\\]")
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We use regex to find palindroms and then apply it to all strings and apply the logic on
whether it must o must not appear.

```{r get-solution-1}
is_palindrome <- function(string) {
  str_detect(string, "(?=(.)(.)\\2\\1)(?!\\1\\1\\1\\1)")
} 

check_ip <- function(tokens) {
  is_p <- is_palindrome(tokens)
  is_n <- str_detect(tokens, "\\[")
  any(is_p[!is_n]) &&
    all(!is_p[is_n])
}

sum(map_lgl(puzzle_data, check_ip))
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

We extract first all palindromes in the respective parts, and check then whether they can 
be found in the reverted form in the respective parts of the IP.

```{r get-solution-2}
extract_palindromes <- function(string) {
  res <- str_match_all(string, "(?=(.)(.)(\\1))(?!\\1\\1\\1)")
  map(res, function(matches) {
    if (nrow(matches) == 0L) {
      character(0)
    } else {
      apply(matches[, -1L, drop = FALSE], 1L, paste, collapse = "")
    }
  })
}

reverse_palindromes <- function(palindrome) {
  str_replace(palindrome, "(.)(.)(.)", "\\2\\1\\2")
}

check_ip_ssl <- function(tokens) {
  all_p <- extract_palindromes(tokens)
  is_n <- str_detect(tokens, "\\[")
  n_rp <- reverse_palindromes(unlist(all_p[is_n]))
  any(unlist(all_p[!is_n]) %in% unlist(n_rp))
}

sum(map_lgl(puzzle_data, check_ip_ssl))
```
