---
params:
   task_nr: 11
   year: 2015
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2015"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{js sql-highlighter, file = "Prism_CPP.js", echo = FALSE}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(tibble)
library(magrittr)
library(dplyr)
library(stringr)
library(knitr)
library(purrr)
library(cli)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block 
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We work with letter positions rather than the letters themselves as it makes calculation 
way easier.

We use a function to validate a given password and another to increment it. We then 
increment the password until we reach a valid password.

The solution was first implemented in Rm but was slow-ish (it took more than 2 minutes to
calculate the password). Hence, we re-implemented it using C++.

The algorithm could be tremendously improved by not increasing passwords by one step at 
a time, but jumping invalid passwords right away.

However, the C++ solution was blasting fast anyways, so we did not bother to further 
improve the solution.

```{Rcpp cpp-algorithm, file = "pwgen.cpp", cache = TRUE}
```

```{r get-solution-1}
(res <- find_next_password(puzzle_data))
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

All we need to do is to apply the same function to the previously generated password.

```{r get-solution-2}
 find_next_password(res)
```


# R Legacy Solution

```{r legacy-code, eval = FALSE}
validate_password <- function(code) {
  distance <- diff(code)
  rld <- rle(distance)
  rl <- rle(code)
  idx_iol <- match(c("i", "o", "l"), letters)
  any(rld$values == 1L & rld$lengths >= 2L) && ## 3 consecutive
    !any(idx_iol %in% code) && ## no i, o or l
    (any(rl$lengths >= 4) || sum(rl$lengths >= 2) >= 2) ## at least 2 pairs
}

increment_password <- function(code) {
  n <- i <- length(code)
  done <- FALSE
  while (!done) {
    new_digit <- max((code[i] + 1L) %% 27L, 1L)
    done <- new_digit != 1L
    code[i] <- new_digit
    i <- if_else(i - 1 == 0, n, i - 1)
  }
  code
}

find_next_valid_password <- function(password) {
  cli_progress_bar("Looking for valid passwords")
  code <- str_split(password, "") %>% 
    extract2(1L) %>% 
    match(letters)
  is_invalid <- TRUE
  n <- length(code)
  idx_iol <- match(c("i", "o", "l"), letters)
  while (is_invalid) {
    cli_progress_update()
    i <- n
    done <- FALSE
    ## increment
    while (!done) {
      new_digit <- max((code[i] + 1L) %% 27L, 1L)
      done <- new_digit != 1L
      code[i] <- new_digit
      i <- if_else(i - 1L == 0L, n, i - 1L)
    }
    ## check
    distance <- diff(code)
    rld <- rle(distance)
    rl <- rle(code)
    is_invalid <- !(
      any(rld$values == 1L & rld$lengths >= 2L) && ## 3 consecutive
      !any(idx_iol %in% code) && ## no i, o or l
      (any(rl$lengths >= 4) || sum(rl$lengths >= 2) >= 2) ## at least 2 pairs
    )
  }
  cli_progress_done()
  letters[code] %>% 
    paste(collapse = "")
}

find_next_valid_password(puzzle_data)
```

