---
params:
   task_nr: 16
   year: 2017
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```

```{js cpp-highlighter, file = "Prism_CPP.js", echo = FALSE}
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  res <- text_block %>% 
    str_split(",") %>% 
    extract2(1L)
  str_sub(res, 2L) %>% 
    str_split("/") %>% 
    map(~ c(.x, NA)[1:2]) %>% 
    do.call(rbind, .) %>% 
    set_colnames(paste0("arg", 1:2)) %>% 
    as_tibble() %>% 
    mutate(op = str_sub(res, 1L, 1L), .before = 1L) %>% 
    mutate(orig_op = res)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We implement a simple iterative approach, executing one line after the other.

```{r get-solution-1, cache = TRUE}
dance <- function(ops, n_dancers = 16L, dancers = letters[1:n_dancers]) {
  p <- function(arg1, arg2) {
    pos <- which(dancers %in% c(arg1, arg2))
    dancers[pos] <<- dancers[rev(pos)]
  }
  x <- function(arg1, arg2) {
    arg1 <- as.integer(arg1) + 1L
    arg2 <- as.integer(arg2) + 1L
    dancers[c(arg1, arg2)] <<- dancers[c(arg2, arg1)]
  }
  s <- function(arg1, arg2) {
    dancers <<- dancers[(seq_along(dancers) - 1L + 
                           n_dancers - as.integer(arg1)) %% n_dancers + 1L]
  }
  for (i in 1:nrow(ops)) {
    call <- ops %>% 
      slice(i) %>% 
      select(op:arg2) %>% 
      as.list()
    do.call(call[[1L]], call[-1L])
  }
  dancers
}
dance(puzzle_data) %>% 
  paste(collapse = "")

```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For the second part we want to detect cycles. If we find a cycle and the remaining 
iterations are a multiple of the cycle length we can skip right away. 

Again `R` is too slow for this, so we fall - once more - back to `C++`. The R algorithm 
can be found in the appendix.

```{Rcpp cpp-algorithm, file = "dance.cpp", cache = TRUE}
```

```{r get-solution-2, cache = TRUE}
dance(puzzle_data %>% 
        pull(orig_op), 1000000000)
```

## R - Version

```{r dance-r-version, eval = FALSE}
repeat_dance <- function(ops, nr_dances) {
  seen <- list()
  state <- letters[1:16]
  for (i in 1:nr_dances) {
    state <- dance(ops, dancers = state)
    state_id <- paste(state, collapse = "")
    if (!is.null(seen[[state_id]])) {
      if ((nr_dances - i) %% (i - seen[[state_id]]) == 0L) {
        break
      }
    }
    seen[[state_id]]
  }
  state_id
}
```

