---
params:
   task_nr: 19
   year: 2022
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: TRUE
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    str_extract_all("\\d+") %>% 
    do.call(rbind, .) %>% 
    set_colnames(c("id", "ore_cost_ore", "clay_cost_ore", "obsidian_cost_ore", 
                   "obsidian_cost_clay", "geode_cost_ore", "geode_cost_obsidian")) %>% 
    as_tibble() %>% 
    mutate(across(everything(), as.integer)) %>% 
    mutate(costs = list(
      list(ore = c(ore = ore_cost_ore, clay = 0L, obsidian = 0L),
                        clay = c(ore = clay_cost_ore, clay = 0L, obsidian = 0L),
                        obsidian = c(ore = obsidian_cost_ore, clay = obsidian_cost_clay,
                                     obsidian = 0L),
                        geode = c(ore = geode_cost_ore, clay = 0L, 
                                  obsidian = geode_cost_obsidian))
      ),
      .by = id) %>% 
    pull(costs)
           
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

```{r get-solution-1}
find_best_blueprint <- function(bps) {
  solve_blueprint <- function(bp, bp_nr) {
    cat("Solving Blueprint <", bp_nr, ">...\n", sep = "")
    time_limit <- 24L
    robots <- c(ore = 1L, clay = 0L, obsidian = 0L, geode = 0L)
    resources <- c(ore = 0L, clay = 0L, obsidian = 0L, geode = 0L)
    max_geode <- 0L
    memo <- new.env(parent = emptyenv())
    max_cost <- do.call(rbind, bp) %>% 
      apply(2L, max)
    max_cost["geode"] <- Inf
    get_upper_bound <- function(time, robots, resources) {
      resources["geode"] + # amount of geode we have so far
        robots["geode"] * time + # amount of geode the existing robots will produce
        time * (time - 1L) / 2L # max output if we could build a geode robot eahc turn
    }
    
    cap_resources <- function(resources, time, max_cost) {
      capped <- pmin(resources[names(max_cost)], max_cost * time)
      c(capped, resources["geode"])
    }
    rec <- 0L
    
    dfs <- function(time, robots, resources) {
            rec <<- rec + 1L
      if (rec  %% 10000L == 0L) {
        cat("Recursion:", rec, "\tMemo Size:", length(ls(memo)), 
            "\tBest:", max_geode, "\r", sep = "")
      }
      if (time == 0L) {
        max_geode <<- max(max_geode, resources["geode"])
        return(resources["geode"])
      }
      if (get_upper_bound(time, robots, resources) <= max_geode) {
        return(0L)
      }
      
      capped_resources <- cap_resources(resources, time, max_cost)
      key <- paste(time, paste(robots, collapse = ","), 
                   paste(capped_resources, collapse = ","),
                   sep = "|")
      if (exists(key, memo, inherits = FALSE)) {
        return(memo[[key]])
      }
      max_local <- 0L
      produced <- resources + robots
      robot_types <- rev(names(robots))
      for (robot_type in robot_types) {
        if (robot_type != "geode" && robots[robot_type] >= max_cost[robot_type]) {
          next
        }
        robot_cost <- bp[[robot_type]]
        if (all(resources[-4L] >= robot_cost)) {
          new_resources <- produced - c(robot_cost, 0L)
          new_robots <- robots
          new_robots[robot_type] <- new_robots[robot_type] + 1L
          if (robot_type == "geode") {
            max_local <- max(max_local, Recall(time - 1L, new_robots, new_resources))
            memo[[key]] <<- max_local
            return(max_local)
          }
          max_local <- max(
            max_local,
            Recall(time - 1L, new_robots, new_resources)
          )
        }  
      }
      max_local <- max(
        max_local,
        Recall(time - 1L, robots, produced)
      )
      memo[[key]] <<- max_local
      return(max_local)
    }
    res <- dfs_cpp(time_limit, robots, resources, max_cost, bp) #dfs(time_limit, robots, resources)
    cat("\nBest Solution:", max_geode, "\n")
    res
  }
  imap_int(bps, solve_blueprint) %>% 
    max()
}


```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For the second part we fall back to a graph. We first construct the cube with the 
appropriate padding. Then, we label all neighbors of the solid blocks as `surface`. Then,
we look at the induced subgraph, where we remove all `solid` nodes. This will split the
original cube into components. For each component we check whether there is at least
one `padding` node, which indicates that the component has a connection to the exterior.

With this classification, we subset all `surface` nodes, which are also on the outside and
count the number of `solid` neighbors per `surface` node and sum these figures,
which represents the number of surfaces.

```{r get-solution-2}
flag_surface_nodes <- function(G) {
  ttype <- as.character(V(G)$type)
  allowed <- ttype != "solid"
  H <- induced_subgraph(G, vids = V(G)[allowed])
  ttype_H <- as.character(V(H)$type)
  comp <- components(H)
  has_pad_by_comp <- tapply(ttype_H == "padding", comp$membership, any)
  has_pad_H <- has_pad_by_comp[as.character(comp$membership)]
  orig_vids <- as.integer(V(G)[allowed])
  flag <- rep(FALSE, vcount(G))
  surface_mask_H <- ttype_H == "surface"
  flag[orig_vids[surface_mask_H]] <- has_pad_H[surface_mask_H]
  flag
}

construct_cube <- function(coords) {
  coords <- coords + 1L
  max_d <- apply(coords, 2L, max) + 2L
  G <- make_lattice(max_d)
  n <- vcount(G)
  i <- seq_len(n) - 1L
  x <- i %% max_d[1L]
  y <- (i %/% max_d[1L]) %% max_d[2L]
  z <- i %/% (max_d[1L] * max_d[2L])
  V(G)$name <- sprintf("%d/%d/%d", x, y, z)
  is_pad <- (x == 0L | x == (max_d[1L] - 1L) |
               y == 0L | y == (max_d[2L] - 1L) |
               z == 0L | z == (max_d[3L] - 1L))
  V(G)$type <- "empty"
  V(G)$type[is_pad]<- "padding"
  solids <- apply(coords, 1L, paste, collapse = "/") 
  V(G)[solids]$type <- "solid"
  nbs <- neighborhood(G, nodes = solids) %>% 
    reduce(union) %>% 
    setdiff(V(G)[solids])
  V(G)[nbs]$type <- "surface"
  V(G)$is_outer <- flag_surface_nodes(G)
  G
}

G <- construct_cube(puzzle_data)

sum(
  sapply(V(G)[V(G)$is_outer], function(v) {
    sum(V(G)[neighbors(G, v)]$type == "solid")
  })
)

```