---
params:
   task_nr: 4
   year: 2023
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   year <- basename(dirname(dirname(inputFile)))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}

```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/", params$year, "/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L)
  }
  text_block[nzchar(text_block)] %>% 
    str_remove_all("Card +\\d+: ") %>% 
    str_split(" +\\| +") %>% 
    map(function(game) {
      res <- str_split(game, " +")
      list(winning = as.integer(res[[1L]]),
           numbers = as.integer(res[[2L]]))
    })
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We need to count the number of matches and take two to the power of this number to get the
value of each card. Eventually we sum up these figures to get the final answer.

```{r get-solution-1}
calculate_points <- function(cards) {
  map_int(cards, function(.x) {
    matches <- sum(.x$numbers %in% .x$winning)
    sign(matches) * 2L ^ (matches - 1L) ## sign(0) = 0 
  }) %>% 
    sum()
}
calculate_points(puzzle_data)
```

## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

For the second part, we calculate the number of matching numbers and iterate over this
vector. We add as many cards as we have copies of the winning card and eventually sum the
number of cards to get the total number of cards.

```{r get-solution-2}
calculate_cards <- function(cards) {
  n <- length(cards)
  nr_wins <- map_int(cards, ~ sum(.x$numbers %in% .x$winning))
  nr_cards <- rep(1L, n)
  for (i in seq_len(n)) {
    idx <- seq(i + 1L, length.out = nr_wins[i])
    idx <- idx[idx <= n]
    nr_cards[idx] <- nr_cards[idx] + nr_cards[i]
  }
  sum(nr_cards)
}

calculate_cards(puzzle_data)
```
