---
params:
   task_nr: 20
title: '`r paste("<span>Task", params$task_nr, "<span class = \"back\"><a href = \"index.html\">", fontawesome::fa("home"), "</a></span></span>")`'
author: 
  name: "Thorn Thaler"
  email: "thorn.thaler@thothal.at"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
   force_year <- "2024"
   
   year <- dplyr::coalesce(force_year, format(Sys.Date(), "%Y"))
   
   o_file <- sprintf("%s_%s", year, 
                     tools::file_path_sans_ext(basename(inputFile)))
                     
   rmarkdown::render(inputFile, encoding = encoding, 
                     output_file = o_file, output_dir = here::here("docs"))
   })
output: 
  rmdformats::downcute:
    highlight: tango
    use_bookdown: TRUE
    mathjax: NULL
    lightbox: TRUE
    gallery: TRUE
editor_options: 
  chunk_output_type: console
---

```{css custom-css, echo = FALSE}
.page-content  .figure {
  width: 100%;
}

.back {
  font-size: 2rem;
}

.day-desc {
  border: 2px dotted;
  border-radius: 6px;
  border-color: var(--blockquote-border-color);
  padding: 0 20px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.align = "center",
                      dev.args = list(bg = "transparent"))
## Define a function to write text in typewriter font
tt <- function(x) {
  if (knitr::is_latex_output()) {
    sprintf("\\texttt{%s}", as.character(x))
  } else if (knitr::is_html_output()) {
    sprintf("<tt>%s</tt>", as.character(x))
  } else {
    x
  }
}
```

# Setup

## Libraries

```{r libs, warning = FALSE}
library(httr)
library(here)
library(xml2)
library(magrittr)
library(dplyr)
library(purrr)
library(stringr)
library(stringi)
library(knitr)
library(cli)
library(bit64)
library(igraph)
```

## Retrieve Data from `r tt("AoC")`


```{r get-data, cache = FALSE, messages = FALSE}
session_cookie <- set_cookies(session = keyring::key_get("AoC-GitHub-Cookie"))
base_url <- paste0("https://adventofcode.com/2024/day/", params$task_nr)
puzzle <- GET(base_url,
              session_cookie) %>% 
  content(encoding = "UTF-8") %>% 
  xml_find_all("///article") %>% 
  lapply(as.character)

parse_puzzle_data <- function(text_block = readClipboard()) {
  if (length(text_block) == 1L) {
    text_block <- text_block %>% 
      str_split("\n") %>% 
      extract2(1L) %>% 
      keep(nzchar)
  }
  text_block %>% 
    str_split("") %>% 
    do.call(rbind, .)
}

puzzle_data <- local({
  GET(paste0(base_url, "/input"),
      session_cookie) %>% 
    content(encoding = "UTF-8") %>% 
    parse_puzzle_data()
})
```

```{r example-data, eval = interactive(), echo = FALSE}
example_data <- parse_puzzle_data()
```

# Puzzle Day `r params$task_nr`

## Part 1

### Description

```{r show-puzzle-a, echo = FALSE, results = "asis"}
cat(puzzle[[1]])
```

### Solution

We create a full lattice graph, where we set the weight through walls to `Inf`. Then we
calculate the shortest path from start to end, store the distance to the end at each node
and iterate over each node. For each node we get all neighbors which were not yet visited 
and are exactly 2 fields away. We calculate the distance saved by substracting the 
distance to the end at the neighbor and the current node - 2. In the example maze below, 
we see that a shortcut from field `84` to `78` would save us `84 - 78 - 2 = 4` fields
(we need to account for the one wall field which we are now traversing as well).

We eventually count all gains, which are greater than 100.

```{r get-solution-1, cache = TRUE}
make_race <- function(data) {
  dd <- dim(data)
  G <- make_lattice(dd)
  V(G)$name <- expand.grid(col = 1:dd[2L],
                           row = dd[1L]:1) %>% 
    apply(1, \(r) paste(rev(r), collapse = "/"))
  walls <- which(data == "#", arr.ind = TRUE)
  start <- which(data == "S", arr.ind = TRUE)
  end <- which(data == "E", arr.ind = TRUE)
  V(G)$type <- case_when(
    V(G)$name %in% apply(walls, 1, paste, collapse = "/") ~ "wall",
    V(G)$name %in% apply(start, 1, paste, collapse = "/") ~ "start",
    V(G)$name %in% apply(end, 1, paste, collapse = "/") ~ "end",
    TRUE ~ "track"
  )
  V(G)$shape = if_else(V(G)$type == "wall", "square", "circle")
  V(G)$color = case_when(
    V(G)$type == "wall" ~ "darkgray",
    V(G)$type == "start" ~ "forestgreen",
    V(G)$type == "end" ~ "firebrick",
    TRUE ~ "beige")
  V(G)$label.color <- if_else(V(G)$type %in% c("start", "end"), "white", "black")
  wghts <- ends(G, E(G), FALSE) %>% 
    apply(2, \(v) V(G)[v]$type == "wall") %>% 
    rowSums()
  E(G)$weight <- if_else(wghts == 0L, 1L, Inf)
  G %>% 
    set_graph_attr("layout", layout_on_grid(G)) %>% 
    set_vertex_attr("size", value = 10L)
}

calculate_shortcuts <- function(G, min_saving = 100L) {
  ref <- shortest_paths(G, V(G)[type == "start"], V(G)[type == "end"])$vpath[[1L]]
  V(G)[ref]$distance <- rev(seq_along(ref) - 1L)
  V(G)$frame.width <- 1
  V(G)[ref]$frame.width <- 2
  V(G)[ref]$name <- V(G)[ref]$distance
  V(G)$calculated <- FALSE
  res <- 0L
  for (node_idx in ref) {
    node <- V(G)[node_idx]
    nbs <-  neighborhood(G, 2L, node, mindist = 2L)[[1L]][type != "wall" & 
                                                            !calculated]
    if (length(nbs) > 0) {
      shortcut_gain <- abs(node$distance - nbs$distance) - 2L
      res <- res + sum(shortcut_gain >= min_saving)
    }
    V(G)[node_idx]$calculated <- TRUE
  }
  res
}

G <- make_race(puzzle_data)
calculate_shortcuts(G, 100)
```

```{r example-maze, echo = FALSE, fig.width = 10, fig.height = 10}
example_maze <- structure(
  c("#", "#", "#", "#", "#", "#", "#", "#", "#", "#", "#", "#", "#", "#", "#", "#", ".", 
    ".", "S", "#", "#", "#", "#", "#", ".", ".", ".", ".", ".", "#", "#", ".", "#", "#", 
    "#", "#", "#", "#", "#", ".", "#", "#", "#", ".", "#", "#", ".", ".", ".", "#", "#", 
    "#", ".", ".", ".", "#", ".", ".", ".", "#", "#", "#", "#", ".", "#", "#", "#", ".", 
    "#", "#", "#", ".", "#", "#", "#", "#", ".", ".", ".", "#", "#", "#", "E", "#", "#", 
    "#", ".", ".", ".", "#", "#", ".", "#", "#", "#", "#", "#", "#", "#", "#", "#", "#", 
    "#", ".", "#", "#", ".", ".", ".", ".", ".", ".", ".", "#", ".", ".", ".", ".", ".", 
    "#", "#", "#", "#", "#", "#", "#", "#", ".", "#", ".", "#", "#", "#", "#", "#", "#", 
    ".", ".", ".", ".", ".", ".", ".", "#", ".", ".", ".", ".", ".", "#", "#", ".", "#", 
    "#", "#", "#", "#", "#", "#", "#", "#", "#", "#", ".", "#", "#", ".", "#", ".", ".", 
    ".", "#", ".", ".", ".", "#", ".", ".", ".", "#", "#", ".", "#", ".", "#", ".", "#", 
    ".", "#", ".", "#", ".", "#", "#", "#", "#", ".", ".", ".", "#", ".", ".", ".", "#", 
    ".", ".", ".", "#", "#", "#", "#", "#", "#", "#", "#", "#", "#", "#", "#", "#", "#", 
    "#", "#", "#", "#"), dim = c(15L, 15L))

ex <- make_race(example_maze)
ref <- shortest_paths(ex, V(ex)[type == "start"], V(ex)[type == "end"])$vpath[[1L]]
V(ex)[ref]$distance <- rev(seq_along(ref) - 1L)
V(ex)$frame.width <- 1
V(ex)[ref]$frame.width <- 2
V(ex)[ref]$name <- V(ex)[ref]$distance
plot(ex)
```


## Part 2

### Description

```{r show-puzzle-b, echo = FALSE, results = "asis", eval = length(puzzle) > 1}
cat(puzzle[[2]])
```

### Solution

The same approach as before, but this time we look at all fields at most 20 fields away.
We need to substract the distance from shortcut entry to exit to account for the fields
used by this shortcut (similar to the -2 factor before).

The algos are a bit of an overkill, and using much lower level logic would prove useful,
but the fun of working with graphs won in this case.

```{r get-solution-2, cache = TRUE}
is_valid_shortcut <- function(path) {
  types <- path$type %>% 
    rle() %>% 
    extract2("values")
  sum(types == "wall") == 1
}

calculate_shortcuts_long <- function(G, min_saving = 100L) {
  ref <- shortest_paths(G, V(G)[type == "start"], V(G)[type == "end"])$vpath[[1L]]
  V(G)[ref]$distance <- rev(seq_along(ref) - 1L)
  V(G)$frame.width <- 1
  V(G)[ref]$frame.width <- 2
  V(G)[ref]$name <- V(G)[ref]$distance
  V(G)$calculated <- FALSE
  res <- 0L
  for (node_idx in ref) {
    node <- V(G)[node_idx]
        #if (node$name == 84) browser()
    nbs <-  neighborhood(G, 20L, node, mindist = 2L)[[1L]][type != "wall" & 
                                                            !calculated]
    
    #paths <- shortest_paths(G, node, nbs, weight = NA)$vpath
    #is_shortcut <- vapply(paths, is_valid_shortcut, logical(1L))
    shortcuts <- nbs#[is_shortcut]
    if (length(shortcuts) > 0) {
      shortcut_gain <- abs(node$distance - shortcuts$distance) - 
        distances(G, node, shortcuts, weights = NA)
      res <- res + sum(shortcut_gain >= min_saving)
    }
    V(G)[node_idx]$calculated <- TRUE
  }
  res
}
calculate_shortcuts_long(G, 100)
```

